---
title: "CIMIC Combined Patients Analysis"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Read in Objects from I.L. Analysis

```{r}

reading_path <- "D:/OneDrive - University of Florida/Gbadamosi Lab/Iasmim Lopes/CIMIC_combined/test3/objects"

combined_TPM_lognorm <- readRDS(paste0(reading_path, "/Combined_TPM_lognorm.rds"))
master_induction_df <- readRDS(paste0(reading_path, "/master_induction_df.rds"))
clustered_plot_df <- readRDS(paste0(reading_path, "/clustered_plot_df.rds"))
formatted_individual_deltas_matrix <- readRDS(paste0(reading_path, "/clustering_matrix.rds"))
formatted_individual_deltas_df <- as.data.frame(formatted_individual_deltas_matrix )
sting_gas_df <- readRDS(paste0(reading_path, "/clustering_matrix_1_sting_gas.rds"))


# merge to get sting and gas
library(dplyr)
library(tibble)

# -------------------------------------------------
# 1. Convert rownames in sting_gas_df to column
# -------------------------------------------------
sting_gas_df2 <- sting_gas_df %>%
  rownames_to_column("sample_id")

# -------------------------------------------------
# 2. Select only STING-related genes
# -------------------------------------------------
sting_subset <- sting_gas_df2 %>%
  select(sample_id, CGAS, STING1)

# -------------------------------------------------
# 3. Merge into main dataframe
# -------------------------------------------------
clustered_plot_df <- clustered_plot_df %>%
  left_join(sting_subset, by = "sample_id")

# -------------------------------------------------
# 4. (Optional sanity check)
# -------------------------------------------------
stopifnot(
  all(clustered_plot_df$sample_id %in% sting_gas_df2$sample_id)
)

# View result
clustered_plot_df$STING1
```



# Read in Objects from MG

```{r}

reading_path <- "D:/OneDrive - University of Florida/Gbadamosi Lab/Mohammed Gbadamosi/3_Manuscripts/CIMS BC PIpeline/final_cimic_analysis"

library(data.table)
library(dplyr)
library(tidyverse)
library(tibble)

combined_TPM_lognorm <- fread(paste0(reading_path, "/nki_smc_combine_TPM_lognorm.csv")) %>% as.data.frame()
master_induction_df <- fread(paste0(reading_path, "/master_induction_df_nki_smc_k2.csv")) %>% as.data.frame()
clustered_plot_df <- fread(paste0(reading_path, "/clustered_plot_df_nki_smc_k2.csv")) %>% as.data.frame()
formatted_individual_deltas_matrix <- fread(paste0(reading_path, "/nki_smc_combine_initial_clustering_mat.csv")) %>% as.data.frame() %>%
  column_to_rownames("V1") %>%
  as.matrix()
formatted_individual_deltas_df <- as.data.frame(formatted_individual_deltas_matrix )



```

# Get all Gene Sets of Interest
```{r}

library(msigdbr)
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(rstatix)


msig_df <- msigdbr::msigdbr(species = "Homo sapiens") %>% as.data.frame()

    # Default CIM Genesets from manuscript
    all_gene_set_names <- c(
      "GOBP_ADAPTIVE_IMMUNE_RESPONSE",
      "GOBP_ANTIGEN_PROCESSING_AND_PRESENTATION",
      "GOBP_B_CELL_ACTIVATION",
      "GOBP_CELLULAR_RESPONSE_TO_STRESS",
      "GOBP_CYTOKINE_PRODUCTION_INVOLVED_IN_INFLAMMATORY_RESPONSE",
      "GOBP_INFLAMMASOME_MEDIATED_SIGNALING_PATHWAY",
      "GOBP_INFLAMMATORY_CELL_APOPTOTIC_PROCESS",
      "GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY",
      "GOBP_INTRINSIC_APOPTOTIC_SIGNALING_PATHWAY_IN_RESPONSE_TO_ENDOPLASMIC_RETICULUM_STRESS",
      "GOBP_LEUKOCYTE_CHEMOTAXIS_INVOLVED_IN_INFLAMMATORY_RESPONSE",
      "GOBP_LEUKOCYTE_MIGRATION_INVOLVED_IN_INFLAMMATORY_RESPONSE",
      "GOBP_RESPONSE_TO_ENDOPLASMIC_RETICULUM_STRESS",
      "GOBP_T_CELL_ACTIVATION",
      "HALLMARK_INFLAMMATORY_RESPONSE",
      "HALLMARK_UNFOLDED_PROTEIN_RESPONSE",
      "REACTOME_CELLULAR_RESPONSE_TO_CHEMICAL_STRESS",
      "REACTOME_CELLULAR_RESPONSE_TO_MITOCHONDRIAL_STRESS",
      "WP_MOLECULAR_PATHWAY_FOR_OXIDATIVE_STRESS",
      "WP_OXIDATIVE_STRESS_RESPONSE"
    )
    
  # Build a named list: each element is the vector of gene symbols for that set
    all_gene_sets <- setNames(
      lapply(all_gene_set_names, function(gs) {
        gs_df <- msig_df[msig_df$gs_name == gs, , drop = FALSE]
        gs_df$gene_symbol
      }),
      all_gene_set_names

    )
 
 remove_prefixes <- c(
  # B‑cell (heavy, light, constant, joining, diversity, …)
  "IGHV", "IGHD", "IGHJ", "IGHC", "IGHM", "IGHA", "IGHE", "IGHG",
  "IGLV", "IGLC", "IGLJ",
  "IGKV", "IGKC", "IGKJ",
  # T‑cell receptor chains
  "TRAV", "TRBV", "TRDV", "TRGV",
  "TRAJ", "TRBJ", "TRGJ", "TRDJ",
  "TRAC", "TRBC", "TRGC", "TRDC"
)

# Build a single regex that matches any of the prefixes at the start
pattern <- paste0("^(", paste(remove_prefixes, collapse = "|"), ")")

# -----------------------------------------------------------------
# 2. Apply the filter to the *named list* `all_gene_sets`
#    (each element is a character vector of gene symbols)
# -----------------------------------------------------------------


  all_gene_sets <- lapply(all_gene_sets, function(gene_vec) {

  # keep everything else
  gene_vec[!grepl(pattern, gene_vec, perl = TRUE)]
})



  # all_gene_sets is a named list:
#   names(all_gene_sets) → gene‑set IDs (e.g. "GOBP_ADAPTIVE_IMMUNE_RESPONSE")
#   each element          → character vector of gene symbols
# -------------------------------------------------
# 2) Convert the list → long format (one row per gene‑set / gene pair)
# -------------------------------------------------
# 2) Convert the list → long format (one row per gene‑set / gene pair)
gene_set_long <- enframe(
  all_gene_sets,
  name = "gene_set_ID",
  value = "genes"
) %>%
  tidyr::unnest_longer(col = genes) %>%
  dplyr::rename(gene = genes)

# -------------------------------------------------
# 3) Add a position index inside each gene‑set (1,2,3,…)
# -------------------------------------------------
gene_set_long <- gene_set_long %>% 
  group_by(gene_set_ID) %>% 
  mutate(pos = row_number()) %>%    # position of the gene inside the set
  ungroup()

# -------------------------------------------------
# 4) Wide‑reshape: one row per gene‑set, columns ID1, ID2, …
# -------------------------------------------------
gene_set_wide <- gene_set_long %>% 
  pivot_wider(names_from   = pos,
              values_from  = gene,
              names_prefix = "V") %>%      # → ID1, ID2, …
  # optional: keep column order tidy
  dplyr::select(gene_set_ID, starts_with("V"))


all_unique_gene_ids <- unique(unlist(all_gene_sets))



death_programs <- list(
  Apoptosis = c(
    "BAX", "BAK1", "BBC3", "PMAIP1", "BCL2L11",
    "APAF1", "CASP9", "CASP3", "CASP7",
    "FAS", "TNFRSF10B", "CASP8", "FADD", "BID",
    "DIABLO", "CYCS", "TP53AIP1", "BAD", "BMF", "HRK"
  ),
  Necroptosis = c(
    "RIPK1", "RIPK3", "MLKL", "ZBP1", "TICAM1", "TNFRSF1A", "FADD", "CASP8"
  ),
  Pyroptosis = c(
    "GSDME", "GSDMD", "CASP1", "CASP4", "CASP5",
    "NLRP3", "AIM2", "PYCARD", "IL1B", "IL18", "CASP3", "NLRC4"
  ),
  PANoptosis = c(
    "ZBP1", "AIM2", "RIPK3", "RIPK1", "CASP8",
    "CASP1", "FADD", "PYCARD", "IRF1"
  ),
  Ferroptosis = c(
    "ACSL4", "LPCAT3", "TFRC", "SAT1", "PTGS2"
  )
)

all_gene_sets <- c(death_programs, all_gene_sets)
all_gene_set_names_cimic <- all_gene_set_names
all_gene_set_names_death <- names(death_programs)

```



# Make ssGSEA Function

```{r}

# ssGSEA Function
ssgsea = function(X, gene_sets, alpha = 0.25, scale = T, norm = F, single = T) {
    row_names = rownames(X)
    num_genes = nrow(X)
    gene_sets = lapply(gene_sets, function(genes) {which(row_names %in% genes)})

    # Ranks for genes
    R = matrixStats::colRanks(X, preserveShape = T, ties.method = 'average')

    # Calculate enrichment score (es) for each sample (column)
    es = apply(R, 2, function(R_col) {
        gene_ranks = order(R_col, decreasing = TRUE)

        # Calc es for each gene set
        es_sample = sapply(gene_sets, function(gene_set_idx) {
            # pos: match (within the gene set)
            # neg: non-match (outside the gene set)
            indicator_pos = gene_ranks %in% gene_set_idx
            indicator_neg = !indicator_pos

            rank_alpha  = (R_col[gene_ranks] * indicator_pos) ^ alpha

            step_cdf_pos = cumsum(rank_alpha)    / sum(rank_alpha)
            step_cdf_neg = cumsum(indicator_neg) / sum(indicator_neg)

            step_cdf_diff = step_cdf_pos - step_cdf_neg

            # Normalize by gene number
            if (scale) step_cdf_diff = step_cdf_diff / num_genes

            # Use ssGSEA or not
            if (single) {
                sum(step_cdf_diff)
            } else {
                step_cdf_diff[which.max(abs(step_cdf_diff))]
            }
        })
        unlist(es_sample)
    })

    if (length(gene_sets) == 1) es = matrix(es, nrow = 1)

    # Normalize by absolute diff between max and min
    if (norm) es = es / diff(range(es))

    # Prepare output
    rownames(es) = names(gene_sets)
    colnames(es) = colnames(X)
    return(es)
}




```


# Extract out samples and run ssGSEA and Z-score (ALLWAYS REMEMBER TO PIVOT FOR DIFFERENT DATA SETS; NEO

```{r}

nki_smc_ssgsea_mat <- combined_TPM_lognorm %>%
  remove_rownames() %>%
  column_to_rownames("gene_id") %>%
  as.matrix() %>%
  {
    matrix(as.numeric(.), nrow = nrow(.), dimnames = dimnames(.))
  }


# Extract columns with "_B_" in their names (Baseline)
nki_smc_ssgsea_B <- nki_smc_ssgsea_mat[, grepl(
  "_B",
  base::colnames(nki_smc_ssgsea_mat)
)]

# Extract columns with "_S_" in their names (Stimulated/Post)
nki_smc_ssgsea_S <- nki_smc_ssgsea_mat[, grepl(
  "_S",
  base::colnames(nki_smc_ssgsea_mat)
)]

# get results for baseline and surgery
system.time(assign(
  'ssgsea_es_nki_smc_baseline',
  ssgsea(nki_smc_ssgsea_B, all_gene_sets, scale = TRUE, norm = FALSE)
))

# get results for baseline and surgery
system.time(assign(
  'ssgsea_es_nki_smc_surgery',
  ssgsea(nki_smc_ssgsea_S, all_gene_sets, scale = TRUE, norm = FALSE)
))


# Get Deltas
ssgsea_es_nki_smc_delta_pre_post <- ssgsea_es_nki_smc_surgery -
  ssgsea_es_nki_smc_baseline


# # Z Score if needed
# # rows = gene sets, cols = patients; just for visual in this case
# ssgsea_es_nki_smc_delta_zcol <- t(apply(ssgsea_es_nki_smc_delta_pre_post, 2, function(x) {
#   (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
# }))
#
# # Convert to data frame
# ssgsea_es_nki_smc_delta_zcol <- as.data.frame(t(ssgsea_es_nki_smc_delta_zcol))

# Row-wise z-score: standardize each gene set across patients; gives a sort of GSVA in this case
ssgsea_es_nki_smc_delta_zrow <- t(apply(
  ssgsea_es_nki_smc_delta_pre_post,
  1,
  function(x) {
    (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  }
))


# Convert to data frame
ssgsea_es_nki_smc_delta_zrow <- as.data.frame(ssgsea_es_nki_smc_delta_zrow)
ssgsea_es_nki_smc_delta_pre_post <- as.data.frame(
  ssgsea_es_nki_smc_delta_pre_post
)


# Set cluster as factor to preserve order
ssgsea_nki_smc_CIM_res_df_zscore <- ssgsea_es_nki_smc_delta_zrow %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  inner_join(
    clustered_plot_df %>%
      dplyr::select(sample_id, cluster_assignments) %>%
      # This regex removes "delta_" from the start AND "_Rep1" from the end
      # dplyr::mutate(sample_id = gsub("^delta_|_Rep1$", "", sample_id)),
      # old regex
      #dplyr::mutate(sample_id = sub("^delta_", "", sample_id)),
      #for neo
      dplyr::mutate(
        sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
          sub("^delta_", "", .) %>% # remove leading "delta_"
          sub("_delta$", "_T4_S_Rep1", .)), # replace trailing "_delta" with "_S"
        
      # new one
      # dplyr::mutate(
      #   sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
      #     sub("^delta_", "", .) %>% # remove leading "delta_"
      #     sub("_delta$", "_S", .) # replace trailing "_delta" with "_S"
      # ),
    by = "sample_id"
  ) %>%
  dplyr::mutate(
    cluster_assignments = factor(
      cluster_assignments,
      levels = sort(unique(cluster_assignments))
    ),
    CIMIC_Cluster = dplyr::case_when(
      cluster_assignments == 1 ~ "Dys-CIM",
      cluster_assignments == 2 ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )


# No Z socre
ssgsea_nki_smc_CIM_res_df <- ssgsea_es_nki_smc_delta_pre_post %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  inner_join(
    clustered_plot_df %>%
      dplyr::select(sample_id, cluster_assignments) %>%
      # This regex removes "delta_" from the start AND "_Rep1" from the end
      # dplyr::mutate(sample_id = gsub("^delta_|_Rep1$", "", sample_id)),
      # old regex
      #dplyr::mutate(sample_id = sub("^delta_", "", sample_id)),
      
            #for neo
      dplyr::mutate(
        sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
          sub("^delta_", "", .) %>% # remove leading "delta_"
          sub("_delta$", "_T4_S_Rep1", .)), # replace trailing "_delta" with "_S"
        
      # new one
      dplyr::mutate(
        sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
          sub("^delta_", "", .) %>% # remove leading "delta_"
          sub("_delta$", "_S", .) # replace trailing "_delta" with "_S"
      ),
    by = "sample_id"
  ) %>%
  dplyr::mutate(
    cluster_assignments = factor(
      cluster_assignments,
      levels = sort(unique(cluster_assignments))
    ),
    CIMIC_Cluster = dplyr::case_when(
      cluster_assignments == 1 ~ "Dys-CIM",
      cluster_assignments == 2 ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )


# -------------------------------------------------
# 1. ES matrix (pathways × samples)
# -------------------------------------------------
ssgsea_nki_smc_es_matrix <- ssgsea_nki_smc_CIM_res_df_zscore %>%
  dplyr::select(-c(cluster_assignments, CIMIC_Cluster)) %>% # drop helper columns
  tibble::column_to_rownames("sample_id") %>% # rows = samples
  t() %>% # transpose → rows = pathways
  as.matrix()


# -------------------------------------------------
# 1. Convert the ES matrix to a tidy (long) format
# -------------------------------------------------
ssgsea_nki_smc_es_long <- ssgsea_nki_smc_es_matrix %>% # ES matrix: pathways × samples
  as.data.frame() %>%
  mutate(pathway = rownames(.)) %>% # keep pathway name
  pivot_longer(
    cols = -pathway,
    names_to = "sample",
    values_to = "ES"
  ) %>%
  # Join the cluster information (CIMIC_Cluster) for each sample
  left_join(
    ssgsea_nki_smc_CIM_res_df %>% # contains sample_id, CIMIC_Cluster, etc.
      dplyr::select(sample_id, CIMIC_Cluster),
    by = c("sample" = "sample_id")
  )


# -------------------------------------------------
# 2. Convert the matrix to a tidy (long) data frame
# -------------------------------------------------
ssgsea_nki_smc_es_long <- as.data.frame(ssgsea_nki_smc_es_matrix) %>%
  tibble::rownames_to_column("pathway") %>% # keep pathway name
  tidyr::pivot_longer(
    cols = -pathway, # all sample columns
    names_to = "sample",
    values_to = "ES"
  ) %>%
  dplyr::left_join(
    ssgsea_nki_smc_CIM_res_df %>%
      dplyr::select(sample_id, CIMIC_Cluster),
    by = c("sample" = "sample_id")
  )

# -------------------------------------------------
# 2. Compute the mean ES per pathway *by cluster*
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster <- ssgsea_nki_smc_es_long %>%
  group_by(pathway, CIMIC_Cluster) %>%
  summarise(
    avg_ES = mean(ES, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(CIMIC_Cluster = paste0(CIMIC_Cluster, "_mean")) %>% # e.g. "Dys‑CIM_mean"
  pivot_wider(
    names_from = CIMIC_Cluster,
    values_from = avg_ES
  ) %>%
  column_to_rownames("pathway")

# -------------------------------------------------
# 3. How many distinct clusters are present?
# -------------------------------------------------
n_clusters <- length(unique(ssgsea_nki_smc_es_long$CIMIC_Cluster))

# -------------------------------------------------
# 4. Statistical test per pathway (Wilcoxon for 2 clusters,
#    Kruskal‑Wallis otherwise)
# -------------------------------------------------
if (n_clusters == 2) {
  stats <- ssgsea_nki_smc_es_long %>%
    group_by(pathway) %>%
    rstatix::wilcox_test(ES ~ CIMIC_Cluster) %>%
    dplyr::select(pathway, p = p)
} else {
  stats <- ssgsea_nki_smc_es_long %>%
    group_by(pathway) %>%
    rstatix::kruskal_test(ES ~ CIMIC_Cluster) %>%
    dplyr::select(pathway, p = p)
}

# -------------------------------------------------
# 5. Merge the statistics with the mean‑ES table
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster <- ssgsea_nki_smc_avg_es_per_cluster %>%
  rownames_to_column("pathway") %>%
  left_join(stats, by = "pathway") %>%
  column_to_rownames("pathway")

# -------------------------------------------------
# 6. Multiple‑testing correction (FDR)
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster$adj_p <-
  p.adjust(ssgsea_nki_smc_avg_es_per_cluster$p, method = "fdr")


# get dataset merged
ssgsea_nki_smc_CIM_res_df_for_merge <- ssgsea_nki_smc_CIM_res_df_zscore %>% 
  dplyr::mutate(
        sample_id = sample_id %>% 
          # sub("_S$", "_delta", .)
        gsub( "_T4_S_Rep1", "_delta", .) # ####for neo
      )


# Merge
clustered_plot_df_for_merge <- merge(
  x = clustered_plot_df,
  y = ssgsea_nki_smc_CIM_res_df_for_merge,
  by = "sample_id"
)

clustered_plot_df_for_merge$cluster_assignments <- clustered_plot_df_for_merge$cluster_assignments.x

clustered_plot_df_for_merge <- clustered_plot_df_for_merge %>%
  mutate(
    Avg_MHC1 = rowMeans(
      dplyr::select(., `HLA-A`, `HLA-B`, `HLA-C`), # keep only the three HLA columns
      na.rm = TRUE
    )
  )










# Z-score Analysis ------------------------------------------------- 
all_programs <- all_gene_sets
df <- clustered_plot_df_for_merge
gene_names <- combined_TPM_lognorm$gene_id

# -----------------------------
# 3. Keep genes present for programs
# -----------------------------
all_cols <- colnames(df)
all_programs_present <- lapply(all_programs, function(genes) intersect(genes, all_cols))
print(all_programs_present)

# -----------------------------
# 4. Gene-wise z-score across samples
# -----------------------------

gene_cols <- gene_names 




zscore_safe <- function(x) {
  s <- sd(x, na.rm = TRUE)
  m <- mean(x, na.rm = TRUE)
  if (is.na(s) || s == 0) rep(0, length(x)) else (x - m) / s
}

df_z <- df %>%
  mutate(across(all_of(gene_cols), zscore_safe))


# -----------------------------
# 5. Score each program
# -----------------------------
score_program_z <- function(data, genes, program_name) {

  # ensure genes actually exist in data
  genes <- intersect(genes, colnames(data))

  if (length(genes) == 0) {
    return(
      data.frame(
        sample_id = data$sample_id,
        cimic_cluster = data$CIMIC_Cluster,
        Program = paste0(program_name, "_z"),
        Score = rep(NA_real_, nrow(data))
      )
    )
  }
  
  score_vec <- rowMeans(as.matrix(data[, genes, drop = FALSE]), na.rm = TRUE)
  
  data.frame(
    sample_id = data$sample_id,
    cimic_cluster = data$CIMIC_Cluster,
    Program = paste0(program_name, "_z"),
    Score = score_vec
  )
}

program_scores_all <- purrr::imap_dfr(all_programs_present, ~score_program_z(df_z, .x, .y))



library(tidyr)
library(dplyr)

program_scores_wide <- program_scores_all %>%
  dplyr::select(sample_id, Program, Score) %>%
  tidyr::pivot_wider(
    names_from = Program,
    values_from = Score
  )

clustered_plot_df_for_merge <- clustered_plot_df_for_merge %>%
  left_join(program_scores_wide, by = "sample_id")



```



# Extract out samples and run ssGSEA and Z-score (ALLWAYS REMEMBER TO PIVOT FOR DIFFERENT DATA SETS; nki

```{r}

nki_smc_ssgsea_mat <- combined_TPM_lognorm %>%
  remove_rownames() %>%
  column_to_rownames("gene_id") %>%
  as.matrix() %>%
  {
    matrix(as.numeric(.), nrow = nrow(.), dimnames = dimnames(.))
  }


# Extract columns with "_B_" in their names (Baseline)
nki_smc_ssgsea_B <- nki_smc_ssgsea_mat[, grepl(
  "_B",
  base::colnames(nki_smc_ssgsea_mat)
)]

# Extract columns with "_S_" in their names (Stimulated/Post)
nki_smc_ssgsea_S <- nki_smc_ssgsea_mat[, grepl(
  "_S",
  base::colnames(nki_smc_ssgsea_mat)
)]

# get results for baseline and surgery
system.time(assign(
  'ssgsea_es_nki_smc_baseline',
  ssgsea(nki_smc_ssgsea_B, all_gene_sets, scale = TRUE, norm = FALSE)
))

# get results for baseline and surgery
system.time(assign(
  'ssgsea_es_nki_smc_surgery',
  ssgsea(nki_smc_ssgsea_S, all_gene_sets, scale = TRUE, norm = FALSE)
))


# Get Deltas
ssgsea_es_nki_smc_delta_pre_post <- ssgsea_es_nki_smc_surgery -
  ssgsea_es_nki_smc_baseline


# # Z Score if needed
# # rows = gene sets, cols = patients; just for visual in this case
# ssgsea_es_nki_smc_delta_zcol <- t(apply(ssgsea_es_nki_smc_delta_pre_post, 2, function(x) {
#   (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
# }))
#
# # Convert to data frame
# ssgsea_es_nki_smc_delta_zcol <- as.data.frame(t(ssgsea_es_nki_smc_delta_zcol))

# Row-wise z-score: standardize each gene set across patients; gives a sort of GSVA in this case
ssgsea_es_nki_smc_delta_zrow <- t(apply(
  ssgsea_es_nki_smc_delta_pre_post,
  1,
  function(x) {
    (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
  }
))


# Convert to data frame
ssgsea_es_nki_smc_delta_zrow <- as.data.frame(ssgsea_es_nki_smc_delta_zrow)
ssgsea_es_nki_smc_delta_pre_post <- as.data.frame(
  ssgsea_es_nki_smc_delta_pre_post
)


# Set cluster as factor to preserve order
ssgsea_nki_smc_CIM_res_df_zscore <- ssgsea_es_nki_smc_delta_zrow %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  inner_join(
    clustered_plot_df %>%
      dplyr::select(sample_id, cluster_assignments) %>%
      # This regex removes "delta_" from the start AND "_Rep1" from the end
      # dplyr::mutate(sample_id = gsub("^delta_|_Rep1$", "", sample_id)),
      # old regex
      #dplyr::mutate(sample_id = sub("^delta_", "", sample_id)),
      #for neo
      # dplyr::mutate(
      #   sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
      #     sub("^delta_", "", .) %>% # remove leading "delta_"
      #     sub("_delta$", "_T4_S_Rep1", .)), # replace trailing "_delta" with "_S"
      #   
      # new one
      dplyr::mutate(
        sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
          sub("^delta_", "", .) %>% # remove leading "delta_"
          sub("_delta$", "_S", .) # replace trailing "_delta" with "_S"
      ),
    by = "sample_id"
  ) %>%
  dplyr::mutate(
    cluster_assignments = factor(
      cluster_assignments,
      levels = sort(unique(cluster_assignments))
    ),
    CIMIC_Cluster = dplyr::case_when(
      cluster_assignments == 1 ~ "Dys-CIM",
      cluster_assignments == 2 ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )


# No Z socre
ssgsea_nki_smc_CIM_res_df <- ssgsea_es_nki_smc_delta_pre_post %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  inner_join(
    clustered_plot_df %>%
      dplyr::select(sample_id, cluster_assignments) %>%
      # This regex removes "delta_" from the start AND "_Rep1" from the end
      # dplyr::mutate(sample_id = gsub("^delta_|_Rep1$", "", sample_id)),
      # old regex
      #dplyr::mutate(sample_id = sub("^delta_", "", sample_id)),
      
            #for neo
      # dplyr::mutate(
      #   sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
      #     sub("^delta_", "", .) %>% # remove leading "delta_"
      #     sub("_delta$", "_T4_S_Rep1", .)), # replace trailing "_delta" with "_S"
        
      # new one
      dplyr::mutate(
        sample_id = sample_id %>% # original IDs like "delta_2046_S" or "2046_delta"
          sub("^delta_", "", .) %>% # remove leading "delta_"
          sub("_delta$", "_S", .) # replace trailing "_delta" with "_S"
      ),
    by = "sample_id"
  ) %>%
  dplyr::mutate(
    cluster_assignments = factor(
      cluster_assignments,
      levels = sort(unique(cluster_assignments))
    ),
    CIMIC_Cluster = dplyr::case_when(
      cluster_assignments == 1 ~ "Dys-CIM",
      cluster_assignments == 2 ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )


# -------------------------------------------------
# 1. ES matrix (pathways × samples)
# -------------------------------------------------
ssgsea_nki_smc_es_matrix <- ssgsea_nki_smc_CIM_res_df_zscore %>%
  dplyr::select(-c(cluster_assignments, CIMIC_Cluster)) %>% # drop helper columns
  tibble::column_to_rownames("sample_id") %>% # rows = samples
  t() %>% # transpose → rows = pathways
  as.matrix()


# -------------------------------------------------
# 1. Convert the ES matrix to a tidy (long) format
# -------------------------------------------------
ssgsea_nki_smc_es_long <- ssgsea_nki_smc_es_matrix %>% # ES matrix: pathways × samples
  as.data.frame() %>%
  mutate(pathway = rownames(.)) %>% # keep pathway name
  pivot_longer(
    cols = -pathway,
    names_to = "sample",
    values_to = "ES"
  ) %>%
  # Join the cluster information (CIMIC_Cluster) for each sample
  left_join(
    ssgsea_nki_smc_CIM_res_df %>% # contains sample_id, CIMIC_Cluster, etc.
      dplyr::select(sample_id, CIMIC_Cluster),
    by = c("sample" = "sample_id")
  )


# -------------------------------------------------
# 2. Convert the matrix to a tidy (long) data frame
# -------------------------------------------------
ssgsea_nki_smc_es_long <- as.data.frame(ssgsea_nki_smc_es_matrix) %>%
  tibble::rownames_to_column("pathway") %>% # keep pathway name
  tidyr::pivot_longer(
    cols = -pathway, # all sample columns
    names_to = "sample",
    values_to = "ES"
  ) %>%
  dplyr::left_join(
    ssgsea_nki_smc_CIM_res_df %>%
      dplyr::select(sample_id, CIMIC_Cluster),
    by = c("sample" = "sample_id")
  )

# -------------------------------------------------
# 2. Compute the mean ES per pathway *by cluster*
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster <- ssgsea_nki_smc_es_long %>%
  group_by(pathway, CIMIC_Cluster) %>%
  summarise(
    avg_ES = mean(ES, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(CIMIC_Cluster = paste0(CIMIC_Cluster, "_mean")) %>% # e.g. "Dys‑CIM_mean"
  pivot_wider(
    names_from = CIMIC_Cluster,
    values_from = avg_ES
  ) %>%
  column_to_rownames("pathway")

# -------------------------------------------------
# 3. How many distinct clusters are present?
# -------------------------------------------------
n_clusters <- length(unique(ssgsea_nki_smc_es_long$CIMIC_Cluster))

# -------------------------------------------------
# 4. Statistical test per pathway (Wilcoxon for 2 clusters,
#    Kruskal‑Wallis otherwise)
# -------------------------------------------------
if (n_clusters == 2) {
  stats <- ssgsea_nki_smc_es_long %>%
    group_by(pathway) %>%
    rstatix::wilcox_test(ES ~ CIMIC_Cluster) %>%
    dplyr::select(pathway, p = p)
} else {
  stats <- ssgsea_nki_smc_es_long %>%
    group_by(pathway) %>%
    rstatix::kruskal_test(ES ~ CIMIC_Cluster) %>%
    dplyr::select(pathway, p = p)
}

# -------------------------------------------------
# 5. Merge the statistics with the mean‑ES table
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster <- ssgsea_nki_smc_avg_es_per_cluster %>%
  rownames_to_column("pathway") %>%
  left_join(stats, by = "pathway") %>%
  column_to_rownames("pathway")

# -------------------------------------------------
# 6. Multiple‑testing correction (FDR)
# -------------------------------------------------
ssgsea_nki_smc_avg_es_per_cluster$adj_p <-
  p.adjust(ssgsea_nki_smc_avg_es_per_cluster$p, method = "fdr")


# get dataset merged
ssgsea_nki_smc_CIM_res_df_for_merge <- ssgsea_nki_smc_CIM_res_df_zscore %>% 
  dplyr::mutate(
        sample_id = sample_id %>% 
          sub("_S$", "_delta", .)
        # gsub( "_T4_S_Rep1", "_delta", .) # ####for neo
      )


# Merge
clustered_plot_df_for_merge <- merge(
  x = clustered_plot_df,
  y = ssgsea_nki_smc_CIM_res_df_for_merge,
  by = "sample_id"
)

clustered_plot_df_for_merge$cluster_assignments <- clustered_plot_df_for_merge$cluster_assignments.x

clustered_plot_df_for_merge <- clustered_plot_df_for_merge %>%
  mutate(
    Avg_MHC1 = rowMeans(
      dplyr::select(., `HLA-A`, `HLA-B`, `HLA-C`), # keep only the three HLA columns
      na.rm = TRUE
    )
  )










# Z-score Analysis ------------------------------------------------- 
all_programs <- all_gene_sets
df <- clustered_plot_df_for_merge
gene_names <- combined_TPM_lognorm$gene_id

# -----------------------------
# 3. Keep genes present for programs
# -----------------------------
all_cols <- colnames(df)
all_programs_present <- lapply(all_programs, function(genes) intersect(genes, all_cols))
print(all_programs_present)

# -----------------------------
# 4. Gene-wise z-score across samples
# -----------------------------

gene_cols <- gene_names 




zscore_safe <- function(x) {
  s <- sd(x, na.rm = TRUE)
  m <- mean(x, na.rm = TRUE)
  if (is.na(s) || s == 0) rep(0, length(x)) else (x - m) / s
}

df_z <- df %>%
  mutate(across(all_of(gene_cols), zscore_safe))


# -----------------------------
# 5. Score each program
# -----------------------------
score_program_z <- function(data, genes, program_name) {

  # ensure genes actually exist in data
  genes <- intersect(genes, colnames(data))

  if (length(genes) == 0) {
    return(
      data.frame(
        sample_id = data$sample_id,
        cimic_cluster = data$CIMIC_Cluster,
        Program = paste0(program_name, "_z"),
        Score = rep(NA_real_, nrow(data))
      )
    )
  }
  
  score_vec <- rowMeans(as.matrix(data[, genes, drop = FALSE]), na.rm = TRUE)
  
  data.frame(
    sample_id = data$sample_id,
    cimic_cluster = data$CIMIC_Cluster,
    Program = paste0(program_name, "_z"),
    Score = score_vec
  )
}

program_scores_all <- purrr::imap_dfr(all_programs_present, ~score_program_z(df_z, .x, .y))



library(tidyr)
library(dplyr)

program_scores_wide <- program_scores_all %>%
  dplyr::select(sample_id, Program, Score) %>%
  tidyr::pivot_wider(
    names_from = Program,
    values_from = Score
  )

clustered_plot_df_for_merge <- clustered_plot_df_for_merge %>%
  left_join(program_scores_wide, by = "sample_id")



```


# UPDATED VOLC PLOT

```{r}


library(ggplot2)
library(dplyr)
library(ggrepel)

# Define the two clusters of interest and other params
clusters_oi <- c("cluster1", "cluster2")
slope_cutoff <- log2(1.5)
pval_cutoff  <- 0.05


ind_cutoff <- log2(1.50)


penetrance_target_cutoff <- 0.20   # ≥20% induced in target cluster
penetrance_other_cutoff  <- 0.20   # ≤20% induced in other cluster

# Build the column names dynamically
diff_col      <- paste0(clusters_oi[1], "_", clusters_oi[2], "_differential")
diff_col_rev  <- paste0(clusters_oi[2], "_", clusters_oi[1], "_differential")
pval_col      <- "stat_test_adj_p"
mean_col	  <- paste0(clusters_oi[1], "_mean")
mean_col_rev	<- paste0(clusters_oi[2], "_mean")

pen_target <- paste0(clusters_oi[1], "_penetrance")
pen_other  <- paste0(clusters_oi[2], "_penetrance")

C1_class <- "Dys-CIM"
C2_class <- "Fun-CIM"


ind_plot_category_1 <- paste0("Induced in ",       C1_class, " only")
ind_plot_category_2 <- paste0("Induced in ",       C2_class, " only")
ind_plot_category_3 <- paste0("Highly induced in ", C1_class)
ind_plot_category_4 <- paste0("Highly induced in ", C2_class)
ind_plot_category_5 <- paste0("Highly repressed in ", C1_class)
ind_plot_category_6 <- paste0("Highly repressed in ", C2_class)
ind_plot_category_NS <- "NS"



plot_data <- master_induction_df %>%
    dplyr::select(gene_id,
                  C1 = all_of(mean_col),
                  C2 = all_of(mean_col_rev),
                  pval = all_of(pval_col),
                  diff = all_of(diff_col)) %>%
    dplyr::mutate(
        significant = pval < pval_cutoff & abs(diff) > slope_cutoff,
        category = dplyr::case_when(
            !significant                                      ~ ind_plot_category_NS,
            C1 >  0 & C2 <  0                                ~ ind_plot_category_1,
            C1 <  0 & C2 >  0                                ~ ind_plot_category_2,
            C1 >  0 & C2 >  0 & (C1 - C2) > slope_cutoff    ~ ind_plot_category_3,
            C1 >  0 & C2 >  0 & (C2 - C1) > slope_cutoff    ~ ind_plot_category_4,
            C1 <  0 & C2 <  0 & (C2 - C1) > slope_cutoff    ~ ind_plot_category_5,
            C1 <  0 & C2 <  0 & (C1 - C2) > slope_cutoff    ~ ind_plot_category_6,
            TRUE                                              ~ ind_plot_category_NS
        )
    )


category_counts <- plot_data %>%
    dplyr::count(category) %>%
    dplyr::mutate(label_with_n = paste0(category, " (N = ", n, ")"))

# Build a named vector to recode legend labels
n_labels <- setNames(category_counts$label_with_n, category_counts$category)

n_genes <- 15
available_genes <- unique(unlist(all_gene_sets[all_gene_set_names_cimic]))

top_up <- plot_data %>%
    dplyr::filter(gene_id %in% available_genes, significant,
                  category %in% c(ind_plot_category_1, ind_plot_category_3)) %>%
    dplyr::mutate(dist_from_diag = C1 - C2) %>%
    dplyr::arrange(dplyr::desc(dist_from_diag)) %>%
    dplyr::slice_head(n = n_genes) %>%
    dplyr::pull(gene_id)

top_down <- plot_data %>%
    dplyr::filter(gene_id %in% available_genes, significant,
                  category %in% c(ind_plot_category_2, ind_plot_category_4)) %>%
    dplyr::mutate(dist_from_diag = C2 - C1) %>%
    dplyr::arrange(dplyr::desc(dist_from_diag)) %>%
    dplyr::slice_head(n = n_genes) %>%
    dplyr::pull(gene_id)

top_genes <- c(top_up, top_down)


# Add label column to plot_data
plot_data <- plot_data %>%
    dplyr::mutate(label = dplyr::if_else(gene_id %in% top_genes, gene_id, NA_character_))


size_n <- 18



ind_graph <- ggplot(plot_data, aes(x = C1, y = C2)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "black", linewidth = 1.5) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "black", linewidth = 1.5) +
    geom_abline(slope = 1, intercept =  slope_cutoff, linetype = "dashed", color = "steelblue", linewidth = 1.5) +
    geom_abline(slope = 1, intercept = -slope_cutoff, linetype = "dashed", color = "firebrick", , linewidth = 1.5) +
    #geom_abline(slope = 1, intercept = 0, linetype = "solid", color = "grey40", linewidth = 0.5) +
    geom_point(aes(color = category, shape = significant), size = 1.5, alpha = 0.9) +
    ggrepel::geom_label_repel(
        data = subset(plot_data, !is.na(label)),
        aes(label = label),
        size          = 4.5,
        fontface      = "bold",
        color         = "black",
        fill          = "white",
        label.size    = 0.5,
        box.padding   = 0.4,
        point.padding = 0.3,
        segment.color = "black",
        segment.size  = 0.4,
        max.overlaps  = Inf
    ) +
    scale_shape_manual(
        name   = "Significance",
        values = c("TRUE" = 16, "FALSE" = 17),
        labels = c("TRUE" = "p < 0.05", "FALSE" = "NS")
    ) +
    scale_color_manual(
        name = "Category",
        values = setNames(
            c("black", "#E41A1C", "#377EB8", "#FF7F00", "#984EA3", "#4DAF4A", "#A65628"),
            c(ind_plot_category_NS, ind_plot_category_1, ind_plot_category_2,
              ind_plot_category_3,  ind_plot_category_4, ind_plot_category_5,
              ind_plot_category_6)
        ),
        labels = n_labels
    ) +
    labs(
        x = paste0("Dys-CIM mean induction\n(Δlog2[TPM+1])"),
        y = paste0("Fun-CIM mean induction\n(Δlog2[TPM+1])")
    ) +
    ggplot2::coord_cartesian(xlim = c(-3, 3), ylim = c(-3, 3)) +
    ggplot2::scale_x_continuous(breaks = seq(-3, 3, 1)) +
    ggplot2::scale_y_continuous(breaks = seq(-3, 3, 1)) +
    theme_classic(base_size = 14) +
    theme(
        axis.title.x = element_text(face = "bold", size = size_n, color = "black"),
        axis.title.y = element_text(face = "bold", size = size_n, color = "black"),
        axis.text.x  = element_text(face = "bold", size = size_n, color = "black"),
        axis.text.y  = element_text(face = "bold", size = size_n, color = "black"),
        legend.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1.0),
        axis.line    = element_blank()
    )

print(ind_graph)

# 
# ind_graph <- ind_graph +
#   annotate("label",
#     x = 2.5,  y = -2.5, label = ind_plot_category_1,
#     fontface = "bold", size = 3.5, fill = "#E41A1C", color = "white", alpha = 0.8
#   ) +
#   annotate("label",
#     x = -2.5, y = 2.5,  label = ind_plot_category_2,
#     fontface = "bold", size = 3.5, fill = "#377EB8", color = "white", alpha = 0.8
#   ) +
#   annotate("label",
#     x = 2.5,  y = 1.5,  label = ind_plot_category_3,
#     fontface = "bold", size = 3.5, fill = "#FF7F00", color = "white", alpha = 0.8
#   ) +
#   annotate("label",
#     x = -2.5, y = -1.5, label = ind_plot_category_4,
#     fontface = "bold", size = 3.5, fill = "#984EA3", color = "white", alpha = 0.8
#   ) +
#   annotate("label",
#     x = -2.5, y = -2.5, label = ind_plot_category_5,
#     fontface = "bold", size = 3.5, fill = "#4DAF4A", color = "white", alpha = 0.8
#   ) +
#   annotate("label",
#     x = 2.5,  y = 2.5,  label = ind_plot_category_6,
#     fontface = "bold", size = 3.5, fill = "#A65628", color = "white", alpha = 0.8
#   )

induced_C1_only <- plot_data %>%
    dplyr::filter(category == ind_plot_category_1)

induced_C2_only <- plot_data %>%
    dplyr::filter(category == ind_plot_category_2)
```

# Heatmap of genesets by cimic cluster

```{r}
library(dplyr)
library(tibble)
library(stringr)
library(ComplexHeatmap)
library(circlize)
library(grid)

# -------------------------------------------------
# 1. Prepare sample-level matrix
#    rows = gene sets, cols = samples
# -------------------------------------------------
genesets_of_interest <- all_gene_set_names

heatmap_df <- clustered_plot_df_for_merge %>%
  dplyr::select(sample_id, CIMIC_Cluster, all_of(genesets_of_interest)) %>%
  dplyr::mutate(
    CIMIC_Cluster = factor(CIMIC_Cluster, levels = c("Dys-CIM", "Fun-CIM"))
  ) %>%
  dplyr::arrange(CIMIC_Cluster)

ssgsea_heatmap_matrix <- heatmap_df %>%
  dplyr::select(-CIMIC_Cluster) %>%
  tibble::column_to_rownames("sample_id") %>%
  as.matrix() %>%
  t()

# optional: clean row names
rownames(ssgsea_heatmap_matrix) <- rownames(ssgsea_heatmap_matrix) %>%
  str_remove_all("GOBP_|REACTOME_|HALLMARK_") %>%
  str_replace_all("_", " ")

# -------------------------------------------------
# 2. Column annotation
# -------------------------------------------------
cluster_colours <- c("Dys-CIM" = "#D55E00", "Fun-CIM" = "#0072B2")

top_ha <- HeatmapAnnotation(
  CIMIC_Cluster = heatmap_df$CIMIC_Cluster,
  col = list(CIMIC_Cluster = cluster_colours),
  show_annotation_name = TRUE
)

# -------------------------------------------------
# 3. Color function
# -------------------------------------------------
col_fun <- circlize::colorRamp2(
  c(-2, 0, 2),
  c("navy", "white", "firebrick3")
)

# -------------------------------------------------
# 4. Draw heatmap
# -------------------------------------------------
ht <- Heatmap(
  ssgsea_heatmap_matrix,
  name = "ssGSEA ES",
  top_annotation = top_ha,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  col = col_fun,
  column_split = heatmap_df$CIMIC_Cluster,
  rect_gp = gpar(col = "black", lwd = 1),
  border = TRUE,
  column_gap = unit(3, "mm"),
  width = unit(16, "cm"),
  height = unit(12, "cm"),
  show_heatmap_legend = TRUE,
  heatmap_legend_param = list(title = "ES"),
  row_names_max_width = unit(8, "cm"),
  row_names_gp = gpar(fontsize = 10)
)

draw(ht)

```


# Averaged Heatmap ssGSEA 

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)


str_wrap_length = 40

# -----------------------------
# Averaged Heatmap with stars above columns
# -----------------------------
genesets_of_interest <- all_gene_set_names


cluster_heatmap_df <- clustered_plot_df_for_merge %>%
  dplyr::select(CIMIC_Cluster, all_of(genesets_of_interest)) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "Program",
    values_to = "Score"
  ) %>%
  mutate(
    Score = as.numeric(Score),
    #CIMIC_Cluster = factor(CIMIC_Cluster, levels = c("Dys-CIM", "Fun-CIM")),
    Program = Program %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = str_wrap_length)
  ) %>%
  group_by(CIMIC_Cluster, Program) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop")


# -----------------------------
# 2. P-VALUE CALCULATION (ADAPTIVE)
# -----------------------------
pval_df <- clustered_plot_df_for_merge %>%
  dplyr::select(CIMIC_Cluster, all_of(genesets_of_interest)) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "GeneSet",
    values_to = "Score"
  ) %>%
  mutate(CIMIC_Cluster = factor(CIMIC_Cluster)) %>%
  group_by(GeneSet) %>%
  summarise(
    n_groups = n_distinct(CIMIC_Cluster),
    
    p = tryCatch({
      if (n_groups == 2) {
        wilcox.test(Score ~ CIMIC_Cluster)$p.value
      } else {
        kruskal.test(Score ~ CIMIC_Cluster)$p.value
      }
    }, error = function(e) NA_real_),
    
    .groups = "drop"
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    p.adj.signif = case_when(
      is.na(p.adj) ~ "NA",
      p.adj <= 0.0001 ~ "****",
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01 ~ "**",
      p.adj <= 0.05 ~ "*",
      TRUE ~ ""
    )
  )


# -----------------------------
# 3. SIGNIFICANCE LABEL TABLE
# -----------------------------
sig_df <- pval_df %>%
  mutate(
    Program = GeneSet %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = str_wrap_length),
    signif_label = p.adj.signif
  ) %>%
  dplyr::select(Program, signif_label)



program_order <- cluster_heatmap_df %>%
  distinct(Program) %>%
  pull(Program)

cluster_heatmap_df <- cluster_heatmap_df %>%
  mutate(Program = factor(Program, levels = rev(program_order)))

sig_df <- sig_df %>%
  mutate(Program = factor(Program, levels = rev(program_order)))

star_df <- star_df <- cluster_heatmap_df %>%
  distinct(Program) %>%
  left_join(sig_df, by = "Program") %>%
  mutate(x = 2.6)


p_cluster_heatmap <- ggplot(
  cluster_heatmap_df,
  aes(x = CIMIC_Cluster, y = Program, fill = mean_score)
) +
  geom_tile(color = "black", linewidth = 0.8, width = 1) +
  annotate(
    "text",
    x = 2.85,
    y = seq_along(levels(cluster_heatmap_df$Program)),
    label = rev(star_df$signif_label),
    size = 20,
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean Δ Z-score ssGSEA"
  ) +
  scale_x_discrete(
    limits = c("Dys-CIM", "Fun-CIM"),
    labels = c("Dys-CIM" = "Dys\nCIM", "Fun-CIM" = "Fun\nCIM"),
    expand = expansion(add = c(0.3, 1.2))
  ) +
  # scale_y_discrete(position = "right") +
  labs(
    x = NULL,
    y = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_classic(base_size = 24) +
  theme(
    axis.text.x = element_text(face = "bold", size = 45),
    axis.text.y = element_text(
      face = "bold",
      size = 28,
      margin = margin(r = 0)
    ),
    legend.text = element_text(face = "bold", size = 24),
    axis.line = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20)
  )

p_cluster_heatmap
```




# Boxplots of genesets ssGSEA  

```{r}
library(ggplot2)
library(dplyr)
library(stringr)
library(ggpubr)
library(tidyr)
library(stringr) # <-- provides str_wrap()
library(rstatix)


# Gene sets of interest TOGGLE here 
genesets_of_interest <- all_gene_set_names


clustered_plot_df_for_merge$CIMIC_Cluster <- factor(
  clustered_plot_df_for_merge$CIMIC_Cluster,
  levels = c("Dys-CIM", "Fun-CIM")
)


# Prepare data
plot_df <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "GeneSet", values_to = "ES") %>%
  mutate(
    ES = as.numeric(ES),
    GeneSet = str_remove_all(GeneSet, "GOBP_|REACTOME_|HALLMARK_") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = 25)
  )

# Compute max per GeneSet for stat_compare_means positioning
y_positions <- plot_df %>%
  group_by(GeneSet) %>%
  summarise(y_pos = max(ES, na.rm = TRUE) + 0.1)

# Generate all pairwise comparisons
comparisons <- combn(unique(plot_df$CIMIC_Cluster), 2, simplify = FALSE)

# Faceted boxplot with stat_compare_means

# Faceted boxplot with stat_compare_means
ggplot(plot_df, aes(x = CIMIC_Cluster, y = ES, fill = CIMIC_Cluster)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, lwd = 1.2) +
  geom_jitter(width = 0.2, size = 0.75, alpha = 0.7,) +
  facet_wrap(~GeneSet, scales = "free_y", ncol = 4) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    size = 15,
    fontface = "bold",
    hide.ns = TRUE,
    vjust = 1.2
  ) +
  labs(
    x = "Cluster",
    y = "Z-Score of ΔssGSEA"
    #title = "Gene Set Enrichment Across Subtypes with Significance"
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 18),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_text(colour = "black", size = 24),
    axis.title.y = element_text(colour = "black", size = 24),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )


library(dplyr)
library(tidyr)
library(rstatix)

# Gene sets of interest

# Prepare long dataframe (same as plotting)
plot_df <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "GeneSet", values_to = "ES") %>%
  mutate(ES = as.numeric(ES))

# Perform pairwise Wilcoxon tests per GeneSet
pval_table <- plot_df %>%
  group_by(GeneSet) %>%
  wilcox_test(ES ~ CIMIC_Cluster) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  ungroup()

# Clean gene set labels (optional, matching your plot)
pval_table <- pval_table %>%
  mutate(
    GeneSet = str_remove_all(GeneSet, "GOBP_|REACTOME_|HALLMARK_") %>%
      str_replace_all("_", " ")
  )

# Save object for downstream use
boxplot_pvalues <- pval_table

boxplot_pvalues


# -------------------------------------------------
# 1. Choose the gene‑set you want to plot
# -------------------------------------------------
# Use any name that is present in `all_gene_set_names_death`
# (e.g. "Apoptosis", "Ferroptosis", "Necroptosis","Apoptosis"   "Necroptosis" "Pyroptosis"  "PANoptosis"  "Ferroptosis")
single_gene_set <- "GOBP_INFLAMMATORY_CELL_APOPTOTIC_PROCESS"

single_gene_set_label <- single_gene_set %>%
  str_remove_all("GOBP_|REACTOME_|HALLMARK_") %>%
  str_replace_all("_", " ") %>%
  str_wrap(width = 25)

# -------------------------------------------------
# 2. Prepare a tidy data frame for that gene-set only
# -------------------------------------------------
plot_df_one <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", single_gene_set))) %>%
  pivot_longer(
    -CIMIC_Cluster,
    names_to = "GeneSet",
    values_to = "ES"
  ) %>%
  mutate(
    ES = as.numeric(ES),
    GeneSet = single_gene_set_label
  )

# -------------------------------------------------
# 3. Compute y‑position for the significance label
# -------------------------------------------------
y_pos_one <- max(plot_df_one$ES, na.rm = TRUE) + 0.1

# -------------------------------------------------
# 4. Plot – single panel, Wilcoxon annotation
# -------------------------------------------------
ggplot(plot_df_one,
       aes(x = CIMIC_Cluster, y = ES, fill = CIMIC_Cluster)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, lwd = 2.25, width = 0.75) +
  geom_jitter(width = 0.2, size = 2.5, alpha = 0.7) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    size = 30,
    fontface = "bold",
    hide.ns = TRUE,
    vjust = 1.2
  ) +
  labs(
    x = "Cluster",
    y = "Z‑Score of ΔssGSEA",
    title = single_gene_set_label
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(face = "bold", colour = "black", size = 45),
    axis.text.y = element_text(face = "bold", colour = "black", size = 45),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold", colour = "black", size = 45),
    axis.title = element_text(colour = "black", size = 45),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 45),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )


```

# Boxplots for cell death programs ssGSEA 

```{r}



library(ggplot2)
library(dplyr)
library(stringr)
library(ggpubr)
library(tidyr)
library(stringr) # <-- provides str_wrap()
library(rstatix)


# Gene sets of interest TOGGLE here 
genesets_of_interest <- all_gene_set_names_death 


clustered_plot_df_for_merge$CIMIC_Cluster <- factor(
  clustered_plot_df_for_merge$CIMIC_Cluster,
  levels = c("Dys-CIM", "Fun-CIM")
)


# Prepare data
plot_df <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "GeneSet", values_to = "ES") %>%
  mutate(
    ES = as.numeric(ES),
    GeneSet = str_remove_all(GeneSet, "GOBP_|REACTOME_|HALLMARK_") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = 25)
  ) %>% as.data.frame()

# Compute max per GeneSet for stat_compare_means positioning
y_positions <- plot_df %>%
  group_by(GeneSet) %>%
  summarise(y_pos = max(ES, na.rm = TRUE) + 0.1)

# Generate all pairwise comparisons
comparisons <- combn(unique(plot_df$CIMIC_Cluster), 2, simplify = FALSE)

# Faceted boxplot with stat_compare_means

# Faceted boxplot with stat_compare_means
ggplot(plot_df, aes(x = CIMIC_Cluster, y = ES, fill = CIMIC_Cluster)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, lwd = 1.2) +
  geom_jitter(width = 0.2, size = 0.75, alpha = 0.7,) +
  facet_wrap(~GeneSet, scales = "free_y", ncol = 4) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    size = 15,
    fontface = "bold",
    hide.ns = TRUE,
    vjust = 1.2
  ) +
  labs(
    x = "Cluster",
    y = "Z-Score of ΔssGSEA"
    #title = "Gene Set Enrichment Across Subtypes with Significance"
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 18),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_text(colour = "black", size = 24),
    axis.title.y = element_text(colour = "black", size = 24),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )


library(dplyr)
library(tidyr)
library(rstatix)

# Gene sets of interest

# Prepare long dataframe (same as plotting)
plot_df <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "GeneSet", values_to = "ES") %>%
  mutate(ES = as.numeric(ES)) %>% as.data.frame()

# Perform pairwise Wilcoxon tests per GeneSet
pval_table <- plot_df %>%
  group_by(GeneSet) %>%
  wilcox_test(ES ~ CIMIC_Cluster) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance("p.adj") %>%
  ungroup()

# Clean gene set labels (optional, matching your plot)
pval_table <- pval_table %>%
  mutate(
    GeneSet = str_remove_all(GeneSet, "GOBP_|REACTOME_|HALLMARK_") %>%
      str_replace_all("_", " ")
  )

# Save object for downstream use
boxplot_pvalues <- pval_table

boxplot_pvalues

# Individual box

library(ggplot2)
library(dplyr)
library(stringr)
library(ggpubr)
library(tidyr)
library(rstatix)

# -------------------------------------------------
# 1. Choose the gene‑set you want to plot
# -------------------------------------------------
# Use any name that is present in `all_gene_set_names_death`
# (e.g. "Apoptosis", "Ferroptosis", "Necroptosis","Apoptosis"   "Necroptosis" "Pyroptosis"  "PANoptosis"  "Ferroptosis")
single_gene_set <- "Pyroptosis"

# -------------------------------------------------
# 2. Prepare a tidy data frame for that gene‑set only
# -------------------------------------------------
plot_df_one <- clustered_plot_df_for_merge %>%
  select(all_of(c("CIMIC_Cluster", single_gene_set))) %>%      # keep cluster + the chosen column
  pivot_longer(-CIMIC_Cluster,
               names_to = "GeneSet",
               values_to = "ES") %>%
  mutate(
    ES = as.numeric(ES),
    # clean the label to match the style used in the faceted plots
    GeneSet = str_remove_all(GeneSet, "GOBP_|REACTOME_|HALLMARK_") %>%
              str_replace_all("_", " ")
  ) %>% as.data.frame()

# -------------------------------------------------
# 3. Compute y‑position for the significance label
# -------------------------------------------------
y_pos_one <- max(plot_df_one$ES, na.rm = TRUE) + 0.1

# -------------------------------------------------
# 4. Plot – single panel, Wilcoxon annotation
# -------------------------------------------------
ggplot(plot_df_one,
       aes(x = CIMIC_Cluster, y = ES, fill = CIMIC_Cluster)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, lwd = 2.25, width = 0.75) +
  geom_jitter(width = 0.2, size = 2.5, alpha = 0.7) +
  stat_compare_means(
    method = "wilcox.test",
    label = "p.signif",
    size = 30,
    fontface = "bold",
    hide.ns = TRUE,
    vjust = 1.2
  ) +
  labs(
    x = "Cluster",
    y = "Z‑Score of ΔssGSEA",
    title = single_gene_set
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(face = "bold", colour = "black", size = 45),
    axis.text.y = element_text(face = "bold", colour = "black", size = 45),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(face = "bold", colour = "black", size = 45),
    axis.title = element_text(colour = "black", size = 45),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "none",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 45),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )

```







# Boxplots using Z-score 

```{r}

library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)



# -----------------------------
# 6. Summary
# -----------------------------
library(dplyr)

# 1. Summary stats per cluster
program_summary <- program_scores_all %>%
  mutate(cimic_cluster = factor(cimic_cluster)) %>%
  group_by(cimic_cluster, Program) %>%
  summarise(
    mean_score   = mean(Score, na.rm = TRUE),
    median_score = median(Score, na.rm = TRUE),
    sd_score     = sd(Score, na.rm = TRUE),
    n            = dplyr::n(),
    .groups = "drop"
  )

# 2. P-values per program (handles k=2 or k>=3)
program_pvals <- program_scores_all %>%
  mutate(cimic_cluster = factor(cimic_cluster)) %>%
  #dplyr::filter(Program %in% paste0(all_gene_set_names_death, "_z")) %>%
  group_by(Program) %>%
  summarise(
    n_clusters = n_distinct(cimic_cluster),
    
    p_value = tryCatch({
      if (n_clusters == 2) {
        wilcox.test(Score ~ cimic_cluster)$p.value
      } else if (n_clusters > 2) {
        kruskal.test(Score ~ cimic_cluster)$p.value
      } else {
        NA_real_
      }
    }, error = function(e) NA_real_),
    
    .groups = "drop"
  ) %>%
  mutate(
    FDR = p.adjust(p_value, method = "fdr")
  )

# 3. Merge
program_summary_final <- program_summary %>%
  left_join(program_pvals, by = "Program")


# -----------------------------
# 7. Faceted boxplot by cluster
# -----------------------------

names_of_int_z_Score <- c(all_gene_set_names_death)

program_scores_plot_df <- program_scores_all %>%
  
  dplyr::filter(Program %in% paste0(names_of_int_z_Score, "_z")) %>%
  
  mutate(
    cimic_cluster = factor(cimic_cluster),
    
    # clean labels FIRST (as character)
    Program = Program %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
      str_replace_all("_", " "),
    
    # then convert to factor with correct ordering
    Program = factor(
      Program,
      levels = unique(
        paste0(names_of_int_z_Score, "_z") %>%
          str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
          str_replace_all("_", " ")
      )
    )
  )

pval_df_all <- program_scores_plot_df %>%
  group_by(Program) %>%
  summarise(
    n_groups = n_distinct(cimic_cluster),
    
    y.position = max(Score, na.rm = TRUE) * 1.1,
    
    pval = tryCatch({
      if (n_groups == 2) {
        wilcox.test(Score ~ cimic_cluster)$p.value
      } else if (n_groups > 2) {
        kruskal.test(Score ~ cimic_cluster)$p.value
      } else {
        NA_real_
      }
    }, error = function(e) NA_real_),
    
    p_label = ifelse(is.na(pval), "p = NA",
                     ifelse(pval < 0.001, "p < 0.001",
                            paste0("p = ", signif(pval, 3)))),
    
    .groups = "drop"
  ) %>%
  mutate(
    FDR = p.adjust(pval, method = "fdr"),
    
    fdr_label = ifelse(is.na(FDR), "FDR = NA",
                        ifelse(FDR < 0.001, "FDR < 0.001",
                               paste0("FDR = ", signif(FDR, 3))))
  ) %>% add_significance("FDR") %>%
add_significance("pval")


p_all_programs_p <- ggplot(program_scores_plot_df, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 1) +
  geom_text(
    data = pval_df_all,
    aes(x = 1.5, y = y.position, label = FDR.signif),
    inherit.aes = FALSE,
    size = 12,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  #scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    title = "Program scores by CIMiC cluster",
    y = "Z-score of\nProgram Activation",
    x = NULL
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 18),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_text(colour = "black", size = 24),
    axis.title.y = element_text(colour = "black", size = 24),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )



p_all_programs_p

# Keep only significant programs
sig_programs <- pval_df_all %>%
  filter(!is.na(FDR), FDR < 0.05) %>%
  pull(Program)

program_scores_plot_df_sig <- program_scores_plot_df %>%
  filter(Program %in% sig_programs)

pval_df_sig <- pval_df_all %>%
  filter(Program %in% sig_programs)

p_all_programs_sig <- ggplot(program_scores_plot_df_sig, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 1) +
  geom_text(
    data = pval_df_sig,
    aes(x = 1.5, y = y.position, label = FDR.signif),
    inherit.aes = FALSE,
    size = 12,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  #scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    title = "Significant program scores by CIMiC cluster",
    y = "Z-score of\nProgram Activation",
    x = NULL
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 18),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_text(colour = "black", size = 24),
    axis.title.y = element_text(colour = "black", size = 24),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "top",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )

p_all_programs_sig



# -------------------------------------------------
# 1. Choose the gene‑set you want to plot
# -------------------------------------------------
# Use any name that is present in `all_gene_set_names_death`
# (e.g. "Apoptosis", "Ferroptosis", "Necroptosis","Apoptosis"   "Necroptosis" "Pyroptosis"  "PANoptosis"  "Ferroptosis")
single_gene_set <- "Ferroptosis"

# -------------------------------------------------
# 2. Prepare a tidy data frame for that gene‑set only
# -------------------------------------------------
program_scores_plot_df_sig <- program_scores_plot_df %>%
  filter(Program %in% single_gene_set) %>% mutate(Program = stringr::str_wrap(Program, width = 20))

pval_df_sig <- pval_df_all %>%
  filter(Program %in% single_gene_set) %>% mutate(Program = stringr::str_wrap(Program, width = 20))

# -------------------------------------------------
# 3. Compute y‑position for the significance label
# -------------------------------------------------
y_pos_one <- max(program_scores_plot_df_sig$Score, na.rm = TRUE) + 0.1

# -------------------------------------------------
# 4. Plot – single panel, Wilcoxon annotation
# -------------------------------------------------
p_program_individual <- ggplot(program_scores_plot_df_sig, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 1) +
  geom_text(
    data = pval_df_sig,
    aes(x = 1.5, y = y_pos_one, label = FDR.signif),
    inherit.aes = FALSE,
    size = 15,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  #scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    y = "Z-score of\nProgram Activation",
    x = NULL
  ) +
  theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 32),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_text(colour = "black", size = 24, face = "bold"), 
    axis.text.y = element_text(colour = "black", size = 24, face = "bold"),
    axis.title.y = element_text(colour = "black", size = 24, face = "bold"),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "none",
    #plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )


p_program_individual



```


# Averaged Heatmap Z-score 

```{r}

str_wrap_length = 40

# -----------------------------
# Averaged Heatmap with stars above columns
# -----------------------------
genesets_of_interest <- paste0(all_gene_set_names, "_z")

library(dplyr)
library(tidyr)
library(stringr)

# -----------------------------
# 1. HEATMAP SUMMARY (UNCHANGED)
# -----------------------------
cluster_heatmap_df <- clustered_plot_df_for_merge %>%
  dplyr::select(CIMIC_Cluster, all_of(genesets_of_interest)) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "Program",
    values_to = "Score"
  ) %>%
  mutate(
    Score = as.numeric(Score),
    CIMIC_Cluster = factor(CIMIC_Cluster),
    Program = Program %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = str_wrap_length)
  ) %>%
  group_by(CIMIC_Cluster, Program) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop")


# -----------------------------
# 2. P-VALUE CALCULATION (ADAPTIVE)
# -----------------------------
pval_df <- clustered_plot_df_for_merge %>%
  dplyr::select(CIMIC_Cluster, all_of(genesets_of_interest)) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "GeneSet",
    values_to = "Score"
  ) %>%
  mutate(CIMIC_Cluster = factor(CIMIC_Cluster)) %>%
  group_by(GeneSet) %>%
  summarise(
    n_groups = n_distinct(CIMIC_Cluster),
    
    p = tryCatch({
      if (n_groups == 2) {
        wilcox.test(Score ~ CIMIC_Cluster)$p.value
      } else {
        kruskal.test(Score ~ CIMIC_Cluster)$p.value
      }
    }, error = function(e) NA_real_),
    
    .groups = "drop"
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    p.adj.signif = case_when(
      is.na(p.adj) ~ "NA",
      p.adj <= 0.0001 ~ "****",
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01 ~ "**",
      p.adj <= 0.05 ~ "*",
      TRUE ~ ""
    )
  )


# -----------------------------
# 3. SIGNIFICANCE LABEL TABLE
# -----------------------------
sig_df <- pval_df %>%
  mutate(
    Program = GeneSet %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
      str_replace_all("_", " ") %>%
      str_wrap(width = str_wrap_length),
    signif_label = p.adj.signif
  ) %>%
  dplyr::select(Program, signif_label)




program_order <- cluster_heatmap_df %>%
  distinct(Program) %>%
  pull(Program)

cluster_heatmap_df <- cluster_heatmap_df %>%
  mutate(Program = factor(Program, levels = rev(program_order)))

sig_df <- sig_df %>%
  mutate(Program = factor(Program, levels = rev(program_order)))

star_df <- star_df <- cluster_heatmap_df %>%
  distinct(Program) %>%
  left_join(sig_df, by = "Program") %>%
  mutate(x = 2.6)


p_cluster_heatmap <- ggplot(
  cluster_heatmap_df,
  aes(x = CIMIC_Cluster, y = Program, fill = mean_score)
) +
  geom_tile(color = "black", linewidth = 0.8, width = 1) +
  annotate(
    "text",
    x = 2.85,
    y = seq_along(levels(cluster_heatmap_df$Program)),
    label = rev(star_df$signif_label),
    size = 20,
    fontface = "bold"
  ) +
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean Δ Z-score ssGSEA"
  ) +
  scale_x_discrete(
    limits = c("Dys-CIM", "Fun-CIM"),
    labels = c("Dys-CIM" = "Dys\nCIM", "Fun-CIM" = "Fun\nCIM"),
    expand = expansion(add = c(0.3, 1.2))
  ) +
  # scale_y_discrete(position = "right") +
  labs(
    x = NULL,
    y = NULL
  ) +
  coord_cartesian(clip = "off") +
  theme_classic(base_size = 24) +
  theme(
    axis.text.x = element_text(face = "bold", size = 45),
    axis.text.y = element_text(
      face = "bold",
      size = 28,
      margin = margin(r = 0)
    ),
    legend.text = element_text(face = "bold", size = 24),
    axis.line = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 10, r = 20, b = 10, l = 20)
  )

p_cluster_heatmap


```

# Analyzing individual genes induction

```{r}

## -------------------------------------------------
## 0. Packages -------------------------------------------------
## -------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)      # ggarrange / annotate_figure
library(rstatix)     # wilcox_test, kruskal_test, etc.
library(stringr)     # str_wrap
library(purrr)




compare_induction_boxplot <- function(df, gene) {
  # -------------------------------------------------
  # 1. Locate the column that actually holds the gene values
  #    (covers PDCD1, PDCD1.x, PDCD1.y, etc.)
  # -------------------------------------------------
  gene_col <- df %>%
    dplyr::select(dplyr::matches(paste0("^", gene, "(\\.x|\\.y)?$"))) %>%
    names()
  if (length(gene_col) == 0) {
    stop("Gene column '", gene, "' not found in the data frame")
  }
  gene_col <- gene_col[[1]]   # use the first match

  # -------------------------------------------------
  # 2. Build a tidy data frame with a numeric expression column
  # -------------------------------------------------
  df2 <- df %>%
    dplyr::mutate(
      Expression = as.numeric(as.character(.data[[gene_col]]))
    ) %>%
    dplyr::select(CIMIC_Cluster, Expression)   # keep only what we need

  # -------------------------------------------------
  # 3. Run the Wilcoxon rank‑sum test (rstatix) and keep the p‑value
  # -------------------------------------------------
  pval <- df2 %>%
    rstatix::wilcox_test(Expression ~ CIMIC_Cluster) %>%
    dplyr::pull(p)

  ptxt <- paste0("P = ", format.pval(pval, digits = 4))

  # -------------------------------------------------
  # 4. Plot (use `annotate()` for the single p‑value label)
  # -------------------------------------------------
  ggplot(df2,
         aes(x = CIMIC_Cluster,
             y = Expression,
             fill = CIMIC_Cluster)) +
    geom_boxplot(width = 0.6,
                 outlier.shape = NA,
                 alpha = 0.8) +
    geom_jitter(width = 0.15,
                size = 1.5,
                alpha = 0.4) +
    annotate("text",
             x = 1.5,
             y = max(df2$Expression, na.rm = TRUE) * 1.1,
             label = ptxt,
             size = 8,
             fontface = "bold") +
    scale_fill_manual(values = c("Dys-CIM" = "#2980B9",
                                 "Fun-CIM" = "#C0392B")) +
    labs(title = paste0(gene),
         y = "Δ(log2[TPM+1])",
         x = NULL) +
    theme_classic(base_size = 14) +
    theme(
      plot.title      = element_text(hjust = 0.5,
                                     face = "bold",
                                     size = 16),
      legend.position = "none",
      axis.text.x     = element_text(face = "bold", size = 20),
      axis.text.y     = element_text(face = "bold", size = 20),
      axis.title.y    = element_text(face = "bold", size = 20),
      panel.border    = element_rect(colour = "black",
                                     fill = NA,
                                     linewidth = 1.5),
      axis.line       = element_line(colour = "black",
                                     linewidth = 1.5)
    )
}




genes_of_interest <- c("PDCD1", "CD274", "IRF1")
ind_plots <- purrr::map(genes_of_interest,
                    ~ compare_induction_boxplot(clustered_plot_df_for_merge, .x))




```



# Analyze indivdual gene penetrance 
```{r}



## -------------------------------------------------
## 0. Packages -------------------------------------------------
## -------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)      # ggarrange / annotate_figure
library(rstatix)     # wilcox_test, kruskal_test, etc.
library(stringr)     # str_wrap
library(purrr)



# -------------------------------------------------
# 1. Penetrance (percentage of positive induction) -----------------
# -------------------------------------------------
compare_penetrance_plot <- function(df, gene) {
  # -------------------------------------------------
  # 1.1  Find the exact column that holds the gene values
  # -------------------------------------------------
  gene_col <- df %>%
    dplyr::select(dplyr::matches(paste0("^", gene, "(\\.x|\\.y)?$"))) %>%
    names()
  if (length(gene_col) == 0) {
    stop("Gene column '", gene, "' not found in the data frame")
  }
  gene_col <- gene_col[[1]]

  # -------------------------------------------------
  # 1.2  Build a tidy frame with a binary “Induction” flag
  # -------------------------------------------------
  df2 <- df %>%
    dplyr::mutate(
      Expression = as.numeric(as.character(.data[[gene_col]])),
      Induction  = ifelse(Expression > 0,
                          "Positive Induction",
                          "Negative/No Induction")
    ) %>%
    dplyr::select(CIMIC_Cluster, Induction)

  # -------------------------------------------------
  # 1.3  Fisher or χ² test on the 2×2 table
  # -------------------------------------------------
  tab <- df2 %>%
    dplyr::count(CIMIC_Cluster, Induction) %>%
    tidyr::pivot_wider(names_from = Induction,
                       values_from = n,
                       values_fill = 0)

  mat <- as.matrix(tab[ , c("Negative/No Induction", "Positive Induction")])
  pval <- if (any(mat < 5) || sum(mat) < 50) {
    stats::fisher.test(mat)$p.value
  } else {
    stats::chisq.test(mat)$p.value
  }
  ptxt <- paste0("P = ", format.pval(pval, digits = 4))

  # -------------------------------------------------
  # 1.4  Prepare data for the stacked‑bar plot
  # -------------------------------------------------
  bar_df <- df2 %>%
    dplyr::group_by(CIMIC_Cluster, Induction) %>%
    dplyr::summarise(n = n(), .groups = "drop") %>%
    dplyr::group_by(CIMIC_Cluster) %>%
    dplyr::mutate(
      Total = sum(n),
      Prop  = n / Total
    ) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(
      CIMIC_Cluster,
      factor(Induction,
             levels = c("Negative/No Induction", "Positive Induction"))
    )

  # -------------------------------------------------
  # 1.5  Plot -------------------------------------------------
  # -------------------------------------------------
  ggplot(bar_df,
         aes(x = CIMIC_Cluster,
             y = Prop,
             fill = Induction)) +
    geom_bar(stat = "identity",
             position = "fill",
             width = 0.7,
             aes(fill = factor(Induction,
                               levels = c("Negative/No Induction",
                                          "Positive Induction")))) +
    geom_text(aes(y = Prop,
                  label = ifelse(Prop > 0,
                                 paste0(round(Prop * 100), "%"),
                                 "")),
              position = position_stack(vjust = 0.5),
              colour = "black",
              size = 8,
              fontface = "bold") +
    geom_text(aes(x = 1.5, y = 1.06, label = ptxt),
              inherit.aes = FALSE,
              size = 8,
              fontface = "bold") +
    geom_text(aes(x = CIMIC_Cluster,
                  y = -0.05,
                  label = paste0("N = ", Total)),
              inherit.aes = FALSE,
              size = 6,
              fontface = "bold") +
    labs(title = paste0(gene),
         y = "Percentage",
         fill = "Induction") +
    scale_fill_manual(values = c("Positive Induction" = "#3fb949ff",
                                 "Negative/No Induction" = "#f76661ff")) +
    scale_y_continuous(labels = scales::percent,
                       limits = c(-0.1, 1.12),
                       breaks = seq(0, 1, 0.2)) +
    scale_x_discrete(expand = expansion(add = 0.5)) +
    theme_classic(base_size = 18) +
    theme(
      plot.title      = element_text(hjust = 0.5,
                                     face = "bold",
                                     size = 18),
      legend.position = "top",
      panel.border    = element_rect(colour = "black",
                                     fill = NA,
                                     linewidth = 1.5),
      axis.line.x     = element_line(colour = "black",
                                     linewidth = 1.5),
      axis.line.y     = element_line(colour = "black",
                                     linewidth = 1.5),
      axis.text.x     = element_text(face = "bold",
                                     size = 20),
      axis.text.y     = element_text(face = "bold",
                                     size = 20),
      axis.title.y    = element_text(face = "bold",
                                     size = 20),
      legend.text     = element_text(size = 18),
      legend.title    = element_text(size = 18,
                                     face = "bold")
    )
}

genes_of_interest <- c("PDCD1", "CD274", "IRF1")
pen_plots <- purrr::map(genes_of_interest,
                    ~ compare_penetrance_plot(clustered_plot_df_for_merge, .x))

```




# Analyze PBSC for individual Genes

```{r}

# -------------------------------------------------
# 2. Point‑biserial correlation (binary cluster vs expression) -----
# -------------------------------------------------
compare_pointbiserial_plot <- function(df, gene) {
  # -------------------------------------------------
  # 2.1  Locate the correct gene column
  # -------------------------------------------------
  gene_col <- df %>%
    dplyr::select(dplyr::matches(paste0("^", gene, "(\\.x|\\.y)?$"))) %>%
    names()
  if (length(gene_col) == 0) {
    stop("Gene column '", gene, "' not found in the data frame")
  }
  gene_col <- gene_col[[1]]

  # -------------------------------------------------
  # 2.2  Build a tidy data frame with a numeric expression column
  # -------------------------------------------------
  df2 <- df %>%
    dplyr::mutate(
      Expression = as.numeric(as.character(.data[[gene_col]])),
      Cluster_bin = ifelse(CIMIC_Cluster == "Fun-CIM", 1, 0)
    ) %>%
    dplyr::select(CIMIC_Cluster, Cluster_bin, Expression)

  # -------------------------------------------------
  # 2.3  Pearson correlation (point‑biserial)
  # -------------------------------------------------
  cor_res <- cor.test(df2$Cluster_bin, df2$Expression,
                      method = "pearson")
  rtxt <- sprintf("r = %.3f\nP = %s",
                  cor_res$estimate,
                  format.pval(cor_res$p.value, digits = 4))

  # -------------------------------------------------
  # 2.4  Plot -------------------------------------------------
  # -------------------------------------------------
  ggplot(df2,
         aes(x = Cluster_bin,
             y = Expression,
             colour = CIMIC_Cluster)) +
    geom_jitter(width = 0.1,
                size = 2,
                alpha = 0.6) +
    geom_smooth(method = "lm",
                se = FALSE,
                colour = "black",
                linewidth = 1.2) +
    scale_x_continuous(breaks = c(0, 1),
                       labels = c("Dys-CIM", "Fun-CIM")) +
    scale_colour_manual(values = c("Dys-CIM" = "#C0392B",
                                   "Fun-CIM" = "#2980B9")) +
    annotate("text",
             x = 0.5,
             y = max(df2$Expression, na.rm = TRUE) * 1.1,
             label = rtxt,
             size = 5,
             fontface = "bold") +
    labs(title = paste0(gene),
         x = NULL,
         y = "Δ(log2[TPM+1])") +
    theme_classic(base_size = 14) +
    theme(
      plot.title      = element_text(hjust = 0.5,
                                     face = "bold",
                                     size = 18),
      legend.position = "none",
      axis.text.x     = element_text(face = "bold",
                                     size = 18),
      axis.text.y     = element_text(face = "bold",
                                     size = 18),
      axis.title.y    = element_text(face = "bold",
                                     size = 18),
      panel.border    = element_rect(colour = "black",
                                     fill = NA,
                                     linewidth = 1.5),
      axis.line       = element_line(colour = "black",
                                     linewidth = 1.5)
    )
}

genes_of_interest <- c("PDCD1", "CD274",  "PDCD1LG2",   # PD-L2
  "CTLA4",
  "LAG3",
  "HAVCR2",     # TIM-3
  "TIGIT",       # VISTA
  "BTLA")
pbs_plots <- purrr::map(genes_of_interest,
                    ~ compare_pointbiserial_plot(clustered_plot_df_for_merge, .x))







```



# Individual Gene three plot 

```{r}


genes_of_interest <- c("PDCD1", "CD274", "IRF1")



# optional: combine the three panels for a single gene
make_three_panel <- function(gene) {
  ind  <- compare_induction_boxplot(clustered_plot_df_for_merge, gene)
  pen  <- compare_penetrance_plot(clustered_plot_df_for_merge, gene)
  pb   <- compare_pointbiserial_plot(clustered_plot_df_for_merge, gene)

  ggpubr::ggarrange(ind, pen, pb,
                    ncol = 3, nrow = 1,
                    common.legend = FALSE) %>%
    ggpubr::annotate_figure(
      top = ggpubr::text_grob(paste0(gene, " – Summary"),
                              face = "bold", size = 18)
    )
}

# Example: three‑panel view for PDCD1
make_three_panel("PDCD1")


```





# Individual Gene two plot 

```{r}





# optional: combine the three panels for a single gene
make_two_panel <- function(gene) {
  ## 1️⃣  Create the two plots (they still carry their titles)
  ind <- compare_induction_boxplot(clustered_plot_df_for_merge, gene)
  pen <- compare_penetrance_plot(clustered_plot_df_for_merge, gene)

  ## 2️⃣  Remove the titles from each plot
  ind <- ind + theme(plot.title = element_blank())
  pen <- pen + theme(plot.title = element_blank(), 
    axis.title.x = element_blank(), legend.position = "none")

  ## 3️⃣  Combine them side‑by‑side
  ggpubr::ggarrange(
    ind, pen,
    ncol = 2, nrow = 1,
    common.legend = FALSE          # keep the legend if you like it
  ) %>%
    ggpubr::annotate_figure(
      top = ggpubr::text_grob(
        paste0(gene),
        face = "bold", size = 18
      )
    )
}

# Example: 2‑panel view for PDCD1
# make_two_panel("PDCD1")

genes_of_interest <- c("PDCD1", "CD274", "Avg_MHC1", "CALR", "NLRC5", "STAT1")
two_panel_plots <- purrr::map(
  genes_of_interest,
  ~ make_two_panel(.x)      # .x is the gene name
)


classical_tcell_inhibitory_checkpoints <- c(
  "PDCD1",      # PD-1
  "CD274",      # PD-L1
  "PDCD1LG2",   # PD-L2
  "CTLA4",      
  "LAG3",
  "HAVCR2",     # TIM-3
  "TIGIT",
  "BTLA",
  "CD160",
  "CD244"
)


two_panel_plots <- purrr::map(
  classical_tcell_inhibitory_checkpoints,
  ~ make_two_panel(.x)      # .x is the gene name
)


classical_icd_genes <- c(
  "CALR",     # Calreticulin (ecto-CALR exposure signal)
  "HMGB1",    # High mobility group box 1 (DAMP released during ICD)
  "HSP90AA1", # Heat shock protein 90
  "HSP90AB1",
  "HSPA1A",   # HSP70
  "HSPA1B",
  "HSPA5",    # ER stress / unfolded protein response
  "ANXA1",    # Annexin A1 (DC recruitment signal)
  "P2RX7",    # ATP sensing receptor
  "PANX1",    # ATP release channel
  "TLR4",     # HMGB1 receptor on dendritic cells
  "EIF2AK3",  # PERK (ER stress signaling)
  "ATF4",     # Integrated stress response
  "DDIT3",    # CHOP
  "CXCL10",   # Type I IFN chemokine
  "IFNB1",     # Type I interferon
  "IFNAR1",
  "IFNAR2",
  "IFNA1",
  "IFNG",
  "IFNGR1",
  "IFNGR2"
)


two_panel_plots_icd <- purrr::map(
  classical_icd_genes,
  ~ make_two_panel(.x)      # .x is the gene name
)


```















# Cell Death between clusters

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(purrr)

# -----------------------------
# 1. Define gene sets
# -----------------------------
death_programs <- list(
  Apoptosis = c(
    "BAX", "BAK1", "BBC3", "PMAIP1", "BCL2L11",
    "APAF1", "CASP9", "CASP3", "CASP7",
    "FAS", "TNFRSF10B", "CASP8", "FADD", "BID",
    "DIABLO", "CYCS", "TP53AIP1"
  ),
  Necroptosis = c(
    "RIPK1", "RIPK3", "MLKL", "ZBP1", "TICAM1"
  ),
  Pyroptosis = c(
    "GSDME", "GSDMD", "CASP1", "CASP4", "CASP5",
    "NLRP3", "AIM2", "PYCARD", "IL1B", "IL18"
  ),
  PANoptosis = c(
    "ZBP1", "AIM2", "RIPK3", "RIPK1", "CASP8",
    "CASP1", "FADD", "IRF1", "PYCARD"
  ),
  Ferroptosis = c(
    "ACSL4", "LPCAT3", "TFRC", "SAT1", "PTGS2"
  )
)

other_programs <- list(
  Senescence = c(
    "CDKN1A", "CDKN2A", "CDKN2B", "GLB1", "SERPINE1",
    "IL6", "CXCL8", "CXCL1", "CCL2", "MMP1", "MMP3"
  ),
  ICD_Markers = c(
    "CALR", "HMGB1", "ANXA1", "PANX1",
    "CXCL10", "CXCL11", "IFNB1", "CCL5"
  ),
  Cytosolic_DNA_Sensing = c(
    "CGAS", "MB21D1", "TMEM173", "STING1", "TBK1", "IKBKE",
    "IRF3", "IRF7", "IFI16", "AIM2", "ZBP1",
    "IFNB1", "CXCL10", "CXCL11", "CCL5", "ISG15", "IFIT1", "IFIT3", "MX1", "OAS1"
  ),
  ER_Stress_ISR = c(
    "DDIT3", "ATF4", "ATF3", "XBP1", "ERN1", "EIF2AK3",
    "HSPA5", "DNAJB9", "PPP1R15A", "EDEM1", "HERPUD1", "TRIB3"
  ),
  Proteostasis = c(
    "HSP90AA1", "HSP90AB1", "HSPA1A", "HSPA1B", "HSPA8", "HSPB1",
    "DNAJA1", "DNAJB1", "BAG3", "STIP1",
    "HSPA5", "CANX", "CALR", "PDIA4", "PDIA6", "ERP44",
    "HERPUD1", "EDEM1", "SEL1L", "SYVN1",
    "PSMB5", "PSMD14", "UBC", "SQSTM1",
    "MAP1LC3B", "ATG5", "ATG7", "BECN1", "LAMP1", "CTSD",
    "HMOX1", "NQO1", "TXN", "TXNRD1", "GCLC"
  ),
  Immune_Checkpoints = c(
    "CD274", "PDCD1LG2", "CD47", "LGALS9", "PVR", "NECTIN2",
    "IDO1", "VTCN1", "CD276", "HHLA2"
  ),
  AntiTumoral_Cytokines = c(
    "CXCL9", "CXCL10", "CXCL11", "CCL5",
    "IFNB1", "IFNA1", "IFNG", "TNF",
    "IL12A", "IL12B", "IL15", "CSF2"
  ),
  ProTumoral_Cytokines = c(
    "CXCL1", "CXCL2", "CXCL3", "CXCL5", "CXCL8",
    "CCL20", "CCL22", "CCL28",
    "IL6", "IL1B", "TGFB1", "CSF1", "VEGFA"
  )
)

all_programs <- c(death_programs, other_programs, all_gene_sets[all_gene_set_names_cimic])

# -----------------------------
# 2. Metadata
# -----------------------------
df <- clustered_plot_df_for_merge %>%
  mutate(
    sample_id = as.character(sample_id),
    cimic_cluster = as.factor(cluster_assignments.x),
    model_id = as.character(base_id)
  )

# -----------------------------
# 3. Keep genes present
# -----------------------------
all_gene_cols <- colnames(df)
all_programs_present <- lapply(all_programs, function(genes) intersect(genes, all_gene_cols))
print(all_programs_present)

# -----------------------------
# 4. Gene-wise z-score across samples
# -----------------------------
meta_cols <- c(
  "sample_id", "UMAP1", "UMAP2", "PACMAP1", "PACMAP2",
  "PC1", "PC2", "PC3", "base_id", "cluster_assignments.x", 
  "cluster_assignments.y", "cimic_cluster", "model_id", "CIMIC_Cluster"
)

gene_cols <- setdiff(colnames(df), meta_cols)

zscore_safe <- function(x) {
  s <- sd(x, na.rm = TRUE)
  m <- mean(x, na.rm = TRUE)
  if (is.na(s) || s == 0) rep(0, length(x)) else (x - m) / s
}

df_z <- df %>%
  mutate(across(all_of(gene_cols), zscore_safe))

# -----------------------------
# 5. Score each program
# -----------------------------
score_program_z <- function(data, genes, program_name) {
  if (length(genes) == 0) {
    return(
      data.frame(
        sample_id = data$sample_id,
        cimic_cluster = data$cimic_cluster,
        model_id = data$model_id,
        Program = program_name,
        Score = NA_real_
      )
    )
  }

  score_vec <- rowMeans(as.matrix(data[, genes, drop = FALSE]), na.rm = TRUE)

  data.frame(
    sample_id = data$sample_id,
    cimic_cluster = data$cimic_cluster,
    model_id = data$model_id,
    Program = program_name,
    Score = score_vec
  )
}

program_scores_all <- purrr::imap_dfr(all_programs_present, ~score_program_z(df_z, .x, .y))

# -----------------------------
# 6. Summary
# -----------------------------
program_summary <- program_scores_all %>%
  group_by(model_id, cimic_cluster, Program) %>%
  summarise(
    mean_score = mean(Score, na.rm = TRUE),
    median_score = median(Score, na.rm = TRUE),
    sd_score = sd(Score, na.rm = TRUE),
    n = dplyr::n(),
    .groups = "drop"
  )

print(program_summary)

cimic_cluster_COLORS <- c("1" = "#1b9e77", "2" = "#d95f02")

# -----------------------------
# 7. Faceted boxplot by cluster
# -----------------------------
program_scores_plot_df <- program_scores_all %>%
  mutate(
    cimic_cluster = factor(cimic_cluster),
    Program = factor(Program, levels = unique(names(all_programs)))
  )

pval_df_all <- program_scores_plot_df %>%
  group_by(Program) %>%
  summarise(
    y.position = max(Score, na.rm = TRUE) * 1.12,
    pval = tryCatch(wilcox.test(Score ~ cimic_cluster)$p.value, error = function(e) NA_real_),
    p_label = ifelse(
      is.na(pval), "p = NA",
      ifelse(pval < 0.001, "p < 0.001", paste0("p = ", signif(pval, 3)))
    ),
    .groups = "drop"
  )

p_all_programs_p <- ggplot(program_scores_plot_df, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  geom_text(
    data = pval_df_all,
    aes(x = 1.5, y = y.position, label = p_label),
    inherit.aes = FALSE,
    size = 4.5,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    title = "Program scores by CIMiC cluster",
    y = "Program z-score",
    x = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "none",
    axis.text.x = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold", size = 14),
    axis.title.y = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold", size = 12),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5),
    axis.line = element_line(colour = "black", linewidth = 1.5)
  )

print(p_all_programs_p)

# -----------------------------
# 8. Significant-only boxplots
# -----------------------------
sig_programs <- pval_df_all %>%
  filter(!is.na(pval), pval < 0.05) %>%
  pull(Program)

program_scores_plot_df_sig <- program_scores_plot_df %>%
  filter(Program %in% sig_programs)

pval_df_sig <- pval_df_all %>%
  filter(Program %in% sig_programs)

p_all_programs_sig <- ggplot(program_scores_plot_df_sig, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  geom_text(
    data = pval_df_sig,
    aes(x = 1.5, y = y.position, label = p_label),
    inherit.aes = FALSE,
    size = 4.5,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    title = "Significant program scores by CIMiC cluster",
    y = "Program z-score",
    x = NULL
  ) +
  theme_classic(base_size = 22) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "none",
    axis.text.x = element_text(face = "bold", size = 14),
    axis.text.y = element_text(face = "bold", size = 14),
    axis.title.y = element_text(face = "bold", size = 16),
    strip.text = element_text(face = "bold", size = 12),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5),
    axis.line = element_line(colour = "black", linewidth = 1.5)
  )

print(p_all_programs_sig)

# -----------------------------
# 9. Heatmap by model + cluster
# -----------------------------
program_heatmap_df <- program_scores_all %>%
  group_by(model_id, cimic_cluster, Program) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    model_cluster = paste0(model_id, "_C", cimic_cluster),
    Program = factor(Program, levels = unique(names(all_programs)))
  )

p_heatmap <- ggplot(program_heatmap_df, aes(x = Program, y = model_cluster, fill = mean_score)) +
  geom_tile(color = "white", linewidth = 0.8) +
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean/nz-score"
  ) +
  labs(
    title = "Program landscape by model and CIMiC cluster",
    x = NULL,
    y = NULL
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 11),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
  )

print(p_heatmap)

# -----------------------------
# 10. Cluster heatmap with FDR
# -----------------------------
cluster_heatmap_df <- program_scores_all %>%
  group_by(cimic_cluster, Program) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    cimic_cluster = factor(cimic_cluster),
    Program = factor(Program, levels = unique(names(all_programs)))
  )

program_pvals <- program_scores_all %>%
  mutate(cimic_cluster = factor(cimic_cluster)) %>%
  group_by(Program) %>%
  summarise(
    p_value = tryCatch(wilcox.test(Score ~ cimic_cluster)$p.value, error = function(e) NA_real_),
    .groups = "drop"
  ) %>%
  mutate(
    FDR = p.adjust(p_value, method = "fdr"),
    fdr_label = case_when(
      is.na(FDR) ~ "FDR=NA",
      FDR < 0.001 ~ "FDR<0.001",
      TRUE ~ paste0("FDR=", signif(FDR, 2))
    )
  )

program_labels <- setNames(
  paste0(program_pvals$Program, "/n", program_pvals$fdr_label),
  program_pvals$Program
)

p_cluster_heatmap <- ggplot(cluster_heatmap_df, aes(x = Program, y = cimic_cluster, fill = mean_score)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = round(mean_score, 2)), size = 4, fontface = "bold") +
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean/nz-score"
  ) +
  scale_x_discrete(labels = program_labels) +
  labs(
    title = "Mean program z-score by CIMiC cluster",
    x = NULL,
    y = "CIMiC cluster"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold", size = 11),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 14),
    axis.line = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
  )

p_cluster_heatmap


```




# Correlation analysis looking at correlation with clusters

```{r}



# Correlation with Cluster Assignment #################

library(dplyr)
library(purrr)
library(broom)
library(ggplot2)
library(tidyr)

# Gene set
human_CIM_genes <- unique(unlist(all_gene_sets[all_gene_set_names_cimic]))

# Initialize lists
all_cor_results <- list()
sig_cor_results <- list()

# choose your dataset
ds_data <- clustered_plot_df   # <- change if needed

# Ensure numeric cluster
ds_data <- ds_data %>%
  mutate(cluster_numeric = as.numeric(cluster_assignments))

# ---- Gene correlation ----
expr_mat <- ds_data %>% dplyr::select(any_of(human_CIM_genes))

cor_df <- purrr::map_dfr(
  colnames(expr_mat),
  function(gene){
    test <- cor.test(expr_mat[[gene]], ds_data$cluster_numeric, method = "spearman")
    
    tibble(
      gene = gene,
      cor = test$estimate,
      cor_pval = test$p.value,
      direction = ifelse(test$estimate > 0,
                         "Fun-CIM",
                         "Dys-CIM")
    )
  }
)

# Adjust p-values
cor_df <- cor_df %>%
  mutate(cor_padj = p.adjust(cor_pval, method = "fdr")) 


# -----------------------------
# 2. MEAN EXPRESSION BY CLUSTER
# -----------------------------
ds_data$group <- factor(ds_data$cluster_numeric)
groups <- levels(ds_data$group)

cluster_means <- map_dfr(colnames(expr_mat), function(gene) {
  
  expr <- ds_data[[gene]]
  
  # ---- means per group (dynamic) ----
  mean_vals <- tapply(expr, ds_data$group, mean, na.rm = TRUE)
  
  mean_df <- as.data.frame(as.list(mean_vals))
  names(mean_df) <- paste0("mean_cluster", names(mean_vals))
  
  # ensure consistency for 2-group downstream ops
  mean_cluster1 <- if ("1" %in% names(mean_vals)) mean_vals[["1"]] else NA
  mean_cluster2 <- if ("2" %in% names(mean_vals)) mean_vals[["2"]] else NA
  
  # ---- differential ----
  cluster2_vs_1 <- mean_cluster2 - mean_cluster1
  cluster1_vs_2 <- mean_cluster1 - mean_cluster2
  
  # ---- test selection ----
  df_tmp <- data.frame(expr = expr, group = ds_data$group)
  
  if (length(groups) == 2) {
    test <- wilcox.test(expr ~ group, data = df_tmp)
    test_name <- "wilcox_pval"
  } else {
    test <- kruskal.test(expr ~ group, data = df_tmp)
    test_name <- "kruskal_pval"
  }
  
  tibble(
    gene = gene,
    !!!mean_df,
    cluster2_vs_1 = cluster2_vs_1,
    cluster1_vs_2 = cluster1_vs_2,
    expr_pval = test$p.value
  )
})


# -----------------------------
# 4. FDR CORRECTION
# -----------------------------
cluster_means <- cluster_means %>%
  mutate(
    expr_padj = p.adjust(expr_pval, method = "fdr")
  )


final_cor_df <- cor_df %>%
  left_join(cluster_means, by = "gene")


# ---- Significant subset ----

corr_thresh <- 0.6
ge_thresh <- 0

sig_cor_df <- final_cor_df %>%
  filter(cor_padj < 0.05, abs(cor) >= corr_thresh, abs(cluster2_vs_1) >= log2(ge_thresh)) %>%
  mutate(direction = ifelse(cor > 0, "CIM Functional", "CIM Dysfunctional"))






# ---- Visualization ----

library(dplyr)

top50_each_direction <- sig_cor_df %>%
  
  # define direction explicitly from biology
  mutate(
    direction = case_when(
      cluster2_vs_1 > 0 ~ "Fun-CIM",
      cluster2_vs_1 < 0 ~ "Dys-CIM"
    )
  ) %>%
  
  group_by(direction) %>%
  
  # rank within each direction
    # rank by effect size first
  dplyr::arrange(desc(abs(cluster2_vs_1)), .by_group = TRUE) %>%
  
  slice_head(n = 50) %>%

  ungroup()

top50_each_direction <- top50_each_direction %>%
  dplyr::group_by(direction) %>%
  dplyr::arrange(gene, .by_group = TRUE) %>%
  dplyr::ungroup()
  
ggplot(top50_each_direction, aes(x = gene, y = abs(cor), fill = direction)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.7) +
    coord_flip() +
    # --- Add the horizontal line here ---
    geom_hline(yintercept = 0.5, linetype = "longdash", color = "black", linewidth = 1) +
    facet_wrap(~direction, scales = "free_y") +
    labs(
        # x = "Gene",
        y = "Absolute Correlation (|r|)", # Updated label for clarity
        # fill = "Dataset",
    ) +
    theme_classic(base_size = 18) +
  theme(
    axis.text.y = element_text(size = 18, colour = "black", face = "bold"),
    axis.text.x = element_text(size = 18, colour = "black", face = "bold"),
    strip.text.x = element_text(size = 18, face = "bold") 
  )


# Extraction ###############


# Separate positive vs negative
positive_genes <- sig_cor_df %>%
  filter(cor > 0) %>%
  distinct(gene) %>%
  pull(gene)

negative_genes <- sig_cor_df %>%
  filter(cor < 0) %>%
  distinct(gene) %>%
  pull(gene)

# Check
length(positive_genes)
length(negative_genes)

# ORA ANALYSIS ###############


# overrepresentation analyses 
  
library(msigdbr)
library(clusterProfiler)
library(openxlsx)
library(dplyr)
library(org.Hs.eg.db)
library(org.Mm.eg.db)


library(paletteer)
library(enrichplot)
library(patchwork)
library(tidyverse)
library(stringr)


combined_shared_genes <- master_induction_df$gene_id



universe_genes <- combined_shared_genes
padj_cutoff <- 0.05


# ORA
shared_neg_ORA <- enrichGO(
  gene          = negative_genes,
  OrgDb         = org.Hs.eg.db,
  ont           = "ALL",
  pAdjustMethod = "BH",
  pvalueCutoff  = padj_cutoff,
  qvalueCutoff  = padj_cutoff,
  universe      = universe_genes,
  keyType       = "SYMBOL"
)

# ORA
shared_pos_ORA <- enrichGO(
  gene          = positive_genes,
  OrgDb         = org.Hs.eg.db,
  ont           = "ALL",
  pAdjustMethod = "BH",
  pvalueCutoff  = padj_cutoff,
  qvalueCutoff  = padj_cutoff,
  universe      = universe_genes,
  keyType       = "SYMBOL"
)



# reactome 
library(ReactomePA)


# Load necessary packages
library(clusterProfiler)
library(org.Hs.eg.db)

# Convert gene symbols to Entrez IDs
universe_genes_entrez <- bitr(
  universe_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)


# Convert gene symbols to Entrez IDs
negative_genes_entrez <- bitr(
  negative_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)



shared_neg_reactome <- enrichPathway(
  negative_genes_entrez$ENTREZID,
  organism = "human",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.2,
  universe = universe_genes_entrez$ENTREZID,
  minGSSize = 10,
  maxGSSize = 500,
  readable = FALSE
)


# Convert gene symbols to Entrez IDs
positive_genes_entrez <- bitr(
  positive_genes,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)


shared_pos_reactome <- enrichPathway(
  positive_genes_entrez$ENTREZID,
  organism = "human",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.2,
  universe = universe_genes_entrez$ENTREZID,
  minGSSize = 10,
  maxGSSize = 500,
  readable = FALSE
)


# VISUALIZE ORA ##########


    pos_ORA_simplified <- clusterProfiler::simplify(
      shared_pos_ORA,
      cutoff = 0.7,   # similarity threshold (0.7 is common, range 0–1)
      by = "p.adjust", # keep the term with the lowest adjusted p-value
      select_fun = min,
      measure = "Wang" # semantic similarity measure; Wang is common for GO
    )



    neg_ORA_simplified <- clusterProfiler::simplify(
      shared_neg_ORA,
      cutoff = 0.7,   # similarity threshold (0.7 is common, range 0–1)
      by = "p.adjust", # keep the term with the lowest adjusted p-value
      select_fun = min,
      measure = "Wang" # semantic similarity measure; Wang is common for GO
    )

    
    n_count <- 15
    
    # Top terms
    top_res_c1_pos_go<- pos_ORA_simplified@result %>%
      dplyr::filter(ONTOLOGY == "BP") %>%
      slice_min(order_by = p.adjust,
                n = n_count,
                with_ties = FALSE) %>%
      mutate(Description = str_wrap(Description, width = 50))

 
    
    
    p1_pos_ora_go <- ggplot(top_res_c1_pos_go, aes(x = Description, y = Count)) +
      geom_col(aes(fill = p.adjust), color = "black") +
      coord_flip() +
      #facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free") +
      labs(
        #title = paste0("positive", "\nTop ", n_count, " Enriched Terms per ONTOLOGY"),
        x = NULL,
        y = "Gene Count",
        fill = "Adj_p"
      ) +
      scale_fill_gradient(low = "red",
                          high = "blue",
                          trans = "reverse") +
      theme_minimal(base_size = 14) +
      scale_fill_gradient(low = "red", high = "blue", trans = "reverse", 
                        labels = scales::label_scientific(digits = 2)) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        strip.background = element_rect(fill = "white", colour = "black"),
        strip.text.y = element_text(size = 18, angle = 0, face = "bold"),
        axis.text.y = element_text(size = 16, colour = "black", , face = "bold"),
        axis.text.x = element_text(size = 18, colour = "black", face = "bold"),
        axis.title.x = element_text(size = 18, colour = "black", face = "bold"),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18)
        #plot.title = element_text(hjust = 0.5)
    )

    
highlight_terms <- c(
  "protein folding",
  "positive regulation of execution phase of apoptosis",
  "proteasome assembly",
  "intrinsic apoptotic signaling pathway",
  "protein-DNA complex organization",
  "RNA catabolic process"
)

top_neg_go <- neg_ORA_simplified@result %>%
  dplyr::filter(ONTOLOGY == "BP")

# top 15 by significance
top15 <- top_neg_go %>%
  slice_min(order_by = p.adjust, n = n_count-length(highlight_terms), with_ties = FALSE)

# forced terms (even if outside top 15)
forced <- top_neg_go %>%
  filter(Description %in% highlight_terms)

# combine + remove duplicates
top_res_c1_neg_go <- bind_rows(top15, forced) %>%
  distinct(Description, .keep_all = TRUE) %>%
  mutate(Description = str_wrap(Description, width = 40))
 
    
    
    p1_neg_ora_go <- ggplot(top_res_c1_neg_go, aes(x = Description, y = Count)) +
      geom_col(aes(fill = p.adjust), color = "black") +
      coord_flip() +
      #facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free") +
      labs(
        #title = paste0("negative", "\nTop ", n_count, " Enriched Terms per ONTOLOGY"),
        x = NULL,
        y = "Gene Count",
        fill = "Adj_p"
      ) +
       scale_fill_gradient(low = "red", high = "blue", trans = "reverse", 
                        labels = scales::label_scientific(digits = 2)) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        strip.background = element_rect(fill = "white", colour = "black"),
        strip.text.y = element_text(size = 18, angle = 0, face = "bold"),
        axis.text.y = element_text(size = 16, colour = "black", , face = "bold"),
        axis.text.x = element_text(size = 18, colour = "black", face = "bold"),
        axis.title.x = element_text(size = 18, colour = "black", face = "bold"),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18)
        #plot.title = element_text(hjust = 0.5)
    )

# Reactome VISUALIZATION ###############
    
        # Top terms
    top_res_c1_pos_react <- shared_pos_reactome@result %>%
      slice_min(order_by = p.adjust,
                n = n_count,
                with_ties = FALSE) %>%
      mutate(Description = str_wrap(Description, width = 50))

 
    
    
    p1_pos_ora_react <- ggplot(top_res_c1_pos_react, aes(x = Description, y = Count)) +
      geom_col(aes(fill = p.adjust), color = "black") +
      coord_flip() +
      # facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free") +
      labs(
        #title = paste0("positive", "\nTop ", n_count, " Enriched Terms per ONTOLOGY"),
        x = NULL,
        y = "Gene Count",
        fill = "Adj_p"
      ) +
      scale_fill_gradient(low = "red",
                          high = "blue",
                          trans = "reverse") +
      theme_minimal(base_size = 14) +
      scale_fill_gradient(low = "red", high = "blue", trans = "reverse", 
                        labels = scales::label_scientific(digits = 2)) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        strip.background = element_rect(fill = "white", colour = "black"),
        strip.text.y = element_text(size = 18, angle = 0, face = "bold"),
        axis.text.y = element_text(size = 16, colour = "black", , face = "bold"),
        axis.text.x = element_text(size = 18, colour = "black", face = "bold"),
        axis.title.x = element_text(size = 18, colour = "black", face = "bold"),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18)
        #plot.title = element_text(hjust = 0.5)
    )


    
    # Top terms
    top_res_c1_neg_react <- shared_neg_reactome@result %>%
      slice_min(order_by = p.adjust,
                n = n_count,
                with_ties = FALSE) %>%
      mutate(Description = str_wrap(Description, width = 50))

 
    
    
    p1_neg_ora_react <- ggplot(top_res_c1_neg_react, aes(x = Description, y = Count)) +
      geom_col(aes(fill = p.adjust), color = "black") +
      coord_flip() +
      # facet_grid(ONTOLOGY ~ ., scales = "free_y", space = "free") +
      labs(
        #title = paste0("negative", "\nTop ", n_count, " Enriched Terms per ONTOLOGY"),
        x = NULL,
        y = "Gene Count",
        fill = "Adj_p"
      ) +
       scale_fill_gradient(low = "red", high = "blue", trans = "reverse", 
                        labels = scales::label_scientific(digits = 2)) +
    theme_minimal(base_size = 14) +
    theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
        strip.background = element_rect(fill = "white", colour = "black"),
        strip.text.y = element_text(size = 18, angle = 0, face = "bold"),
        axis.text.y = element_text(size = 16, colour = "black", , face = "bold"),
        axis.text.x = element_text(size = 18, colour = "black", face = "bold"),
        axis.title.x = element_text(size = 18, colour = "black", face = "bold"),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18)
        #plot.title = element_text(hjust = 0.5)
    )

go_pos <- pos_ORA_simplified@result %>% dplyr::filter(ONTOLOGY == "BP")
go_neg <- neg_ORA_simplified@result %>% dplyr::filter(ONTOLOGY == "BP")

react_pos <- shared_pos_reactome@result
react_neg <- shared_neg_reactome@result
```

# Correlation analyses 


```{r}


library(dplyr)



library(dplyr)

deltas_df <- formatted_individual_deltas_df %>% t() %>% as.data.frame()
all_unique_gene_ids <- unique(unlist(all_gene_sets))


GE_Correlation_Data_Frame <- deltas_df

# Add baseline + surgery expression (remove gene_id column from TPM df)
GE_Correlation_Data_Frame <- cbind(
  GE_Correlation_Data_Frame,
  combined_TPM_lognorm[ , -1]
)

GE_Correlation_Data_Frame <- GE_Correlation_Data_Frame %>% rownames_to_column ("gene_id")

colnames_vec <- colnames(GE_Correlation_Data_Frame)[
  colnames(GE_Correlation_Data_Frame) != "gene_id"
]

# delta columns
delta_cols <- colnames_vec[grep("^delta_", colnames_vec)]

# BIOPSY (_B_)
BIOPSY_cols <- colnames_vec[grep("_B_", colnames_vec)]

# SURGERY (_S_) but not delta
SURGERY_cols <- colnames_vec[
  grepl("_S_", colnames_vec) &
  !grepl("^delta_", colnames_vec)
]


# 3️⃣ Correlation Per Sample (Delta vs Biopsy)
Corr_Results <- data.frame(
  sample_id_full = character(),
  DELTA_BIOPSY_Corr_Value = numeric(),
  DELTA_BIOPSY_P_Value = numeric(),
  SURGERY_BIOPSY_Corr_Value = numeric(),
  SURGERY_BIOPSY_P_Value = numeric(),
  stringsAsFactors = FALSE
)

# 🔁 Loop per delta sample
for(i in seq_along(delta_cols)) {

  delta_col <- delta_cols[i]
# IMPROVED REGEX:
  # Captures everything after 'delta_' and before '_S_Rep' as the ID
  # Captures 'RepX' at the very end
  sample_id  <- sub("^delta_(.*)_S_Rep\\d+$", "\\1", delta_col)
  rep_number <- sub("^delta_.*_S_(Rep\\d+)$", "\\1", delta_col)
  
  sample_id_full <- paste0(sample_id, "_", rep_number)

  # Match baseline (Biopsy) and surgery using exact anchors
  # We look for [ID]_B_[Rep] and [ID]_S_[Rep]
  BIOPSY_match <- BIOPSY_cols[
    grep(paste0("^", sample_id, "_B_", rep_number, "$"), BIOPSY_cols)
  ]

  SURGERY_match <- SURGERY_cols[
    grep(paste0("^", sample_id, "_S_", rep_number, "$"), SURGERY_cols)
  ]

  # Check if Biopsy exists to perform the Delta vs Biopsy correlation
  if(length(BIOPSY_match) == 1) {

    # DELTA vs BIOPSY
    DELTA_BIOPSY_corr <- cor.test(
      GE_Correlation_Data_Frame[[delta_col]],
      GE_Correlation_Data_Frame[[BIOPSY_match]],
      method = "spearman",
      exact = FALSE
    )

    # SURGERY vs BIOPSY
    if(length(SURGERY_match) == 1) {
      SURGERY_BIOPSY_corr <- cor.test(
        GE_Correlation_Data_Frame[[SURGERY_match]],
        GE_Correlation_Data_Frame[[BIOPSY_match]],
        method = "spearman",
        exact = FALSE
      )
      SUR_val <- SURGERY_BIOPSY_corr$estimate
      SUR_p   <- SURGERY_BIOPSY_corr$p.value
    } else {
      SUR_val <- NA
      SUR_p   <- NA
    }

    # Append results
    Corr_Results <- Corr_Results %>%
      add_row(
        sample_id_full = sample_id_full,
        DELTA_BIOPSY_Corr_Value = as.numeric(DELTA_BIOPSY_corr$estimate),
        DELTA_BIOPSY_P_Value = as.numeric(DELTA_BIOPSY_corr$p.value),
        SURGERY_BIOPSY_Corr_Value = as.numeric(SUR_val),
        SURGERY_BIOPSY_P_Value = as.numeric(SUR_p)
      )
  }
}


Corr_Results <- Corr_Results %>%
  mutate(
    DELTA_BIOPSY_FDR = p.adjust(DELTA_BIOPSY_P_Value, method = "BH"),
    SURGERY_BIOPSY_FDR = p.adjust(SURGERY_BIOPSY_P_Value, method = "BH"),
    abs_DELTA_BIOPSY = abs(DELTA_BIOPSY_Corr_Value)
  )


# 4️⃣ Correlation Per Gene (Across Samples)


library(dplyr)
library(purrr)

# 1. Prepare Matrices with Explicit Row Names
# We use the gene_id column to set the row names for each matrix
gene_names <- GE_Correlation_Data_Frame$gene_id

mat_delta <- as.matrix(GE_Correlation_Data_Frame[, delta_cols])
rownames(mat_delta) <- gene_names

mat_bio <- as.matrix(GE_Correlation_Data_Frame[, BIOPSY_cols])
rownames(mat_bio) <- gene_names

mat_surg <- as.matrix(GE_Correlation_Data_Frame[, SURGERY_cols])
rownames(mat_surg) <- gene_names

# 2. Iterate using Gene Names (Safe and Explicit)
Corr_Results_By_Gene <- map_dfr(gene_names, function(gene) {
  
  # Access data directly by the gene name string
  # This removes any risk of index mismatching
  v_delta <- mat_delta[gene, ]
  v_bio   <- mat_bio[gene, ]
  v_surg  <- mat_surg[gene, ]
  
  # Correlation: Delta vs Biopsy
  res_db <- tryCatch({
    # Use 'complete.obs' or na.action to handle NAs; check length for statistical validity
    if(sum(!is.na(v_delta) & !is.na(v_bio)) > 3) {
      c_test <- cor.test(v_delta, v_bio, method = "spearman", exact = FALSE, na.action = "na.exclude")
      list(val = c_test$estimate, p = c_test$p.value)
    } else { list(val = NA, p = NA) }
  }, error = function(e) list(val = NA, p = NA))
  
  # Correlation: Surgery vs Biopsy
  res_sb <- tryCatch({
    if(sum(!is.na(v_surg) & !is.na(v_bio)) > 3) {
      c_test <- cor.test(v_surg, v_bio, method = "spearman", exact = FALSE, na.action = "na.exclude")
      list(val = c_test$estimate, p = c_test$p.value)
    } else { list(val = NA, p = NA) }
  }, error = function(e) list(val = NA, p = NA))
  
  # Construct the result row
  data.frame(
    gene_id = gene,
    DELTA_BIOPSY_Corr_Value = as.numeric(res_db$val),
    DELTA_BIOPSY_P_Value = as.numeric(res_db$p),
    SURGERY_BIOPSY_Corr_Value = as.numeric(res_sb$val),
    SURGERY_BIOPSY_P_Value = as.numeric(res_sb$p),
    stringsAsFactors = FALSE
  )
})

# Filter out rows that didn't meet the sample size (length > 3) criteria
Corr_Results_By_Gene <- Corr_Results_By_Gene %>% filter(!is.na(DELTA_BIOPSY_Corr_Value))

Corr_Results_By_Gene <- Corr_Results_By_Gene %>%
  mutate(
    DELTA_BIOPSY_FDR = p.adjust(DELTA_BIOPSY_P_Value, method = "BH"),
    SURGERY_BIOPSY_FDR = p.adjust(SURGERY_BIOPSY_P_Value, method = "BH"),
    abs_DELTA_BIOPSY = abs(DELTA_BIOPSY_Corr_Value)
  )

# Corr_Results_By_Gene <- data.frame(
#   gene_id = character(),
#   DELTA_BIOPSY_Corr_Value = numeric(),
#   DELTA_BIOPSY_P_Value = numeric(),
#   SURGERY_BIOPSY_Corr_Value = numeric(),
#   SURGERY_BIOPSY_P_Value = numeric(),
#   stringsAsFactors = FALSE
# )



# gene_list <- GE_Correlation_Data_Frame$gene_id

# for (gene in gene_list) {

#   gene_row <- GE_Correlation_Data_Frame[
#     GE_Correlation_Data_Frame$gene_id == gene, ]

#   df_delta <- as.numeric(gene_row[delta_cols])
#   df_bio   <- as.numeric(gene_row[BIOPSY_cols])
#   df_surg  <- as.numeric(gene_row[SURGERY_cols])

#   # Remove NAs consistently
#   keep <- complete.cases(df_delta, df_bio)
#   df_delta <- df_delta[keep]
#   df_bio   <- df_bio[keep]

#   if(length(df_delta) > 3){

#     DELTA_BIOPSY_corr <- cor.test(
#       df_delta,
#       df_bio,
#       method = "spearman",
#       exact = FALSE
#     )

#     # Surgery vs biopsy
#     keep2 <- complete.cases(df_surg, df_bio)
#     df_surg2 <- df_surg[keep2]
#     df_bio2  <- df_bio[keep2]

#     if(length(df_surg2) > 3){
#       SURGERY_BIOPSY_corr <- cor.test(
#         df_surg2,
#         df_bio2,
#         method = "spearman",
#         exact = FALSE
#       )

#       surg_val <- SURGERY_BIOPSY_corr$estimate
#       surg_p   <- SURGERY_BIOPSY_corr$p.value
#     } else {
#       surg_val <- NA
#       surg_p   <- NA
#     }

#     Corr_Results_By_Gene <- Corr_Results_By_Gene %>%
#       add_row(
#         gene_id = gene,
#         DELTA_BIOPSY_Corr_Value = DELTA_BIOPSY_corr$estimate,
#         DELTA_BIOPSY_P_Value = DELTA_BIOPSY_corr$p.value,
#         SURGERY_BIOPSY_Corr_Value = surg_val,
#         SURGERY_BIOPSY_P_Value = surg_p
#       )
#   }
# }





# Make Scatter Plots
library(ggrepel)
library(ggplot2)
library(dplyr)


# Get Scatterplot data
Scatterplot_data <- data.frame(
  Corr_Value = c(Corr_Results$DELTA_BIOPSY_Corr_Value, Corr_Results$SURGERY_BIOPSY_Corr_Value),
  FDR = c(Corr_Results$DELTA_BIOPSY_FDR, Corr_Results$SURGERY_BIOPSY_FDR),
  Sample = rep(Corr_Results$sample_id_full, 2), 
  Condition = rep(c("DELTA_BIOPSY_Correlation", "SURGERY_BIOPSY_Correlation"), each = nrow(Corr_Results))
)


#replace 0 with -
Scatterplot_data$FDR <- replace(Scatterplot_data$FDR,
                                    Scatterplot_data$FDR  == 0, 2.225074e-308)

ggplot(data=Scatterplot_data, aes(x = Corr_Value,
                              y = -log10(FDR),
                              color = Condition)) +
        geom_point(size = 3) + 
        theme_minimal() +
        scale_color_manual(values=c("DELTA_BIOPSY_Correlation" = "blue", "SURGERY_BIOPSY_Correlation" = "red")) +
        geom_vline(xintercept = 0.3, col="red", linewidth = 1.2) +
        geom_vline(xintercept = -0.3, col="red", linewidth = 1.2)+ 
        geom_vline(xintercept = 0.5, col="orange", linewidth = 1.2) +
        geom_vline(xintercept = -0.5, col="orange", linewidth = 1.2)+ 
        geom_hline(yintercept = -log10(0.05), col="red", linewidth = 1.2, linetype='dotted') +
        labs(x = "Spearman Coefficient", y = "-log10(P Value)") +
        coord_cartesian(ylim = c(0, 325), xlim = c(-1,1))+
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5), 
          legend.position = "top",
          axis.text = element_text(colour = "black", size = (24)),
          axis.text.x = element_text(colour = "black", size = (24)),
          axis.text.y = element_text(colour = "black", size = (24)),
          axis.title = element_text(colour = "black", size = (30)),
          axis.ticks = element_line(size = 1.5),
          panel.border = element_rect(colour = "black", fill=NA, size=1))		
  



# Make Scatter Plots
library(ggrepel)
library(ggplot2)
library(dplyr)


# Get Scatterplot data
Scatterplot_data <- data.frame(
  Corr_Value = c(Corr_Results_By_Gene$DELTA_BIOPSY_Corr_Value, Corr_Results_By_Gene$SURGERY_BIOPSY_Corr_Value),
  FDR = c(Corr_Results_By_Gene$DELTA_BIOPSY_FDR, Corr_Results_By_Gene$SURGERY_BIOPSY_FDR),
  Condition = rep(c("DELTA_BIOPSY_Correlation", "SURGERY_BIOPSY_Correlation"), each = nrow(Corr_Results_By_Gene)),
  Gene = rep(Corr_Results_By_Gene$gene_id, 2)
)




#replace 0 with 1e-300
Scatterplot_data$FDR <- replace(Scatterplot_data$FDR,
                                    Scatterplot_data$FDR  == 0, 2.225074e-308)

ggplot(data=Scatterplot_data, aes(x = Corr_Value,
                              y = -log10(FDR),
                              color = Condition)) +
        geom_point(size = 3) + 
        theme_minimal() +
        scale_color_manual(values=c("DELTA_BIOPSY_Correlation" = "blue", "SURGERY_BIOPSY_Correlation" = "red")) +
        geom_vline(xintercept = 0.3, col="red", linewidth = 1.2) +
        geom_vline(xintercept = -0.3, col="red", linewidth = 1.2)+ 
        geom_vline(xintercept = 0.5, col="orange", linewidth = 1.2) +
        geom_vline(xintercept = -0.5, col="orange", linewidth = 1.2)+ 
        labs(x = "Spearman Coefficient", y = "-log10(P Value)") +
        coord_cartesian(ylim = c(0, 325), xlim = c(-1,1))+
    theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5), 
          legend.position = "top",
          axis.text = element_text(colour = "black", size = (24)),
          axis.text.x = element_text(colour = "black", size = (24)),
          axis.text.y = element_text(colour = "black", size = (24)),
          axis.title = element_text(colour = "black", size = (30)),
          axis.ticks = element_line(size = 1.5),
          panel.border = element_rect(colour = "black", fill=NA, size=1))		
  

#Box plot to look at distribution

library(EnvStats)
library(patchwork)

all_unique_gene_ids <- unique(unlist(all_gene_sets))

# Get Scatterplot data
Scatterplot_data <- data.frame(
  Corr_Value = c(Corr_Results_By_Gene$DELTA_BIOPSY_Corr_Value, Corr_Results_By_Gene$SURGERY_BIOPSY_Corr_Value),
  FDR = c(Corr_Results_By_Gene$DELTA_BIOPSY_FDR, Corr_Results_By_Gene$SURGERY_BIOPSY_FDR),
  Condition = rep(c("DELTA_BIOPSY_Correlation", "SURGERY_BIOPSY_Correlation"), each = nrow(Corr_Results_By_Gene)),
  Gene = rep(Corr_Results_By_Gene$gene_id, 2)
)




#replace 0 with 1e-300
Scatterplot_data$FDR <- replace(Scatterplot_data$FDR,
                                    Scatterplot_data$FDR  == 0, 2.225074e-308)

Boxplot_data <- Scatterplot_data %>%
  mutate(abs_Corr_Value = abs(Corr_Value))  %>%
  mutate(log10FDR = -log10(FDR)) %>% 
  filter(Gene %in% all_unique_gene_ids)

Boxplot_data <- Boxplot_data %>%
  mutate(Significance = if_else(FDR < 0.05, "Significant", "Not Significant"))

 plot1 <- ggplot(Boxplot_data, aes(x = Condition, y = abs_Corr_Value)) +
  geom_boxplot(width = 0.5, size = 1.0, alpha = 0.25, outlier.shape = NA, fill = "blue") +
  geom_dotplot(binaxis='y', binwidth = 0.0025, position=position_jitter(0.2))+ 
  theme_minimal() +
  labs(x = "Condition",
       y = "| Correlation Value |") +
        geom_hline(yintercept = 0.3, col="red", linewidth = 1.2) +
        geom_hline(yintercept = 0.5, col="orange", linewidth = 1.2)+ 
  theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5), 
          legend.position = "top",
          axis.text = element_text(colour = "black", size = (20)),
          axis.text.x = element_text(colour = "black", size = (18)),
          axis.text.y = element_text(colour = "black", size = (18)),
          axis.title = element_text(colour = "black", size = (24)),
          axis.ticks = element_line(size = 1.5),
          panel.border = element_rect(colour = "black", fill=NA, size=1)) +			
       stat_n_text(color = "black",
                size = 6,
                vjust = 0.5) + 
  scale_y_continuous(limits = c(-0.2, 1.2), 
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1.0), name="| Correlation Value | ") +
   scale_x_discrete(labels = c("Delta ~ Pre-Treat", "Post-Treat ~ Pre-Treat"))

plot2 <- ggplot(Boxplot_data, aes(x = Condition, y = log10FDR)) +
  geom_boxplot(width = 0.5, size = 1.0, alpha = 0.25, outlier.shape = NA, fill = "red") +
  geom_dotplot(binaxis='y', binwidth = 0.025, position=position_jitter(0.2))+ 
  theme_minimal() +
  labs(x = "Condition",
       y = "-log10(FDR)") +
  geom_hline(yintercept = -log10(0.05), col="red", linewidth = 1.2) +
  theme_classic() + 
    theme(plot.title = element_text(hjust = 0.5), 
          legend.position = "top",
          axis.text = element_text(colour = "black", size = (20)),
          axis.text.x = element_text(colour = "black", size = (18)),
          axis.text.y = element_text(colour = "black", size = (18)),
          axis.title = element_text(colour = "black", size = (24)),
          axis.ticks = element_line(size = 1.5),
          panel.border = element_rect(colour = "black", fill=NA, size=1)) +			
       stat_n_text(color = "black",
                size = 6,
                vjust = 0.5) + 
  scale_y_continuous(limits = c(-2, 10), 
                     breaks = c(0, 2, 4, 6, 8), name="-log10(p)") +
  scale_x_discrete(labels = c("Delta ~ Pre-Treat", "Post-Treat ~ Pre-Treat"))


plot2_bar <- ggplot(Boxplot_data, aes(x = Condition, fill = Significance)) +
  geom_bar(position = "fill", color = "black", size = 0.8) + 
  theme_classic() + 
  labs(x = "Condition",
       y = "Percentage") +
  scale_fill_manual(values = c("Significant" = "#ef8a62", "Not Significant" = "#6d8df7ff")) +
  scale_x_discrete(labels = c("Delta ~ Pre-Treat", "Post-Treat ~ Pre-Treat")) +
    scale_y_continuous(labels = scales::percent) +
  theme(plot.title = element_text(hjust = 0.5, size = 22), 
        legend.position = "bottom",
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        axis.text = element_text(colour = "black", size = 20),
        axis.title = element_text(colour = "black", size = 24),
        axis.ticks = element_line(size = 1.5),
        panel.border = element_rect(colour = "black", fill = NA, size = 1))



plot1 | plot2
plot2_bar | plot1

```



# Heatmap of Chemokines/Cytokines

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(matrixStats)
library(tibble)
library(rstatix)

# -------------------------------------------------
# 1. CHEMOIMMUNOMODULATION GENE SETS
# -------------------------------------------------
gene_categories <- list(

  "CCL28", "CCL2", "IL2", "LTA", "CX3CL1", 
                  "CXCL9", "CCL22", "CCL5", "CXCL11", 
                 "CXCL10", "IL6", "CCL20", "XCL1", "XCL2",
                 # Recommended Additions: Anti-Tumor/Th1
  "IL12A", "IL15", "IL18", "CXCR3", "CCR5",
  
  # Recommended Additions: Pro-Tumor/Suppressive
  "IL10", "IL4", "IL13", "IL11", "IL1B", "IL17A", "IL33", "IFNG"

)


genesets_of_interest <- unlist(gene_categories)

genesets_of_interest <- genesets_of_interest[genesets_of_interest %in% names(clustered_plot_df_for_merge)]

# -------------------------------------------------
# 2. LONG FORMAT
# -------------------------------------------------
plot_df <- clustered_plot_df_for_merge %>%
  dplyr::select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "Gene",
    values_to = "Score"
  ) %>%
  mutate(
    Score = as.numeric(Score),
    CIMIC_Cluster = factor(CIMIC_Cluster, levels = c("Dys-CIM", "Fun-CIM")),
    Gene = Gene %>%
      str_replace_all("_", " ")
  ) %>%
  group_by(CIMIC_Cluster, Gene) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop")

# -------------------------------------------------
# 3. STATISTICS (Fun vs Dys per gene)
# -------------------------------------------------
pval_df <- clustered_plot_df_for_merge %>%
  dplyr::select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "Gene", values_to = "Score") %>%
  group_by(Gene) %>%
  summarise(
    p = wilcox.test(Score ~ CIMIC_Cluster)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    signif = case_when(
      p.adj <= 0.0001 ~ "****",
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01 ~ "**",
      p.adj <= 0.05 ~ "*",
      TRUE ~ ""
    ),
    Gene = str_replace_all(Gene, "_", " ")
  )

# -------------------------------------------------
# 4. ORDERING
# -------------------------------------------------
program_order <- plot_df %>%
  distinct(Gene) %>%
  pull(Gene)

plot_df <- plot_df %>%
  mutate(Gene = factor(Gene, levels = rev(program_order)))

pval_df <- pval_df %>%
  mutate(Gene = factor(Gene, levels = rev(program_order)))

star_df <- plot_df %>%
  distinct(Gene) %>%
  left_join(pval_df, by = "Gene") %>%
  mutate(x = 2.6)

# -------------------------------------------------
# 5. PLOT
# -------------------------------------------------
p <- ggplot(plot_df, aes(x = CIMIC_Cluster, y = Gene, fill = mean_score)) +
  
  geom_tile(color = "black", linewidth = 0.8) +
  
  annotate(
    "text",
    x = 2.65,
    y = seq_along(levels(plot_df$Gene)),
    label = rev(star_df$signif),
    size = 10,
    fontface = "bold"
  ) +
  
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean Z-score"
  ) +
  
  scale_x_discrete(
    limits = c("Dys-CIM", "Fun-CIM"),
    labels = c("Dys-CIM" = "Dys\nCIM", "Fun-CIM" = "Fun\nCIM"),
    expand = expansion(add = c(0.3, 1.2))
  ) +
  
  labs(x = NULL, y = NULL) +
  coord_cartesian(clip = "off") +
  
  theme_void(base_size = 18) +
  theme(
    axis.text.x = element_text(face = "bold", size = 24),
    axis.text.y = element_text(face = "bold", size = 20),
    legend.position = "right",
    legend.text = element_text(face = "bold", size = 20),
    plot.margin = margin(10, 20, 10, 20)
  )

p
```





# Viral Mimicry

```{r}


# Viral Mimicry Z-score  ----------------


# Boxplot of Z-score



library(rstatix)
delta_combined_TPM_lognormnorm <- combined_TPM_lognorm


all_programs <- all_gene_sets

viral_mimicry_set <- c(

# dsDNA sensing
"CGAS","STING1","IFI16","AIM2","ZBP1",

# dsRNA sensing
"RIGI","IFIH1","DHX58","TLR3","TLR7","TLR8","TLR9",

# adaptor/signaling
"MAVS","TBK1","IRF3","IRF7",
"STAT1","STAT2",

# antiviral effectors
"IFIT1","IFIT2","IFIT3",
"OAS1","OAS2","OAS3","RNASEL",
"MX1","MX2",
"ISG15","RSAD2","BST2",
"IFI44","IFI44L","IFI6",
"EIF2AK2",

# antigen presentation
"B2M","HLA-A","HLA-B","HLA-C",
"TAP1","TAPBP","PSMB8","PSMB9","NLRC5",

# inflammatory outputs
"CXCL9","CXCL10","CXCL11","CCL5"

)



df <- clustered_plot_df
gene_names <- delta_combined_TPM_lognormnorm$gene_id

all_cols <- colnames(df)
viral_mimicry_set <-  intersect(viral_mimicry_set, all_cols)
print(viral_mimicry_set)

gene_cols <- gene_names 




zscore_safe <- function(x) {
  s <- sd(x, na.rm = TRUE)
  m <- mean(x, na.rm = TRUE)
  if (is.na(s) || s == 0) rep(0, length(x)) else (x - m) / s
}

df_z <- df %>%
  mutate(across(all_of(gene_cols), zscore_safe))



score_program_z <- function(data, genes, program_name) {

  # ensure genes actually exist in data
  genes <- intersect(genes, colnames(data))

  if (length(genes) == 0) {
    return(
      data.frame(
        sample_id = data$sample_id,
        cimic_cluster = data$cluster_assignment,
        Program = paste0(program_name, "_z"),
        Score = rep(NA_real_, nrow(data))
      )
    )
  }
  
  score_vec <- rowMeans(as.matrix(data[, genes, drop = FALSE]), na.rm = TRUE)
  
  data.frame(
    sample_id = data$sample_id,
    cimic_cluster = data$cluster_assignments,
    Program = paste0(program_name, "_z"),
    Score = score_vec
  )
}

viral_mimicry_set_zscore <- score_program_z(
  data = df_z,
  genes = viral_mimicry_set,
  program_name = "Viral_Mimicry"
)




library(tidyr)
library(dplyr)

viral_mimicry_set_zscore_wide <- viral_mimicry_set_zscore %>%
  dplyr::select(sample_id, Program, Score) %>%
  tidyr::pivot_wider(
    names_from = Program,
    values_from = Score
  )

clustered_plot_df_for_merge <- clustered_plot_df_for_merge %>%
  left_join(viral_mimicry_set_zscore_wide, by = "sample_id")




viral_mimicry_plot_df <- viral_mimicry_set_zscore %>%
  
  mutate(
    cimic_cluster = dplyr::case_when(
      cimic_cluster == 1 ~ "Fun-CIM",
      cimic_cluster == 2 ~ "Dys-CIM"
    ),
    
    # clean labels FIRST (as character)
    Program = Program %>%
      str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
      str_replace_all("_", " "),
    
    # then convert to factor with correct ordering
    Program = factor(
      Program,
      levels = unique(
        paste0("Viral_Mimicry_z") %>%
          str_remove_all("GOBP_|REACTOME_|HALLMARK_|_z") %>%
          str_replace_all("_", " ")
      )
    ), 
    cimic_cluster = factor(cimic_cluster, levels = c("Dys-CIM", "Fun-CIM"))
)

pval_df_all <- viral_mimicry_plot_df %>%
  group_by(Program) %>%
  summarise(
    n_groups = n_distinct(cimic_cluster),
    
    y.position = max(Score, na.rm = TRUE) * 1.1,
    
    pval = tryCatch({
      if (n_groups == 2) {
        wilcox.test(Score ~ cimic_cluster)$p.value
      } else if (n_groups > 2) {
        kruskal.test(Score ~ cimic_cluster)$p.value
      } else {
        NA_real_
      }
    }, error = function(e) NA_real_),
    
    p_label = ifelse(is.na(pval), "p = NA",
                     ifelse(pval < 0.001, "p < 0.001",
                            paste0("p = ", signif(pval, 3)))),
    
    .groups = "drop"
  ) %>%
  mutate(
    FDR = p.adjust(pval, method = "fdr"),
    
    fdr_label = ifelse(is.na(FDR), "FDR = NA",
                        ifelse(FDR < 0.001, "FDR < 0.001",
                               paste0("FDR = ", signif(FDR, 3))))
  ) %>% add_significance("FDR")


p_vm <- ggplot(viral_mimicry_plot_df, aes(x = cimic_cluster, y = Score, fill = cimic_cluster)) +
  geom_boxplot(width = 0.6, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 1) +
  geom_text(
    data = pval_df_all,
    aes(x = 1.5, y = y.position, label = FDR.signif),
    inherit.aes = FALSE,
    size = 12,
    fontface = "bold"
  ) +
  facet_wrap(~ Program, scales = "free_y") +
  #scale_fill_manual(values = cimic_cluster_COLORS) +
  labs(
    y = "Z-score of\nProgram Activation",
    x = NULL
  ) +
    theme_bw(base_line_size = 2) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 32),
    axis.text = element_text(colour = "black", size = 24),
    axis.title.x = element_blank(),
    axis.text.x = element_text(colour = "black", size = 24, face = "bold"), 
    axis.text.y = element_text(colour = "black", size = 24, face = "bold"),
    axis.title.y = element_text(colour = "black", size = 24, face = "bold"),
    axis.title = element_text(colour = "black", size = 24),
    legend.text = element_text(colour = "black", size = 24),
    legend.title = element_text(colour = "black", size = 24),
    legend.position = "none",
    #plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    panel.grid = element_blank(),
    panel.border = element_rect(linewidth = 2)
  )



p_vm


# Heatmap of individual Viral Mimi gemnes ------------


library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(matrixStats)
library(tibble)
library(rstatix)





viral_mimicry_gene_categories <- list(

# dsDNA sensing
"CGAS","STING1","IFI16","AIM2","ZBP1",

# dsRNA sensing
"RIGI","IFIH1","DHX58","TLR3","TLR7","TLR8","TLR9",

# adaptor/signaling
"MAVS","TBK1","IRF3","IRF7",
"STAT1","STAT2",

# antiviral effectors
"IFIT1","IFIT2","IFIT3",
"OAS1","OAS2","OAS3","RNASEL",
"MX1","MX2",
"ISG15","RSAD2","BST2",
"IFI44","IFI44L","IFI6",
"EIF2AK2",

# antigen presentation
"B2M","HLA-A","HLA-B","HLA-C",
"TAP1","TAPBP","PSMB8","PSMB9","NLRC5",

# inflammatory outputs
"CXCL9","CXCL10","CXCL11","CCL5"

)


genesets_of_interest <- unlist(viral_mimicry_gene_categories)

genesets_of_interest <- genesets_of_interest[genesets_of_interest %in% names(clustered_plot_df_for_merge)]


plot_df <- clustered_plot_df_for_merge %>%
  dplyr::select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(
    cols = -CIMIC_Cluster,
    names_to = "Gene",
    values_to = "Score"
  ) %>%
  mutate(
    Score = as.numeric(Score),
    CIMIC_Cluster = factor(CIMIC_Cluster, levels = c("Dys-CIM", "Fun-CIM")),
    Gene = Gene %>%
      str_replace_all("_", " ")
  ) %>%
  group_by(CIMIC_Cluster, Gene) %>%
  summarise(mean_score = mean(Score, na.rm = TRUE), .groups = "drop")


pval_df <- clustered_plot_df_for_merge %>%
  dplyr::select(all_of(c("CIMIC_Cluster", genesets_of_interest))) %>%
  pivot_longer(-CIMIC_Cluster, names_to = "Gene", values_to = "Score") %>%
  group_by(Gene) %>%
  summarise(
    p = wilcox.test(Score ~ CIMIC_Cluster)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p.adj = p.adjust(p, method = "BH"),
    signif = case_when(
      p.adj <= 0.0001 ~ "****",
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01 ~ "**",
      p.adj <= 0.05 ~ "*",
      TRUE ~ ""
    ),
    Gene = str_replace_all(Gene, "_", " ")
  )


program_order <- plot_df %>%
  distinct(Gene) %>%
  pull(Gene)

plot_df <- plot_df %>%
  mutate(Gene = factor(Gene, levels = rev(program_order)))

pval_df <- pval_df %>%
  mutate(Gene = factor(Gene, levels = rev(program_order)))

star_df <- plot_df %>%
  distinct(Gene) %>%
  left_join(pval_df, by = "Gene") %>%
  mutate(x = 2.6)

p <- ggplot(plot_df, aes(x = CIMIC_Cluster, y = Gene, fill = mean_score)) +
  
  geom_tile(color = "black", linewidth = 0.8) +
  
  annotate(
    "text",
    x = 2.65,
    y = seq_along(levels(plot_df$Gene)),
    label = rev(star_df$signif),
    size = 10,
    fontface = "bold"
  ) +
  
  scale_fill_gradient2(
    low = "#313695",
    mid = "white",
    high = "#A50026",
    midpoint = 0,
    name = "Mean\nZ-score\nΔ"
  ) +
  
  scale_x_discrete(
    limits = c("Dys-CIM", "Fun-CIM"),
    labels = c("Dys-CIM" = "Dys\nCIM", "Fun-CIM" = "Fun\nCIM"),
    expand = expansion(add = c(0.3, 1.2))
  ) +
  
  labs(x = NULL, y = NULL) +
  coord_cartesian(clip = "off") +
  
  theme_void(base_size = 18) +
  theme(
    axis.text.x = element_text(face = "bold", size = 24),
    axis.text.y = element_text(face = "bold", size = 16),
    legend.position = "right",
    legend.text = element_text(face = "bold", size = 20),
    plot.margin = margin(10, 20, 10, 20)
  )

p










```




