---
title: "Baseline Individual Immune-Gene Analysis"
author: "Gbadamosi Lab - IL/MG"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    code_folding: show
params:
  data_dir: "data"
  expression_file: "nki_smc_combine_TPM_lognorm.csv"
  cluster_file: "nki_smc_clusters.csv"
  output_dir: "results/baseline_individual_genes"
  save_individual_plots: true
  random_seed: 20260815
---

# Purpose

This R Markdown file compares prespecified individual immune features between
baseline Dys-CIM and Fun-CIM tumors. `Avg_MHC1` is defined as the sample-level
mean of `HLA-A`, `HLA-B`, and `HLA-C` log2(TPM + 1) expression. The contrast is
Dys-CIM minus Fun-CIM. Benjamini-Hochberg adjustment is performed once across
all unique features in `features_to_test`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 300
)

set.seed(params$random_seed)

required_packages <- c(
  "readr", "dplyr", "tidyr", "tibble", "stringr", "purrr",
  "ggplot2", "patchwork", "rlang", "limma", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0L) {
  stop(
    "Missing required R packages: ",
    paste(missing_packages, collapse = ", "),
    call. = FALSE
  )
}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(patchwork)
  library(limma)
})

CIM_COLORS <- c(
  "Dys-CIM" = "#F08A8A",
  "Fun-CIM" = "#4DB6AC"
)

CIM_PLOT_LEVELS <- c("Dys-CIM", "Fun-CIM")
CIM_MODEL_LEVELS <- c("Fun-CIM", "Dys-CIM")

output_dir <- normalizePath(
  params$output_dir,
  winslash = "/",
  mustWork = FALSE
)

table_dir <- file.path(output_dir, "tables")
figure_dir <- file.path(output_dir, "figures")
individual_plot_dir <- file.path(figure_dir, "individual")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figure_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(individual_plot_dir, recursive = TRUE, showWarnings = FALSE)
```

# Feature vector

`HAVCR2` is listed once because it occurs in both source gene sets.

```{r feature-vector}
features_to_test <- c(
  "CXCL13", "GZMB", "IFI6", "HLA-DRA", "HLA-DQA2", "HLA-DRB5",
  "PRF1", "MX1", "HLA-DRB1", "CD82", "LY6E", "IFIT3", "ISG15",
  "IFI44", "IFITM3", "ENTPD1", "OAS1", "IFI44L", "BST2", "GNLY",
  "HAVCR2", "KRT86", "ALOX5AP", "CCL3", "IFI27", "PDCD1", "VSIR",
  "LGALS9", "HHLA2", "CD274", "TIGIT", "LAG3", "PDCD1LG2", "BTLA",
  "CTLA4", "VTCN1", "Avg_MHC1"
)

mhc1_component_genes <- c("HLA-A", "HLA-B", "HLA-C")

stopifnot(
  length(features_to_test) == 37L,
  !anyDuplicated(features_to_test),
  identical(tail(features_to_test, 1L), "Avg_MHC1")
)

knitr::kable(tibble(feature = features_to_test))
```

# Helper functions

```{r helpers}
resolve_input_file <- function(data_dir, filename) {
  candidates <- unique(c(
    file.path(data_dir, filename),
    file.path(getwd(), data_dir, filename),
    file.path(getwd(), filename)
  ))

  existing <- candidates[file.exists(candidates)]

  if (length(existing) == 0L) {
    stop(
      "Could not find input file '", filename, "'. Checked: ",
      paste(candidates, collapse = "; "),
      call. = FALSE
    )
  }

  normalizePath(existing[[1]], winslash = "/", mustWork = TRUE)
}

assert_true <- function(condition, message) {
  if (!isTRUE(condition)) stop(message, call. = FALSE)
}

format_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 0.001) return(format(p, scientific = TRUE, digits = 2))
  sprintf("%.3f", p)
}

make_group_boxplot <- function(data, feature_name, p_value) {
  plot_data <- data |>
    mutate(CIM_group = factor(CIM_group, levels = CIM_PLOT_LEVELS))

  observed_values <- plot_data$expression
  observed_values <- observed_values[is.finite(observed_values)]

  assert_true(
    length(observed_values) > 0L,
    paste0("No finite values were available for ", feature_name, ".")
  )

  y_min <- min(observed_values)
  y_max <- max(observed_values)
  y_range <- y_max - y_min

  if (!is.finite(y_range) || y_range <= 0) {
    y_range <- max(abs(c(y_min, y_max)), 1)
  }

  label_y <- y_max + 0.10 * y_range
  upper_y <- y_max + 0.24 * y_range
  lower_y <- y_min - 0.04 * y_range

  ggplot(plot_data, aes(x = CIM_group, y = expression, fill = CIM_group)) +
    geom_boxplot(
      width = 0.58,
      outlier.shape = NA,
      linewidth = 0.35,
      color = "black"
    ) +
    geom_jitter(
      width = 0.10,
      height = 0,
      size = 1.35,
      alpha = 0.90,
      color = "black"
    ) +
    scale_fill_manual(values = CIM_COLORS, drop = FALSE) +
    annotate(
      "text",
      x = 1.5,
      y = label_y,
      label = paste0("P = ", format_p(p_value)),
      hjust = 0.5,
      vjust = 0,
      size = 3.0
    ) +
    coord_cartesian(ylim = c(lower_y, upper_y), clip = "off") +
    labs(
      title = feature_name,
      x = NULL,
      y = expression("Baseline log"[2] * "(TPM + 1)")
    ) +
    theme_classic(base_size = 9) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 10),
      axis.title.y = element_text(size = 9),
      axis.text.x = element_text(size = 8, color = "black"),
      axis.text.y = element_text(size = 8, color = "black"),
      axis.line = element_line(linewidth = 0.35),
      axis.ticks = element_line(linewidth = 0.35),
      legend.position = "none",
      plot.margin = margin(9, 8, 5.5, 5.5)
    )
}

save_plot_pdf_png <- function(plot_object, output_stem, width, height, dpi = 600) {
  ggsave(
    paste0(output_stem, ".pdf"),
    plot_object,
    width = width,
    height = height,
    units = "in",
    bg = "white"
  )

  ggsave(
    paste0(output_stem, ".png"),
    plot_object,
    width = width,
    height = height,
    units = "in",
    dpi = dpi,
    bg = "white"
  )
}
```

# Import data

```{r import-data}
expression_path <- resolve_input_file(params$data_dir, params$expression_file)
cluster_path <- resolve_input_file(params$data_dir, params$cluster_file)

expression_raw <- readr::read_csv(
  expression_path,
  show_col_types = FALSE,
  name_repair = "minimal"
)

sample_columns <- names(expression_raw)[
  stringr::str_detect(names(expression_raw), "_(B|S)$")
]

assert_true(
  length(sample_columns) > 0L,
  "No expression columns ending in _B or _S were detected."
)

non_sample_columns <- setdiff(names(expression_raw), sample_columns)
preferred_gene_columns <- c(
  "gene", "Gene", "gene_id", "Gene_ID", "symbol", "SYMBOL",
  "gene_symbol", "GeneSymbol", "X1"
)

gene_column <- intersect(preferred_gene_columns, non_sample_columns)

if (length(gene_column) == 0L) {
  assert_true(
    length(non_sample_columns) == 1L,
    paste0(
      "Could not uniquely determine the gene-symbol column. Non-sample columns: ",
      paste(non_sample_columns, collapse = ", ")
    )
  )
  gene_column <- non_sample_columns[[1]]
} else {
  gene_column <- gene_column[[1]]
}

expression_clean <- expression_raw |>
  transmute(
    gene = stringr::str_trim(as.character(.data[[gene_column]])),
    across(all_of(sample_columns), ~ suppressWarnings(as.numeric(.x)))
  ) |>
  filter(!is.na(gene), gene != "") |>
  group_by(gene) |>
  summarise(
    across(all_of(sample_columns), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

expr_matrix <- expression_clean |>
  column_to_rownames("gene") |>
  as.matrix()

storage.mode(expr_matrix) <- "double"

assert_true(!anyDuplicated(rownames(expr_matrix)), "Duplicated genes remain.")
assert_true(!anyNA(expr_matrix), "Expression contains missing or nonnumeric values.")

expression_metadata <- tibble(expression_sample_id = colnames(expr_matrix)) |>
  mutate(
    timepoint_code = stringr::str_extract(expression_sample_id, "[BS]$"),
    timepoint = recode(timepoint_code, B = "Biopsy", S = "Surgery"),
    patient_id = stringr::str_remove(expression_sample_id, "_[BS]$")
  )

cluster_raw <- readr::read_csv(
  cluster_path,
  show_col_types = FALSE,
  col_types = cols(
    sample_id = col_character(),
    cluster_assignments = col_integer(),
    cohort = col_character()
  )
)

required_cluster_columns <- c("sample_id", "cluster_assignments", "cohort")
assert_true(
  all(required_cluster_columns %in% names(cluster_raw)),
  "Cluster input must contain sample_id, cluster_assignments, and cohort."
)

cluster_metadata <- cluster_raw |>
  transmute(
    patient_id = stringr::str_trim(sample_id),
    cluster_assignment = as.integer(cluster_assignments),
    cohort = stringr::str_to_upper(stringr::str_trim(cohort)),
    CIM_group = case_when(
      cluster_assignment == 1L ~ "Dys-CIM",
      cluster_assignment == 2L ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )

assert_true(!anyNA(cluster_metadata$CIM_group), "Invalid cluster assignment.")
assert_true(!anyDuplicated(cluster_metadata$patient_id), "Duplicated patient ID.")

baseline_metadata <- expression_metadata |>
  filter(timepoint == "Biopsy") |>
  left_join(cluster_metadata, by = "patient_id") |>
  arrange(match(expression_sample_id, colnames(expr_matrix)))

assert_true(
  !anyNA(baseline_metadata$CIM_group),
  paste0(
    "Baseline samples without a cluster match: ",
    paste(
      baseline_metadata$expression_sample_id[is.na(baseline_metadata$CIM_group)],
      collapse = ", "
    )
  )
)

baseline_metadata <- baseline_metadata |>
  mutate(CIM_group = factor(CIM_group, levels = CIM_MODEL_LEVELS))

sample_summary <- baseline_metadata |>
  count(CIM_group, name = "n") |>
  mutate(CIM_group = as.character(CIM_group))

knitr::kable(sample_summary)
```

# Feature coverage and matrix

```{r feature-coverage}
individual_genes <- setdiff(features_to_test, "Avg_MHC1")
required_expression_genes <- unique(c(individual_genes, mhc1_component_genes))

feature_coverage <- tibble(
  gene = required_expression_genes,
  present_in_expression = gene %in% rownames(expr_matrix)
)

knitr::kable(feature_coverage)
readr::write_csv(
  feature_coverage,
  file.path(table_dir, "Baseline_feature_coverage.csv")
)

assert_true(
  all(feature_coverage$present_in_expression),
  paste0(
    "Missing required genes: ",
    paste(feature_coverage$gene[!feature_coverage$present_in_expression], collapse = ", ")
  )
)

baseline_expression_matrix <- expr_matrix[
  ,
  baseline_metadata$expression_sample_id,
  drop = FALSE
]

baseline_expression_matrix <- rbind(
  baseline_expression_matrix,
  Avg_MHC1 = colMeans(
    baseline_expression_matrix[mhc1_component_genes, , drop = FALSE],
    na.rm = FALSE
  )
)

feature_matrix <- baseline_expression_matrix[
  features_to_test,
  ,
  drop = FALSE
]
```

# Baseline limma analysis

The empirical Bayes model is fitted to the complete transcriptome plus
`Avg_MHC1`. Raw P values are then extracted for the 37 prespecified features,
and BH adjustment is applied across those 37 tests only.

```{r baseline-limma}
group_vector <- factor(
  baseline_metadata$CIM_group,
  levels = CIM_MODEL_LEVELS
)

design <- model.matrix(~ 0 + group_vector)
colnames(design) <- c("Fun_CIM", "Dys_CIM")

fit <- limma::lmFit(baseline_expression_matrix, design)
contrast_matrix <- limma::makeContrasts(
  Dys_minus_Fun = Dys_CIM - Fun_CIM,
  levels = design
)
fit <- limma::contrasts.fit(fit, contrast_matrix)
fit <- limma::eBayes(fit, trend = TRUE, robust = TRUE)

limma_all <- limma::topTable(
  fit,
  coef = "Dys_minus_Fun",
  number = Inf,
  sort.by = "none",
  adjust.method = "none",
  confint = TRUE
) |>
  rownames_to_column("feature") |>
  as_tibble()

feature_expression_long <- feature_matrix |>
  t() |>
  as.data.frame(check.names = FALSE) |>
  rownames_to_column("expression_sample_id") |>
  as_tibble() |>
  pivot_longer(
    cols = all_of(features_to_test),
    names_to = "feature",
    values_to = "expression"
  ) |>
  left_join(
    baseline_metadata |>
      select(expression_sample_id, patient_id, cohort, CIM_group),
    by = "expression_sample_id"
  ) |>
  mutate(
    feature = factor(feature, levels = features_to_test),
    CIM_group = factor(CIM_group, levels = CIM_PLOT_LEVELS)
  )

group_statistics <- feature_expression_long |>
  group_by(feature, CIM_group) |>
  summarise(
    n = sum(is.finite(expression)),
    mean = mean(expression, na.rm = TRUE),
    sd = sd(expression, na.rm = TRUE),
    median = median(expression, na.rm = TRUE),
    IQR = IQR(expression, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    feature = as.character(feature),
    CIM_group = as.character(CIM_group)
  ) |>
  pivot_wider(
    names_from = CIM_group,
    values_from = c(n, mean, sd, median, IQR),
    names_glue = "{.value}_{CIM_group}"
  )

baseline_statistics <- limma_all |>
  filter(feature %in% features_to_test) |>
  transmute(
    feature,
    difference_Dys_minus_Fun = logFC,
    conf_low = CI.L,
    conf_high = CI.R,
    average_expression = AveExpr,
    moderated_t = t,
    p_value = P.Value,
    B_statistic = B
  ) |>
  mutate(
    feature = factor(feature, levels = features_to_test)
  ) |>
  arrange(feature) |>
  mutate(
    feature = as.character(feature),
    p_adjust_BH = p.adjust(p_value, method = "BH"),
    significant_BH = !is.na(p_adjust_BH) & p_adjust_BH < 0.05,
    direction = case_when(
      significant_BH & difference_Dys_minus_Fun > 0 ~ "Dys-CIM higher",
      significant_BH & difference_Dys_minus_Fun < 0 ~ "Fun-CIM higher",
      TRUE ~ "NS"
    )
  ) |>
  left_join(group_statistics, by = "feature") |>
  select(
    feature,
    n_Dys_CIM = `n_Dys-CIM`,
    mean_Dys_CIM = `mean_Dys-CIM`,
    sd_Dys_CIM = `sd_Dys-CIM`,
    median_Dys_CIM = `median_Dys-CIM`,
    IQR_Dys_CIM = `IQR_Dys-CIM`,
    n_Fun_CIM = `n_Fun-CIM`,
    mean_Fun_CIM = `mean_Fun-CIM`,
    sd_Fun_CIM = `sd_Fun-CIM`,
    median_Fun_CIM = `median_Fun-CIM`,
    IQR_Fun_CIM = `IQR_Fun-CIM`,
    difference_Dys_minus_Fun,
    conf_low,
    conf_high,
    average_expression,
    moderated_t,
    p_value,
    p_adjust_BH,
    significant_BH,
    direction,
    B_statistic
  )

assert_true(
  nrow(baseline_statistics) == length(features_to_test),
  "Not all requested features were returned by limma."
)

knitr::kable(baseline_statistics, digits = 4)

readr::write_csv(
  baseline_statistics,
  file.path(table_dir, "Baseline_individual_gene_statistics.csv")
)

readr::write_csv(
  feature_expression_long,
  file.path(table_dir, "Baseline_individual_gene_expression.csv")
)
```

# Individual boxplots

The displayed `P` values are BH-adjusted across all 37 features. No legend or
caption is added to the plots.

```{r individual-boxplots, fig.width=12, fig.height=20.5}
baseline_plots <- setNames(features_to_test, features_to_test) |>
  purrr::map(function(current_feature) {
    current_p <- baseline_statistics |>
      filter(feature == current_feature) |>
      pull(p_adjust_BH)

    make_group_boxplot(
      feature_expression_long |>
        filter(feature == current_feature),
      feature_name = current_feature,
      p_value = current_p
    )
  })

if (isTRUE(params$save_individual_plots)) {
  purrr::iwalk(baseline_plots, function(plot_object, feature_name) {
    save_plot_pdf_png(
      plot_object,
      file.path(individual_plot_dir, paste0(feature_name, "_baseline")),
      width = 2.5,
      height = 2.9
    )
  })
}

baseline_multipanel <- patchwork::wrap_plots(
  baseline_plots,
  ncol = 5
)

print(baseline_multipanel)

save_plot_pdf_png(
  baseline_multipanel,
  file.path(figure_dir, "Baseline_individual_gene_multipanel"),
  width = 12,
  height = 20.5
)

ggsave(
  file.path(figure_dir, "Baseline_individual_gene_multipanel.tiff"),
  baseline_multipanel,
  width = 12,
  height = 20.5,
  units = "in",
  dpi = 600,
  compression = "lzw",
  bg = "white"
)
```

# Reproducibility

```{r session-info}
writeLines(
  capture.output(sessionInfo()),
  file.path(output_dir, "sessionInfo.txt")
)

sessionInfo()
```
