---
title: "CIM ssGSEA Kaplan–Meier Plot Generator"
author: "Gbadamosi Lab - IL/MG"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
params:
  data_dir: "C:/Users/yasmi/OneDrive - University of Florida/Gbadamosi Lab/Iasmim Lopes/01.CIMIC/CIM_ssGSEA"
  scores_file: "CIM_ssGSEA_survival_adjusted_cont_median_corr060/02_Scores/sample_level_ssGSEA_scores.csv"
  metabric_clinical_file: "data/METABRIC_clinical_data.xlsx"
  scanb_clinical_file: "data/SCANB_clinical_harmonized_OS_DRFi_RFi.rds"
  output_dir: "CIM_ssGSEA_KM_results"
  run_batch: true
  save_png: true
  save_pdf: true
  width: 6
  height: 5
  dpi: 600
  break_x_by: 48
  min_group_n: 5
  min_events: 3
  alpha: 0.05
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE,
                     fig.width = 6, fig.height = 5, dpi = 600)
options(stringsAsFactors = FALSE, scipen = 999)
```

# Purpose

This standalone R Markdown document creates publication-quality median-split
Kaplan–Meier plots from the Fun-CIM and Dys-CIM scores already calculated by
the dedicated ssGSEA survival pipeline. It does **not** recalculate ssGSEA.

The primary continuous-score Cox models remain the preferred inferential
analysis. These median-split Kaplan–Meier analyses are secondary descriptive
visualizations. For each cohort and signature, one median cutoff is calculated
from all matched patients with a finite score before endpoint-specific survival
missingness is removed. Consequently, a patient's High/Low group remains fixed
across endpoints within the same cohort and signature.

The central function is:

```r
plot_single_ssgsea_km(
  signature = "Dys-CIM",
  dataset = "SCANB",
  endpoint = "RFi",
  title = "Dys-CIM — SCAN-B RFI",
  save_plot = TRUE,
  display = TRUE,
  width = 7,
  height = 6.6,
  dpi = 600
)
```

# Packages and directories

```{r packages-and-directories}
required_packages <- c(
  "dplyr", "tidyr", "purrr", "tibble", "stringr", "readr", "readxl",
  "survival", "broom", "ggplot2", "survminer", "grid", "scales",
  "openxlsx"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop(
    "Missing required packages: ", paste(missing_packages, collapse = ", "),
    ". Install them before knitting this document."
  )
}

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(stringr)
  library(ggplot2)
  library(survival)
  library(survminer)
})

data_dir <- normalizePath(params$data_dir, winslash = "/", mustWork = FALSE)
input_path <- function(x) {
  if (grepl("^[A-Za-z]:|^/", x)) x else file.path(data_dir, x)
}

files <- c(
  scores = input_path(params$scores_file),
  metabric_clinical = input_path(params$metabric_clinical_file),
  scanb_clinical = input_path(params$scanb_clinical_file)
)
missing_files <- names(files)[!file.exists(files)]
if (length(missing_files)) {
  stop(
    "Missing input file(s): ",
    paste(paste0(missing_files, " = ", files[missing_files]), collapse = "; ")
  )
}

output_dir <- if (grepl("^[A-Za-z]:|^/", params$output_dir)) {
  params$output_dir
} else {
  file.path(data_dir, params$output_dir)
}

subdirs <- setNames(
  file.path(output_dir, c("01_QC", "02_Statistics", "03_Individual_KM",
                          "04_Batch_Manifest", "05_Reproducibility")),
  c("qc", "statistics", "plots", "batch", "reproducibility")
)
invisible(lapply(c(output_dir, subdirs), dir.create, recursive = TRUE,
                 showWarnings = FALSE))

input_manifest <- tibble(
  input = names(files),
  path = unname(files),
  exists = file.exists(files),
  size_bytes = unname(file.info(files)$size),
  modified = format(unname(file.info(files)$mtime), "%Y-%m-%d %H:%M:%S")
)
readr::write_csv(input_manifest,
                 file.path(subdirs["qc"], "input_file_manifest.csv"))
```

# Endpoint and aesthetic configuration

```{r configuration}
endpoint_config <- list(
  METABRIC = tibble::tribble(
    ~endpoint, ~time_col,     ~event_col,   ~y_label,
    "OS",      "OS_MONTHS",  "OS_EVENT",   "OS probability",
    "DSS",     "DSS_MONTHS", "DSS_EVENT",  "DSS probability",
    "RFI",     "RFI_MONTHS", "RFI_EVENT",  "RFI probability"
  ),
  SCANB = tibble::tribble(
    ~endpoint, ~time_col,      ~event_col,    ~y_label,
    "OS",      "OS_MONTHS",   "OS_EVENT",    "OS probability",
    "DRFi",    "DRFI_MONTHS", "DRFI_EVENT",  "DRFi probability",
    "RFi",     "RFI_MONTHS",  "RFI_EVENT",   "RFI probability"
  )
)

dataset_aesthetics <- tibble::tribble(
  ~dataset,    ~display_name, ~y_min, ~y_max, ~pval_y,
  "METABRIC", "METABRIC",     0.00,   1.00,    0.20,
  "SCANB",    "SCAN-B",       0.75,   1.00,    0.80
)

group_palette <- c("Low" = "#0072B2", "High" = "#CC0000")

analysis_parameters <- tibble(
  parameter = c(
    "group_definition", "median_population", "cox_contrast", "cox_ties",
    "curve_test", "plot_width_inches", "plot_height_inches", "png_dpi",
    "x_axis_break_months", "METABRIC_y_axis", "SCANB_y_axis"
  ),
  value = c(
    "High >= median; Low < median",
    "one fixed median per cohort and signature before endpoint filtering",
    "High versus Low", "efron", "two-sided log-rank",
    params$width, params$height, params$dpi, params$break_x_by,
    "0.00 to 1.00", "0.75 to 1.00"
  )
)
readr::write_csv(analysis_parameters,
                 file.path(subdirs["qc"], "analysis_parameters.csv"))
```

# Helpers

```{r helpers}
safe_numeric <- function(x) suppressWarnings(as.numeric(as.character(x)))
clean_id <- function(x) stringr::str_squish(as.character(x))

get_first_available <- function(df, candidates, numeric = FALSE) {
  hits <- candidates[candidates %in% names(df)]
  if (!length(hits)) {
    value <- if (numeric) rep(NA_real_, nrow(df)) else rep(NA_character_, nrow(df))
    return(list(value = value, source = NA_character_))
  }
  usable <- vapply(hits, function(candidate) {
    value <- df[[candidate]]
    if (numeric) value <- safe_numeric(value)
    any(!is.na(value) & str_squish(as.character(value)) != "")
  }, logical(1))
  source <- if (any(usable)) hits[which(usable)[1]] else hits[1]
  value <- df[[source]]
  if (numeric) value <- safe_numeric(value)
  list(value = value, source = source)
}

standardize_event <- function(x, variable, dataset) {
  if (is.logical(x)) return(as.integer(x))
  z <- str_to_lower(str_squish(as.character(x)))
  event_map <- c(
    "0" = 0, "1" = 1, "false" = 0, "true" = 1, "no" = 0, "yes" = 1,
    "alive" = 0, "living" = 0, "censored" = 0,
    "dead" = 1, "deceased" = 1, "event" = 1,
    "0:living" = 0, "1:deceased" = 1,
    "0:disease free" = 0, "1:recurred" = 1,
    "died of disease" = 1, "died of other causes" = 0
  )
  invalid <- setdiff(unique(z[!is.na(z) & z != ""]), names(event_map))
  if (length(invalid)) {
    stop(dataset, " ", variable, " has unmapped event values: ",
         paste(invalid, collapse = ", "))
  }
  unname(event_map[z])
}

format_p_value <- function(p, digits = 4) {
  if (!is.finite(p)) return("p = NA")
  threshold <- 10^(-digits)
  if (p < threshold) {
    paste0("p < ", format(threshold, scientific = FALSE, nsmall = digits))
  } else {
    paste0("p = ", formatC(p, format = "f", digits = digits))
  }
}

safe_file_tag <- function(x) {
  x %>% str_replace_all("—", "-") %>%
    str_replace_all("[^A-Za-z0-9._-]+", "_") %>%
    str_replace_all("_+", "_") %>%
    str_remove("^_") %>% str_remove("_$")
}

save_ggsurvplot <- function(km_object, filename, width, height, dpi, type) {
  if (type == "png") {
    grDevices::png(filename, width = width * dpi, height = height * dpi,
                   units = "px", res = dpi, bg = "white")
  } else if (type == "pdf") {
    if (capabilities("cairo")) {
      grDevices::cairo_pdf(filename, width = width, height = height,
                           onefile = TRUE, bg = "white")
    } else {
      grDevices::pdf(filename, width = width, height = height,
                     onefile = TRUE, bg = "white")
    }
  } else {
    stop("Unknown output type: ", type)
  }
  on.exit(grDevices::dev.off(), add = TRUE)
  print_ggsurvplot_clean(km_object, newpage = TRUE)
  invisible(filename)
}

# survminer can emit a harmless ggplot2 warning about an unused colour label
# when the composite plot includes a risk table. Muffle only that known warning;
# all other warnings remain visible.
print_ggsurvplot_clean <- function(km_object, newpage = TRUE) {
  withCallingHandlers(
    print(km_object, newpage = newpage),
    warning = function(w) {
      if (grepl("Ignoring unknown labels", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
  invisible(km_object)
}
```

# Read scores and clinical data

```{r read-and-harmonize-inputs}
scores_raw <- readr::read_csv(files[["scores"]], show_col_types = FALSE,
                              name_repair = "minimal")
required_score_columns <- c("cohort", "sample_id", "signature", "ssgsea_score")
missing_score_columns <- setdiff(required_score_columns, names(scores_raw))
if (length(missing_score_columns)) {
  stop("The score file is missing: ", paste(missing_score_columns, collapse = ", "))
}

scores <- scores_raw %>% transmute(
  dataset = case_when(
    str_to_upper(str_squish(as.character(cohort))) == "METABRIC" ~ "METABRIC",
    str_to_upper(str_replace_all(str_squish(as.character(cohort)), "-", "")) == "SCANB" ~ "SCANB",
    TRUE ~ NA_character_
  ),
  sample_id = clean_id(sample_id),
  signature = case_when(
    str_to_lower(str_squish(as.character(signature))) == "fun-cim" ~ "Fun-CIM",
    str_to_lower(str_squish(as.character(signature))) == "dys-cim" ~ "Dys-CIM",
    TRUE ~ NA_character_
  ),
  ssgsea_score = safe_numeric(ssgsea_score),
  cim_z = if ("cim_z" %in% names(scores_raw)) safe_numeric(scores_raw$cim_z) else NA_real_
)

invalid_scores <- scores %>% filter(
  is.na(dataset) | is.na(signature) | is.na(sample_id) | sample_id == "" |
    !is.finite(ssgsea_score)
)
if (nrow(invalid_scores)) {
  readr::write_csv(invalid_scores,
                   file.path(subdirs["qc"], "invalid_score_rows.csv"))
  stop(nrow(invalid_scores), " invalid rows were found in the ssGSEA score file.")
}

duplicate_scores <- scores %>% count(dataset, sample_id, signature, name = "n") %>%
  filter(n > 1L)
readr::write_csv(duplicate_scores,
                 file.path(subdirs["qc"], "duplicate_score_rows.csv"))
if (nrow(duplicate_scores)) {
  stop("Duplicate dataset × sample_id × signature score rows were detected.")
}
if (!all(c("Fun-CIM", "Dys-CIM") %in% unique(scores$signature))) {
  stop("Both Fun-CIM and Dys-CIM must be present in the score file.")
}

metabric_raw <- readxl::read_xlsx(files[["metabric_clinical"]]) %>% as.data.frame()
scanb_raw <- readRDS(files[["scanb_clinical"]]) %>% as.data.frame()

harmonize_km_clinical <- function(df, dataset) {
  id_candidates <- if (dataset == "METABRIC") {
    c("PATIENT_ID", "Patient_ID", "patient_id")
  } else {
    c("GEX.assay", "GEX_assay", "sample_id")
  }
  id <- get_first_available(df, id_candidates)
  if (is.na(id$source)) stop(dataset, " clinical ID column was not found.")

  alias_map <- if (dataset == "METABRIC") {
    list(
      OS_MONTHS = c("OS_MONTHS", "OS_months"),
      OS_EVENT = c("OS_EVENT", "OS_event", "OS_STATUS"),
      DSS_MONTHS = c("DSS_MONTHS", "DSS_months"),
      DSS_EVENT = c("DSS_EVENT", "DSS_event", "VITAL_STATUS"),
      RFI_MONTHS = c("RFI_MONTHS", "RFI_months", "RFS_MONTHS"),
      RFI_EVENT = c("RFI_EVENT", "RFI_event", "RFS_STATUS")
    )
  } else {
    list(
      OS_MONTHS = c("OS_MONTHS", "OS_months"),
      OS_EVENT = c("OS_EVENT", "OS_event"),
      DRFI_MONTHS = c("DRFI_MONTHS", "DRFi_months", "DRFI_months"),
      DRFI_EVENT = c("DRFI_EVENT", "DRFi_event", "DRFI_event"),
      RFI_MONTHS = c("RFI_MONTHS", "RFi_months", "RFI_months"),
      RFI_EVENT = c("RFI_EVENT", "RFi_event", "RFI_event")
    )
  }

  out <- tibble(sample_id = clean_id(id$value))
  audit <- tibble(
    dataset, harmonized_column = "sample_id", source_column = id$source,
    nonmissing_n = sum(!is.na(out$sample_id) & out$sample_id != "")
  )
  for (target in names(alias_map)) {
    found <- get_first_available(
      df, alias_map[[target]], numeric = str_ends(target, "_MONTHS")
    )
    out[[target]] <- found$value
    audit <- bind_rows(audit, tibble(
      dataset, harmonized_column = target,
      source_column = ifelse(is.na(found$source), "not_available", found$source),
      nonmissing_n = sum(!is.na(found$value))
    ))
  }
  for (event_col in names(alias_map)[str_ends(names(alias_map), "_EVENT")]) {
    out[[event_col]] <- standardize_event(out[[event_col]], event_col, dataset)
  }
  if (any(is.na(out$sample_id) | out$sample_id == "") ||
      anyDuplicated(out$sample_id)) {
    duplicate_ids <- out %>% count(sample_id, name = "n") %>%
      filter(is.na(sample_id) | sample_id == "" | n > 1L)
    readr::write_csv(duplicate_ids,
      file.path(subdirs["qc"], paste0(dataset, "_duplicate_clinical_ids.csv")))
    stop(dataset, " clinical data contain duplicated sample IDs.")
  }
  attr(out, "source_audit") <- audit
  out
}

metabric_clinical <- harmonize_km_clinical(metabric_raw, "METABRIC")
scanb_clinical <- harmonize_km_clinical(scanb_raw, "SCANB")
clinical_source_audit <- bind_rows(
  attr(metabric_clinical, "source_audit"),
  attr(scanb_clinical, "source_audit")
)
readr::write_csv(clinical_source_audit,
                 file.path(subdirs["qc"], "clinical_source_audit.csv"))

clinical_long <- bind_rows(
  mutate(metabric_clinical, dataset = "METABRIC", .before = 1),
  mutate(scanb_clinical, dataset = "SCANB", .before = 1)
)

score_clinical_long <- scores %>%
  left_join(clinical_long, by = c("dataset", "sample_id"),
            relationship = "many-to-one")

join_audit <- scores %>% distinct(dataset, sample_id) %>%
  left_join(clinical_long %>% distinct(dataset, sample_id) %>%
              mutate(clinical_match = TRUE),
            by = c("dataset", "sample_id"), relationship = "many-to-one") %>%
  mutate(clinical_match = coalesce(clinical_match, FALSE))
readr::write_csv(join_audit,
                 file.path(subdirs["qc"], "score_clinical_join_audit.csv"))
readr::write_csv(filter(join_audit, !clinical_match),
                 file.path(subdirs["qc"], "unmatched_score_sample_ids.csv"))
if (any(!join_audit$clinical_match)) {
  stop(sum(!join_audit$clinical_match),
       " score samples did not match their cohort-specific clinical file.")
}

median_cutoff_table <- score_clinical_long %>%
  filter(is.finite(ssgsea_score)) %>%
  group_by(dataset, signature) %>%
  summarise(
    median_cutoff = median(ssgsea_score),
    score_n = n(),
    score_min = min(ssgsea_score),
    score_max = max(ssgsea_score),
    .groups = "drop"
  )
readr::write_csv(median_cutoff_table,
                 file.path(subdirs["qc"], "median_cutoffs.csv"))

score_clinical_long <- score_clinical_long %>%
  left_join(select(median_cutoff_table, dataset, signature, median_cutoff),
            by = c("dataset", "signature"), relationship = "many-to-one") %>%
  mutate(
    median_group = if_else(ssgsea_score >= median_cutoff, "High", "Low"),
    median_group = factor(median_group, levels = c("Low", "High"))
  )
```

# Manual and batch KM function

```{r km-function}
plot_single_ssgsea_km <- function(
    signature,
    dataset,
    endpoint,
    title = NULL,
    median_cutoff = NULL,
    save_plot = TRUE,
    display = TRUE,
    output_dir = subdirs[["plots"]],
    filename = NULL,
    width = 6,
    height = 5,
    dpi = 600,
    max_time = NULL,
    break_x_by = params$break_x_by
) {
  signature <- match.arg(signature, c("Fun-CIM", "Dys-CIM"))
  dataset <- match.arg(dataset, c("METABRIC", "SCANB"))

  valid_endpoints <- endpoint_config[[dataset]]$endpoint
  if (!endpoint %in% valid_endpoints) {
    stop(
      "Endpoint '", endpoint, "' is not available for ", dataset,
      ". Valid options: ", paste(valid_endpoints, collapse = ", "), "."
    )
  }

  endpoint_row <- endpoint_config[[dataset]] %>% filter(.data$endpoint == .env$endpoint)
  aesthetic_row <- dataset_aesthetics %>% filter(.data$dataset == .env$dataset)
  time_col <- endpoint_row$time_col[[1]]
  event_col <- endpoint_row$event_col[[1]]

  base_data <- score_clinical_long %>%
    filter(.data$dataset == .env$dataset,
           .data$signature == .env$signature)
  if (!nrow(base_data)) stop("No score data were found for ", dataset, " ", signature, ".")

  cutoff_source <- if (is.null(median_cutoff)) "Prespecified cohort-signature median" else "User-defined cutoff"
  cutoff <- if (is.null(median_cutoff)) {
    unique(base_data$median_cutoff)
  } else {
    as.numeric(median_cutoff)
  }
  if (length(cutoff) != 1L || !is.finite(cutoff)) {
    stop("The selected median cutoff must be one finite numeric value.")
  }

  df_plot <- base_data %>% transmute(
    dataset, sample_id, signature, ssgsea_score, cim_z,
    time_var = safe_numeric(.data[[time_col]]),
    event_var = standardize_event(.data[[event_col]], event_col, dataset),
    ssGSEA_group = factor(
      if_else(ssgsea_score >= cutoff, "High", "Low"),
      levels = c("Low", "High")
    )
  ) %>% filter(
    is.finite(time_var), time_var > 0,
    event_var %in% c(0, 1),
    is.finite(ssgsea_score), !is.na(ssGSEA_group)
  )

  group_counts <- df_plot %>% group_by(ssGSEA_group) %>% summarise(
    n = n(), events = sum(event_var == 1), censored = sum(event_var == 0),
    .groups = "drop"
  ) %>% complete(
    ssGSEA_group = factor(c("Low", "High"), levels = c("Low", "High")),
    fill = list(n = 0L, events = 0L, censored = 0L)
  )
  if (any(group_counts$n < params$min_group_n)) {
    stop("At least one ssGSEA group has fewer than ", params$min_group_n,
         " endpoint-eligible patients.")
  }
  if (sum(group_counts$events) < params$min_events) {
    stop("Fewer than ", params$min_events, " events are available for this plot.")
  }

  cox_model <- survival::coxph(
    survival::Surv(time_var, event_var) ~ ssGSEA_group,
    data = df_plot, ties = "efron", x = TRUE, y = TRUE
  )
  cox_result <- broom::tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term == "ssGSEA_groupHigh")
  if (nrow(cox_result) != 1L) stop("The High-versus-Low Cox coefficient was not estimable.")

  logrank_model <- survival::survdiff(
    survival::Surv(time_var, event_var) ~ ssGSEA_group,
    data = df_plot, rho = 0
  )
  logrank_df <- length(logrank_model$n) - 1L
  logrank_p <- stats::pchisq(logrank_model$chisq, df = logrank_df,
                            lower.tail = FALSE)
  km_fit <- survival::survfit(
    survival::Surv(time_var, event_var) ~ ssGSEA_group,
    data = df_plot
  )

  cox_p_text <- format_p_value(cox_result$p.value[[1]], digits = 4)
  subtitle_text <- sprintf(
    "Cox HR (High vs Low) = %.2f (%.2f–%.2f), Wald %s",
    cox_result$estimate[[1]], cox_result$conf.low[[1]],
    cox_result$conf.high[[1]], cox_p_text
  )
  logrank_text <- paste0("Log-rank ", format_p_value(logrank_p, digits = 4))

  if (is.null(title)) {
    title <- paste0(signature, " — ", aesthetic_row$display_name[[1]], " ", endpoint)
  }
  observed_max_time <- max(df_plot$time_var, na.rm = TRUE)
  x_axis_max <- if (is.null(max_time)) observed_max_time else as.numeric(max_time)
  if (length(x_axis_max) != 1L || !is.finite(x_axis_max) || x_axis_max <= 0) {
    stop("max_time must be NULL or one positive finite number.")
  }
  pval_x <- 0.15 * x_axis_max

  km_plot <- survminer::ggsurvplot(
    km_fit,
    data = df_plot,
    palette = group_palette,
    conf.int = FALSE,
    censor.shape = "|",
    censor.size = 2,
    xlab = "Time (months)",
    ylab = endpoint_row$y_label[[1]],
    xlim = c(0, x_axis_max),
    ylim = c(aesthetic_row$y_min[[1]], aesthetic_row$y_max[[1]]),
    break.x.by = break_x_by,
    pval = logrank_text,
    pval.coord = c(pval_x, aesthetic_row$pval_y[[1]]),
    pval.size = 5,
    pval.fontface = "plain",
    legend = "top",
    legend.title = "ssGSEA group",
    legend.labs = c("Low", "High"),
    risk.table = TRUE,
    risk.table.title = "Number at risk",
    risk.table.height = 0.25,
    risk.table.fontsize = 5,
    risk.table.y.text.col = FALSE,
    risk.table.y.text = TRUE,
    tables.theme = theme_bw(base_size = 13) +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text.x = element_text(size = 12, face = "bold", color = "black"),
        axis.text.y = element_text(size = 12, face = "bold", color = "black"),
        axis.title.x = element_text(size = 12, face = "bold", color = "black"),
        axis.title.y = element_blank(),
        plot.title = element_text(face = "plain", size = 12, hjust = 0,
                                  margin = margin(b = 2))
      ),
    ggtheme = theme_bw(base_size = 16) +
      theme(
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        axis.text = element_text(size = 13, color = "black", face = "bold"),
        axis.title = element_text(size = 13, face = "bold", color = "black"),
        legend.position = c(0.15, 0.90),
        legend.justification = c(0, 1),
        legend.direction = "horizontal",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.key = element_blank(),
        legend.title = element_blank(),
        legend.text = element_text(size = 12, face = "bold"),
        legend.spacing.x = grid::unit(0.2, "cm"),
        legend.margin = margin(0, 0, 0, 0),
        plot.title = element_text(face = "bold", size = 13, hjust = 0.5,
                                  margin = margin(t = 2, b = 2)),
        plot.subtitle = element_text(face = "bold", size = 11, hjust = 0.5,
                                     color = "black", margin = margin(t = 2, b = 5)),
        plot.margin = margin(t = 5, r = 10, b = 5, l = 10)
      )
  )

  km_plot$plot <- km_plot$plot + labs(title = title, subtitle = subtitle_text)
  statistics <- tibble(
    dataset, dataset_display = aesthetic_row$display_name[[1]], signature, endpoint,
    cutoff = cutoff, cutoff_source = cutoff_source,
    n = nrow(df_plot), events = sum(df_plot$event_var == 1),
    low_n = group_counts$n[group_counts$ssGSEA_group == "Low"],
    low_events = group_counts$events[group_counts$ssGSEA_group == "Low"],
    high_n = group_counts$n[group_counts$ssGSEA_group == "High"],
    high_events = group_counts$events[group_counts$ssGSEA_group == "High"],
    cox_contrast = "High vs Low", hazard_ratio = cox_result$estimate[[1]],
    conf_low = cox_result$conf.low[[1]], conf_high = cox_result$conf.high[[1]],
    cox_wald_p = cox_result$p.value[[1]], logrank_chisq = unname(logrank_model$chisq),
    logrank_df = logrank_df, logrank_p = logrank_p,
    max_observed_time = observed_max_time, plotted_max_time = x_axis_max
  )

  saved_files <- character()
  if (isTRUE(save_plot)) {
    dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
    if (is.null(filename)) {
      filename <- paste(signature, aesthetic_row$display_name[[1]], endpoint, sep = "__")
    }
    filename <- safe_file_tag(tools::file_path_sans_ext(basename(filename)))
    plot_data_file <- file.path(output_dir, paste0(filename, "__plot_data.csv"))
    statistics_file <- file.path(output_dir, paste0(filename, "__statistics.csv"))
    readr::write_csv(df_plot, plot_data_file)
    readr::write_csv(statistics, statistics_file)
    saved_files <- c(saved_files, plot_data_file, statistics_file)

    if (isTRUE(params$save_png)) {
      png_file <- file.path(output_dir, paste0(filename, ".png"))
      save_ggsurvplot(km_plot, png_file, width, height, dpi, "png")
      saved_files <- c(saved_files, png_file)
    }
    if (isTRUE(params$save_pdf)) {
      pdf_file <- file.path(output_dir, paste0(filename, ".pdf"))
      save_ggsurvplot(km_plot, pdf_file, width, height, dpi, "pdf")
      saved_files <- c(saved_files, pdf_file)
    }
  }

  if (isTRUE(display)) print_ggsurvplot_clean(km_plot, newpage = TRUE)

  invisible(list(
    plot = km_plot,
    survfit = km_fit,
    cox_model = cox_model,
    logrank_model = logrank_model,
    plotting_data = df_plot,
    group_counts = group_counts,
    statistics = statistics,
    median_cutoff = cutoff,
    saved_files = saved_files
  ))
}
```

# Manual plotting

The function can be called after running all chunks through `km-function`.
Manual calls do not require rerunning the batch section.

```{r manual-example, eval=FALSE}
result <- plot_single_ssgsea_km(
  signature = "Dys-CIM",
  dataset = "SCANB",
  endpoint = "RFi",
  title = "Dys-CIM — SCAN-B RFI",
  save_plot = TRUE,
  display = TRUE,
  width = 6,
  height = 5,
  dpi = 600
)

result$statistics
result$median_cutoff
result$group_counts
```

Additional examples:

```{r more-manual-examples, eval=FALSE}
# Fun-CIM, METABRIC overall survival
plot_single_ssgsea_km(
  signature = "Fun-CIM",
  dataset = "METABRIC",
  endpoint = "OS",
  save_plot = TRUE,
  display = TRUE
)

# Dys-CIM, METABRIC RFI, using the prespecified median
plot_single_ssgsea_km(
  signature = "Dys-CIM",
  dataset = "METABRIC",
  endpoint = "RFI",
  title = "Dys-CIM — METABRIC RFI",
  save_plot = TRUE,
  display = TRUE
)

# Optional exploratory user-defined cutoff. The output table identifies this
# explicitly as a user-defined cutoff.
#plot_single_ssgsea_km(
#  signature = "Fun-CIM",
#  dataset = "SCANB",
#  endpoint = "OS",
#  median_cutoff = 0.42,
#  title = "Fun-CIM — SCAN-B OS (exploratory cutoff)",
 # save_plot = TRUE,
 # display = TRUE
#)
```

# Optional batch generation

When `run_batch: true`, all 12 prespecified plots are generated using the same
manual plotting function and the same fixed cohort–signature medians.

```{r batch-generation}
batch_design <- bind_rows(
  tidyr::crossing(
    dataset = "METABRIC",
    signature = c("Fun-CIM", "Dys-CIM"),
    endpoint = c("OS", "DSS", "RFI")
  ),
  tidyr::crossing(
    dataset = "SCANB",
    signature = c("Fun-CIM", "Dys-CIM"),
    endpoint = c("OS", "DRFi", "RFi")
  )
) %>% mutate(
  title = paste0(
    signature, " — ", if_else(dataset == "SCANB", "SCAN-B", dataset), " ",
    if_else(endpoint == "RFi", "RFI", endpoint)
  )
)

batch_results <- list()
batch_manifest <- batch_design %>% mutate(
  status = "not_run", warning_text = NA_character_,
  n = NA_integer_, events = NA_integer_, hazard_ratio = NA_real_,
  conf_low = NA_real_, conf_high = NA_real_, cox_wald_p = NA_real_,
  logrank_p = NA_real_, cutoff = NA_real_
)

if (isTRUE(params$run_batch)) {
  for (i in seq_len(nrow(batch_design))) {
    specification <- batch_design[i, ]
    result <- tryCatch(
      plot_single_ssgsea_km(
        signature = specification$signature,
        dataset = specification$dataset,
        endpoint = specification$endpoint,
        title = specification$title,
        save_plot = TRUE,
        display = FALSE,
        width = params$width,
        height = params$height,
        dpi = params$dpi
      ),
      error = function(e) e
    )
    if (inherits(result, "error")) {
      batch_manifest$status[i] <- "failed"
      batch_manifest$warning_text[i] <- conditionMessage(result)
    } else {
      batch_results[[paste(specification$dataset, specification$signature,
                           specification$endpoint, sep = "__")]] <- result
      stat <- result$statistics
      batch_manifest$status[i] <- "completed"
      batch_manifest$n[i] <- stat$n
      batch_manifest$events[i] <- stat$events
      batch_manifest$hazard_ratio[i] <- stat$hazard_ratio
      batch_manifest$conf_low[i] <- stat$conf_low
      batch_manifest$conf_high[i] <- stat$conf_high
      batch_manifest$cox_wald_p[i] <- stat$cox_wald_p
      batch_manifest$logrank_p[i] <- stat$logrank_p
      batch_manifest$cutoff[i] <- stat$cutoff
    }
  }
}

readr::write_csv(batch_design,
                 file.path(subdirs["batch"], "batch_plot_design.csv"))
readr::write_csv(batch_manifest,
                 file.path(subdirs["batch"], "batch_plot_manifest.csv"))

all_batch_statistics <- if (length(batch_results)) {
  map_dfr(batch_results, "statistics")
} else {
  tibble()
}
readr::write_csv(all_batch_statistics,
                 file.path(subdirs["statistics"], "all_KM_Cox_logrank_results.csv"))
```

# Endpoint eligibility and group audits

```{r endpoint-audits}
endpoint_eligibility <- map_dfr(names(endpoint_config), function(dataset) {
  map_dfr(seq_len(nrow(endpoint_config[[dataset]])), function(i) {
    endpoint_row <- endpoint_config[[dataset]][i, ]
    map_dfr(c("Fun-CIM", "Dys-CIM"), function(signature) {
      d <- score_clinical_long %>%
        filter(.data$dataset == .env$dataset,
               .data$signature == .env$signature) %>%
        mutate(
          time = safe_numeric(.data[[endpoint_row$time_col]]),
          event = standardize_event(.data[[endpoint_row$event_col]],
                                    endpoint_row$event_col, dataset),
          eligible = is.finite(time) & time > 0 & event %in% c(0, 1)
        )
      d %>% group_by(median_group) %>% summarise(
        dataset = .env$dataset,
        signature = .env$signature,
        endpoint = endpoint_row$endpoint[[1]],
        assigned_n = n(), eligible_n = sum(eligible),
        excluded_endpoint_n = sum(!eligible),
        events = sum(event[eligible] == 1), .groups = "drop"
      )
    })
  })
})
readr::write_csv(endpoint_eligibility,
                 file.path(subdirs["qc"], "endpoint_group_eligibility.csv"))

excluded_endpoint_samples <- map_dfr(names(endpoint_config), function(dataset) {
  map_dfr(seq_len(nrow(endpoint_config[[dataset]])), function(i) {
    endpoint_row <- endpoint_config[[dataset]][i, ]
    score_clinical_long %>% filter(.data$dataset == .env$dataset) %>%
      mutate(
        time = safe_numeric(.data[[endpoint_row$time_col]]),
        event = standardize_event(.data[[endpoint_row$event_col]],
                                  endpoint_row$event_col, dataset),
        endpoint = endpoint_row$endpoint,
        exclusion_reason = case_when(
          !is.finite(time) | time <= 0 ~ "missing_or_nonpositive_time",
          !event %in% c(0, 1) | is.na(event) ~ "missing_or_invalid_event",
          TRUE ~ NA_character_
        )
      ) %>% filter(!is.na(exclusion_reason)) %>%
      select(dataset, signature, endpoint, sample_id, median_group,
             exclusion_reason)
  })
})
readr::write_csv(excluded_endpoint_samples,
                 file.path(subdirs["qc"], "excluded_endpoint_samples.csv"))
```

# Workbook and reproducibility

```{r workbook-and-reproducibility}
workbook <- openxlsx::createWorkbook()
add_table_sheet <- function(name, data) {
  sheet <- substr(gsub("[\\[\\]*?/\\\\:]", "_", name), 1, 31)
  openxlsx::addWorksheet(workbook, sheet)
  if (!ncol(data)) data <- tibble(note = "No rows were generated")
  openxlsx::writeDataTable(workbook, sheet, as.data.frame(data),
                           tableStyle = "TableStyleMedium2")
  openxlsx::freezePane(workbook, sheet, firstRow = TRUE)
  if (ncol(data)) openxlsx::setColWidths(workbook, sheet,
                                         cols = seq_len(ncol(data)), widths = "auto")
}
add_table_sheet("KM Cox logrank", all_batch_statistics)
add_table_sheet("Batch manifest", batch_manifest)
add_table_sheet("Median cutoffs", median_cutoff_table)
add_table_sheet("Endpoint eligibility", endpoint_eligibility)
add_table_sheet("Join audit", join_audit)
add_table_sheet("Clinical sources", clinical_source_audit)
add_table_sheet("Parameters", analysis_parameters)
openxlsx::saveWorkbook(
  workbook,
  file.path(output_dir, "CIM_ssGSEA_KM_results.xlsx"),
  overwrite = TRUE
)

writeLines(capture.output(sessionInfo()),
           file.path(subdirs["reproducibility"], "sessionInfo.txt"))
writeLines(capture.output(tools::md5sum(files)),
           file.path(subdirs["reproducibility"], "input_md5_checksums.txt"))

completion_check <- tibble(
  check = c(
    "both_cohorts_present", "both_signatures_present", "median_cutoffs_unique",
    "all_score_samples_matched", "manual_function_available",
    "batch_attempted_when_requested", "workbook_written"
  ),
  passed = c(
    all(c("METABRIC", "SCANB") %in% unique(scores$dataset)),
    all(c("Fun-CIM", "Dys-CIM") %in% unique(scores$signature)),
    !anyDuplicated(median_cutoff_table[c("dataset", "signature")]),
    all(join_audit$clinical_match),
    exists("plot_single_ssgsea_km", mode = "function"),
    !isTRUE(params$run_batch) ||
      (nrow(batch_manifest) == 12L && all(batch_manifest$status == "completed")),
    file.exists(file.path(output_dir, "CIM_ssGSEA_KM_results.xlsx"))
  )
)
readr::write_csv(completion_check,
                 file.path(subdirs["reproducibility"], "completion_check.csv"))
if (any(!completion_check$passed)) {
  warning("One or more completion checks failed; inspect completion_check.csv.")
}
```

# Interpretation safeguards

- `High` means an ssGSEA score greater than or equal to the fixed
  cohort–signature median; `Low` means a score below that median.
- The Cox subtitle reports the univariable High-versus-Low hazard ratio, 95%
  confidence interval, and Wald p-value.
- The annotation inside the curve panel reports the two-sided log-rank p-value.
- The Cox model, log-rank test, KM curves, and risk table use the exact same
  endpoint-eligible patient set.
- The continuous ssGSEA Cox analysis should remain primary because median
  splitting reduces information and statistical power.
- METABRIC RFI and SCAN-B RFi are recurrence-focused contextual counterparts.
  METABRIC DSS and SCAN-B DRFi are cohort-specific and should not be presented
  as equivalent endpoints.
- A manually supplied cutoff is exploratory and is labeled as user-defined in
  its exported statistics table.
