---
title: "CIMIC Immune Deconvolution Pipeline — Compact Figure Outputs"
author: "Gbadamosi Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
params:
  project_root: ""
  dataset_subdir: "Datasets/NKI_SMC"
  results_subdir: "Results/nki_smc_immunedeconv_k3"
  expression_file: "nki_smc_combine_TPM.csv"
  cluster_file: "nki_smc_combine_k3.csv"
  cibersortx_file: "CIBERSORTx_smc_nki_relative2.csv"
  run_immunedeconv: true
  run_cibersortx_import: true
  make_plots: true
  make_heatmaps: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  error = FALSE,
  fig.align = "center"
)
```

# Overview

This compact R Markdown document runs the CIMIC immune-infiltration workflow for the **combined NKI + SMC paired neoadjuvant dataset**. NKI and SMC are retained as metadata for provenance/QC.

1. Loads one combined, non-log-transformed TPM matrix (`nki_smc_combine_TPM.csv`).
2. Loads the combined CIMIC cluster table (`nki_smc_combine_clustered_plot_df.csv`).
3. Harmonizes sample IDs into one canonical patient identifier.
4. Infers source cohort for metadata only (`OB_*` = SMC; all other IDs = NKI).
5. Maps numeric CIMIC cluster IDs to **C1**, **C2**, and **C3**.
6. Performs explicit sample/cluster/timepoint quality control before analysis.
7. Runs every `immunedeconv` method **once on the complete combined NKI+SMC TPM matrix**.
8. Imports one combined CIBERSORTx relative-mode file and removes CIBERSORTx QC columns (`P-value`, `Correlation`, `RMSE`) before immune-cell results are reshaped and merged.
9. Aggregates technical replicates, where present, to the patient level.
10. Performs pooled unpaired comparisons across C1, C2, and C3 at biopsy and surgery, and paired biopsy-to-surgery analyses within each cluster.
11. Generates publication-style plots, heatmaps, run logs, and reproducibility information.

The primary unit for downstream statistical analysis is the patient.

# Repository layout

The Rmd is intended to live in `oncoimmunology_paper/figure_generation_scripts/`. It discovers its own location at render time, treats the parent directory as the project root, and constructs all input/output paths from the repository structure. No user-specific OneDrive, Windows, macOS, or Linux path is stored in the script.

```text
oncoimmunology_paper/
├── Datasets/
│   └── NKI_SMC/
│       ├── nki_smc_combine_TPM.csv
│       ├── nki_smc_combine_clustered_plot_df.csv
│       └── CIBERSORTx_smc_nki_relative.csv  # optional
├── figure_generation_scripts/
│   └── CIM_Immunedeconv_pipeline_v3.1.Rmd
└── Results/
    └── nki_smc_immunodeconv/          # created automatically
        ├── figures/
        ├── logs/
        ├── tmp/
        └── bioc_cache/
```

`project_root` can be supplied explicitly when rendering in an unusual environment, but its default is an empty string, which activates automatic project-root detection from the Rmd location.

Input requirements:

- The expression matrix must contain genes in rows and samples in columns.
- A gene identifier column named `gene_id`, `gene`, `symbol`, or another configured candidate must be present.
- Expression values must be TPM and **not log transformed**.
- Combined TPM sample IDs must end in `_B` or `_S`, for example `2046_B`, `2046_S`, `OB_0139_B`, and `OB_0139_S`.
- The cluster file must contain `sample_id` and either `cluster_assignment` or `cluster_assignments`. Cluster-table IDs such as `2046_delta` and `OB_0139_delta` are converted to canonical patient IDs by removing the `_delta` suffix.
- The combined CIBERSORTx file must contain a `Mixture` column plus immune-cell columns. `P-value`, `Correlation`, and `RMSE` are explicitly removed before reshaping/combining results. Canonical IDs such as `2046_B` and `OB_0139_S` are supported directly; older prefixed/replicate forms are also harmonized.

# Reproducibility notes

This workflow uses a fixed random seed and saves package/session information at the end of each run. For a GitHub release, using `renv` is strongly recommended so collaborators can restore the same R package versions.

```r
install.packages("renv")
renv::init()
renv::snapshot()
```

On another machine:

```r
renv::restore()
```

# Package setup

This document does not install packages during analysis. Install dependencies once before running the workflow.

```{r packages}
required_packages <- c(
  "data.table", "dplyr", "tidyr", "tibble", "stringr", "purrr",
  "ggplot2", "readr", "ComplexHeatmap", "circlize", "withr",
  "FSA", "grid", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    "\nInstall them before running this R Markdown. ",
    "For Bioconductor packages, use BiocManager."
  )
}

if (!requireNamespace("immunedeconv", quietly = TRUE)) {
  stop(
    "Package 'immunedeconv' is not installed.\n",
    "Install it with:\n",
    "remotes::install_github('omnideconv/immunedeconv')"
  )
}

suppressPackageStartupMessages({
  library(data.table)
  library(immunedeconv)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(readr)
  library(ComplexHeatmap)
  library(circlize)
  library(withr)
  library(FSA)
  library(grid)
})

cat("All required packages loaded successfully.\n")
```

# Configuration


```{r config}
set.seed(2025)
options(stringsAsFactors = FALSE, scipen = 999)

# ---- Repository-aware path resolution -------------------------------------
# Expected location of this Rmd:
#   oncoimmunology_paper/figure_generation_scripts/
# Project root:
#   oncoimmunology_paper/
#
# No machine-specific absolute paths are required. If project_root is left
# empty (recommended), the project root is inferred as the parent directory
# of this Rmd. An explicit project_root can still be supplied to
# rmarkdown::render() for non-standard execution environments.

resolve_rmd_path <- function() {
  # Preferred during knitting/rendering.
  input <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) "")
  if (!is.null(input) && length(input) == 1 && nzchar(input) && file.exists(input)) {
    return(normalizePath(input, winslash = "/", mustWork = TRUE))
  }

  # Helpful when sourced interactively from RStudio.
  if (requireNamespace("rstudioapi", quietly = TRUE) &&
      rstudioapi::isAvailable()) {
    ctx_path <- tryCatch(rstudioapi::getSourceEditorContext()$path,
                         error = function(e) "")
    if (nzchar(ctx_path) && file.exists(ctx_path)) {
      return(normalizePath(ctx_path, winslash = "/", mustWork = TRUE))
    }
  }

  return(NA_character_)
}

RMD_PATH <- resolve_rmd_path()

if (!is.null(params$project_root) &&
    length(params$project_root) == 1 &&
    !is.na(params$project_root) &&
    nzchar(params$project_root)) {
  PROJECT_DIR <- normalizePath(
    params$project_root, winslash = "/", mustWork = TRUE
  )
} else {
  if (is.na(RMD_PATH)) {
    stop(
      "Could not determine the Rmd location automatically. ",
      "Render the Rmd normally from the repository or provide ",
      "params = list(project_root = '/path/to/oncoimmunology_paper').",
      call. = FALSE
    )
  }
  SCRIPT_DIR <- dirname(RMD_PATH)
  PROJECT_DIR <- normalizePath(
    file.path(SCRIPT_DIR, ".."), winslash = "/", mustWork = TRUE
  )
}

is_absolute_path <- function(x) {
  grepl("^[A-Za-z]:[/\\\\]|^/|^~", x)
}

resolve_param_path <- function(base_dir, x, must_work = FALSE) {
  candidate <- if (is_absolute_path(x)) x else file.path(base_dir, x)
  normalizePath(candidate, winslash = "/", mustWork = must_work)
}

# dataset_subdir/results_subdir may be either repository-relative or absolute.
# Absolute paths are used directly so they are never appended to PROJECT_DIR.
DATA_DIR <- resolve_param_path(PROJECT_DIR, params$dataset_subdir, must_work = TRUE)
RESULTS_DIR <- resolve_param_path(PROJECT_DIR, params$results_subdir, must_work = FALSE)

FILES <- list(
  combined_expr = file.path(DATA_DIR, params$expression_file),
  clusters = file.path(DATA_DIR, params$cluster_file),
  cibersortx = file.path(DATA_DIR, params$cibersortx_file)
)

# CIBERSORTx file-name fallback for older repository copies. The configured
# YAML name is always preferred; fallback is used only when it is absent.
if (!file.exists(FILES$cibersortx)) {
  ciber_candidates <- file.path(
    DATA_DIR,
    c(
      "CIBERSORTx_smc_nki_relative.csv",
      "CIBERSORTx_smc_nki_relative1.csv"
    )
  )
  ciber_found <- ciber_candidates[file.exists(ciber_candidates)]
  if (length(ciber_found) >= 1) {
    FILES$cibersortx <- ciber_found[1]
    message("Using detected CIBERSORTx input: ", basename(FILES$cibersortx))
  }
}

DIRS <- list(
  project = PROJECT_DIR,
  data = DATA_DIR,
  results = RESULTS_DIR,
  plots = file.path(RESULTS_DIR, "figures"),
  tables = file.path(RESULTS_DIR, "tables"),
  # Keep transient/cache paths short to reduce Windows path-length problems.
  logs = file.path(RESULTS_DIR, "log"),
  tmp = file.path(PROJECT_DIR, ".tmp_cim"),
  cache = file.path(PROJECT_DIR, ".cache_cim")
)

if (!dir.exists(DATA_DIR)) {
  stop(
    "Expected data directory was not found:\n  ", DATA_DIR,
    "\n\nExpected repository layout: ",
    "<project>/Datasets/NKI_SMC",
    call. = FALSE
  )
}

invisible(lapply(
  DIRS[c("results", "plots", "tables", "logs", "tmp", "cache")],
  dir.create, recursive = TRUE, showWarnings = FALSE
))

cat("Project root: ", PROJECT_DIR, "\n", sep = "")
cat("Input data:   ", DATA_DIR, "\n", sep = "")
cat("Results:      ", RESULTS_DIR, "\n", sep = "")

Sys.setenv(TMPDIR = DIRS$tmp)
Sys.setenv(TEMP = DIRS$tmp)
Sys.setenv(TMP = DIRS$tmp)
options(ExperimentHub.cache = DIRS$cache)

LOG_FILE <- file.path(
  DIRS$logs,
  paste0("run_log_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".txt")
)

METHODS_TO_RUN <- c(
  "quantiseq", "epic", "xcell", "mcp_counter",
  "estimate", "timer", "consensus_tme"
)

TUMOR_INDICATION <- "BRCA"
GENE_COL_CANDIDATES <- c("gene_id", "gene", "hgnc_symbol", "symbol", "Gene", "GeneSymbol")
INDEX_COL_PATTERNS <- c("^Unnamed(:|\\.)?", "^X$")

METHOD_ORDER <- c(
  "CIBERSORTx", "quantiseq", "epic", "mcp_counter", "estimate",
  "xcell", "timer", "consensus_tme"
)

AGGREGATE_XCELL_TCELLS <- TRUE
AGGREGATE_CIBERSORT_TCELLS <- TRUE
CIBERSORT_AGGREGATION_STRATEGY <- "refined" # "refined" or "simple"

# Numeric CIMIC cluster IDs -> biological labels.
USE_CLUSTER_MAPPING <- TRUE
CLUSTER_MAPPING <- c(
  "1" = "C1",
  "2" = "C2",
  "3" = "C3"
)

cluster_colors <- c(
  "C1" = "#4A90C4",
  "C2" = "#F08A8A",
  "C3" = "#A8CDED"
)

# Paired figures encode timepoint with color while keeping the paired lines gray.
TIMEPOINT_COLORS <- c(
  "Biopsy" = "#1F77B4",
  "Surgery" = "#9467BD"
)

# Minimum sample-size safeguards for inferential tests.
MIN_UNPAIRED_GROUP_N <- 2L
MIN_PAIRED_N <- 3L

missing_cluster_colors <- base::setdiff(unique(unname(CLUSTER_MAPPING)), names(cluster_colors))
if (length(missing_cluster_colors) > 0) {
  stop("Missing cluster colors for: ", paste(missing_cluster_colors, collapse = ", "))
}

METHOD_COLORS <- c(
  "quantiseq" = "#1F77B4",
  "epic" = "#FF7F0E",
  "xcell" = "#2CA02C",
  "mcp_counter" = "#D62728",
  "estimate" = "#9467BD",
  "timer" = "#8C564B",
  "consensus_tme" = "#E377C2",
  "CIBERSORTx" = "#FFD700"
)

# Plot aesthetics shared by unpaired and paired figures.
# CIM trajectory colors remain the primary visual encoding in both plot types.
PLOT_WIDTH_UNPAIRED <- 5
PLOT_HEIGHT_UNPAIRED <- 4.6
PLOT_WIDTH_PAIRED <- PLOT_WIDTH_UNPAIRED
PLOT_HEIGHT_PAIRED <- PLOT_HEIGHT_UNPAIRED
PLOT_DPI <- 300

# Multipanel/publication typography: individual panels will be reduced when assembled,
# so text and points are intentionally larger at export size.

PLOT_BOX_ALPHA <- 0.70
PLOT_BOX_WIDTH <- 0.58
PLOT_BOX_LINEWIDTH <- 0.80
PLOT_UNPAIRED_POINT_SIZE <- 2.4
PLOT_UNPAIRED_POINT_ALPHA <- 0.65
PLOT_JITTER_WIDTH <- 0.15
PLOT_PAIRED_POINT_SIZE <- 3.2
PLOT_PAIRED_POINT_ALPHA <- 0.85
PLOT_PAIRED_LINE_ALPHA <- 0.70
PLOT_PAIRED_LINE_WIDTH <- 0.9
PLOT_PAIRED_LINE_COLOR <- "gray65"
PLOT_P_ANNOTATION_SIZE <- 6.5

# Compact-output option. Aggregated plots are always generated when make_plots = TRUE.
# Set TRUE only when original (non-T-cell-aggregated) plots are also needed.
MAKE_ORIGINAL_PLOTS <- FALSE
# MAKE_ORIGINAL_PLOTS <- TRUE
```

# Helper functions

```{r helpers}
validate_file <- function(path, required = TRUE) {
  if (!file.exists(path)) {
    msg <- paste0("File not found: ", path)
    if (required) stop(msg, call. = FALSE) else warning(msg, call. = FALSE)
    return(FALSE)
  }
  TRUE
}

sanitize_filename <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
safe_stem <- function(...) sanitize_filename(paste(..., sep = "_"))
method_dir_name <- function(x) sanitize_filename(x)
cluster_file_label <- function(x) {
  x <- as.character(x)
  dplyr::case_when(
    x == "Fun-CIM" ~ "FunCIM",
    x == "Dys-CIM" ~ "DysCIM",
    TRUE ~ sanitize_filename(x)
  )
}

method_value_type <- function(m) {
  switch(m,
    "quantiseq" = "Cell Fraction",
    "epic" = "Cell Fraction",
    "mcp_counter" = "Abundance Score",
    "xcell" = "Enrichment Score",
    "timer" = "Infiltration Score",
    "consensus_tme" = "ssGSEA Score",
    "estimate" = "ESTIMATE Score",
    "CIBERSORTx" = "Cell Fraction",
    "Score"
  )
}

auto_ylims <- function(values, ymin_floor = 0) {
  v <- values[is.finite(values)]
  if (!length(v)) return(c(ymin_floor, ymin_floor + 0.01))
  ymin <- ymin_floor
  ymax <- max(v, na.rm = TRUE)
  yr <- ymax - ymin
  if (yr <= 0) yr <- max(abs(ymax) * 0.15, 0.01)
  c(ymin, ymax + 0.18 * yr)
}

safe_zscore <- function(x) {
  x <- as.numeric(x)
  s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(0, length(x)))
  as.numeric((x - mean(x, na.rm = TRUE)) / s)
}

apply_cluster_mapping <- function(df, cluster_col = "cluster") {
  # Apply the user-defined cluster mapping to the analysis cluster column.
  # This is the central rule used throughout the workflow.
  # Current k=3 CIMIC mapping: 1 = C1; 2 = C2; 3 = C3.
  # All downstream statistics, plots, paired analyses, and heatmaps use these
  # mapped cluster labels rather than source cohort.
  if (!USE_CLUSTER_MAPPING) return(df)
  if (!cluster_col %in% names(df)) return(df)

  original_cluster <- as.character(df[[cluster_col]])
  mapped_cluster <- dplyr::recode(
    original_cluster,
    !!!CLUSTER_MAPPING,
    .default = original_cluster
  )

  mapped_levels <- unique(unname(CLUSTER_MAPPING))
  extra_levels <- unique(mapped_cluster[!mapped_cluster %in% mapped_levels])
  final_levels <- unique(c(mapped_levels, extra_levels))

  if (!"cluster_original" %in% names(df)) {
    df$cluster_original <- original_cluster
  }

  df[[cluster_col]] <- factor(mapped_cluster, levels = final_levels)
  df
}

# Backward-compatible alias used by earlier versions of the workflow.
apply_cluster_names <- apply_cluster_mapping

theme_pub <- function() {
  theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 24, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 19, face = "bold"),
      axis.text = element_text(size = 19, face = "bold"),
      axis.title = element_text(size = 23, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 15)),
      axis.ticks = element_line(linewidth = 1.1),
      axis.line = element_line(linewidth = 1),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      legend.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 16),
      strip.text = element_text(size = 18, face = "bold"),
      strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
      plot.margin = margin(t = 10, r = 10, b = 50, l = 10)
    )
}
```

# Sample ID harmonization

All input sources are converted to the same canonical patient identifier before joining metadata. This prevents cohort prefixes, SMC year codes, timepoint suffixes, and CIBERSORTx replicate suffixes from silently breaking joins.

```{r id-parsing}
normalize_sample_id <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_trim() %>%
    stringr::str_replace_all('^"|"$', "")
}

canonicalize_smc_patient <- function(x) {
  # CIBERSORTx SMC IDs may contain a two-digit collection/year component,
  # e.g. OB_15_0139, whereas the combined TPM/cluster files use OB_0139.
  # Any many-to-one collision created by this rule is checked explicitly below.
  stringr::str_replace(x, "^OB_[0-9]{2}_([0-9]{4})$", "OB_\\1")
}

get_timepoint <- function(sample_id) {
  sample_id <- normalize_sample_id(sample_id)
  dplyr::case_when(
    stringr::str_detect(sample_id, "_B(?:_Rep[0-9]+)?$") ~ "Biopsy",
    stringr::str_detect(sample_id, "_S(?:_Rep[0-9]+)?$") ~ "Surgery",
    TRUE ~ NA_character_
  )
}

get_patient_key_from_expression <- function(sample_id) {
  sample_id %>%
    normalize_sample_id() %>%
    stringr::str_replace("_(B|S)$", "") %>%
    canonicalize_smc_patient()
}

get_patient_key_from_cluster <- function(cluster_sample_id) {
  cluster_sample_id %>%
    normalize_sample_id() %>%
    stringr::str_replace("^delta_", "") %>%
    stringr::str_replace("_delta$", "") %>%
    canonicalize_smc_patient()
}

get_patient_key_from_cibersortx <- function(sample_id) {
  sample_id %>%
    normalize_sample_id() %>%
    stringr::str_replace("^(NKI|SMC)_", "") %>%
    stringr::str_replace("_(B|S)_Rep[0-9]+$", "") %>%
    stringr::str_replace("_(B|S)$", "") %>%
    canonicalize_smc_patient()
}

infer_dataset <- function(patient_key) {
  dplyr::if_else(stringr::str_detect(patient_key, "^OB_"), "SMC", "NKI")
}
```

# Validate inputs

```{r validate-inputs}
validate_file(FILES$combined_expr)
validate_file(FILES$clusters)

if (params$run_cibersortx_import) {
  validate_file(FILES$cibersortx, required = FALSE)
}
```

# Load combined CIMIC cluster table

The cluster file must contain `sample_id` and either `cluster_assignment` or `cluster_assignments`. Other dimensional-reduction columns present in the clustered plotting file are ignored.

```{r load-clusters}
clusters_raw <- readr::read_csv(FILES$clusters, show_col_types = FALSE)

if (!"sample_id" %in% colnames(clusters_raw)) {
  stop("Cluster file is missing required column: sample_id", call. = FALSE)
}

cluster_col <- dplyr::case_when(
  "cluster_assignment" %in% colnames(clusters_raw) ~ "cluster_assignment",
  "cluster_assignments" %in% colnames(clusters_raw) ~ "cluster_assignments",
  TRUE ~ NA_character_
)

if (is.na(cluster_col)) {
  stop(
    "Cluster file must contain either 'cluster_assignment' or 'cluster_assignments'.",
    call. = FALSE
  )
}

clusters <- clusters_raw %>%
  dplyr::transmute(
    cluster_sample_id = as.character(sample_id),
    patient_key = get_patient_key_from_cluster(cluster_sample_id),
    cluster = as.character(.data[[cluster_col]])
  )

if (any(is.na(clusters$patient_key) | clusters$patient_key == "")) {
  stop("One or more cluster rows produced an empty patient_key.", call. = FALSE)
}

missing_mapping <- setdiff(
  unique(clusters$cluster),
  c(names(CLUSTER_MAPPING), unname(CLUSTER_MAPPING))
)
if (USE_CLUSTER_MAPPING && length(missing_mapping) > 0) {
  stop(
    "Cluster IDs missing from CLUSTER_MAPPING: ",
    paste(missing_mapping, collapse = ", "),
    call. = FALSE
  )
}

clusters <- apply_cluster_mapping(clusters)

dup_pat <- clusters %>% count(patient_key) %>% filter(n > 1)
if (nrow(dup_pat) > 0) {
  stop(
    "Cluster file contains duplicate canonical patient IDs: ",
    paste(dup_pat$patient_key, collapse = ", "),
    call. = FALSE
  )
}

k <- dplyr::n_distinct(stats::na.omit(clusters$cluster))
message("Detected ", k, " mapped CIM group(s).")
knitr::kable(head(clusters, 10), caption = "Harmonized cluster table preview")
```

# Load combined TPM expression matrix

The input is expected to be non-log TPM. Index-like columns generated by spreadsheet/Python exports (for example `Unnamed: 0`) are removed automatically before selecting the gene identifier column.

```{r load-expression}
load_combined_expr_matrix <- function(path) {
  validate_file(path)
  message("Loading combined NKI + SMC TPM matrix: ", path)

  df <- data.table::fread(path, data.table = FALSE, check.names = FALSE)

  # Remove non-biological row-index columns created by CSV export.
  # Important: data.table::fread() renames an EMPTY first CSV header ("") to
  # a generic name such as V1. Therefore, matching only "Unnamed: 0"/"X"
  # is not sufficient for pandas-exported files like the current TPM matrix.
  #
  # We remove:
  #   1) explicitly index-like names (Unnamed..., X), and
  #   2) generic V1/V2/... columns ONLY when their values are exactly 1:nrow(df).
  # This prevents a legitimate biological/sample column from being dropped solely
  # because its name happens to resemble a generic fread name.
  is_sequential_index <- function(x, n) {
    y <- suppressWarnings(as.numeric(x))
    length(y) == n &&
      all(is.finite(y)) &&
      identical(as.numeric(y), as.numeric(seq_len(n)))
  }

  named_index_cols <- unique(unlist(lapply(
    INDEX_COL_PATTERNS,
    function(p) grep(p, names(df), value = TRUE)
  )))

  generic_index_cols <- names(df)[
    grepl("^V[0-9]+$", names(df)) &
      vapply(df, is_sequential_index, logical(1), n = nrow(df))
  ]

  # Also catch the common case where the first column is a sequential index and
  # the next column is the recognized gene identifier, regardless of its name.
  first_col_index <- character(0)
  if (ncol(df) >= 2 &&
      names(df)[2] %in% GENE_COL_CANDIDATES &&
      is_sequential_index(df[[1]], nrow(df))) {
    first_col_index <- names(df)[1]
  }

  index_cols <- unique(c(named_index_cols, generic_index_cols, first_col_index))

  if (length(index_cols) > 0) {
    message("Removing non-biological index column(s): ", paste(index_cols, collapse = ", "))
    df <- df[, setdiff(names(df), index_cols), drop = FALSE]
  }

  gene_col <- intersect(GENE_COL_CANDIDATES, names(df))
  if (length(gene_col) == 0) {
    stop(
      "No recognized gene identifier column found. Expected one of: ",
      paste(GENE_COL_CANDIDATES, collapse = ", "),
      call. = FALSE
    )
  }
  gene_col <- gene_col[1]

  # At this point every non-gene column must be a biological sample column.
  # This catches export/index problems before numeric conversion.
  unexpected_non_sample <- setdiff(names(df), gene_col)
  unexpected_non_sample <- unexpected_non_sample[!grepl("_(B|S)$", unexpected_non_sample)]
  if (length(unexpected_non_sample) > 0) {
    stop(
      "Unexpected non-sample column(s) remain after input cleaning: ",
      paste(unexpected_non_sample, collapse = ", "),
      "\nCheck the CSV header/export format. Expected only the gene column plus sample IDs ending in _B/_S.",
      call. = FALSE
    )
  }

  genes <- trimws(as.character(df[[gene_col]]))
  if (any(is.na(genes) | genes == "")) {
    stop("The gene identifier column contains missing/empty values.", call. = FALSE)
  }
  if (anyDuplicated(genes)) {
    dup <- unique(genes[duplicated(genes)])
    stop(
      "Duplicate gene identifiers detected. Resolve duplicates before deconvolution. Examples: ",
      paste(head(dup, 10), collapse = ", "),
      call. = FALSE
    )
  }

  expr_df <- df[, setdiff(names(df), gene_col), drop = FALSE]
  if (ncol(expr_df) == 0) stop("No sample columns found in the expression matrix.", call. = FALSE)
  if (anyDuplicated(names(expr_df))) stop("Duplicate expression sample columns detected.", call. = FALSE)

  # Standardize sample-column names before conversion. This prevents accidental
  # whitespace from breaking downstream ID matching.
  original_sample_names <- names(expr_df)
  sample_names <- stringr::str_trim(original_sample_names)

  if (any(is.na(sample_names) | sample_names == "")) {
    stop("One or more expression sample columns have empty names.", call. = FALSE)
  }
  if (anyDuplicated(sample_names)) {
    dup <- unique(sample_names[duplicated(sample_names)])
    stop(
      "Duplicate expression sample IDs after trimming whitespace: ",
      paste(dup, collapse = ", "),
      call. = FALSE
    )
  }
  names(expr_df) <- sample_names

  # Fail early if a non-sample column survived input cleaning.
  bad_sample_names <- sample_names[!grepl("_(B|S)$", sample_names)]
  if (length(bad_sample_names) > 0) {
    stop(
      "Unexpected expression sample column name(s): ",
      paste(head(bad_sample_names, 20), collapse = ", "),
      "\nExpected sample IDs ending in _B or _S, e.g. 2046_B or OB_0139_S.",
      call. = FALSE
    )
  }

  # Strict numeric conversion: fail rather than silently introducing NAs.
  # Explicitly restore list names because some conversion workflows can drop
  # them and cause generic matrix names such as V1, V2, ... .
  numeric_df <- lapply(expr_df, function(x) suppressWarnings(as.numeric(x)))
  names(numeric_df) <- names(expr_df)

  bad_cols <- names(expr_df)[vapply(seq_along(expr_df), function(i) {
    original <- expr_df[[i]]
    converted <- numeric_df[[i]]
    any(is.na(converted) & !is.na(original) & trimws(as.character(original)) != "")
  }, logical(1))]

  if (length(bad_cols) > 0) {
    stop(
      "Non-numeric expression values detected in sample column(s): ",
      paste(bad_cols, collapse = ", "),
      call. = FALSE
    )
  }

  # Construct the matrix explicitly so sample IDs are preserved exactly.
  mat <- do.call(cbind, numeric_df)
  rownames(mat) <- genes
  colnames(mat) <- names(expr_df)
  storage.mode(mat) <- "numeric"

  if (!identical(colnames(mat), names(expr_df))) {
    stop("Expression sample names were altered during matrix construction.", call. = FALSE)
  }

  if (any(!is.finite(mat))) {
    stop("Expression matrix contains NA/NaN/Inf values after numeric conversion.", call. = FALSE)
  }

  message("Loaded ", nrow(mat), " genes x ", ncol(mat), " samples.")
  message("First sample IDs: ", paste(head(colnames(mat), 6), collapse = ", "))
  mat
}

check_nonlog_tpm <- function(mat, label = "Combined TPM") {
  if (any(!is.finite(mat))) stop(label, ": non-finite expression values detected.", call. = FALSE)

  mn <- min(mat, na.rm = TRUE)
  mx <- max(mat, na.rm = TRUE)
  q99 <- as.numeric(stats::quantile(mat, 0.99, na.rm = TRUE))

  if (mn < 0) {
    stop(label, ": negative values detected; TPM should be non-negative.", call. = FALSE)
  }
  if (mx < 30) {
    warning(label, ": maximum value < 30. Verify that the matrix is not log transformed.", call. = FALSE)
  }

  tibble(min = mn, q99 = q99, max = mx)
}

expr_combined <- load_combined_expr_matrix(FILES$combined_expr)
tpm_qc <- check_nonlog_tpm(expr_combined)
knitr::kable(tpm_qc, caption = "Combined TPM scale check")
```

# Build expression metadata and sample-matching QC

```{r metadata}
build_expression_metadata <- function(sample_ids) {
  tibble(sample_id = sample_ids) %>%
    mutate(
      Timepoint = factor(get_timepoint(sample_id), levels = c("Biopsy", "Surgery")),
      patient_key = get_patient_key_from_expression(sample_id),
      Patient_ID = patient_key,
      dataset = infer_dataset(patient_key)
    ) %>%
    left_join(clusters %>% select(patient_key, cluster), by = "patient_key") %>%
    mutate(cluster = factor(cluster, levels = levels(clusters$cluster))) %>%
    select(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster)
}

meta_all <- build_expression_metadata(colnames(expr_combined))

bad_timepoint <- meta_all %>% filter(is.na(Timepoint))
if (nrow(bad_timepoint) > 0) {
  stop(
    "Expression sample IDs with unrecognized timepoint suffixes: ",
    paste(bad_timepoint$sample_id, collapse = ", "),
    "\nExpected IDs ending in _B or _S.",
    call. = FALSE
  )
}

# The expression matrix should contain at most one biological sample per
# canonical patient and timepoint. If canonicalization creates duplicates,
# stop rather than averaging distinct samples as if they were technical reps.
expr_patient_timepoint_dups <- meta_all %>%
  dplyr::count(Patient_ID, Timepoint, name = "n_samples") %>%
  dplyr::filter(n_samples > 1)

if (nrow(expr_patient_timepoint_dups) > 0) {
  dup_details <- meta_all %>%
    dplyr::semi_join(expr_patient_timepoint_dups, by = c("Patient_ID", "Timepoint")) %>%
    dplyr::group_by(Patient_ID, Timepoint) %>%
    dplyr::summarise(sample_ids = paste(sample_id, collapse = ";"), .groups = "drop")

  stop(
    "Multiple expression samples map to the same canonical patient/timepoint. ",
    "Resolve these IDs before deconvolution. Examples: ",
    paste(utils::head(paste0(dup_details$Patient_ID, "/", dup_details$Timepoint, "=", dup_details$sample_ids), 10), collapse = " | "),
    call. = FALSE
  )
}

unmatched_expr <- meta_all %>% filter(is.na(cluster))
unmatched_cluster <- clusters %>% anti_join(meta_all %>% distinct(patient_key), by = "patient_key")

metadata_summary <- meta_all %>%
  mutate(has_cluster = !is.na(cluster)) %>%
  group_by(dataset) %>%
  summarise(
    n_samples = n(),
    n_patients = n_distinct(patient_key),
    n_biopsy = sum(Timepoint == "Biopsy"),
    n_surgery = sum(Timepoint == "Surgery"),
    n_with_cluster = sum(has_cluster),
    .groups = "drop"
  )

pair_qc <- meta_all %>%
  filter(!is.na(cluster)) %>%
  distinct(dataset, patient_key, Timepoint) %>%
  count(dataset, patient_key, name = "n_timepoints") %>%
  group_by(dataset) %>%
  summarise(
    n_clustered_patients = n(),
    n_complete_pairs = sum(n_timepoints == 2),
    n_incomplete_pairs = sum(n_timepoints != 2),
    .groups = "drop"
  )

# compact output: omitted sample_metadata_all.csv
# compact output: omitted qc_expression_samples_without_cluster.csv
# compact output: omitted qc_cluster_patients_without_expression.csv
# compact output: omitted qc_metadata_summary.csv
# compact output: omitted qc_pair_summary.csv

knitr::kable(metadata_summary, caption = "Expression metadata matching summary")
knitr::kable(pair_qc, caption = "Biopsy/surgery pair summary")

if (nrow(unmatched_expr) > 0) {
  warning(nrow(unmatched_expr), " expression sample(s) do not have a cluster assignment and will be excluded.", call. = FALSE)
}

meta_retained <- meta_all %>% filter(!is.na(cluster), !is.na(Timepoint))
if (nrow(meta_retained) == 0) stop("No expression samples matched cluster and timepoint metadata.", call. = FALSE)

# Patients expected to contribute to paired analyses based on the expression
# metadata alone. Metric-specific paired n may be lower if a deconvolution
# method/cell type has no valid value at one or both timepoints.
expected_paired_patients <- meta_retained %>%
  distinct(Patient_ID, cluster, Timepoint) %>%
  group_by(Patient_ID, cluster) %>%
  summarise(n_timepoints = n_distinct(Timepoint), .groups = "drop") %>%
  filter(n_timepoints == 2) %>%
  select(Patient_ID, cluster) %>%
  apply_cluster_mapping() %>%
  mutate(cluster = droplevels(cluster))

# compact output: expected paired patients retained in memory

keep_samples <- meta_retained$sample_id
expr_retained <- expr_combined[, keep_samples, drop = FALSE]
meta_retained <- meta_retained %>% arrange(match(sample_id, colnames(expr_retained)))
stopifnot(identical(meta_retained$sample_id, colnames(expr_retained)))

# Retain source cohort only as metadata; do NOT split the expression matrix.
meta_combined <- meta_retained
expr_analysis <- expr_retained[, meta_combined$sample_id, drop = FALSE]

if (ncol(expr_analysis) == 0) {
  stop("No retained samples are available for combined immunedeconv analysis.", call. = FALSE)
}

stopifnot(identical(meta_combined$sample_id, colnames(expr_analysis)))

cat("Combined immunedeconv analysis set:\n")
cat("  Samples:", ncol(expr_analysis), "\n")
cat("  Patients:", dplyr::n_distinct(meta_combined$patient_key), "\n")
cat("  NKI samples:", sum(meta_combined$dataset == "NKI"), "\n")
cat("  SMC samples:", sum(meta_combined$dataset == "SMC"), "\n")
cat("  C1 samples:", sum(meta_combined$cluster == "C1"), "\n")
cat("  C2 samples:", sum(meta_combined$cluster == "C2"), "\n")
cat("  C3 samples:", sum(meta_combined$cluster == "C3"), "\n")
```

# Run immunedeconv

The complete retained NKI+SMC TPM matrix is analyzed in a single `immunedeconv` run per method. NKI/SMC source cohort is retained only as metadata for provenance and QC; it does not define the deconvolution or statistical groups.

```{r immunedeconv-functions}
make_indications <- function(mat, method) {
  if (method %in% c("timer", "consensus_tme")) {
    setNames(rep(tolower(TUMOR_INDICATION), ncol(mat)), colnames(mat))
  } else {
    NULL
  }
}

deconv_to_long <- function(res, method) {
  if (is.null(res)) return(NULL)

  if (method == "estimate") {
    long <- res %>%
      as.data.frame() %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id")

    long %>%
      pivot_longer(-sample_id, names_to = "Cell_Type", values_to = "Value") %>%
      mutate(Method = method, Value = as.numeric(Value))
  } else {
    if (!"cell_type" %in% colnames(res)) return(NULL)

    res %>%
      column_to_rownames("cell_type") %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id") %>%
      pivot_longer(-sample_id, names_to = "Cell_Type", values_to = "Value") %>%
      mutate(Method = method, Value = as.numeric(Value))
  }
}

run_methods <- function(expr, meta, analysis_label = "NKI+SMC combined") {
  out <- list()
  log <- list()

  for (m in METHODS_TO_RUN) {
    message("Running ", analysis_label, " / ", m)
    ind <- make_indications(expr, m)
    convergence_warning <- FALSE

    res <- withr::with_dir(DIRS$tmp, {
      if (m == "estimate") {
        tryCatch(immunedeconv::deconvolute_estimate(expr), error = function(e) { message("ERROR: ", e$message); NULL })
      } else if (m == "epic") {
        tryCatch(
          withCallingHandlers(
            immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE),
            warning = function(w) {
              if (grepl("didn't fully converge", w$message, ignore.case = TRUE)) {
                convergence_warning <<- TRUE
                message("EPIC partial convergence warning detected.")
              }
              invokeRestart("muffleWarning")
            }
          ),
          error = function(e) { message("ERROR: ", e$message); NULL }
        )
      } else if (m == "quantiseq") {
        tryCatch(immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      } else if (!is.null(ind)) {
        tryCatch(immunedeconv::deconvolute(expr, m, indications = ind, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      } else {
        tryCatch(immunedeconv::deconvolute(expr, m, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      }
    })

    if (m == "epic" && !is.null(res) && "cell_type" %in% colnames(res)) {
      n_samples_returned <- ncol(res) - 1
      rate <- n_samples_returned / ncol(expr)
      if (rate < 0.30) {
        message("Discarding EPIC output because <30% of samples converged.")
        res <- NULL
      }
    }

    long <- deconv_to_long(res, m)

    if (!is.null(long)) {
      # immunedeconv should preserve input sample names. Check this explicitly
      # before downstream aggregation so altered IDs cannot silently create NAs.
      unexpected_ids <- setdiff(unique(long$sample_id), meta$sample_id)
      if (length(unexpected_ids) > 0) {
        warning(
          analysis_label, " / ", m,
          ": result sample IDs not found in metadata: ",
          paste(head(unexpected_ids, 20), collapse = ", "),
          call. = FALSE
        )
      }

      long <- long %>%
        left_join(
          meta %>% select(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster),
          by = "sample_id"
        )

      n_missing_meta <- sum(is.na(long$Patient_ID) | is.na(long$Timepoint) | is.na(long$cluster))
      if (n_missing_meta > 0) {
        warning(
          analysis_label, " / ", m, ": ", n_missing_meta,
          " result row(s) lack metadata and will be removed downstream.",
          call. = FALSE
        )
      }
    }

    out[[m]] <- long
    log[[m]] <- tibble(
      dataset = analysis_label,
      Method = m,
      status = ifelse(is.null(long), "FAILED", "OK"),
      n_samples = ifelse(is.null(long), 0, n_distinct(long$sample_id)),
      n_celltypes = ifelse(is.null(long), 0, n_distinct(long$Cell_Type)),
      notes = case_when(
        is.null(long) ~ "Method failed or returned no usable result",
        m == "epic" & convergence_warning ~ "Partial convergence",
        TRUE ~ "Success"
      )
    )
  }

  list(immune_long = bind_rows(out), run_log = bind_rows(log))
}
```

```{r run-immunedeconv}
if (params$run_immunedeconv) {
  res_combined <- run_methods(expr_analysis, meta_combined, "NKI+SMC combined")
  immune_long_raw <- res_combined$immune_long
  run_summary <- res_combined$run_log
} else {
  immune_long_raw <- tibble()
  run_summary <- tibble()
}
```

# Import combined CIBERSORTx results

The combined CIBERSORTx file is treated as one external deconvolution result for the same pooled NKI+SMC sample set. Source cohort is inferred only for metadata/QC. The non-cellular CIBERSORTx columns `P-value`, `Correlation`, and `RMSE` are removed before conversion to long format.

```{r cibersortx-import}
build_cibersortx_metadata <- function(sample_ids) {
  meta <- tibble(sample_id = as.character(sample_ids)) %>%
    mutate(
      Timepoint = factor(get_timepoint(sample_id), levels = c("Biopsy", "Surgery")),
      patient_key = get_patient_key_from_cibersortx(sample_id),
      Patient_ID = patient_key,
      dataset = infer_dataset(patient_key)
    ) %>%
    left_join(clusters %>% select(patient_key, cluster), by = "patient_key") %>%
    mutate(cluster = factor(cluster, levels = levels(clusters$cluster)))

  bad_tp <- meta %>% filter(is.na(Timepoint))
  if (nrow(bad_tp) > 0) {
    stop(
      "Unrecognized CIBERSORTx timepoint ID(s): ",
      paste(bad_tp$sample_id, collapse = ", "),
      call. = FALSE
    )
  }

  # Multiple rows per canonical patient/timepoint are permitted only when they
  # are explicit technical replicates of the same raw sample base (_RepN).
  ciber_collision <- meta %>%
    dplyr::mutate(raw_base = stringr::str_remove(sample_id, "_Rep[0-9]+$")) %>%
    dplyr::group_by(Patient_ID, Timepoint) %>%
    dplyr::summarise(
      n_rows = dplyr::n(),
      n_raw_bases = dplyr::n_distinct(raw_base),
      raw_ids = paste(sample_id, collapse = ";"),
      .groups = "drop"
    ) %>%
    dplyr::filter(n_rows > 1, n_raw_bases > 1)

  if (nrow(ciber_collision) > 0) {
    stop(
      "Distinct CIBERSORTx sample IDs collapse to the same canonical patient/timepoint. ",
      "These are not simple _RepN technical replicates. Examples: ",
      paste(utils::head(paste0(ciber_collision$Patient_ID, "/", ciber_collision$Timepoint, "=", ciber_collision$raw_ids), 10), collapse = " | "),
      call. = FALSE
    )
  }

  meta
}

read_cibersort_noqc_long <- function(path, expression_sample_ids = NULL) {
  if (!file.exists(path)) {
    warning("CIBERSORTx file not found: ", path, call. = FALSE)
    return(NULL)
  }

  cs <- data.table::fread(path, data.table = FALSE, check.names = FALSE)

  if (!"Mixture" %in% colnames(cs)) {
    stop("Combined CIBERSORTx file must contain a 'Mixture' column: ", path, call. = FALSE)
  }

  cs$Mixture <- trimws(as.character(cs$Mixture))
  if (anyDuplicated(cs$Mixture)) {
    dup <- unique(cs$Mixture[duplicated(cs$Mixture)])
    stop("Duplicate CIBERSORTx Mixture IDs: ", paste(dup, collapse = ", "), call. = FALSE)
  }

  # These are CIBERSORTx model-quality metrics, not immune-cell estimates.
  # Drop them BEFORE selecting/pivoting immune-cell columns.
  qc_cols <- c(
    "P-value", "P.value", "P value",
    "Correlation", "RMSE",
    "Absolute score", "Absolute.Score"
  )
  qc_present <- intersect(qc_cols, colnames(cs))
  if (length(qc_present) > 0) {
    message("Removing CIBERSORTx QC column(s): ", paste(qc_present, collapse = ", "))
    cs <- cs %>% dplyr::select(-dplyr::all_of(qc_present))
  }

  value_cols <- setdiff(colnames(cs), "Mixture")
  if (length(value_cols) == 0) {
    stop("No CIBERSORTx immune-cell columns remain after QC-column removal.", call. = FALSE)
  }

  bad_value_cols <- value_cols[vapply(value_cols, function(nm) {
    x <- cs[[nm]]
    y <- suppressWarnings(as.numeric(x))
    any(is.na(y) & !is.na(x) & trimws(as.character(x)) != "")
  }, logical(1))]

  if (length(bad_value_cols) > 0) {
    stop(
      "Non-numeric CIBERSORTx immune-cell column(s): ",
      paste(bad_value_cols, collapse = ", "),
      call. = FALSE
    )
  }

  meta <- build_cibersortx_metadata(cs$Mixture)

  unmatched_cluster <- meta %>% filter(is.na(cluster))
  if (nrow(unmatched_cluster) > 0) {
    warning(
      nrow(unmatched_cluster),
      " CIBERSORTx sample(s) do not match the CIM cluster table and will be excluded: ",
      paste(unmatched_cluster$sample_id, collapse = ", "),
      call. = FALSE
    )
  }

  # For the current combined files, Mixture IDs should match the combined TPM
  # sample IDs exactly. Keep this as an explicit reproducibility/QC check.
  if (!is.null(expression_sample_ids)) {
    canonical_cs <- paste0(meta$patient_key, ifelse(meta$Timepoint == "Biopsy", "_B", "_S"))
    missing_from_ciber <- setdiff(expression_sample_ids, canonical_cs)
    extra_in_ciber <- setdiff(canonical_cs, expression_sample_ids)

    if (length(missing_from_ciber) > 0 || length(extra_in_ciber) > 0) {
      warning(
        "CIBERSORTx/TPM sample-set mismatch. Missing from CIBERSORTx: ",
        ifelse(length(missing_from_ciber), paste(missing_from_ciber, collapse = ", "), "none"),
        "; extra in CIBERSORTx: ",
        ifelse(length(extra_in_ciber), paste(extra_in_ciber, collapse = ", "), "none"),
        call. = FALSE
      )
    }
  }

  cs_long <- cs %>%
    rename(sample_id = Mixture) %>%
    pivot_longer(
      cols = all_of(value_cols),
      names_to = "Cell_Type",
      values_to = "Value"
    ) %>%
    mutate(Method = "CIBERSORTx", Value = as.numeric(Value)) %>%
    left_join(
      meta %>% select(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster),
      by = "sample_id"
    ) %>%
    filter(!is.na(cluster), !is.na(Timepoint), is.finite(Value))

  cs_long
}

if (params$run_cibersortx_import) {
  ciber_combined_long <- read_cibersort_noqc_long(
    FILES$cibersortx,
    expression_sample_ids = colnames(expr_analysis)
  )

  if (!is.null(ciber_combined_long)) {
    immune_long_raw <- bind_rows(immune_long_raw, ciber_combined_long)
    run_summary <- bind_rows(
      run_summary,
      tibble(
        dataset = "NKI+SMC combined",
        Method = "CIBERSORTx",
        status = "OK",
        n_samples = n_distinct(ciber_combined_long$sample_id),
        n_celltypes = n_distinct(ciber_combined_long$Cell_Type),
        notes = "Imported from one combined CIBERSORTx relative-mode file; QC columns removed before reshaping"
      )
    )
  }
}

# Save a compact cross-source sample-matching table for reproducibility.
if (exists("ciber_combined_long") && !is.null(ciber_combined_long)) {
  ciber_sample_qc <- ciber_combined_long %>%
    distinct(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster) %>%
    mutate(
      canonical_sample_id = paste0(
        patient_key,
        ifelse(Timepoint == "Biopsy", "_B", "_S")
      )
    ) %>%
    select(canonical_sample_id, ciber_sample_id = sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster) %>%
    full_join(
      meta_combined %>%
        transmute(
          canonical_sample_id = sample_id,
          expression_present = TRUE
        ),
      by = "canonical_sample_id"
    ) %>%
    mutate(
      cibersortx_present = !is.na(ciber_sample_id),
      expression_present = dplyr::coalesce(expression_present, FALSE)
    ) %>%
    arrange(canonical_sample_id)
  # Compact pipeline: CIBERSORTx QC retained in memory only.
}

if (nrow(immune_long_raw) == 0 || !all(c("Method", "Cell_Type", "Value") %in% names(immune_long_raw))) {
  stop(
    "No usable immune-deconvolution results are available. ",
    "Check run_immunedeconv/run_cibersortx_import and the method run summary.",
    call. = FALSE
  )
}

# Compact pipeline: no intermediate RDS written.
# compact output: omitted run_summary_methods.csv
knitr::kable(run_summary, caption = "Method run summary")
```

# Patient-level aggregation

Technical replicates are averaged to create one value per patient, method, cell type, timepoint, and cluster.

```{r patient-aggregation}
# Clean long-format immune deconvolution results
immune_long_clean <- immune_long_raw %>%
  dplyr::filter(
    !is.na(Patient_ID),
    !is.na(Timepoint),
    !is.na(cluster),
    is.finite(Value)
  )

# Apply biological CIM-state labels before patient-level aggregation.
# Biological mapping used in this workflow:
#   1 -> C1
#   2 -> C2
#   3 -> C3
immune_long_clean <- apply_cluster_mapping(immune_long_clean)

# Average technical replicates at the patient level
immune_patient <- immune_long_clean %>%
  dplyr::group_by(Method, Cell_Type, Patient_ID, Timepoint, cluster) %>%
  dplyr::summarise(
    dataset = dplyr::first(dataset),
    Value = mean(Value, na.rm = TRUE),
    n_replicates = dplyr::n(),
    .groups = "drop"
  )

# Compact pipeline: original patient-level object retained in memory only.

# Store method-specific lower y-axis limits for plotting
method_ymin_original <- immune_patient %>%
  dplyr::group_by(Method) %>%
  dplyr::summarise(
    ymin = ifelse(
      min(Value, na.rm = TRUE) < 0,
      min(Value, na.rm = TRUE),
      0
    ),
    .groups = "drop"
  )

# Quick summary
immune_patient_summary <- immune_patient %>%
  dplyr::count(Method, cluster, Timepoint, name = "n_patient_level_values")

knitr::kable(
  immune_patient_summary,
  caption = "Patient-level immune deconvolution values after cluster mapping"
)
```

# T-cell aggregation

Different deconvolution algorithms report T-cell populations at different levels of granularity. This section optionally creates broader CD4+ and CD8+ categories to simplify downstream interpretation.

Aggregation is performed after patient-level averaging and before statistical analysis.

## xCell aggregation

When `AGGREGATE_XCELL_TCELLS = TRUE`, all xCell populations matching `T cell CD4+` are summed into `T cell CD4+ (aggregated)`, and all xCell populations matching `T cell CD8+` are summed into `T cell CD8+ (aggregated)`.

## CIBERSORTx aggregation

CIBERSORTx supports two strategies:

- `refined`: creates `T cells CD4+ (general)` from naive, memory resting, and memory activated CD4 T cells. Follicular helper T cells are kept separate. This is recommended when you want to preserve a distinct Tfh population.
- `simple`: creates `T cells CD4+ (total)` from naive, memory resting, memory activated, and follicular helper T cells. This is useful for broad CD4 versus CD8 summaries.

```{r tcell-aggregation}
aggregate_xcell_tcells <- function(immune_patient_df) {
  if (!AGGREGATE_XCELL_TCELLS) return(list(immune_patient_aggregated = immune_patient_df, aggregation_details = NULL))

  xcell_data <- immune_patient_df %>% filter(Method == "xcell")
  non_xcell <- immune_patient_df %>% filter(Method != "xcell")
  if (nrow(xcell_data) == 0) return(list(immune_patient_aggregated = immune_patient_df, aggregation_details = NULL))

  cd4_aggregated <- xcell_data %>%
    filter(grepl("T cell CD4\\+", Cell_Type)) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      dataset = dplyr::first(dataset),
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD4+ (aggregated)")

  cd8_aggregated <- xcell_data %>%
    filter(grepl("T cell CD8\\+", Cell_Type)) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      dataset = dplyr::first(dataset),
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD8+ (aggregated)")

  other_xcell <- xcell_data %>% filter(!grepl("T cell CD4\\+", Cell_Type), !grepl("T cell CD8\\+", Cell_Type))

  immune_patient_aggregated <- bind_rows(
    non_xcell,
    other_xcell,
    cd4_aggregated %>% select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_aggregated %>% select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )

  aggregation_details <- bind_rows(
    cd4_aggregated %>% select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_aggregated %>% select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )

  list(immune_patient_aggregated = immune_patient_aggregated, aggregation_details = aggregation_details)
}

aggregate_cibersort_tcells <- function(df, strategy = "refined") {
  if (!AGGREGATE_CIBERSORT_TCELLS) return(list(aggregated = df, details = NULL))
  if (nrow(df) == 0) return(list(aggregated = df, details = NULL))

  if (strategy == "refined") {
    cd4_subtypes <- c("T cells CD4 naive", "T cells CD4 memory resting", "T cells CD4 memory activated")
    cd4_name <- "T cells CD4+ (general)"
    cd8_subtypes <- c("T cells CD8")
    cd8_name <- "T cells CD8+ (total)"
  } else if (strategy == "simple") {
    cd4_subtypes <- c("T cells CD4 naive", "T cells CD4 memory resting", "T cells CD4 memory activated", "T cells follicular helper")
    cd4_name <- "T cells CD4+ (total)"
    cd8_subtypes <- c("T cells CD8")
    cd8_name <- "T cells CD8+ (total)"
  } else {
    stop("CIBERSORT_AGGREGATION_STRATEGY must be 'refined' or 'simple'.")
  }

  cd4_agg <- df %>%
    filter(Cell_Type %in% cd4_subtypes) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      dataset = dplyr::first(dataset),
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd4_name)

  cd8_agg <- df %>%
    filter(Cell_Type %in% cd8_subtypes) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      dataset = dplyr::first(dataset),
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd8_name)

  aggregated_subtypes <- c(cd4_subtypes, cd8_subtypes)
  other_cells <- df %>% filter(!Cell_Type %in% aggregated_subtypes)

  aggregated <- bind_rows(
    other_cells,
    cd4_agg %>% select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_agg %>% select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )

  details <- bind_rows(
    cd4_agg %>% select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_agg %>% select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )

  list(aggregated = aggregated, details = details)
}

xcell_agg <- aggregate_xcell_tcells(immune_patient)
immune_after_xcell <- xcell_agg$immune_patient_aggregated

if (!is.null(xcell_agg$aggregation_details)) {
  write.csv(xcell_agg$aggregation_details, file.path(DIRS$tables, "xcell_tcell_aggregation_details.csv"), row.names = FALSE)
}

ciber_df <- immune_after_xcell %>% filter(Method == "CIBERSORTx")
non_ciber <- immune_after_xcell %>% filter(Method != "CIBERSORTx")
ciber_agg <- aggregate_cibersort_tcells(ciber_df, CIBERSORT_AGGREGATION_STRATEGY)

immune_patient_aggregated <- bind_rows(non_ciber, ciber_agg$aggregated)
immune_patient_aggregated <- apply_cluster_mapping(immune_patient_aggregated)

if (!is.null(ciber_agg$details)) {
  write.csv(ciber_agg$details, file.path(DIRS$tables, "cibersort_tcell_aggregation_details.csv"), row.names = FALSE)
}

# Compact pipeline: aggregated patient-level object retained in memory only.
# compact output: omitted immune_patient_level_AGGREGATED.csv

method_ymin_aggregated <- immune_patient_aggregated %>%
  group_by(Method) %>%
  summarise(ymin = if (min(Value, na.rm = TRUE) < 0) min(Value, na.rm = TRUE) else 0, .groups = "drop")
```

# Patient-level integrity checks ------------------------------------------------
assert_unique_patient_values <- function(df, context = "patient-level table") {
  required <- c("Method", "Cell_Type", "Patient_ID", "Timepoint", "cluster", "Value")
  missing <- base::setdiff(required, names(df))
  if (length(missing) > 0) {
    stop(context, " is missing required columns: ", paste(missing, collapse = ", "), call. = FALSE)
  }

  dup <- df %>%
    dplyr::filter(!is.na(Patient_ID), !is.na(Timepoint), !is.na(cluster)) %>%
    dplyr::count(Method, Cell_Type, Patient_ID, Timepoint, cluster, name = "n_rows") %>%
    dplyr::filter(n_rows > 1)

  if (nrow(dup) > 0) {
    stop(
      context, " contains duplicate patient-level values. ",
      "Technical replicates must be resolved before statistics/plots. Examples: ",
      paste(utils::head(paste0(dup$Method, "/", dup$Cell_Type, "/", dup$Patient_ID, "/", dup$Timepoint), 10), collapse = " | "),
      call. = FALSE
    )
  }

  invisible(TRUE)
}

assert_unique_patient_values(immune_patient, "ORIGINAL immune patient table")
assert_unique_patient_values(immune_patient_aggregated, "AGGREGATED immune patient table")

# Unpaired statistics

For each method, cell type, and timepoint, the pipeline compares immune estimates across CIMIC clusters.

- If `k = 2`, the global test is Wilcoxon rank-sum.
- If `k >= 3`, the global test is Kruskal-Wallis, followed by Dunn post hoc testing with Benjamini-Hochberg FDR correction.

```{r unpaired-stats}
summarise_by_cluster <- function(df) {
  df %>%
    group_by(cluster) %>%
    summarise(n = n(), mean = mean(Value, na.rm = TRUE), sd = sd(Value, na.rm = TRUE), .groups = "drop") %>%
    mutate(cluster = as.character(cluster))
}

empty_skip_tbl <- function() {
  tibble(analysis = character(), Method = character(), Cell_Type = character(), Timepoint = character(), reason = character(), k = integer(), n_total = integer(), n_unique_value = integer(), min_cluster_n = integer())
}

make_skip_row <- function(Method, Cell_Type, Timepoint, analysis, reason, k = NA_integer_, n_total = NA_integer_, n_unique_value = NA_integer_, min_cluster_n = NA_integer_) {
  tibble(analysis = analysis, Method = as.character(Method), Cell_Type = as.character(Cell_Type), Timepoint = as.character(Timepoint), reason = reason, k = as.integer(k), n_total = as.integer(n_total), n_unique_value = as.integer(n_unique_value), min_cluster_n = as.integer(min_cluster_n))
}

unpaired_global_one <- function(df, Method, Cell_Type, Timepoint) {
  df <- df %>%
    dplyr::filter(is.finite(Value), !is.na(cluster), !is.na(Patient_ID))

  # Statistics are strictly patient-level: one value per patient for the
  # current Method x Cell_Type x Timepoint.
  dup <- df %>%
    dplyr::count(Patient_ID, name = "n_rows") %>%
    dplyr::filter(n_rows > 1)
  if (nrow(dup) > 0) {
    stop(
      "Duplicate patients reached unpaired statistics for ",
      Method, " / ", Cell_Type, " / ", as.character(Timepoint),
      ": ", paste(dup$Patient_ID, collapse = ", "),
      call. = FALSE
    )
  }

  kk <- dplyr::n_distinct(df$cluster)
  n_total <- nrow(df)
  n_unique <- dplyr::n_distinct(df$Value)
  group_sizes <- table(df$cluster)
  min_n <- if (kk > 0) min(as.integer(group_sizes)) else NA_integer_

  if (kk < 2) {
    return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_global", "k<2", kk, n_total, n_unique, min_n)))
  }

  summ <- summarise_by_cluster(df)

  if (!is.na(min_n) && min_n < MIN_UNPAIRED_GROUP_N) {
    return(list(
      out = tibble(
        Method = as.character(Method), Cell_Type = as.character(Cell_Type),
        Timepoint = as.character(Timepoint), test = ifelse(kk == 2, "wilcox_rank_sum", "kruskal"),
        p_value = NA_real_, n_clusters = kk, note = paste0("skipped_min_group_n<", MIN_UNPAIRED_GROUP_N),
        summary = list(summ)
      ),
      skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_global", "group_too_small", kk, n_total, n_unique, min_n)
    ))
  }

  if (n_unique < 2) {
    return(list(
      out = tibble(
        Method = as.character(Method), Cell_Type = as.character(Cell_Type),
        Timepoint = as.character(Timepoint), test = ifelse(kk == 2, "wilcox_rank_sum", "kruskal"),
        p_value = NA_real_, n_clusters = kk, note = "skipped_no_variation", summary = list(summ)
      ),
      skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_global", "no_variation", kk, n_total, n_unique, min_n)
    ))
  }

  p <- if (kk == 2) {
    # Two independent CIM groups -> two-sided Wilcoxon rank-sum test.
    # exact = FALSE is intentional because deconvolution outputs commonly
    # contain ties/zeros; R's exact calculation is not valid with ties.
    # IMPORTANT: do not pass paired=FALSE to the formula interface.
    # stats::wilcox.test.formula() does not accept a paired argument and can
    # therefore return an error/NA when wrapped in tryCatch. With two CIM
    # groups, the formula interface is intrinsically an independent-samples
    # Wilcoxon rank-sum test.
    tryCatch(
      stats::wilcox.test(
        Value ~ cluster,
        data = df,
        alternative = "two.sided",
        exact = FALSE,
        correct = TRUE
      )$p.value,
      error = function(e) {
        warning(
          "Unpaired Wilcoxon failed for ", Method, " / ", Cell_Type, " / ",
          as.character(Timepoint), ": ", conditionMessage(e)
        )
        NA_real_
      }
    )
  } else {
    tryCatch(
      stats::kruskal.test(Value ~ cluster, data = df)$p.value,
      error = function(e) NA_real_
    )
  }

  list(
    out = tibble(
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      Timepoint = as.character(Timepoint),
      test = ifelse(kk == 2, "wilcox_rank_sum", "kruskal"),
      p_value = p,
      n_clusters = kk,
      note = NA_character_,
      summary = list(summ)
    ),
    skip = NULL
  )
}

unpaired_dunn_one <- function(df, Method, Cell_Type, Timepoint, drop_small_groups_n = 2) {
  df2 <- df %>%
    filter(is.finite(Value), !is.na(cluster)) %>%
    group_by(cluster) %>%
    filter(n() >= drop_small_groups_n) %>%
    ungroup()

  kk <- n_distinct(df2$cluster)
  n_total <- nrow(df2)
  n_unique <- n_distinct(df2$Value)
  min_n <- if (kk > 0) min(as.integer(table(df2$cluster))) else NA_integer_

  if (kk < 3) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "k<3_after_filter", kk, n_total, n_unique, min_n)))
  if (n_unique < 2) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "no_variation", kk, n_total, n_unique, min_n)))

  cl_summ <- summarise_by_cluster(df2) %>% rename(n_group = n, mean_group = mean, sd_group = sd)
  dt <- tryCatch(suppressWarnings(FSA::dunnTest(Value ~ cluster, data = df2, method = "bh")$res), error = function(e) NULL)

  if (is.null(dt)) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "dunn_error", kk, n_total, n_unique, min_n)))

  desired_pairwise_order <- c("C1 - C2", "C3 - C2", "C1 - C3")

  out <- as_tibble(dt) %>%
    rename(comparison = Comparison, z = Z, p_unadj = P.unadj, fdr = P.adj) %>%
    mutate(
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      Timepoint = as.character(Timepoint),
      comparison_original = as.character(comparison),
      z = as.numeric(z),
      z = case_when(
        comparison_original %in% c("C2 - C1", "C2 - C3", "C3 - C1") ~ -z,
        TRUE ~ z
      ),
      comparison = case_when(
        comparison_original %in% c("C1 - C2", "C2 - C1") ~ "C1 - C2",
        comparison_original %in% c("C3 - C2", "C2 - C3") ~ "C3 - C2",
        comparison_original %in% c("C1 - C3", "C3 - C1") ~ "C1 - C3",
        TRUE ~ comparison_original
      ),
      cluster1 = sub(" -.*$", "", comparison),
      cluster2 = sub("^.*- ", "", comparison)
    ) %>%
    left_join(cl_summ %>% rename(cluster1 = cluster, n1 = n_group, mean1 = mean_group, sd1 = sd_group), by = "cluster1") %>%
    left_join(cl_summ %>% rename(cluster2 = cluster, n2 = n_group, mean2 = mean_group, sd2 = sd_group), by = "cluster2") %>%
    mutate(
      comparison = factor(comparison, levels = desired_pairwise_order, ordered = TRUE)
    ) %>%
    arrange(comparison) %>%
    mutate(comparison = as.character(comparison)) %>%
    select(-comparison_original) %>%
    relocate(Method, Cell_Type, Timepoint, comparison, cluster1, cluster2, n1, mean1, sd1, n2, mean2, sd2, z, p_unadj, fdr)

  list(out = out, skip = NULL)
}

compute_unpaired_all <- function(immune_patient_df, out_prefix, drop_small_groups_n = 2, write_outputs = FALSE) {
  # Safety step: all unpaired statistics use mapped CIM groups.
  # Numeric cluster IDs are mapped first; pooled analyses compare C1, C2, and C3 regardless of NKI/SMC source cohort.
  immune_patient_df <- apply_cluster_mapping(immune_patient_df) %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  keys <- immune_patient_df %>% dplyr::distinct(Method, Cell_Type, Timepoint) %>% dplyr::arrange(Timepoint, Method, Cell_Type)

  global_rows <- list()
  dunn_rows <- list()
  skip_rows <- list()

  for (i in seq_len(nrow(keys))) {
    key <- keys[i, ]
    method_i <- key$Method[[1]]
    cell_type_i <- key$Cell_Type[[1]]
    timepoint_i <- key$Timepoint[[1]]

    df_sub <- immune_patient_df %>%
      dplyr::filter(
        .data$Method == .env$method_i,
        .data$Cell_Type == .env$cell_type_i,
        .data$Timepoint == .env$timepoint_i,
        is.finite(.data$Value),
        !is.na(.data$Patient_ID),
        !is.na(.data$cluster)
      )

    g <- unpaired_global_one(df_sub, method_i, cell_type_i, timepoint_i)
    if (!is.null(g$out)) global_rows[[length(global_rows) + 1]] <- g$out
    if (!is.null(g$skip)) skip_rows[[length(skip_rows) + 1]] <- g$skip

    d <- unpaired_dunn_one(df_sub, method_i, cell_type_i, timepoint_i, drop_small_groups_n)
    if (!is.null(d$out)) dunn_rows[[length(dunn_rows) + 1]] <- d$out
    if (!is.null(d$skip)) skip_rows[[length(skip_rows) + 1]] <- d$skip
  }

  global_tbl <- if (length(global_rows)) bind_rows(global_rows) else tibble()
  dunn_tbl <- if (length(dunn_rows)) bind_rows(dunn_rows) else tibble()
  skipped <- if (length(skip_rows)) bind_rows(skip_rows) else empty_skip_tbl()

  global_tbl <- global_tbl %>% mutate(FDR = p.adjust(p_value, method = "BH"))

  global_wide <- global_tbl %>%
    tidyr::unnest(summary) %>%
    pivot_wider(
      id_cols = c(Method, Cell_Type, Timepoint, test, p_value, FDR, n_clusters, note),
      names_from = cluster,
      values_from = c(n, mean, sd),
      names_glue = "{cluster}_{.value}"
    ) %>%
    arrange(Timepoint, Method, Cell_Type)

  if (isTRUE(write_outputs)) {
    write.csv(global_wide, file.path(DIRS$tables, paste0(out_prefix, "_unpaired_global.csv")), row.names = FALSE)
    write.csv(dunn_tbl, file.path(DIRS$tables, paste0(out_prefix, "_unpaired_dunn.csv")), row.names = FALSE)
  }

  list(global = global_wide, dunn = dunn_tbl, skipped = skipped)
}

unpaired_original <- compute_unpaired_all(immune_patient, "ORIGINAL", drop_small_groups_n = 2, write_outputs = FALSE)
unpaired_aggregated <- compute_unpaired_all(immune_patient_aggregated, "AGGREGATED", drop_small_groups_n = 2, write_outputs = TRUE)
```

# Paired statistics

Within each CIMIC cluster, the pipeline performs paired Wilcoxon signed-rank tests comparing surgery versus biopsy for each method and cell type.

```{r paired-stats}
make_skip_row_paired <- function(Method, Cell_Type, cluster, reason, n_pairs = NA_integer_, n_unique_delta = NA_integer_) {
  tibble(analysis = "paired", Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster), reason = reason, n_pairs = n_pairs, n_unique_delta = n_unique_delta)
}

paired_one <- function(df, Method, Cell_Type, cluster_label, expected_pairs_df) {
  expected_cluster <- expected_pairs_df %>%
    dplyr::filter(as.character(cluster) == as.character(cluster_label)) %>%
    dplyr::distinct(Patient_ID)

  df <- df %>%
    dplyr::filter(!is.na(Patient_ID), !is.na(Timepoint), !is.na(cluster))

  # At this stage technical replicates should already have been collapsed.
  dup <- df %>%
    dplyr::count(Patient_ID, Timepoint, name = "n_rows") %>%
    dplyr::filter(n_rows > 1)
  if (nrow(dup) > 0) {
    stop(
      "Duplicate patient/timepoint rows reached paired statistics for ",
      Method, " / ", Cell_Type, " / ", as.character(cluster_label),
      ": ", paste0(dup$Patient_ID, "(", dup$Timepoint, ")", collapse = ", "),
      call. = FALSE
    )
  }

  observed_availability <- df %>%
    dplyr::mutate(
      Timepoint_chr = as.character(Timepoint),
      valid_value = is.finite(Value)
    ) %>%
    dplyr::group_by(Patient_ID) %>%
    dplyr::summarise(
      has_valid_biopsy = any(Timepoint_chr == "Biopsy" & valid_value, na.rm = TRUE),
      has_valid_surgery = any(Timepoint_chr == "Surgery" & valid_value, na.rm = TRUE),
      .groups = "drop"
    )

  pair_audit <- expected_cluster %>%
    dplyr::left_join(observed_availability, by = "Patient_ID") %>%
    dplyr::mutate(
      has_valid_biopsy = tidyr::replace_na(has_valid_biopsy, FALSE),
      has_valid_surgery = tidyr::replace_na(has_valid_surgery, FALSE),
      included_pair = has_valid_biopsy & has_valid_surgery,
      exclusion_reason = dplyr::case_when(
        included_pair ~ NA_character_,
        !has_valid_biopsy & !has_valid_surgery ~ "missing_or_nonfinite_both_timepoints",
        !has_valid_biopsy ~ "missing_or_nonfinite_biopsy",
        !has_valid_surgery ~ "missing_or_nonfinite_surgery",
        TRUE ~ "incomplete_pair"
      ),
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      cluster = as.character(cluster_label)
    )

  wide <- df %>%
    dplyr::filter(is.finite(Value)) %>%
    dplyr::select(Patient_ID, Timepoint, Value) %>%
    tidyr::pivot_wider(names_from = Timepoint, values_from = Value) %>%
    dplyr::filter(is.finite(Biopsy), is.finite(Surgery))

  unexpected_wide <- wide %>%
    dplyr::anti_join(expected_cluster, by = "Patient_ID")
  if (nrow(unexpected_wide) > 0) {
    stop(
      "Unexpected patient(s) entered paired analysis for ",
      Method, " / ", Cell_Type, " / ", as.character(cluster_label),
      ": ", paste(unexpected_wide$Patient_ID, collapse = ", "),
      call. = FALSE
    )
  }

  n_candidates <- nrow(pair_audit)
  n_pairs_actual <- sum(pair_audit$included_pair)
  n_dropped <- n_candidates - n_pairs_actual

  if (nrow(wide) != n_pairs_actual) {
    stop(
      "Paired-count consistency check failed for ", Method, " / ", Cell_Type,
      " / ", as.character(cluster_label), ": wide table has ", nrow(wide),
      " complete pairs but audit counted ", n_pairs_actual, ".",
      call. = FALSE
    )
  }

  if (nrow(wide) == 0) {
    return(list(
      stats = NULL,
      wide = NULL,
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "no_complete_pairs", 0),
      pair_audit = pair_audit
    ))
  }

  wide <- wide %>%
    dplyr::mutate(
      Delta = Surgery - Biopsy,
      nonzero_delta = Delta != 0
    )

  n_pairs <- nrow(wide)
  n_zero_delta <- sum(!wide$nonzero_delta)
  n_nonzero_delta <- sum(wide$nonzero_delta)
  n_unique_delta <- dplyr::n_distinct(wide$Delta)

  stats_base <- tibble(
    Method = as.character(Method),
    Cell_Type = as.character(Cell_Type),
    cluster = as.character(cluster_label),
    n_candidates = n_candidates,
    n_pairs = n_pairs,
    n_dropped = n_dropped,
    n_zero_delta = n_zero_delta,
    n_nonzero_delta = n_nonzero_delta,
    mean_biopsy = mean(wide$Biopsy),
    sd_biopsy = stats::sd(wide$Biopsy),
    mean_surgery = mean(wide$Surgery),
    sd_surgery = stats::sd(wide$Surgery),
    mean_delta = mean(wide$Delta),
    sd_delta = stats::sd(wide$Delta)
  )

  if (n_pairs < MIN_PAIRED_N) {
    return(list(
      stats = stats_base %>% dplyr::mutate(p_value = NA_real_, note = paste0("skipped_n_pairs<", MIN_PAIRED_N)),
      wide = wide %>% dplyr::mutate(Method = as.character(Method), Cell_Type = as.character(Cell_Type), cluster = as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "n_pairs_too_small", n_pairs, n_unique_delta),
      pair_audit = pair_audit
    ))
  }

  # Wilcoxon signed-rank discards zero differences from the rank calculation.
  # Only an all-zero delta vector makes the signed-rank test undefined.
  if (n_nonzero_delta == 0) {
    return(list(
      stats = stats_base %>% dplyr::mutate(p_value = NA_real_, note = "skipped_all_zero_deltas"),
      wide = wide %>% dplyr::mutate(Method = as.character(Method), Cell_Type = as.character(Cell_Type), cluster = as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "all_zero_deltas", n_pairs, n_unique_delta),
      pair_audit = pair_audit
    ))
  }

  p <- tryCatch(
    stats::wilcox.test(
      wide$Surgery,
      wide$Biopsy,
      paired = TRUE,
      alternative = "two.sided",
      exact = FALSE,
      correct = TRUE
    )$p.value,
    error = function(e) NA_real_
  )

  list(
    stats = stats_base %>% dplyr::mutate(p_value = p, note = NA_character_),
    wide = wide %>% dplyr::mutate(Method = as.character(Method), Cell_Type = as.character(Cell_Type), cluster = as.character(cluster_label)),
    skip = NULL,
    pair_audit = pair_audit
  )
}

compute_paired_all <- function(immune_patient_df, out_prefix, write_outputs = FALSE) {
  # Safety step: paired tests are stratified by mapped CIM groups.
  immune_patient_df <- apply_cluster_mapping(immune_patient_df) %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  groups <- immune_patient_df %>%
    dplyr::group_by(Method, Cell_Type, cluster) %>%
    dplyr::group_split(.keep = TRUE)

  paired_results <- purrr::map(
    groups,
    ~ paired_one(
      .x,
      .x$Method[[1]],
      .x$Cell_Type[[1]],
      .x$cluster[[1]],
      expected_pairs_df = expected_paired_patients
    )
  )

  stats_tbl <- purrr::map(paired_results, "stats") %>% bind_rows()
  wide_tbl <- purrr::map(paired_results, "wide") %>% bind_rows()
  skip_tbl <- purrr::map(paired_results, "skip") %>% bind_rows()
  pair_audit_tbl <- purrr::map(paired_results, "pair_audit") %>% bind_rows()

  stats_tbl <- stats_tbl %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(Method, cluster, Cell_Type)

  if (nrow(skip_tbl) == 0) {
    skip_tbl <- tibble(analysis = character(), Method = character(), Cell_Type = character(), cluster = character(), reason = character(), n_pairs = integer(), n_unique_delta = integer())
  }

  pair_completeness_summary <- pair_audit_tbl %>%
    group_by(Method, Cell_Type, cluster) %>%
    summarise(
      n_candidates = n(),
      n_pairs_used = sum(included_pair),
      n_dropped = sum(!included_pair),
      n_missing_biopsy = sum(exclusion_reason == "missing_or_nonfinite_biopsy", na.rm = TRUE),
      n_missing_surgery = sum(exclusion_reason == "missing_or_nonfinite_surgery", na.rm = TRUE),
      n_missing_both = sum(exclusion_reason == "missing_or_nonfinite_both_timepoints", na.rm = TRUE),
      dropped_patient_ids = paste(Patient_ID[!included_pair], collapse = ";"),
      .groups = "drop"
    ) %>%
    arrange(Method, cluster, Cell_Type)

  if (isTRUE(write_outputs)) {
    write.csv(stats_tbl, file.path(DIRS$tables, paste0(out_prefix, "_paired_stats.csv")), row.names = FALSE)
    # Patient-level exclusions/invalid pairs are retained in CSVs, not on figures.
    write.csv(pair_audit_tbl, file.path(DIRS$tables, paste0(out_prefix, "_paired_pair_completeness_by_patient.csv")), row.names = FALSE)
    write.csv(pair_completeness_summary, file.path(DIRS$tables, paste0(out_prefix, "_paired_pair_completeness_summary.csv")), row.names = FALSE)
  }

  list(
    stats = stats_tbl,
    wide = wide_tbl,
    skipped = skip_tbl,
    pair_audit = pair_audit_tbl,
    pair_completeness = pair_completeness_summary
  )
}

paired_original <- compute_paired_all(immune_patient, "ORIGINAL", write_outputs = FALSE)
paired_aggregated <- compute_paired_all(immune_patient_aggregated, "AGGREGATED", write_outputs = TRUE)
```

# Plots

The plotting layer deliberately includes **all** returned cell types and score endpoints. In particular, score-like outputs such as `immune score`, `cytotoxicity score`, `ImmuneScore`, `StromalScore`, and `ESTIMATEScore` are not filtered out. Plot generation is audited so that any failed or skipped plot is written to a CSV instead of silently disappearing.

```{r plotting-functions}
# Unpaired figures are strictly patient-level. Upstream QC requires exactly one
# value per Patient_ID for each Method × Cell_Type × Timepoint combination; the
# plotting layer does not silently average or discard duplicates. A box may still
# collapse to a horizontal line when Q1 = median = Q3 (IQR = 0), which is a genuine
# feature of the data and is not artificially expanded.

# Choose which value to display on unpaired plots.
PLOT_P_COLUMN <- "p_value"
# PLOT_P_COLUMN <- "FDR"   # Use this instead for adjusted P values.

# ------------------------------------------------------------------------------
# Feature-specific y-axis limits
# ------------------------------------------------------------------------------
# Earlier versions used a method-wide lower bound. That is unnecessarily broad
# for methods that return heterogeneous outputs (cell scores + composite scores)
# and can visually compress individual endpoints. We now compute limits from the
# actual Method x Cell_Type data being plotted.
feature_ylims <- function(values, lower_zero_if_nonnegative = TRUE) {
  v <- as.numeric(values)
  v <- v[is.finite(v)]

  if (!length(v)) return(c(0, 0.01))

  vmin <- min(v)
  vmax <- max(v)
  span <- vmax - vmin

  if (!is.finite(span) || span <= 0) {
    span <- max(abs(vmax) * 0.20, 0.01)
  }

  lower <- if (lower_zero_if_nonnegative && vmin >= 0) 0 else vmin - 0.08 * span
  upper <- vmax + 0.18 * span

  if (!is.finite(lower) || !is.finite(upper) || lower >= upper) {
    lower <- vmin - 0.01
    upper <- vmax + 0.01
  }

  c(lower, upper)
}

compute_shared_ylims <- function(df, method_ymin = NULL) {
  feature_ylims(df$Value)
}

format_plot_p <- function(p) {
  if (length(p) == 0 || is.na(p) || !is.finite(p)) return("P = N/A")
  if (p < 0.0001) return("P < 0.0001")
  paste0("P = ", formatC(p, format = "f", digits = 4))
}

plot_unpaired_one <- function(
    df,
    shared_ylim,
    unpaired_stats,
    p_column = PLOT_P_COLUMN
) {
  # Plot exactly the patient-level observations used by the unpaired test.
  # Duplicates are not averaged here; they indicate an upstream integrity issue.
  df <- df %>%
    dplyr::filter(is.finite(Value), !is.na(cluster), !is.na(Patient_ID))

  dup <- df %>%
    dplyr::count(Patient_ID, name = "n_rows") %>%
    dplyr::filter(n_rows > 1)
  if (nrow(dup) > 0) {
    stop(
      "Duplicate patients reached unpaired plotting: ",
      paste(dup$Patient_ID, collapse = ", "),
      call. = FALSE
    )
  }

  if (dplyr::n_distinct(df$cluster) < 2) return(NULL)

  method_name <- as.character(unique(df$Method)[1])
  cell_type   <- as.character(unique(df$Cell_Type)[1])
  timepoint   <- as.character(unique(df$Timepoint)[1])

  df <- df %>%
    mutate(cluster = factor(as.character(cluster), levels = names(cluster_colors)))

  stat_row <- unpaired_stats %>%
    mutate(Timepoint = as.character(Timepoint)) %>%
    filter(
      Method == method_name,
      Cell_Type == cell_type,
      Timepoint == timepoint
    )

  if (nrow(stat_row) > 1) {
    stop("Multiple unpaired-stat rows matched one plot: ", method_name, " / ", cell_type, " / ", timepoint, call. = FALSE)
  }

  # The figure and the statistical table must use exactly the same patient n.
  if (nrow(stat_row) == 1) {
    for (cl in levels(df$cluster)) {
      n_col <- paste0(cl, "_n")
      if (n_col %in% names(stat_row)) {
        n_plot <- dplyr::n_distinct(df$Patient_ID[as.character(df$cluster) == cl])
        n_stat <- suppressWarnings(as.integer(stat_row[[n_col]][1]))
        if (!is.na(n_stat) && n_plot != n_stat) {
          stop(
            "Unpaired plot/stat n mismatch for ", method_name, " / ", cell_type,
            " / ", timepoint, " / ", cl, ": plot n=", n_plot,
            ", stats n=", n_stat, ".",
            call. = FALSE
          )
        }
      }
    }
  }

  if (!p_column %in% names(stat_row) || nrow(stat_row) == 0) {
    stop(
      "No unpaired statistic matched plot: ", method_name, " / ", cell_type,
      " / ", timepoint, ". Expected column: ", p_column, ".",
      call. = FALSE
    )
  } else {
    p_value <- suppressWarnings(as.numeric(stat_row[[p_column]][1]))
    if (!is.finite(p_value)) {
      stat_note <- if ("note" %in% names(stat_row)) as.character(stat_row$note[1]) else NA_character_
      stop(
        "Unpaired P value is NA/non-finite for ", method_name, " / ", cell_type,
        " / ", timepoint,
        if (!is.na(stat_note) && nzchar(stat_note)) paste0(" (", stat_note, ")") else "",
        ". Check AGGREGATED_unpaired_global.csv and the warnings above.",
        call. = FALSE
      )
    }
  }

  p_label <- format_plot_p(p_value)
  y_position <- shared_ylim[2] - 0.05 * diff(shared_ylim)

  ggplot(df, aes(x = cluster, y = Value, fill = cluster)) +
    geom_boxplot(
      outlier.shape = NA,
      width = PLOT_BOX_WIDTH,
      alpha = PLOT_BOX_ALPHA,
      linewidth = PLOT_BOX_LINEWIDTH,
      color = "gray20"
    ) +
    geom_jitter(
      width = PLOT_JITTER_WIDTH,
      alpha = PLOT_UNPAIRED_POINT_ALPHA,
      size = PLOT_UNPAIRED_POINT_SIZE,
      color = "gray20"
    ) +
    annotate(
      "text",
      x = 1.5,
      y = y_position,
      label = p_label,
      size = PLOT_P_ANNOTATION_SIZE,
      fontface = "bold"
    ) +
    scale_fill_manual(values = cluster_colors, name = "Cluster", drop = FALSE) +
    theme_pub() +
    coord_cartesian(ylim = shared_ylim, clip = "off") +
    labs(
      title = paste0(cell_type, " — ", timepoint),
      x = "",
      y = method_value_type(method_name)
    ) +
    theme(legend.position = "none")
}

save_unpaired_plots <- function(
    immune_patient_df,
    method_ymin,
    version_label,
    unpaired_stats,
    p_column = PLOT_P_COLUMN
) {
  immune_patient_df <- apply_cluster_mapping(immune_patient_df) %>%
    mutate(cluster = droplevels(cluster))

  out_dir <- DIRS$plots
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  combinations <- immune_patient_df %>%
    distinct(Method, Cell_Type, Timepoint) %>%
    arrange(Method, Cell_Type, Timepoint)

  audit_rows <- vector("list", nrow(combinations))

  for (i in seq_len(nrow(combinations))) {
    method_i <- combinations$Method[[i]]
    cell_type_i <- combinations$Cell_Type[[i]]
    timepoint_i <- combinations$Timepoint[[i]]

    # IMPORTANT: use uniquely named environment variables on the RHS.
    # Using `Method`, `Cell_Type`, or `Timepoint` as loop variables here can
    # collide with identically named data columns inside dplyr::filter(),
    # causing a condition such as `.data$Method == Method` to evaluate as
    # column == itself and unintentionally retain every method/cell type.
    plot_data <- immune_patient_df %>%
      filter(
        .data$Method == .env$method_i,
        .data$Cell_Type == .env$cell_type_i,
        .data$Timepoint == .env$timepoint_i,
        is.finite(.data$Value),
        !is.na(.data$cluster),
        !is.na(.data$Patient_ID)
      )

    # Hard QC: an unpaired method/cell-type/timepoint plot must contain at most
    # one observation per patient. Stop rather than silently drawing a malformed
    # figure if duplicate patient rows survive upstream processing.
    dup_qc <- plot_data %>%
      count(Patient_ID, name = "n_rows") %>%
      filter(n_rows > 1)
    if (nrow(dup_qc) > 0) {
      stop(
        "Duplicate patient observations remain in unpaired plot data for ",
        method_i, " / ", cell_type_i, " / ", as.character(timepoint_i),
        ". Patient IDs: ", paste(dup_qc$Patient_ID, collapse = ", "),
        call. = FALSE
      )
    }

    plot_subdir <- if (identical(version_label, "ORIGINAL")) "unpaired_original" else "unpaired"
    method_dir <- file.path(out_dir, method_dir_name(method_i), plot_subdir)
    dir.create(method_dir, recursive = TRUE, showWarnings = FALSE)
    file_png <- file.path(
      method_dir,
      paste0(safe_stem(timepoint_i, cell_type_i), ".png")
    )

    status <- "saved"
    error_message <- NA_character_

    if (nrow(plot_data) == 0) {
      status <- "skipped_no_data"
    } else if (n_distinct(plot_data$cluster) < 2) {
      status <- "skipped_lt2_groups"
    } else {
      tryCatch({
        shared_ylim <- compute_shared_ylims(plot_data, method_ymin)
        plot_object <- plot_unpaired_one(
          df = plot_data,
          shared_ylim = shared_ylim,
          unpaired_stats = unpaired_stats,
          p_column = p_column
        )

        if (is.null(plot_object)) {
          status <- "skipped_plot_null"
        } else {
          ggsave(
            file_png,
            plot_object,
            width = PLOT_WIDTH_UNPAIRED,
            height = PLOT_HEIGHT_UNPAIRED,
            dpi = PLOT_DPI
          )
        }
      }, error = function(e) {
        status <<- "failed"
        error_message <<- conditionMessage(e)
      })
    }

    audit_rows[[i]] <- tibble(
      version = version_label,
      plot_type = "unpaired",
      Method = method_i,
      Cell_Type = cell_type_i,
      Timepoint = as.character(timepoint_i),
      cluster = NA_character_,
      n_values = nrow(plot_data),
      n_patients = n_distinct(plot_data$Patient_ID),
      n_C1 = n_distinct(plot_data$Patient_ID[as.character(plot_data$cluster) == "C1"]),
      n_C2 = n_distinct(plot_data$Patient_ID[as.character(plot_data$cluster) == "C2"]),
      n_C3 = n_distinct(plot_data$Patient_ID[as.character(plot_data$cluster) == "C3"]),
      n_groups = n_distinct(plot_data$cluster),
      output_file = file_png,
      file_exists = file.exists(file_png),
      status = status,
      error_message = error_message
    )
  }

  audit <- bind_rows(audit_rows)
  audit
}

compute_shared_ylims_paired <- function(paired_wide_df, method, cell_type, method_ymin = NULL) {
  all_values <- paired_wide_df %>%
    filter(.data$Method == method, .data$Cell_Type == cell_type) %>%
    select(Biopsy, Surgery) %>%
    unlist(use.names = FALSE) %>%
    as.numeric()

  feature_ylims(all_values)
}

plot_paired_split_one <- function(
    paired_wide_df,
    paired_stats_df,
    method,
    cell_type,
    cluster_level,
    shared_ylim
) {
  d <- paired_wide_df %>%
    filter(
      .data$Method == method,
      .data$Cell_Type == cell_type,
      as.character(.data$cluster) == as.character(cluster_level)
    )

  if (nrow(d) < 3) return(NULL)

  stat_row <- paired_stats_df %>%
    filter(
      .data$Method == method,
      .data$Cell_Type == cell_type,
      as.character(.data$cluster) == as.character(cluster_level)
    )

  pval <- if (nrow(stat_row) > 0) suppressWarnings(as.numeric(stat_row$p_value[[1]])) else NA_real_
  p_label <- format_plot_p(pval)

  # Display only the actual number of complete patient pairs used for this
  # metric/cluster. Exclusion details are retained in the paired-completeness
  # CSV outputs rather than shown on the figure.
  n_pairs_used <- nrow(d)

  if (nrow(stat_row) > 0 && "n_pairs" %in% names(stat_row)) {
    n_pairs_stats <- suppressWarnings(as.integer(stat_row$n_pairs[[1]]))
    if (!is.na(n_pairs_stats) && n_pairs_stats != n_pairs_used) {
      stop(
        "Paired plot-count consistency check failed for ", method, " / ", cell_type,
        " / ", as.character(cluster_level), ": plot has ", n_pairs_used,
        " pairs but paired statistics report ", n_pairs_stats, ".",
        call. = FALSE
      )
    }
  }

  pair_subtitle <- paste0("n = ", n_pairs_used, " pairs")

  plot_df <- d %>%
    select(Patient_ID, Biopsy, Surgery) %>%
    pivot_longer(
      cols = c(Biopsy, Surgery),
      names_to = "Timepoint",
      values_to = "Value"
    ) %>%
    mutate(Timepoint = factor(Timepoint, levels = c("Biopsy", "Surgery")))

  y_position <- shared_ylim[2] - 0.05 * diff(shared_ylim)

  ggplot(plot_df, aes(x = Timepoint, y = Value, group = Patient_ID)) +
    geom_line(
      color = PLOT_PAIRED_LINE_COLOR,
      alpha = PLOT_PAIRED_LINE_ALPHA,
      linewidth = PLOT_PAIRED_LINE_WIDTH
    ) +
    geom_point(
      aes(color = Timepoint),
      size = PLOT_PAIRED_POINT_SIZE,
      alpha = PLOT_PAIRED_POINT_ALPHA
    ) +
    scale_color_manual(values = TIMEPOINT_COLORS, drop = FALSE) +
    theme_pub() +
    coord_cartesian(ylim = shared_ylim, clip = "off") +
    labs(
      title = paste0(cell_type, " — ", as.character(cluster_level)),
      subtitle = pair_subtitle,
      x = "",
      y = method_value_type(method)
    ) +
    annotate(
      "text",
      x = 1.5,
      y = y_position,
      label = p_label,
      size = PLOT_P_ANNOTATION_SIZE,
      fontface = "bold"
    ) +
    theme(legend.position = "none")
}

save_paired_split_plots <- function(paired_obj, method_ymin, version_label) {
  paired_obj$wide <- apply_cluster_mapping(paired_obj$wide) %>%
    mutate(cluster = droplevels(cluster))

  paired_obj$stats <- apply_cluster_mapping(paired_obj$stats) %>%
    mutate(cluster = droplevels(cluster))

  out_dir <- DIRS$plots
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  wide <- paired_obj$wide
  stats <- paired_obj$stats

  combos <- wide %>%
    distinct(Method, Cell_Type, cluster) %>%
    arrange(Method, Cell_Type, cluster)

  audit_rows <- vector("list", nrow(combos))

  for (i in seq_len(nrow(combos))) {
    method_i <- combos$Method[[i]]
    cell_type_i <- combos$Cell_Type[[i]]
    cluster_i <- as.character(combos$cluster[[i]])

    # Use explicit environment variables to prevent dplyr data-mask name
    # collisions (column == column), which would otherwise mix methods,
    # cell types, and CIM trajectories in a single paired figure.
    d <- wide %>%
      filter(
        .data$Method == .env$method_i,
        .data$Cell_Type == .env$cell_type_i,
        as.character(.data$cluster) == .env$cluster_i
      )

    plot_subdir <- if (identical(version_label, "ORIGINAL")) "paired_original" else "paired"
    method_dir <- file.path(out_dir, method_dir_name(method_i), plot_subdir)
    dir.create(method_dir, recursive = TRUE, showWarnings = FALSE)
    file_png <- file.path(
      method_dir,
      paste0(safe_stem(cluster_file_label(cluster_i), cell_type_i), ".png")
    )

    status <- "saved"
    error_message <- NA_character_

    if (nrow(d) < 3) {
      status <- "skipped_lt3_pairs"
    } else {
      tryCatch({
        shared_ylim <- compute_shared_ylims_paired(wide, method_i, cell_type_i, method_ymin)
        p <- plot_paired_split_one(
          paired_wide_df = wide,
          paired_stats_df = stats,
          method = method_i,
          cell_type = cell_type_i,
          cluster_level = cluster_i,
          shared_ylim = shared_ylim
        )

        if (is.null(p)) {
          status <- "skipped_plot_null"
        } else {
          ggsave(
            file_png,
            p,
            width = PLOT_WIDTH_PAIRED,
            height = PLOT_HEIGHT_PAIRED,
            dpi = PLOT_DPI
          )
        }
      }, error = function(e) {
        status <<- "failed"
        error_message <<- conditionMessage(e)
      })
    }

    audit_rows[[i]] <- tibble(
      version = version_label,
      plot_type = "paired",
      Method = method_i,
      Cell_Type = cell_type_i,
      Timepoint = NA_character_,
      cluster = cluster_i,
      n_values = 2L * nrow(d),
      n_pairs_used = nrow(d),
      n_groups = 1L,
      output_file = file_png,
      file_exists = file.exists(file_png),
      status = status,
      error_message = error_message
    )
  }

  audit <- bind_rows(audit_rows)
  audit
}
```

```{r run-plots}
if (params$make_plots) {
  # Optional original-method plots (off by default).
  if (isTRUE(MAKE_ORIGINAL_PLOTS)) {
    save_unpaired_plots(
      immune_patient,
      method_ymin_original,
      "ORIGINAL",
      unpaired_original$global
    )
    save_paired_split_plots(
      paired_original,
      method_ymin_original,
      "ORIGINAL"
    )
  }

  unpaired_plot_audit <- save_unpaired_plots(
    immune_patient_aggregated,
    method_ymin_aggregated,
    "AGGREGATED",
    unpaired_aggregated$global
  )

  paired_plot_audit <- save_paired_split_plots(
    paired_aggregated,
    method_ymin_aggregated,
    "AGGREGATED"
  )

  # Hard check for key cytotoxicity endpoints. If the data exist but the plot was
  # not saved, stop with a targeted message rather than silently finishing.
  cytotox_unpaired_expected <- immune_patient_aggregated %>%
    filter(Cell_Type == "cytotoxicity score") %>%
    distinct(Method, Cell_Type, Timepoint)

  cytotox_paired_expected <- paired_aggregated$wide %>%
    filter(Cell_Type == "cytotoxicity score") %>%
    distinct(Method, Cell_Type, cluster)

  cytotox_unpaired_failed <- unpaired_plot_audit %>%
    filter(Cell_Type == "cytotoxicity score", status != "saved")

  cytotox_paired_failed <- paired_plot_audit %>%
    filter(Cell_Type == "cytotoxicity score", status != "saved")

  if (nrow(cytotox_unpaired_expected) > 0 && nrow(cytotox_unpaired_failed) > 0) {
    warning(
      "One or more cytotoxicity-score unpaired plots were not saved.",
      call. = FALSE
    )
  }

  if (nrow(cytotox_paired_expected) > 0 && nrow(cytotox_paired_failed) > 0) {
    warning(
      "One or more cytotoxicity-score paired plots were not saved.",
      call. = FALSE
    )
  }
}
```

# Heatmaps

The cluster-level heatmap uses mean delta values, where delta is defined as:

```text
Delta = Surgery - Biopsy
```

Mean deltas are z-scored within each deconvolution method.

```{r heatmap-functions}
calculate_zscore_cluster_mean_deltas <- function(paired_wide_df) {
  paired_wide_df <- apply_cluster_mapping(paired_wide_df) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  required_cols <- c("Method", "Cell_Type", "cluster", "Biopsy", "Surgery", "Delta")
  missing_cols <- base::setdiff(required_cols, colnames(paired_wide_df))
  if (length(missing_cols) > 0) stop("Missing required columns: ", paste(missing_cols, collapse = ", "))

  paired_wide_df %>%
    filter(!is.na(cluster)) %>%
    group_by(Method, Cell_Type, cluster) %>%
    summarise(
      n_pairs = n(),
      mean_biopsy = mean(Biopsy, na.rm = TRUE),
      sd_biopsy = sd(Biopsy, na.rm = TRUE),
      mean_surgery = mean(Surgery, na.rm = TRUE),
      sd_surgery = sd(Surgery, na.rm = TRUE),
      mean_delta = mean(Delta, na.rm = TRUE),
      sd_delta = sd(Delta, na.rm = TRUE),
      p_value = if (n() >= 3) tryCatch(wilcox.test(Surgery, Biopsy, paired = TRUE, exact = FALSE)$p.value, error = function(e) NA_real_) else NA_real_,
      .groups = "drop"
    ) %>%
    filter(n_pairs >= 3) %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    group_by(Method) %>%
    mutate(zscore_delta = safe_zscore(mean_delta)) %>%
    ungroup()
}

prepare_method_cell_by_cluster_matrix <- function(zscore_deltas, fill_na = FALSE) {
  zscore_deltas <- zscore_deltas %>%
    mutate(CellType_Method = paste0(Method, ": ", Cell_Type), cluster = as.character(cluster)) %>%
    filter(!is.na(cluster))

  cluster_order <- unique(unname(CLUSTER_MAPPING))
  present_clusters <- unique(as.character(zscore_deltas$cluster))
  clusts_display <- cluster_order[cluster_order %in% present_clusters]
  if (length(clusts_display) == 0) clusts_display <- present_clusters

  mat <- zscore_deltas %>%
    select(CellType_Method, cluster, zscore_delta) %>%
    pivot_wider(names_from = cluster, values_from = zscore_delta) %>%
    column_to_rownames("CellType_Method") %>%
    as.matrix()

  sig_mat <- zscore_deltas %>%
    select(CellType_Method, cluster, FDR) %>%
    pivot_wider(names_from = cluster, values_from = FDR) %>%
    column_to_rownames("CellType_Method") %>%
    as.matrix()

  mat <- mat[, clusts_display, drop = FALSE]
  sig_mat <- sig_mat[, clusts_display, drop = FALSE]
  if (fill_na) mat[is.na(mat)] <- 0

  method_anno <- data.frame(
    CellType_Method = rownames(mat),
    Method = sub(":.*$", "", rownames(mat)),
    stringsAsFactors = FALSE
  )
  method_anno$Method <- factor(method_anno$Method, levels = METHOD_ORDER)
  ord <- order(method_anno$Method, method_anno$CellType_Method)

  list(mat = mat[ord, , drop = FALSE], sig = sig_mat[ord, , drop = FALSE], anno = method_anno[ord, , drop = FALSE])
}

plot_cluster_delta_heatmap <- function(zscore_deltas, output_file, show_significance = TRUE, cluster_rows = TRUE, fill_na = FALSE, na_col = "grey90") {
  prep <- prepare_method_cell_by_cluster_matrix(zscore_deltas, fill_na = fill_na)
  mat <- prep$mat
  sig <- prep$sig
  anno <- prep$anno

  if (nrow(mat) == 0 || ncol(mat) == 0 || all(is.na(mat))) stop("Heatmap matrix is empty or all NA.")

  max_abs <- max(abs(mat), na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs == 0) max_abs <- 2
  max_abs <- max(max_abs, 2)

  col_fun <- circlize::colorRamp2(c(-max_abs, -1, 0, 1, max_abs), c("#0571B0", "#92C5DE", "white", "#F4A582", "#CA0020"))

  methods_present <- unique(as.character(anno$Method))
  method_colors_subset <- METHOD_COLORS[methods_present]
  method_colors_subset <- method_colors_subset[!is.na(method_colors_subset)]

  row_ha <- ComplexHeatmap::rowAnnotation(
    Method = anno$Method,
    col = list(Method = method_colors_subset),
    annotation_name_gp = grid::gpar(fontsize = 11, fontface = "bold")
  )

  cell_fun <- NULL
  if (show_significance) {
    cell_fun <- function(j, i, x, y, width, height, fill) {
      fdr <- sig[i, j]
      if (is.na(fdr)) return(NULL)
      label <- if (fdr < 0.001) "***" else if (fdr < 0.01) "**" else if (fdr < 0.05) "*" else ""
      if (label != "") grid::grid.text(label, x, y, gp = grid::gpar(fontsize = 8, fontface = "bold"))
    }
  }

  display_names <- stringr::str_remove(rownames(mat), "^[^:]+: ")

  ht <- ComplexHeatmap::Heatmap(
    mat,
    name = "Z-score\n(mean Δ)",
    col = col_fun,
    na_col = na_col,
    cell_fun = cell_fun,
    row_names_side = "left",
    row_labels = display_names,
    row_names_gp = grid::gpar(fontsize = 9),
    cluster_rows = cluster_rows,
    show_row_dend = cluster_rows,
    cluster_columns = FALSE,
    column_names_gp = grid::gpar(fontsize = 14, fontface = "bold"),
    column_names_rot = 0,
    column_title = "Cluster-mean zΔ (Surgery - Biopsy)",
    column_title_gp = grid::gpar(fontsize = 16, fontface = "bold"),
    left_annotation = row_ha,
    border = TRUE
  )

  plot_height <- max(10, nrow(mat) * 0.15 + 3)
  png(output_file, width = 12, height = plot_height, units = "in", res = 300)
  ComplexHeatmap::draw(ht, heatmap_legend_side = "right", annotation_legend_side = "bottom")
  dev.off()

  output_file
}
```

```{r run-heatmaps}
if (params$make_heatmaps) {
  zscore_deltas_agg <- calculate_zscore_cluster_mean_deltas(paired_aggregated$wide)

  write.csv(zscore_deltas_agg, file.path(DIRS$tables, "AGGREGATED_zscore_cluster_mean_deltas.csv"), row.names = FALSE)

  prep_tmp <- prepare_method_cell_by_cluster_matrix(zscore_deltas_agg, fill_na = FALSE)
  write.csv(prep_tmp$mat, file.path(DIRS$tables, "AGGREGATED_heatmap_matrix_zscore_delta.csv"))

  heat_file <- file.path(DIRS$plots, "heatmap_zDelta.png")
  plot_cluster_delta_heatmap(zscore_deltas_agg, heat_file, show_significance = TRUE, cluster_rows = TRUE, fill_na = FALSE)
}
```

# Output summary

```{r output-summary}
cat("Analysis complete.\n")
cat("Results folder:", DIRS$results, "\n")
cat("Plots folder:", DIRS$plots, "\n")
```

# Compact output summary

```{r compact-output-summary}
cat("Compact analysis complete.\n")
cat("Tables: ", DIRS$tables, "\n", sep = "")
cat("Figures: ", DIRS$plots, "\n", sep = "")
cat("Original plots enabled: ", MAKE_ORIGINAL_PLOTS, "\n", sep = "")
```
