---
title: "Immune Infiltration Analysis Pipeline"
author: "Gbadamosi Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
params:
  run_immunedeconv: true
  run_cibersortx_import: true
  make_plots: true
  make_heatmaps: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  error = FALSE,
  fig.align = "center"
)
```

# Overview

This R Markdown document runs the combined-cohort immune infiltration analysis for paired neoadjuvant samples. The workflow:

1. Loads TPM expression matrices for each cohort separately.
2. Loads combined CIMIC cluster assignments.
3. Runs `immunedeconv` methods separately by cohort, then merges estimates at the result level.
4. Optionally imports CIBERSORTx relative-mode outputs.
5. Aggregates technical replicates to the patient level.
6. Optionally aggregates selected T-cell subtypes for xCell and CIBERSORTx.
7. Performs unpaired cluster comparisons at biopsy and surgery.
8. Performs paired biopsy-to-surgery tests within each cluster.

The primary unit for downstream statistical analysis is the patient, not the technical replicate.

# Required input files

The pipeline expects a project folder with the following structure. File names can be changed in the configuration section below.

```text
project/
├── data/
│   ├── RNAseq_TPM_avg_NKI.csv
│   ├── RNAseq_TPM_SMC.csv
│   ├── clustered_samples_app_three.csv
│   ├── CIBERSORTx_NKI_rel1.txt        # optional
│   └── CIBERSORTx_SMC_rel1.txt        # optional
├── results/                           # created automatically
├── plots/                             # created automatically
├── logs/                              # created automatically
├── tmp/                               # created automatically
└── bioc_cache/                        # created automatically
```

Input requirements:

- Expression matrices must be genes x samples.
- The first column must contain gene symbols or gene IDs.
- Expression values must be TPM and non-log transformed.
- Cluster file must contain `sample_id` and `cluster_assignments`.
- CIBERSORTx files must contain `Mixture` plus immune-cell columns only. Remove QC columns before running this workflow.
- Sample IDs must contain biopsy/surgery labels using `_B_Rep#` and `_S_Rep#`, for example `NKI_2046_B_Rep1` and `NKI_2046_S_Rep1`.

# Reproducibility notes

This workflow uses a fixed seed for reproducible steps within R. However, package versions can affect deconvolution outputs. For GitHub release, use `renv` to create a lockfile:

```r
install.packages("renv")
renv::init()
renv::snapshot()
```

To restore the same package versions later:

```r
renv::restore()
```

# Package setup

This document does not automatically install packages during analysis. Install packages once before running the workflow.

```{r packages}
required_packages <- c(
  "data.table", "dplyr", "tidyr", "tibble", "stringr", "purrr",
  "ggplot2", "readr", "ComplexHeatmap", "circlize", "withr",
  "FSA", "grid", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Missing required packages: ",
    paste(missing_packages, collapse = ", "),
    "\nInstall them before running this R Markdown. ",
    "For Bioconductor packages, use BiocManager."
  )
}

if (!requireNamespace("immunedeconv", quietly = TRUE)) {
  stop(
    "Package 'immunedeconv' is not installed.\n",
    "Install it with:\n",
    "remotes::install_github('omnideconv/immunedeconv')"
  )
}

suppressPackageStartupMessages({
  library(data.table)
  library(immunedeconv)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(readr)
  library(ComplexHeatmap)
  library(circlize)
  library(withr)
  library(FSA)
  library(grid)
})

cat("All required packages loaded successfully.\n")
```

# Configuration

Edit only this section for a new project or collaborator machine.

```{r config}
set.seed(2025)
options(stringsAsFactors = FALSE, scipen = 999)

# Add here the base directory for the project.
# This folder should contain the data/, results/, plots/, logs/, tmp/, and bioc_cache/ folders.
project_dir <- "C:/Users/lopesdelima.i/OneDrive - University of Florida/Gbadamosi Lab/Iasmim Lopes/01.CIMIC/Immunedeconv2"

PROJECT_DIR <- normalizePath(project_dir, mustWork = TRUE)
DATA_DIR <- file.path(PROJECT_DIR, "data")

FILES <- list(
  nki_expr = file.path(DATA_DIR, "RNAseq_TPM_avg_NKI.csv"),
  smc_expr = file.path(DATA_DIR, "RNAseq_TPM_SMC.csv"),
  clusters = file.path(DATA_DIR, "clustered_samples_app_three.csv"),
  cibersortx_nki = file.path(DATA_DIR, "CIBERSORTx_NKI_rel1.txt"),
  cibersortx_smc = file.path(DATA_DIR, "CIBERSORTx_SMC_rel1.txt")
)

DIRS <- list(
  project = PROJECT_DIR,
  data = DATA_DIR,
  results = file.path(PROJECT_DIR, "results"),
  plots = file.path(PROJECT_DIR, "plots"),
  logs = file.path(PROJECT_DIR, "logs"),
  tmp = file.path(PROJECT_DIR, "tmp"),
  cache = file.path(PROJECT_DIR, "bioc_cache")
)

invisible(lapply(DIRS, dir.create, recursive = TRUE, showWarnings = FALSE))

Sys.setenv(TMPDIR = DIRS$tmp)
Sys.setenv(TEMP = DIRS$tmp)
Sys.setenv(TMP = DIRS$tmp)
options(ExperimentHub.cache = DIRS$cache)

LOG_FILE <- file.path(
  DIRS$logs,
  paste0("run_log_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".txt")
)

METHODS_TO_RUN <- c(
  "quantiseq", "epic", "xcell", "mcp_counter",
  "estimate", "timer", "consensus_tme"
)

TUMOR_INDICATION <- "BRCA"
GENE_COL_CANDIDATES <- c("gene", "gene_id", "hgnc_symbol", "symbol", "Gene", "GeneSymbol")

METHOD_ORDER <- c(
  "CIBERSORTx", "quantiseq", "epic", "mcp_counter", "estimate",
  "xcell", "timer", "consensus_tme"
)

AGGREGATE_XCELL_TCELLS <- TRUE
AGGREGATE_CIBERSORT_TCELLS <- TRUE
CIBERSORT_AGGREGATION_STRATEGY <- "refined" # options: "refined" or "simple"

# ------------------------------------------------------------------------------
# CIMIC cluster mapping
# ------------------------------------------------------------------------------
# Use this to map numeric cluster IDs from the cluster file into biological groups.
# This also supports collapsing multiple clusters into one biological group.
#
# Example for 2 input clusters:
# CLUSTER_MAPPING <- c(
#   "1" = "Fun-CIM",
#   "2" = "Dys-CIM"
# )
#
# Example for 3 input clusters, where clusters 2 and 3 are both Dys-CIM:
# CLUSTER_MAPPING <- c(
#   "1" = "Fun-CIM",
#   "2" = "Dys-CIM",
#   "3" = "Dys-CIM"
# )

USE_CLUSTER_MAPPING <- TRUE

 CLUSTER_MAPPING <- c(
   "2" = "Fun-CIM",
   "1" = "Dys-CIM"
 )

cluster_colors <- c(
  "Fun-CIM" = "#2A9D8F",
  "Dys-CIM" = "#E76F51"#E76F51
)

# If additional biological groups are added to CLUSTER_MAPPING, assign colors here.
missing_cluster_colors <- base::setdiff(unique(unname(CLUSTER_MAPPING)), names(cluster_colors))
if (length(missing_cluster_colors) > 0) {
  stop(
    "Missing cluster colors for: ",
    paste(missing_cluster_colors, collapse = ", ")
  )
}

METHOD_COLORS <- c(
  "quantiseq" = "#1F77B4",
  "epic" = "#FF7F0E",
  "xcell" = "#2CA02C",
  "mcp_counter" = "#D62728",
  "estimate" = "#9467BD",
  "timer" = "#8C564B",
  "consensus_tme" = "#E377C2",
  "CIBERSORTx" = "#FFD700"
)

condition_colors <- c("Biopsy" = "#2166AC", "Surgery" = "#762A83")

PLOT_WIDTH_UNPAIRED <- 5
PLOT_HEIGHT_UNPAIRED <- 6
PLOT_WIDTH_PAIRED <- 6
PLOT_HEIGHT_PAIRED <- 5
PLOT_DPI <- 300
```

# Helper functions

```{r helpers}
validate_file <- function(path, required = TRUE) {
  if (!file.exists(path)) {
    msg <- paste0("File not found: ", path)
    if (required) stop(msg, call. = FALSE) else warning(msg, call. = FALSE)
    return(FALSE)
  }
  TRUE
}

sanitize_filename <- function(x) gsub("[^A-Za-z0-9]+", "_", x)
safe_stem <- function(...) sanitize_filename(paste(..., sep = "_"))

method_value_type <- function(m) {
  switch(m,
    "quantiseq" = "Cell Fraction",
    "epic" = "Cell Fraction",
    "mcp_counter" = "Abundance Score",
    "xcell" = "Enrichment Score",
    "timer" = "Infiltration Score",
    "consensus_tme" = "ssGSEA Score",
    "estimate" = "ESTIMATE Score",
    "CIBERSORTx" = "Cell Fraction",
    "Score"
  )
}

auto_ylims <- function(values, ymin_floor = 0) {
  v <- values[is.finite(values)]
  if (!length(v)) return(c(ymin_floor, ymin_floor + 0.01))
  ymin <- ymin_floor
  ymax <- max(v, na.rm = TRUE)
  yr <- ymax - ymin
  if (yr <= 0) yr <- max(abs(ymax) * 0.15, 0.01)
  c(ymin, ymax + 0.18 * yr)
}

safe_zscore <- function(x) {
  x <- as.numeric(x)
  s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(0, length(x)))
  as.numeric((x - mean(x, na.rm = TRUE)) / s)
}

apply_cluster_mapping <- function(df, cluster_col = "cluster") {
  # Apply the user-defined cluster mapping to the analysis cluster column.
  # This is the central rule used throughout the workflow.
  # Example: if 1 = Fun-CIM and 2/3 = Dys-CIM, all downstream statistics,
  # plots, paired analyses, and heatmaps use two groups: Fun-CIM and Dys-CIM.
  if (!USE_CLUSTER_MAPPING) return(df)
  if (!cluster_col %in% names(df)) return(df)

  original_cluster <- as.character(df[[cluster_col]])
  mapped_cluster <- dplyr::recode(
    original_cluster,
    !!!CLUSTER_MAPPING,
    .default = original_cluster
  )

  mapped_levels <- unique(unname(CLUSTER_MAPPING))
  extra_levels <- unique(mapped_cluster[!mapped_cluster %in% mapped_levels])
  final_levels <- unique(c(mapped_levels, extra_levels))

  if (!"cluster_original" %in% names(df)) {
    df$cluster_original <- original_cluster
  }

  df[[cluster_col]] <- factor(mapped_cluster, levels = final_levels)
  df
}

# Backward-compatible alias used by earlier versions of the workflow.
apply_cluster_names <- apply_cluster_mapping

theme_pub <- function() {
  theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.text = element_text(size = 16, face = "bold"),
      axis.title = element_text(size = 20, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 15)),
      axis.ticks = element_line(linewidth = 1.1),
      axis.line = element_line(linewidth = 1),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      legend.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 16),
      strip.text = element_text(size = 18, face = "bold"),
      strip.background = element_rect(fill = "white", color = "black", linewidth = 1),
      plot.margin = margin(t = 10, r = 10, b = 50, l = 10)
    )
}
```

# Sample ID parsing

The functions below extract timepoint and patient identifiers from sample names. They are intentionally strict to avoid silent metadata errors.

```{r id-parsing}
get_timepoint <- function(sample_id) {
  case_when(
    str_detect(sample_id, "_B_Rep\\d+$") ~ "Biopsy",
    str_detect(sample_id, "_S_Rep\\d+$") ~ "Surgery",
    TRUE ~ NA_character_
  )
}

get_patient_id_with_cohort <- function(sample_id) {
  str_replace(sample_id, "_(B|S)_Rep\\d+$", "")
}

get_patient_key_from_sample <- function(sample_id) {
  sample_id %>%
    str_replace("^(NKI|SMC)_", "") %>%
    str_replace("_(B|S)_Rep\\d+$", "")
}

get_patient_key_from_delta <- function(delta_id) {
  delta_id %>%
    str_replace("^delta_", "") %>%
    str_replace("_(B|S)_Rep\\d+$", "")
}
```

# Validate inputs

```{r validate-inputs}
validate_file(FILES$nki_expr)
validate_file(FILES$smc_expr)
validate_file(FILES$clusters)

if (params$run_cibersortx_import) {
  validate_file(FILES$cibersortx_nki, required = FALSE)
  validate_file(FILES$cibersortx_smc, required = FALSE)
}
```

# Load combined CIMIC cluster table

In some cases, the cluster table is expected to use `delta_*` sample IDs. The pipeline converts these IDs into patient keys and joins the cluster assignment to each biopsy/surgery sample.

```{r load-clusters}
clusters_raw <- readr::read_csv(FILES$clusters, show_col_types = FALSE)

required_cluster_cols <- c("sample_id", "cluster_assignments")
missing_cluster_cols <- base::setdiff(required_cluster_cols, colnames(clusters_raw))
if (length(missing_cluster_cols) > 0) {
  stop("Cluster file is missing columns: ", paste(missing_cluster_cols, collapse = ", "))
}

clusters <- clusters_raw %>%
  dplyr::transmute(
    delta_sample_id = as.character(sample_id),
    patient_key = get_patient_key_from_delta(delta_sample_id),
    cluster = as.character(cluster_assignments)
  ) %>%
  dplyr::mutate(cluster = factor(cluster))

# Apply the biological cluster mapping immediately after loading the cluster file.
# Downstream analyses use this mapped cluster column, not the original numeric K.
clusters <- apply_cluster_mapping(clusters)

dup_pat <- clusters %>%
  dplyr::count(patient_key) %>%
  dplyr::filter(n > 1)

if (nrow(dup_pat) > 0) {
  stop("Cluster file has duplicate patient_key rows. Deduplicate before running.")
}

k <- dplyr::n_distinct(stats::na.omit(clusters$cluster))
message("Detected ", k, " mapped CIM group(s) after applying CLUSTER_MAPPING.")
knitr::kable(head(clusters), caption = "Cluster table preview")
```

# Load expression matrices

```{r load-expression}
load_expr_matrix <- function(path, dataset_label) {
  validate_file(path)
  message("Loading expression data for ", dataset_label, ": ", path)

  df <- data.table::fread(path, data.table = FALSE, check.names = FALSE)
  gene_col <- base::intersect(GENE_COL_CANDIDATES, colnames(df))

  if (length(gene_col) > 0) {
    genes <- make.unique(as.character(df[[gene_col[1]]]))
    df[[gene_col[1]]] <- NULL
    mat <- as.matrix(df)
    rownames(mat) <- genes
  } else {
    message("No named gene column found. Treating first column as row names.")
    df <- read.csv(path, row.names = 1, check.names = FALSE)
    mat <- as.matrix(df)
  }

  storage.mode(mat) <- "numeric"

  if (anyDuplicated(colnames(mat))) {
    stop(dataset_label, ": duplicate sample columns detected.")
  }

  message("Loaded ", dataset_label, ": ", nrow(mat), " genes x ", ncol(mat), " samples.")
  mat
}

check_nonlog_tpm <- function(mat, label) {
  mx <- suppressWarnings(max(mat, na.rm = TRUE))
  mn <- suppressWarnings(min(mat, na.rm = TRUE))
  if (mn < 0) warning(label, ": negative values present; unexpected for TPM.", call. = FALSE)
  if (mx < 30) warning(label, ": maximum value < 30; data may be log-scaled. immunedeconv expects non-log TPM.", call. = FALSE)
  invisible(TRUE)
}

expr_nki <- load_expr_matrix(FILES$nki_expr, "NKI")
expr_smc <- load_expr_matrix(FILES$smc_expr, "SMC")
check_nonlog_tpm(expr_nki, "NKI TPM")
check_nonlog_tpm(expr_smc, "SMC TPM")
```

# Build sample metadata

```{r metadata}
build_metadata_from_samples <- function(sample_ids, dataset_label) {
  tibble(sample_id = sample_ids) %>%
    mutate(
      dataset = dataset_label,
      Timepoint = factor(get_timepoint(sample_id), levels = c("Biopsy", "Surgery")),
      Patient_ID = get_patient_id_with_cohort(sample_id),
      patient_key = get_patient_key_from_sample(sample_id)
    ) %>%
    left_join(clusters, by = "patient_key") %>%
    mutate(cluster = factor(cluster, levels = levels(clusters$cluster))) %>%
    select(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster)
}

meta_nki <- build_metadata_from_samples(colnames(expr_nki), "NKI")
meta_smc <- build_metadata_from_samples(colnames(expr_smc), "SMC")

metadata_summary <- bind_rows(meta_nki, meta_smc) %>%
  mutate(has_timepoint = !is.na(Timepoint), has_cluster = !is.na(cluster)) %>%
  group_by(dataset) %>%
  summarise(
    n_samples = n(),
    n_with_timepoint = sum(has_timepoint),
    n_with_cluster = sum(has_cluster),
    n_retained = sum(has_timepoint & has_cluster),
    .groups = "drop"
  )

knitr::kable(metadata_summary, caption = "Metadata matching summary")

keep_nki <- meta_nki %>% filter(!is.na(cluster), !is.na(Timepoint)) %>% pull(sample_id)
keep_smc <- meta_smc %>% filter(!is.na(cluster), !is.na(Timepoint)) %>% pull(sample_id)

if (length(keep_nki) == 0) stop("No NKI samples matched cluster + timepoint metadata.")
if (length(keep_smc) == 0) stop("No SMC samples matched cluster + timepoint metadata.")

expr_nki <- expr_nki[, keep_nki, drop = FALSE]
expr_smc <- expr_smc[, keep_smc, drop = FALSE]

meta_nki <- meta_nki %>% filter(sample_id %in% colnames(expr_nki)) %>% arrange(match(sample_id, colnames(expr_nki)))
meta_smc <- meta_smc %>% filter(sample_id %in% colnames(expr_smc)) %>% arrange(match(sample_id, colnames(expr_smc)))

stopifnot(identical(meta_nki$sample_id, colnames(expr_nki)))
stopifnot(identical(meta_smc$sample_id, colnames(expr_smc)))
```

# Run immunedeconv

Deconvolution is performed separately for each cohort because preprocessing can differ between cohorts. Results are merged only after method outputs are converted to long format.

```{r immunedeconv-functions}
make_indications <- function(mat, method) {
  if (method %in% c("timer", "consensus_tme")) {
    setNames(rep(tolower(TUMOR_INDICATION), ncol(mat)), colnames(mat))
  } else {
    NULL
  }
}

deconv_to_long <- function(res, method, dataset) {
  if (is.null(res)) return(NULL)

  if (method == "estimate") {
    long <- res %>%
      as.data.frame() %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id")

    long %>%
      pivot_longer(-sample_id, names_to = "Cell_Type", values_to = "Value") %>%
      mutate(Method = method, dataset = dataset, Value = as.numeric(Value))
  } else {
    if (!"cell_type" %in% colnames(res)) return(NULL)

    res %>%
      column_to_rownames("cell_type") %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id") %>%
      pivot_longer(-sample_id, names_to = "Cell_Type", values_to = "Value") %>%
      mutate(Method = method, dataset = dataset, Value = as.numeric(Value))
  }
}

run_methods <- function(expr, meta, dataset_label) {
  out <- list()
  log <- list()

  for (m in METHODS_TO_RUN) {
    message("Running ", dataset_label, " / ", m)
    ind <- make_indications(expr, m)
    convergence_warning <- FALSE

    res <- withr::with_dir(DIRS$tmp, {
      if (m == "estimate") {
        tryCatch(immunedeconv::deconvolute_estimate(expr), error = function(e) { message("ERROR: ", e$message); NULL })
      } else if (m == "epic") {
        tryCatch(
          withCallingHandlers(
            immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE),
            warning = function(w) {
              if (grepl("didn't fully converge", w$message, ignore.case = TRUE)) {
                convergence_warning <<- TRUE
                message("EPIC partial convergence warning detected.")
              }
              invokeRestart("muffleWarning")
            }
          ),
          error = function(e) { message("ERROR: ", e$message); NULL }
        )
      } else if (m == "quantiseq") {
        tryCatch(immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      } else if (!is.null(ind)) {
        tryCatch(immunedeconv::deconvolute(expr, m, indications = ind, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      } else {
        tryCatch(immunedeconv::deconvolute(expr, m, arrays = FALSE), error = function(e) { message("ERROR: ", e$message); NULL })
      }
    })

    if (m == "epic" && !is.null(res) && "cell_type" %in% colnames(res)) {
      n_samples_returned <- ncol(res) - 1
      rate <- n_samples_returned / ncol(expr)
      if (rate < 0.3) {
        message("Discarding EPIC output because <30% of samples converged.")
        res <- NULL
      }
    }

    long <- deconv_to_long(res, m, dataset_label)

    if (!is.null(long)) {
      long <- long %>% left_join(meta %>% select(sample_id, Patient_ID, Timepoint, cluster), by = "sample_id")
    }

    out[[m]] <- long
    log[[m]] <- tibble(
      dataset = dataset_label,
      Method = m,
      status = ifelse(is.null(long), "FAILED", "OK"),
      n_samples = ifelse(is.null(long), 0, n_distinct(long$sample_id)),
      n_celltypes = ifelse(is.null(long), 0, n_distinct(long$Cell_Type)),
      notes = case_when(
        is.null(long) ~ "Method failed or returned no usable result",
        m == "epic" & convergence_warning ~ "Partial convergence",
        TRUE ~ "Success"
      )
    )
  }

  list(immune_long = bind_rows(out), run_log = bind_rows(log))
}
```

```{r run-immunedeconv}
if (params$run_immunedeconv) {
  res_nki <- run_methods(expr_nki, meta_nki, "NKI")
  res_smc <- run_methods(expr_smc, meta_smc, "SMC")

  immune_long_raw <- bind_rows(res_nki$immune_long, res_smc$immune_long)
  run_summary <- bind_rows(res_nki$run_log, res_smc$run_log)
} else {
  immune_long_raw <- tibble()
  run_summary <- tibble()
}
```

# Import CIBERSORTx results

CIBERSORTx is imported from web-output files and merged with the `immunedeconv` results. These files are expected to be relative-mode outputs with QC columns already removed.

```{r cibersortx-import}
read_cibersort_noqc_long <- function(path, dataset_label) {
  if (!file.exists(path)) {
    warning("CIBERSORTx file not found: ", path, call. = FALSE)
    return(NULL)
  }

  cs <- data.table::fread(path, data.table = FALSE, check.names = FALSE)
  if (!"Mixture" %in% colnames(cs)) stop("CIBERSORTx file must contain a Mixture column: ", path)

  cs_long <- cs %>%
    rename(sample_id = Mixture) %>%
    pivot_longer(cols = -sample_id, names_to = "Cell_Type", values_to = "Value") %>%
    mutate(Method = "CIBERSORTx", dataset = dataset_label, Value = as.numeric(Value))

  meta <- build_metadata_from_samples(unique(cs_long$sample_id), dataset_label)

  cs_long %>%
    left_join(meta %>% select(sample_id, Patient_ID, Timepoint, cluster), by = "sample_id")
}

if (params$run_cibersortx_import) {
  ciber_nki_long <- read_cibersort_noqc_long(FILES$cibersortx_nki, "NKI")
  ciber_smc_long <- read_cibersort_noqc_long(FILES$cibersortx_smc, "SMC")

  if (!is.null(ciber_nki_long)) {
    immune_long_raw <- bind_rows(immune_long_raw, ciber_nki_long)
    run_summary <- bind_rows(run_summary, tibble(
      dataset = "NKI", Method = "CIBERSORTx", status = "OK",
      n_samples = n_distinct(ciber_nki_long$sample_id),
      n_celltypes = n_distinct(ciber_nki_long$Cell_Type),
      notes = "Imported from CIBERSORTx relative output"
    ))
  }

  if (!is.null(ciber_smc_long)) {
    immune_long_raw <- bind_rows(immune_long_raw, ciber_smc_long)
    run_summary <- bind_rows(run_summary, tibble(
      dataset = "SMC", Method = "CIBERSORTx", status = "OK",
      n_samples = n_distinct(ciber_smc_long$sample_id),
      n_celltypes = n_distinct(ciber_smc_long$Cell_Type),
      notes = "Imported from CIBERSORTx relative output"
    ))
  }
}

saveRDS(immune_long_raw, file.path(DIRS$results, "immune_long_raw_all_methods.rds"))
write.csv(run_summary, file.path(DIRS$results, "run_summary_methods.csv"), row.names = FALSE)
knitr::kable(run_summary, caption = "Method run summary")
```

# Patient-level aggregation

Technical replicates are averaged to create one value per patient, method, cell type, timepoint, and cluster.

```{r patient-aggregation}
# Clean long-format immune deconvolution results
immune_long_clean <- immune_long_raw %>%
  dplyr::filter(
    !is.na(Patient_ID),
    !is.na(Timepoint),
    !is.na(cluster),
    is.finite(Value)
  )

# Apply biological CIM-state labels before patient-level aggregation.
# Example mapping used in this workflow:
#   1 -> Fun-CIM
#   2 -> Dys-CIM
#   3 -> Dys-CIM
immune_long_clean <- apply_cluster_mapping(immune_long_clean)

# Average technical replicates at the patient level
immune_patient <- immune_long_clean %>%
  dplyr::group_by(Method, Cell_Type, Patient_ID, Timepoint, cluster) %>%
  dplyr::summarise(
    Value = mean(Value, na.rm = TRUE),
    n_replicates = dplyr::n(),
    .groups = "drop"
  )

# Save original patient-level results
saveRDS(
  immune_patient,
  file.path(DIRS$results, "immune_patient_level_ORIGINAL.rds")
)

utils::write.csv(
  immune_patient,
  file.path(DIRS$results, "immune_patient_level_ORIGINAL.csv"),
  row.names = FALSE
)

# Store method-specific lower y-axis limits for plotting
method_ymin_original <- immune_patient %>%
  dplyr::group_by(Method) %>%
  dplyr::summarise(
    ymin = ifelse(
      min(Value, na.rm = TRUE) < 0,
      min(Value, na.rm = TRUE),
      0
    ),
    .groups = "drop"
  )

# Quick summary
immune_patient_summary <- immune_patient %>%
  dplyr::count(Method, cluster, Timepoint, name = "n_patient_level_values")

knitr::kable(
  immune_patient_summary,
  caption = "Patient-level immune deconvolution values after cluster mapping"
)
```

# T-cell aggregation

Different deconvolution algorithms report T-cell populations at different levels of granularity. This section optionally creates broader CD4+ and CD8+ categories to simplify downstream interpretation.

Aggregation is performed after patient-level averaging and before statistical analysis.

## xCell aggregation

When `AGGREGATE_XCELL_TCELLS = TRUE`, all xCell populations matching `T cell CD4+` are summed into `T cell CD4+ (aggregated)`, and all xCell populations matching `T cell CD8+` are summed into `T cell CD8+ (aggregated)`.

## CIBERSORTx aggregation

CIBERSORTx supports two strategies:

- `refined`: creates `T cells CD4+ (general)` from naive, memory resting, and memory activated CD4 T cells. Follicular helper T cells are kept separate. This is recommended when you want to preserve a distinct Tfh population.
- `simple`: creates `T cells CD4+ (total)` from naive, memory resting, memory activated, and follicular helper T cells. This is useful for broad CD4 versus CD8 summaries.

```{r tcell-aggregation}
aggregate_xcell_tcells <- function(immune_patient_df) {
  if (!AGGREGATE_XCELL_TCELLS) return(list(immune_patient_aggregated = immune_patient_df, aggregation_details = NULL))

  xcell_data <- immune_patient_df %>% filter(Method == "xcell")
  non_xcell <- immune_patient_df %>% filter(Method != "xcell")
  if (nrow(xcell_data) == 0) return(list(immune_patient_aggregated = immune_patient_df, aggregation_details = NULL))

  cd4_aggregated <- xcell_data %>%
    filter(grepl("T cell CD4\\+", Cell_Type)) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD4+ (aggregated)")

  cd8_aggregated <- xcell_data %>%
    filter(grepl("T cell CD8\\+", Cell_Type)) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD8+ (aggregated)")

  other_xcell <- xcell_data %>% filter(!grepl("T cell CD4\\+", Cell_Type), !grepl("T cell CD8\\+", Cell_Type))

  immune_patient_aggregated <- bind_rows(
    non_xcell,
    other_xcell,
    cd4_aggregated %>% select(Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_aggregated %>% select(Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )

  aggregation_details <- bind_rows(
    cd4_aggregated %>% select(Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_aggregated %>% select(Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )

  list(immune_patient_aggregated = immune_patient_aggregated, aggregation_details = aggregation_details)
}

aggregate_cibersort_tcells <- function(df, strategy = "refined") {
  if (!AGGREGATE_CIBERSORT_TCELLS) return(list(aggregated = df, details = NULL))
  if (nrow(df) == 0) return(list(aggregated = df, details = NULL))

  if (strategy == "refined") {
    cd4_subtypes <- c("T cells CD4 naive", "T cells CD4 memory resting", "T cells CD4 memory activated")
    cd4_name <- "T cells CD4+ (general)"
    cd8_subtypes <- c("T cells CD8")
    cd8_name <- "T cells CD8+ (total)"
  } else if (strategy == "simple") {
    cd4_subtypes <- c("T cells CD4 naive", "T cells CD4 memory resting", "T cells CD4 memory activated", "T cells follicular helper")
    cd4_name <- "T cells CD4+ (total)"
    cd8_subtypes <- c("T cells CD8")
    cd8_name <- "T cells CD8+ (total)"
  } else {
    stop("CIBERSORT_AGGREGATION_STRATEGY must be 'refined' or 'simple'.")
  }

  cd4_agg <- df %>%
    filter(Cell_Type %in% cd4_subtypes) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd4_name)

  cd8_agg <- df %>%
    filter(Cell_Type %in% cd8_subtypes) %>%
    group_by(Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd8_name)

  aggregated_subtypes <- c(cd4_subtypes, cd8_subtypes)
  other_cells <- df %>% filter(!Cell_Type %in% aggregated_subtypes)

  aggregated <- bind_rows(
    other_cells,
    cd4_agg %>% select(Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_agg %>% select(Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )

  details <- bind_rows(
    cd4_agg %>% select(Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_agg %>% select(Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )

  list(aggregated = aggregated, details = details)
}

xcell_agg <- aggregate_xcell_tcells(immune_patient)
immune_after_xcell <- xcell_agg$immune_patient_aggregated

if (!is.null(xcell_agg$aggregation_details)) {
  write.csv(xcell_agg$aggregation_details, file.path(DIRS$results, "xcell_tcell_aggregation_details.csv"), row.names = FALSE)
}

ciber_df <- immune_after_xcell %>% filter(Method == "CIBERSORTx")
non_ciber <- immune_after_xcell %>% filter(Method != "CIBERSORTx")
ciber_agg <- aggregate_cibersort_tcells(ciber_df, CIBERSORT_AGGREGATION_STRATEGY)

immune_patient_aggregated <- bind_rows(non_ciber, ciber_agg$aggregated)
immune_patient_aggregated <- apply_cluster_mapping(immune_patient_aggregated)

if (!is.null(ciber_agg$details)) {
  write.csv(ciber_agg$details, file.path(DIRS$results, "cibersort_tcell_aggregation_details.csv"), row.names = FALSE)
}

saveRDS(immune_patient_aggregated, file.path(DIRS$results, "immune_patient_level_AGGREGATED.rds"))
write.csv(immune_patient_aggregated, file.path(DIRS$results, "immune_patient_level_AGGREGATED.csv"), row.names = FALSE)

method_ymin_aggregated <- immune_patient_aggregated %>%
  group_by(Method) %>%
  summarise(ymin = if (min(Value, na.rm = TRUE) < 0) min(Value, na.rm = TRUE) else 0, .groups = "drop")
```

# Unpaired statistics

For each method, cell type, and timepoint, the pipeline compares immune estimates across CIMIC clusters.

- If `k = 2`, the global test is Wilcoxon rank-sum.
- If `k >= 3`, the global test is Kruskal-Wallis, followed by Dunn post hoc testing with Benjamini-Hochberg FDR correction.

```{r unpaired-stats}
summarise_by_cluster <- function(df) {
  df %>%
    group_by(cluster) %>%
    summarise(n = n(), mean = mean(Value, na.rm = TRUE), sd = sd(Value, na.rm = TRUE), .groups = "drop") %>%
    mutate(cluster = as.character(cluster))
}

empty_skip_tbl <- function() {
  tibble(analysis = character(), Method = character(), Cell_Type = character(), Timepoint = character(), reason = character(), k = integer(), n_total = integer(), n_unique_value = integer(), min_cluster_n = integer())
}

make_skip_row <- function(Method, Cell_Type, Timepoint, analysis, reason, k = NA_integer_, n_total = NA_integer_, n_unique_value = NA_integer_, min_cluster_n = NA_integer_) {
  tibble(analysis = analysis, Method = as.character(Method), Cell_Type = as.character(Cell_Type), Timepoint = as.character(Timepoint), reason = reason, k = as.integer(k), n_total = as.integer(n_total), n_unique_value = as.integer(n_unique_value), min_cluster_n = as.integer(min_cluster_n))
}

unpaired_global_one <- function(df, Method, Cell_Type, Timepoint) {
  df <- df %>% filter(is.finite(Value), !is.na(cluster))
  kk <- n_distinct(df$cluster)
  n_total <- nrow(df)
  n_unique <- n_distinct(df$Value)
  min_n <- if (kk > 0) min(as.integer(table(df$cluster))) else NA_integer_

  if (kk < 2) {
    return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_global", "k<2", kk, n_total, n_unique, min_n)))
  }

  summ <- summarise_by_cluster(df)

  if (n_unique < 2) {
    return(list(
      out = tibble(Method = as.character(Method), Cell_Type = as.character(Cell_Type), Timepoint = as.character(Timepoint), test = ifelse(kk == 2, "wilcox", "kruskal"), p_value = NA_real_, n_clusters = kk, note = "skipped_no_variation", summary = list(summ)),
      skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_global", "no_variation", kk, n_total, n_unique, min_n)
    ))
  }

  p <- if (kk == 2) {
    tryCatch(stats::wilcox.test(Value ~ cluster, data = df, exact = FALSE)$p.value, error = function(e) NA_real_)
  } else {
    tryCatch(stats::kruskal.test(Value ~ cluster, data = df)$p.value, error = function(e) NA_real_)
  }

  list(
    out = tibble(Method = as.character(Method), Cell_Type = as.character(Cell_Type), Timepoint = as.character(Timepoint), test = ifelse(kk == 2, "wilcox", "kruskal"), p_value = p, n_clusters = kk, note = NA_character_, summary = list(summ)),
    skip = NULL
  )
}

unpaired_dunn_one <- function(df, Method, Cell_Type, Timepoint, drop_small_groups_n = 2) {
  df2 <- df %>%
    filter(is.finite(Value), !is.na(cluster)) %>%
    group_by(cluster) %>%
    filter(n() >= drop_small_groups_n) %>%
    ungroup()

  kk <- n_distinct(df2$cluster)
  n_total <- nrow(df2)
  n_unique <- n_distinct(df2$Value)
  min_n <- if (kk > 0) min(as.integer(table(df2$cluster))) else NA_integer_

  if (kk < 3) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "k<3_after_filter", kk, n_total, n_unique, min_n)))
  if (n_unique < 2) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "no_variation", kk, n_total, n_unique, min_n)))

  cl_summ <- summarise_by_cluster(df2) %>% rename(n_group = n, mean_group = mean, sd_group = sd)
  dt <- tryCatch(suppressWarnings(FSA::dunnTest(Value ~ cluster, data = df2, method = "bh")$res), error = function(e) NULL)

  if (is.null(dt)) return(list(out = NULL, skip = make_skip_row(Method, Cell_Type, Timepoint, "unpaired_dunn", "dunn_error", kk, n_total, n_unique, min_n)))

  out <- as_tibble(dt) %>%
    rename(comparison = Comparison, z = Z, p_unadj = P.unadj, fdr = P.adj) %>%
    mutate(
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      Timepoint = as.character(Timepoint),
      cluster1 = sub(" -.*$", "", comparison),
      cluster2 = sub("^.*- ", "", comparison),
      z = as.numeric(z)
    ) %>%
    left_join(cl_summ %>% rename(cluster1 = cluster, n1 = n_group, mean1 = mean_group, sd1 = sd_group), by = "cluster1") %>%
    left_join(cl_summ %>% rename(cluster2 = cluster, n2 = n_group, mean2 = mean_group, sd2 = sd_group), by = "cluster2") %>%
    relocate(Method, Cell_Type, Timepoint, comparison, cluster1, cluster2, n1, mean1, sd1, n2, mean2, sd2, z, p_unadj, fdr)

  list(out = out, skip = NULL)
}

compute_unpaired_all <- function(immune_patient_df, out_prefix, drop_small_groups_n = 2) {
  # Safety step: all unpaired statistics use mapped CIM groups.
  # Example: if clusters 2 and 3 are both mapped to Dys-CIM, they are treated as one group.
  immune_patient_df <- apply_cluster_mapping(immune_patient_df) %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  keys <- immune_patient_df %>% dplyr::distinct(Method, Cell_Type, Timepoint) %>% dplyr::arrange(Timepoint, Method, Cell_Type)

  global_rows <- list()
  dunn_rows <- list()
  skip_rows <- list()

  for (i in seq_len(nrow(keys))) {
    key <- keys[i, ]
    df_sub <- immune_patient_df %>% filter(Method == key$Method, Cell_Type == key$Cell_Type, Timepoint == key$Timepoint)

    g <- unpaired_global_one(df_sub, key$Method, key$Cell_Type, key$Timepoint)
    if (!is.null(g$out)) global_rows[[length(global_rows) + 1]] <- g$out
    if (!is.null(g$skip)) skip_rows[[length(skip_rows) + 1]] <- g$skip

    d <- unpaired_dunn_one(df_sub, key$Method, key$Cell_Type, key$Timepoint, drop_small_groups_n)
    if (!is.null(d$out)) dunn_rows[[length(dunn_rows) + 1]] <- d$out
    if (!is.null(d$skip)) skip_rows[[length(skip_rows) + 1]] <- d$skip
  }

  global_tbl <- if (length(global_rows)) bind_rows(global_rows) else tibble()
  dunn_tbl <- if (length(dunn_rows)) bind_rows(dunn_rows) else tibble()
  skipped <- if (length(skip_rows)) bind_rows(skip_rows) else empty_skip_tbl()

  global_tbl <- global_tbl %>% mutate(FDR = p.adjust(p_value, method = "BH"))

  global_wide <- global_tbl %>%
    tidyr::unnest(summary) %>%
    pivot_wider(
      id_cols = c(Method, Cell_Type, Timepoint, test, p_value, FDR, n_clusters, note),
      names_from = cluster,
      values_from = c(n, mean, sd),
      names_glue = "{cluster}_{.value}"
    ) %>%
    arrange(Timepoint, Method, Cell_Type)

  write.csv(global_wide, file.path(DIRS$results, paste0(out_prefix, "_unpaired_global.csv")), row.names = FALSE)
  write.csv(dunn_tbl, file.path(DIRS$results, paste0(out_prefix, "_unpaired_dunn.csv")), row.names = FALSE)
  write.csv(skipped, file.path(DIRS$results, paste0(out_prefix, "_unpaired_skipped.csv")), row.names = FALSE)

  list(global = global_wide, dunn = dunn_tbl, skipped = skipped)
}

unpaired_original <- compute_unpaired_all(immune_patient, "ORIGINAL", drop_small_groups_n = 2)
unpaired_aggregated <- compute_unpaired_all(immune_patient_aggregated, "AGGREGATED", drop_small_groups_n = 2)
```

# Paired statistics

Within each CIMIC cluster, the pipeline performs paired Wilcoxon signed-rank tests comparing surgery versus biopsy for each method and cell type.

```{r paired-stats}
make_skip_row_paired <- function(Method, Cell_Type, cluster, reason, n_pairs = NA_integer_, n_unique_delta = NA_integer_) {
  tibble(analysis = "paired", Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster), reason = reason, n_pairs = n_pairs, n_unique_delta = n_unique_delta)
}

paired_one <- function(df, Method, Cell_Type, cluster_label) {
  wide <- df %>%
    filter(is.finite(Value), !is.na(Timepoint), !is.na(Patient_ID)) %>%
    group_by(Patient_ID, Timepoint) %>%
    summarise(Value = mean(Value), .groups = "drop") %>%
    pivot_wider(names_from = Timepoint, values_from = Value) %>%
    filter(is.finite(Biopsy) & is.finite(Surgery))

  if (nrow(wide) == 0) {
    return(list(stats = NULL, wide = NULL, skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "no_complete_pairs", 0)))
  }

  wide <- wide %>% mutate(Delta = Surgery - Biopsy)
  n_pairs <- nrow(wide)
  n_unique_delta <- n_distinct(wide$Delta)

  stats_base <- tibble(
    Method = Method,
    Cell_Type = Cell_Type,
    cluster = as.character(cluster_label),
    n_pairs = n_pairs,
    mean_biopsy = mean(wide$Biopsy),
    sd_biopsy = sd(wide$Biopsy),
    mean_surgery = mean(wide$Surgery),
    sd_surgery = sd(wide$Surgery),
    mean_delta = mean(wide$Delta),
    sd_delta = sd(wide$Delta)
  )

  if (n_pairs < 3) {
    return(list(
      stats = stats_base %>% mutate(p_value = NA_real_, note = "skipped_n_pairs<3"),
      wide = wide %>% mutate(Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "n_pairs<3", n_pairs, n_unique_delta)
    ))
  }

  if (n_unique_delta < 2) {
    return(list(
      stats = stats_base %>% mutate(p_value = NA_real_, note = "skipped_no_variation_in_delta"),
      wide = wide %>% mutate(Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, "no_variation_in_delta", n_pairs, n_unique_delta)
    ))
  }

  p <- tryCatch(stats::wilcox.test(wide$Surgery, wide$Biopsy, paired = TRUE, exact = FALSE)$p.value, error = function(e) NA_real_)

  list(
    stats = stats_base %>% mutate(p_value = p, note = NA_character_),
    wide = wide %>% mutate(Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label)),
    skip = NULL
  )
}

compute_paired_all <- function(immune_patient_df, out_prefix) {
  # Safety step: paired tests are stratified by mapped CIM groups.
  immune_patient_df <- apply_cluster_mapping(immune_patient_df) %>%
    dplyr::filter(!is.na(cluster)) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  groups <- immune_patient_df %>%
    dplyr::group_by(Method, Cell_Type, cluster) %>%
    dplyr::group_split(.keep = TRUE)

  stats_tbl <- purrr::map(groups, ~ paired_one(.x, .x$Method[[1]], .x$Cell_Type[[1]], .x$cluster[[1]])$stats) %>% bind_rows()
  wide_tbl <- purrr::map(groups, ~ paired_one(.x, .x$Method[[1]], .x$Cell_Type[[1]], .x$cluster[[1]])$wide) %>% bind_rows()
  skip_tbl <- purrr::map(groups, ~ paired_one(.x, .x$Method[[1]], .x$Cell_Type[[1]], .x$cluster[[1]])$skip) %>% bind_rows()

  stats_tbl <- stats_tbl %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(Method, cluster, Cell_Type)

  if (nrow(skip_tbl) == 0) {
    skip_tbl <- tibble(analysis = character(), Method = character(), Cell_Type = character(), cluster = character(), reason = character(), n_pairs = integer(), n_unique_delta = integer())
  }

  write.csv(stats_tbl, file.path(DIRS$results, paste0(out_prefix, "_paired_stats.csv")), row.names = FALSE)
  write.csv(wide_tbl, file.path(DIRS$results, paste0(out_prefix, "_paired_patient_deltas.csv")), row.names = FALSE)
  write.csv(skip_tbl, file.path(DIRS$results, paste0(out_prefix, "_paired_skipped.csv")), row.names = FALSE)

  list(stats = stats_tbl, wide = wide_tbl, skipped = skip_tbl)
}

paired_original <- compute_paired_all(immune_patient, "ORIGINAL")
paired_aggregated <- compute_paired_all(immune_patient_aggregated, "AGGREGATED")
```

# Plots

```{r plotting-functions}
# Choose which value to display
PLOT_P_COLUMN <- "p_value"
# PLOT_P_COLUMN <- "FDR"   # Use this instead for adjusted P values


compute_shared_ylims <- function(df, method_ymin) {

  method_name <- unique(df$Method)[1]

  method_min <- method_ymin %>%
    filter(Method == method_name) %>%
    pull(ymin)

  if (!length(method_min) || is.na(method_min)) {
    method_min <- 0
  }

  auto_ylims(
    df$Value[is.finite(df$Value)],
    ymin_floor = method_min
  )
}


plot_unpaired_one <- function(
    df,
    shared_ylim,
    unpaired_stats,
    p_column = PLOT_P_COLUMN
) {

  df <- df %>%
    filter(is.finite(Value), !is.na(cluster))

  if (n_distinct(df$cluster) < 2) {
    return(NULL)
  }

  method_name <- unique(df$Method)[1]
  cell_type   <- unique(df$Cell_Type)[1]
  timepoint   <- unique(df$Timepoint)[1]

  df <- df %>%
    mutate(
      cluster = factor(
        cluster,
        levels = names(cluster_colors)
      )
    )

  # Retrieve the P value already calculated
  stat_row <- unpaired_stats %>%
    filter(
      Method == method_name,
      Cell_Type == cell_type,
      Timepoint == timepoint
    )

  # Also match the dataset when available
  if ("dataset" %in% names(df) && "dataset" %in% names(stat_row)) {
    dataset_name <- unique(df$dataset)[1]

    stat_row <- stat_row %>%
      filter(dataset == dataset_name)
  }

  p_value <- stat_row[[p_column]][1]

  p_label <- if (length(p_value) == 0 || is.na(p_value)) {
    "P = N/A"
  } else if (p_value < 0.0001) {
    "P < 0.0001"
  } else {
    paste0("P = ", formatC(p_value, format = "f", digits = 4))
  }

  y_position <- shared_ylim[2] -
    0.05 * diff(shared_ylim)

  ggplot(df, aes(x = cluster, y = Value, fill = cluster)) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.7
    ) +
    geom_jitter(
      width = 0.15,
      alpha = 0.65,
      size = 2,
      color = "gray20"
    ) +
    annotate(
      "text",
      x = 1.5,
      y = y_position,
      label = p_label,
      size = 5,
      fontface = "bold"
    ) +
    scale_fill_manual(
      values = cluster_colors,
      name = "Cluster"
    ) +
    theme_pub() +
    coord_cartesian(
      ylim = shared_ylim,
      clip = "off"
    ) +
    labs(
      title = paste0(cell_type, " — ", timepoint),
      x = "",
      y = method_value_type(method_name)
    ) +
    theme(
      legend.position = "none"
    )
}


save_unpaired_plots <- function(
    immune_patient_df,
    method_ymin,
    version_label,
    unpaired_stats,
    p_column = PLOT_P_COLUMN
) {

  immune_patient_df <- apply_cluster_mapping(
    immune_patient_df
  ) %>%
    mutate(
      cluster = droplevels(cluster)
    )

  out_dir <- file.path(
    DIRS$plots,
    version_label,
    "unpaired"
  )

  dir.create(
    out_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  combinations <- immune_patient_df %>%
    distinct(Method, Cell_Type, Timepoint)

  purrr::pwalk(
    combinations,
    function(Method, Cell_Type, Timepoint) {

      plot_data <- immune_patient_df %>%
        filter(
          Method == !!Method,
          Cell_Type == !!Cell_Type,
          Timepoint == !!Timepoint
        )

      if (
        nrow(plot_data) == 0 ||
        n_distinct(
          plot_data$cluster[is.finite(plot_data$Value)]
        ) < 2
      ) {
        return(invisible(NULL))
      }

      shared_ylim <- compute_shared_ylims(
        plot_data,
        method_ymin
      )

      plot_object <- plot_unpaired_one(
        df = plot_data,
        shared_ylim = shared_ylim,
        unpaired_stats = unpaired_stats,
        p_column = p_column
      )

      if (is.null(plot_object)) {
        return(invisible(NULL))
      }

      file_png <- file.path(
        out_dir,
        paste0(
          safe_stem(
            "Unpaired",
            Method,
            Timepoint,
            Cell_Type
          ),
          ".png"
        )
      )

      ggsave(
        file_png,
        plot_object,
        width = PLOT_WIDTH_UNPAIRED,
        height = PLOT_HEIGHT_UNPAIRED,
        dpi = PLOT_DPI
      )
    }
  )

  invisible(TRUE)
}

compute_shared_ylims_paired <- function(paired_wide_df, method, cell_type, method_ymin) {
  method_min <- method_ymin %>% filter(Method == method) %>% pull(ymin)
  if (!length(method_min) || is.na(method_min)) method_min <- 0

  all_values <- paired_wide_df %>%
    filter(Method == !!method, Cell_Type == !!cell_type) %>%
    select(Biopsy, Surgery) %>%
    unlist() %>%
    as.numeric()

  auto_ylims(all_values, ymin_floor = method_min)
}

plot_paired_split_one <- function(paired_wide_df, method, cell_type, cluster_level, shared_ylim) {
  d <- paired_wide_df %>% filter(Method == !!method, Cell_Type == !!cell_type, cluster == !!cluster_level)
  if (nrow(d) < 3) return(NULL)

  pval <- tryCatch(wilcox.test(d$Surgery, d$Biopsy, paired = TRUE, exact = FALSE)$p.value, error = function(e) NA_real_)
  p_label <- if (is.na(pval)) "P = N/A" else if (pval < 0.001) "P < 0.001" else sprintf("P = %.4f", pval)

  plot_df <- d %>%
    select(Patient_ID, Biopsy, Surgery) %>%
    pivot_longer(cols = c(Biopsy, Surgery), names_to = "Timepoint", values_to = "Value") %>%
    mutate(Timepoint = factor(Timepoint, levels = c("Biopsy", "Surgery")))

  ggplot(plot_df, aes(x = Timepoint, y = Value, group = Patient_ID)) +
    geom_line(color = "gray65", alpha = 0.7, linewidth = 0.9) +
    geom_point(aes(color = Timepoint), size = 2.8, alpha = 0.85) +
    scale_color_manual(values = condition_colors, name = "Condition") +
    theme_pub() +
    coord_cartesian(ylim = shared_ylim, clip = "off") +
    labs(title = paste0(cell_type, " — ", as.character(cluster_level)), subtitle = paste0("n = ", nrow(d), " pairs"), x = "", y = method_value_type(method)) +
    annotate("text", x = 1.5, y = shared_ylim[2] * 0.97, label = p_label, size = 6, fontface = "bold")
}

save_paired_split_plots <- function(paired_obj, method_ymin, version_label) {
  paired_obj$wide <- apply_cluster_mapping(paired_obj$wide) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  out_dir <- file.path(DIRS$plots, version_label, "paired_split")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

  wide <- paired_obj$wide
  combos <- wide %>% distinct(Method, Cell_Type, cluster)

  purrr::pwalk(combos, function(Method, Cell_Type, cluster) {
    shared_ylim <- compute_shared_ylims_paired(wide, Method, Cell_Type, method_ymin)
    p <- plot_paired_split_one(wide, Method, Cell_Type, cluster, shared_ylim)
    if (is.null(p)) return(invisible(NULL))

    fn_png <- file.path(out_dir, paste0(safe_stem("Paired", Method, Cell_Type, paste0("cluster", cluster)), ".png"))
    ggsave(fn_png, p, width = PLOT_WIDTH_PAIRED, height = PLOT_HEIGHT_PAIRED, dpi = PLOT_DPI)
  })

  invisible(TRUE)
}
```

```{r run-plots}
if (params$make_plots) {

  #save_unpaired_plots(
  # immune_patient,
  #  method_ymin_original,
  #  "ORIGINAL",
  #  unpaired_original$global
  #)

 # save_paired_split_plots(
 #   paired_original,
 #  method_ymin_original,
 #   "ORIGINAL"
 # )

  save_unpaired_plots(
    immune_patient_aggregated,
    method_ymin_aggregated,
    "AGGREGATED",
    unpaired_aggregated$global
  )

  save_paired_split_plots(
    paired_aggregated,
    method_ymin_aggregated,
    "AGGREGATED"
  )

}
```

# Heatmaps

The cluster-level heatmap uses mean delta values, where delta is defined as:

```text
Delta = Surgery - Biopsy
```

Mean deltas are z-scored within each deconvolution method.

```{r heatmap-functions}
calculate_zscore_cluster_mean_deltas <- function(paired_wide_df) {
  paired_wide_df <- apply_cluster_mapping(paired_wide_df) %>%
    dplyr::mutate(cluster = droplevels(cluster))

  required_cols <- c("Method", "Cell_Type", "cluster", "Biopsy", "Surgery", "Delta")
  missing_cols <- base::setdiff(required_cols, colnames(paired_wide_df))
  if (length(missing_cols) > 0) stop("Missing required columns: ", paste(missing_cols, collapse = ", "))

  paired_wide_df %>%
    filter(!is.na(cluster)) %>%
    group_by(Method, Cell_Type, cluster) %>%
    summarise(
      n_pairs = n(),
      mean_biopsy = mean(Biopsy, na.rm = TRUE),
      sd_biopsy = sd(Biopsy, na.rm = TRUE),
      mean_surgery = mean(Surgery, na.rm = TRUE),
      sd_surgery = sd(Surgery, na.rm = TRUE),
      mean_delta = mean(Delta, na.rm = TRUE),
      sd_delta = sd(Delta, na.rm = TRUE),
      p_value = if (n() >= 3) tryCatch(wilcox.test(Surgery, Biopsy, paired = TRUE, exact = FALSE)$p.value, error = function(e) NA_real_) else NA_real_,
      .groups = "drop"
    ) %>%
    filter(n_pairs >= 3) %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    group_by(Method) %>%
    mutate(zscore_delta = safe_zscore(mean_delta)) %>%
    ungroup()
}

prepare_method_cell_by_cluster_matrix <- function(zscore_deltas, fill_na = FALSE) {
  zscore_deltas <- zscore_deltas %>%
    mutate(CellType_Method = paste0(Method, ": ", Cell_Type), cluster = as.character(cluster)) %>%
    filter(!is.na(cluster))

  cluster_order <- unique(unname(CLUSTER_MAPPING))
  present_clusters <- unique(as.character(zscore_deltas$cluster))
  clusts_display <- cluster_order[cluster_order %in% present_clusters]
  if (length(clusts_display) == 0) clusts_display <- present_clusters

  mat <- zscore_deltas %>%
    select(CellType_Method, cluster, zscore_delta) %>%
    pivot_wider(names_from = cluster, values_from = zscore_delta) %>%
    column_to_rownames("CellType_Method") %>%
    as.matrix()

  sig_mat <- zscore_deltas %>%
    select(CellType_Method, cluster, FDR) %>%
    pivot_wider(names_from = cluster, values_from = FDR) %>%
    column_to_rownames("CellType_Method") %>%
    as.matrix()

  mat <- mat[, clusts_display, drop = FALSE]
  sig_mat <- sig_mat[, clusts_display, drop = FALSE]
  if (fill_na) mat[is.na(mat)] <- 0

  method_anno <- data.frame(
    CellType_Method = rownames(mat),
    Method = sub(":.*$", "", rownames(mat)),
    stringsAsFactors = FALSE
  )
  method_anno$Method <- factor(method_anno$Method, levels = METHOD_ORDER)
  ord <- order(method_anno$Method, method_anno$CellType_Method)

  list(mat = mat[ord, , drop = FALSE], sig = sig_mat[ord, , drop = FALSE], anno = method_anno[ord, , drop = FALSE])
}

plot_cluster_delta_heatmap <- function(zscore_deltas, output_file, show_significance = TRUE, cluster_rows = TRUE, fill_na = FALSE, na_col = "grey90") {
  prep <- prepare_method_cell_by_cluster_matrix(zscore_deltas, fill_na = fill_na)
  mat <- prep$mat
  sig <- prep$sig
  anno <- prep$anno

  if (nrow(mat) == 0 || ncol(mat) == 0 || all(is.na(mat))) stop("Heatmap matrix is empty or all NA.")

  max_abs <- max(abs(mat), na.rm = TRUE)
  if (!is.finite(max_abs) || max_abs == 0) max_abs <- 2
  max_abs <- max(max_abs, 2)

  col_fun <- circlize::colorRamp2(c(-max_abs, -1, 0, 1, max_abs), c("#0571B0", "#92C5DE", "white", "#F4A582", "#CA0020"))

  methods_present <- unique(as.character(anno$Method))
  method_colors_subset <- METHOD_COLORS[methods_present]
  method_colors_subset <- method_colors_subset[!is.na(method_colors_subset)]

  row_ha <- ComplexHeatmap::rowAnnotation(
    Method = anno$Method,
    col = list(Method = method_colors_subset),
    annotation_name_gp = grid::gpar(fontsize = 11, fontface = "bold")
  )

  cell_fun <- NULL
  if (show_significance) {
    cell_fun <- function(j, i, x, y, width, height, fill) {
      fdr <- sig[i, j]
      if (is.na(fdr)) return(NULL)
      label <- if (fdr < 0.001) "***" else if (fdr < 0.01) "**" else if (fdr < 0.05) "*" else ""
      if (label != "") grid::grid.text(label, x, y, gp = grid::gpar(fontsize = 8, fontface = "bold"))
    }
  }

  display_names <- stringr::str_remove(rownames(mat), "^[^:]+: ")

  ht <- ComplexHeatmap::Heatmap(
    mat,
    name = "Z-score\n(mean Δ)",
    col = col_fun,
    na_col = na_col,
    cell_fun = cell_fun,
    row_names_side = "left",
    row_labels = display_names,
    row_names_gp = grid::gpar(fontsize = 9),
    cluster_rows = cluster_rows,
    show_row_dend = cluster_rows,
    cluster_columns = FALSE,
    column_names_gp = grid::gpar(fontsize = 14, fontface = "bold"),
    column_names_rot = 0,
    column_title = "Cluster-mean zΔ (Surgery - Biopsy)",
    column_title_gp = grid::gpar(fontsize = 16, fontface = "bold"),
    left_annotation = row_ha,
    border = TRUE
  )

  plot_height <- max(10, nrow(mat) * 0.15 + 3)
  png(output_file, width = 12, height = plot_height, units = "in", res = 300)
  ComplexHeatmap::draw(ht, heatmap_legend_side = "right", annotation_legend_side = "bottom")
  dev.off()

  output_file
}
```

```{r run-heatmaps}
if (params$make_heatmaps) {
  zscore_deltas_agg <- calculate_zscore_cluster_mean_deltas(paired_aggregated$wide)

  write.csv(zscore_deltas_agg, file.path(DIRS$results, "AGGREGATED_zscore_cluster_mean_deltas.csv"), row.names = FALSE)

  prep_tmp <- prepare_method_cell_by_cluster_matrix(zscore_deltas_agg, fill_na = FALSE)
  write.csv(prep_tmp$mat, file.path(DIRS$results, "AGGREGATED_heatmap_matrix_zscore_delta.csv"))
  write.csv(prep_tmp$sig, file.path(DIRS$results, "AGGREGATED_heatmap_matrix_FDR.csv"))

  heat_file <- file.path(DIRS$plots, "AGGREGATED_heatmap_cluster_mean_zDelta.png")
  plot_cluster_delta_heatmap(zscore_deltas_agg, heat_file, show_significance = TRUE, cluster_rows = TRUE, fill_na = FALSE)
}
```

# Output summary

```{r output-summary}
cat("Analysis complete.\n")
cat("Results folder:", DIRS$results, "\n")
cat("Plots folder:", DIRS$plots, "\n")
cat("Temporary folder:", DIRS$tmp, "\n")
cat("Bioconductor cache:", DIRS$cache, "\n")
```

# Session information

This section should be kept in the rendered report for reproducibility.

```{r session-info}
sessionInfo()
```
