---
title: "CIMIC Baseline Analysis: Fun-CIM versus Dys-CIM"
author: " Gbadamosi Lab - IL/MG"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    df_print: paged
---

<!--
PURPOSE
This R Markdown performs an differential-expression analysis of baseline
NKI and SMC samples classified as Fun-CIM or Dys-CIM. It fits a single
CIM-group limma-trend model and runs direction-specific GO Biological Process
and Hallmark overrepresentation analyses.

IMPORTANT
The input expression matrix is assumed to contain log2(TPM + 1), or another
already log-transformed TPM measure. Do not use this workflow on raw counts.
For raw counts, use a count-aware workflow such as limma-voom, edgeR, or DESeq2.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 150
)
options(stringsAsFactors = FALSE, scipen = 999)
```

# Analysis overview

This analysis uses baseline (`_B`) samples only and compares **Fun-CIM minus
Dys-CIM**. Therefore, a positive `logFC` indicates higher expression in
Fun-CIM, whereas a negative `logFC` indicates higher expression in Dys-CIM.

One model is fitted:

1. **CIM-group model:** `expression ~ CIM_group`

This model estimates the overall observed Fun-CIM versus Dys-CIM expression
difference across the combined baseline dataset. ORA is performed independently for
Fun-CIM-higher and Dys-CIM-higher genes. The universe is the set of genes that
passed expression filtering, was tested by limma, and was successfully mapped
to the identifier used for enrichment.

# 1. Configuration

Edit this section when paths, thresholds, colors, or plotting preferences
change. No other section should normally require manual editing.

```{r configuration}
# ---- Input files ------------------------------------------------------------
DATA_DIR <- paste0(
  "C:/Users/lopesdelima.i/OneDrive - University of Florida/",
  "Gbadamosi Lab/Iasmim Lopes/01.CIMIC/",
  "DEGs_baseline/data"
)

CLUSTER_FILE <- file.path(DATA_DIR, "nki_smc_clusters.csv")
EXPRESSION_FILE <- file.path(DATA_DIR, "nki_smc_combine_TPM_lognorm.csv")

# Results are written beside the data directory.
OUTPUT_DIR <- file.path(dirname(DATA_DIR), "results")

# ---- Biological definitions ------------------------------------------------
CLUSTER_MAPPING <- c("1" = "Dys-CIM", "2" = "Fun-CIM")
CIM_LEVELS <- c("Dys-CIM", "Fun-CIM")  # Reference level is Dys-CIM.
COHORT_LEVELS <- c("nki", "smc")

# Fixed manuscript palette. Replace these hexadecimal values if needed.
CIM_COLORS <- c(
  "Fun-CIM" = "#2A9D8F",
  "Dys-CIM" = "#D95F59",
  "Not significant" = "#B8B8B8"
)
COHORT_COLORS <- c("nki" = "#3B6FB6", "smc" = "#E69F00")

# ---- Filtering and significance thresholds --------------------------------
# The matrix is assumed to be ALREADY log-transformed TPM.
MIN_LOG_EXPRESSION <- 1
MIN_SAMPLE_FRACTION <- 0.20
MIN_VARIANCE <- 1e-8

FDR_CUTOFF <- 0.05
LOG2FC_CUTOFF <- log2(1.25)  # A 25% linear fold difference (~0.322 log2FC).

# ---- Figure settings --------------------------------------------------------
N_VOLCANO_LABELS_PER_DIRECTION <- 10
N_HEATMAP_GENES <- 40
N_ORA_TERMS <- 15
ORA_DESCRIPTION_WRAP <- 35
ORA_FIGURE_WIDTH <- 4.5
ORA_FIGURE_HEIGHT <- 3.2
FIGURE_DPI <- 600
RANDOM_SEED <- 1234

set.seed(RANDOM_SEED)
```

# 2. Packages and output folders

The report checks packages but does not install them silently. Install any
reported missing packages before rendering.

```{r packages-and-folders}
required_packages <- c(
  "tidyverse", "limma", "ggrepel", "pheatmap",
  "clusterProfiler", "org.Hs.eg.db", "msigdbr", "AnnotationDbi"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following packages before rendering: ",
    paste(missing_packages, collapse = ", ")
  )
}

suppressPackageStartupMessages({
  library(tidyverse)
  library(limma)
  library(ggrepel)
  library(pheatmap)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(msigdbr)
  library(AnnotationDbi)
})

dir_map <- list(
  qc_tables       = file.path(OUTPUT_DIR, "00_qc", "tables"),
  qc_figures      = file.path(OUTPUT_DIR, "00_qc", "figures"),
  model_tables = file.path(OUTPUT_DIR, "01_CIM_group_model", "DEG_tables"),
  model_figs   = file.path(OUTPUT_DIR, "01_CIM_group_model", "DEG_figures"),
  model_ora    = file.path(OUTPUT_DIR, "01_CIM_group_model", "ORA"),
  logs = file.path(OUTPUT_DIR, "logs")
)
invisible(lapply(dir_map, dir.create, recursive = TRUE, showWarnings = FALSE))

save_plot <- function(plot, filename, folder, width, height) {
  ggplot2::ggsave(
    file.path(folder, paste0(filename, ".png")), plot,
    width = width, height = height, units = "in", dpi = FIGURE_DPI,
    bg = "white"
  )
  ggplot2::ggsave(
    file.path(folder, paste0(filename, ".pdf")), plot,
    width = width, height = height, units = "in", device = cairo_pdf,
    bg = "white"
  )
}
```

# 3. Import and validate inputs

```{r import-inputs}
if (!file.exists(CLUSTER_FILE)) stop("Cluster file not found: ", CLUSTER_FILE)
if (!file.exists(EXPRESSION_FILE)) stop("Expression file not found: ", EXPRESSION_FILE)

cluster_metadata <- readr::read_csv(CLUSTER_FILE, show_col_types = FALSE) |>
  mutate(
    sample_id = as.character(sample_id),
    cluster_assignments = as.character(cluster_assignments),
    cohort = str_to_lower(as.character(cohort))
  )

expression_raw <- readr::read_csv(EXPRESSION_FILE, show_col_types = FALSE)

required_metadata_columns <- c("sample_id", "cluster_assignments", "cohort")
missing_metadata_columns <- setdiff(required_metadata_columns, names(cluster_metadata))
if (length(missing_metadata_columns) > 0) {
  stop("Missing annotation columns: ", paste(missing_metadata_columns, collapse = ", "))
}
if (anyDuplicated(cluster_metadata$sample_id)) {
  stop("Duplicated sample_id values were found in the cluster file.")
}
if (!all(cluster_metadata$cluster_assignments %in% names(CLUSTER_MAPPING))) {
  stop("Unexpected cluster assignment(s): ", paste(
    setdiff(unique(cluster_metadata$cluster_assignments), names(CLUSTER_MAPPING)),
    collapse = ", "
  ))
}
if (!all(cluster_metadata$cohort %in% COHORT_LEVELS)) {
  stop("Unexpected cohort value(s): ", paste(
    setdiff(unique(cluster_metadata$cohort), COHORT_LEVELS), collapse = ", "
  ))
}

# Sample columns are defined by terminal _B or _S suffixes. All other columns
# are candidates for the gene identifier column.
sample_columns <- names(expression_raw)[str_detect(names(expression_raw), "_(B|S)$")]
non_sample_columns <- setdiff(names(expression_raw), sample_columns)

if (length(sample_columns) == 0) stop("No sample columns ending in _B or _S were found.")
if (length(non_sample_columns) != 1) {
  stop(
    "Expected exactly one non-sample gene identifier column, but found: ",
    paste(non_sample_columns, collapse = ", ")
  )
}
GENE_COLUMN <- non_sample_columns[[1]]

# Force sample data to numeric and detect failed conversions.
expression_numeric <- expression_raw |>
  mutate(across(all_of(sample_columns), ~ suppressWarnings(as.numeric(.x))))

if (anyNA(expression_numeric[, sample_columns])) {
  stop("Missing or nonnumeric expression values were found in sample columns.")
}
if (any(!is.finite(as.matrix(expression_numeric[, sample_columns])))) {
  stop("Infinite expression values were found.")
}
if (any(is.na(expression_numeric[[GENE_COLUMN]]) |
        expression_numeric[[GENE_COLUMN]] == "")) {
  stop("Missing gene identifiers were found in column: ", GENE_COLUMN)
}

input_summary <- tibble(
  metric = c(
    "Expression rows", "All expression sample columns", "Baseline columns",
    "Surgery columns", "Annotation rows", "Expression minimum",
    "Expression median", "Expression maximum"
  ),
  value = c(
    nrow(expression_numeric), length(sample_columns),
    sum(str_detect(sample_columns, "_B$")),
    sum(str_detect(sample_columns, "_S$")), nrow(cluster_metadata),
    min(as.matrix(expression_numeric[, sample_columns])),
    median(as.matrix(expression_numeric[, sample_columns])),
    max(as.matrix(expression_numeric[, sample_columns]))
  )
)
readr::write_csv(input_summary, file.path(dir_map$qc_tables, "input_summary.csv"))
knitr::kable(input_summary)
```

The range above must be reviewed. Values that are very large (for example,
hundreds or thousands) may indicate untransformed TPM, in which case this
analysis should be stopped and the transformation status confirmed.

# 4. Normalize sample names and select baseline samples

Only a terminal `_B` or `_S` is removed. Internal underscores in SMC IDs such
as `OB_0139` are preserved.

```{r sample-matching}
expression_sample_metadata <- tibble(
  expression_sample = sample_columns,
  timepoint = case_when(
    str_detect(expression_sample, "_B$") ~ "Baseline",
    str_detect(expression_sample, "_S$") ~ "Surgery",
    TRUE ~ NA_character_
  ),
  sample_id = str_remove(expression_sample, "_(B|S)$")
)

matching_report <- expression_sample_metadata |>
  left_join(cluster_metadata, by = "sample_id") |>
  mutate(
    CIM_group = unname(CLUSTER_MAPPING[cluster_assignments]),
    match_status = if_else(
      is.na(cluster_assignments) | is.na(cohort), "Unmatched", "Matched"
    )
  )

baseline_metadata <- matching_report |>
  filter(timepoint == "Baseline")

unmatched_expression <- matching_report |>
  filter(match_status == "Unmatched")
unmatched_annotation <- cluster_metadata |>
  anti_join(expression_sample_metadata, by = "sample_id")

readr::write_csv(matching_report, file.path(dir_map$qc_tables, "sample_matching.csv"))
readr::write_csv(
  unmatched_expression,
  file.path(dir_map$qc_tables, "unmatched_expression_samples.csv")
)
readr::write_csv(
  unmatched_annotation,
  file.path(dir_map$qc_tables, "unmatched_annotation_samples.csv")
)

if (nrow(unmatched_expression) > 0) {
  stop("Unmatched expression samples were found. Review the exported matching report.")
}
if (nrow(unmatched_annotation) > 0) {
  stop("Annotation samples absent from the expression matrix were found.")
}
if (anyDuplicated(baseline_metadata$expression_sample)) {
  stop("Duplicated baseline expression sample names were found.")
}

baseline_metadata <- baseline_metadata |>
  mutate(
    CIM_group = factor(CIM_group, levels = CIM_LEVELS),
    cohort = factor(cohort, levels = COHORT_LEVELS)
  )

if (anyNA(baseline_metadata$CIM_group) || anyNA(baseline_metadata$cohort)) {
  stop("Missing CIM or cohort assignments remain after matching.")
}

knitr::kable(
  baseline_metadata |>
    count(cohort, CIM_group, name = "n") |>
    tidyr::pivot_wider(names_from = CIM_group, values_from = n, values_fill = 0)
)
```

# 5. Build and filter the expression matrix

Duplicated gene identifiers are collapsed by their mean expression before
filtering. A gene is retained when its expression is at least
`r MIN_LOG_EXPRESSION` in at least `r scales::percent(MIN_SAMPLE_FRACTION)` of
baseline samples and its variance exceeds `r MIN_VARIANCE`.

```{r expression-matrix-and-filtering}
baseline_long <- expression_numeric |>
  dplyr::select(all_of(c(GENE_COLUMN, baseline_metadata$expression_sample))) |>
  dplyr::rename(gene_id = all_of(GENE_COLUMN)) |>
  mutate(gene_id = as.character(gene_id)) |>
  group_by(gene_id) |>
  summarise(across(everything(), mean), .groups = "drop")

baseline_matrix <- baseline_long |>
  column_to_rownames("gene_id") |>
  as.matrix()

# Reorder explicitly: design rows and expression columns must be identical.
baseline_matrix <- baseline_matrix[, baseline_metadata$expression_sample, drop = FALSE]
stopifnot(identical(colnames(baseline_matrix), baseline_metadata$expression_sample))

minimum_samples <- ceiling(ncol(baseline_matrix) * MIN_SAMPLE_FRACTION)
keep_expression <- rowSums(baseline_matrix >= MIN_LOG_EXPRESSION) >= minimum_samples
gene_variance <- apply(baseline_matrix, 1, var)
keep_variance <- is.finite(gene_variance) & gene_variance > MIN_VARIANCE
keep_gene <- keep_expression & keep_variance
filtered_matrix <- baseline_matrix[keep_gene, , drop = FALSE]

filtering_summary <- tibble(
  metric = c(
    "Unique genes before filtering", "Removed for low expression",
    "Removed for insufficient variance among expression-passing genes",
    "Genes retained and tested"
  ),
  n = c(
    nrow(baseline_matrix), sum(!keep_expression),
    sum(keep_expression & !keep_variance), nrow(filtered_matrix)
  )
)
readr::write_csv(filtering_summary, file.path(dir_map$qc_tables, "filtering_summary.csv"))
knitr::kable(filtering_summary)
```

# 6. Sample composition and exploratory quality control

```{r sample-composition, fig.width=6, fig.height=4}
composition_data <- baseline_metadata |>
  count(cohort, CIM_group, name = "n")

p_composition <- ggplot(composition_data, aes(cohort, n, fill = CIM_group)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(label = n), position = position_dodge(width = 0.7), vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = CIM_COLORS[c("Dys-CIM", "Fun-CIM")]) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Cohort", y = "Number of baseline samples", fill = NULL) +
  theme_classic(base_size = 11) +
  theme(legend.position = "top")

save_plot(p_composition, "sample_distribution", dir_map$qc_figures, 5, 3.6)
p_composition
```

```{r expression-distributions, fig.width=8, fig.height=4}
distribution_data <- as_tibble(filtered_matrix, rownames = "gene_id") |>
  pivot_longer(-gene_id, names_to = "expression_sample", values_to = "expression") |>
  left_join(
    baseline_metadata |> dplyr::select(expression_sample, CIM_group, cohort),
    by = "expression_sample"
  )

p_distribution <- ggplot(
  distribution_data,
  aes(expression_sample, expression, fill = CIM_group)
) +
  geom_boxplot(outlier.size = 0.2, linewidth = 0.25) +
  facet_grid(~ cohort, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = CIM_COLORS[c("Dys-CIM", "Fun-CIM")]) +
  labs(x = NULL, y = "Log-normalized TPM", fill = NULL) +
  theme_classic(base_size = 9) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
    legend.position = "top"
  )

save_plot(p_distribution, "expression_distributions", dir_map$qc_figures, 9, 4.5)
p_distribution
```

# 7. Fit the limma-trend CIM-group model

`eBayes(trend = TRUE, robust = TRUE)` is used because the input consists of
continuous log-normalized TPM values and may retain an expression-dependent
mean-variance relationship. This is **limma-trend**, not limma-voom.

```{r model-functions}
fit_cim_group_model <- function(expression_matrix, metadata) {
  design <- model.matrix(~ CIM_group, data = metadata)

  coefficient_name <- "CIM_groupFun-CIM"
  if (!coefficient_name %in% colnames(design)) {
    stop("Expected coefficient was not created: ", coefficient_name)
  }
  if (qr(design)$rank < ncol(design)) stop("The model design matrix is not full rank.")

  fit <- limma::lmFit(expression_matrix, design)
  fit <- limma::eBayes(fit, trend = TRUE, robust = TRUE)

  results <- limma::topTable(
    fit, coef = coefficient_name, number = Inf, sort.by = "none",
    adjust.method = "BH"
  ) |>
    rownames_to_column("gene_id") |>
    as_tibble() |>
    mutate(
      linear_fold_change = 2^logFC,
      percent_difference = 100 * (linear_fold_change - 1),
      direction = case_when(
        logFC > 0 ~ "Fun-CIM higher",
        logFC < 0 ~ "Dys-CIM higher",
        TRUE ~ "No change"
      ),
      significance = case_when(
        adj.P.Val <= FDR_CUTOFF & logFC >= LOG2FC_CUTOFF ~ "Fun-CIM higher",
        adj.P.Val <= FDR_CUTOFF & logFC <= -LOG2FC_CUTOFF ~ "Dys-CIM higher",
        TRUE ~ "Not significant"
      )
    )

  list(fit = fit, design = design, results = results)
}

export_deg_results <- function(results, output_folder) {
  readr::write_csv(results, file.path(output_folder, "all_tested_genes.csv"))
  readr::write_csv(
    filter(results, significance != "Not significant"),
    file.path(output_folder, "significant_DEGs.csv")
  )
  readr::write_csv(
    filter(results, significance == "Fun-CIM higher"),
    file.path(output_folder, "Fun-CIM_higher_DEGs.csv")
  )
  readr::write_csv(
    filter(results, significance == "Dys-CIM higher"),
    file.path(output_folder, "Dys-CIM_higher_DEGs.csv")
  )
}
```

```{r fit-models}
cim_group_model <- fit_cim_group_model(filtered_matrix, baseline_metadata)

export_deg_results(cim_group_model$results, dir_map$model_tables)

deg_counts <- cim_group_model$results |>
  count(significance) |>
  mutate(model = "CIM-group model") |>
  dplyr::select(model, significance, n)

readr::write_csv(deg_counts, file.path(dir_map$model_tables, "DEG_counts.csv"))
knitr::kable(deg_counts)
```

# 8. Volcano plots

The horizontal dotted line represents raw `P = 0.05` for visual reference only.
DEG classification is based on BH-adjusted `P <= r FDR_CUTOFF` and
`|log2FC| >= r round(LOG2FC_CUTOFF, 3)`.

```{r volcano-functions}
make_volcano <- function(results, title) {
  label_data <- results |>
    filter(significance != "Not significant") |>
    group_by(significance) |>
    slice_min(order_by = adj.P.Val, n = N_VOLCANO_LABELS_PER_DIRECTION,
              with_ties = FALSE) |>
    ungroup()

  ggplot(results, aes(logFC, -log10(pmax(P.Value, .Machine$double.xmin)))) +
    geom_point(aes(color = significance), size = 1.25, alpha = 0.75) +
    geom_vline(
      xintercept = c(-LOG2FC_CUTOFF, LOG2FC_CUTOFF),
      linetype = "dashed", color = "grey45", linewidth = 0.4
    ) +
    geom_hline(
      yintercept = -log10(0.05), linetype = "dotted",
      color = "grey45", linewidth = 0.4
    ) +
    ggrepel::geom_text_repel(
      data = label_data, aes(label = gene_id), size = 2.7,
      max.overlaps = Inf, box.padding = 0.35, min.segment.length = 0
    ) +
    scale_color_manual(
      values = c(
        "Fun-CIM higher" = CIM_COLORS[["Fun-CIM"]],
        "Dys-CIM higher" = CIM_COLORS[["Dys-CIM"]],
        "Not significant" = CIM_COLORS[["Not significant"]]
      )
    ) +
    labs(
      title = title, x = expression(log[2]*"FC (Fun-CIM - Dys-CIM)"),
      y = expression(-log[10]*"(raw p-value)"), color = NULL
    ) +
    theme_classic(base_size = 11) +
    theme(legend.position = "top", plot.title = element_text(face = "bold"))
}

volcano_cim_group <- make_volcano(cim_group_model$results, "CIM-group model")
save_plot(volcano_cim_group, "volcano_CIM_group_model", dir_map$model_figs, 5, 4.3)
volcano_cim_group
```

# 9. Heatmap of top CIM-group model DEGs

The heatmap uses the row-scaled expression matrix. Column
annotations show both CIM group and cohort. If the model has no
genes passing the DEG thresholds, its most statistically ranked tested genes
are shown instead; this fallback should be stated when interpreting the figure.

```{r heatmaps, fig.width=9, fig.height=8}
column_annotation <- baseline_metadata |>
  dplyr::select(expression_sample, CIM_group, cohort) |>
  column_to_rownames("expression_sample")

annotation_colors <- list(
  CIM_group = CIM_COLORS[c("Dys-CIM", "Fun-CIM")],
  cohort = COHORT_COLORS
)

make_deg_heatmap <- function(results, model_label, output_folder, filename) {
  significant_results <- results |>
    filter(significance != "Not significant") |>
    arrange(adj.P.Val, desc(abs(logFC)))

  used_fallback <- nrow(significant_results) == 0
  heatmap_source <- if (!used_fallback) {
    significant_results
  } else {
    results |> arrange(adj.P.Val, desc(abs(logFC)))
  }

  heatmap_genes <- head(heatmap_source$gene_id, N_HEATMAP_GENES)
  heatmap_matrix <- filtered_matrix[heatmap_genes, , drop = FALSE]
  heatmap_matrix <- t(scale(t(heatmap_matrix)))
  heatmap_matrix[!is.finite(heatmap_matrix)] <- 0

  draw_heatmap <- function() {
    pheatmap::pheatmap(
      heatmap_matrix,
      annotation_col = column_annotation,
      annotation_colors = annotation_colors,
      color = colorRampPalette(c("#2166AC", "white", "#B2182B"))(100),
      show_colnames = FALSE,
      fontsize_row = 7,
      border_color = NA,
      main = paste0("Top ", model_label, " differential-expression genes")
    )
  }

  png(
    file.path(output_folder, paste0(filename, ".png")),
    width = 8, height = 8, units = "in", res = FIGURE_DPI
  )
  draw_heatmap()
  dev.off()

  pdf(
    file.path(output_folder, paste0(filename, ".pdf")),
    width = 8, height = 8
  )
  draw_heatmap()
  dev.off()

  tibble(
    model = model_label,
    significant_DEGs_available = nrow(significant_results),
    genes_displayed = length(heatmap_genes),
    ranked_gene_fallback_used = used_fallback
  )
}

heatmap_summary <- make_deg_heatmap(
  cim_group_model$results,
  "CIM-group model",
  dir_map$model_figs,
  "top_DEG_heatmap"
)

readr::write_csv(
  heatmap_summary,
  file.path(dir_map$model_tables, "heatmap_gene_selection_summary.csv")
)
knitr::kable(heatmap_summary)
```

# 10. Direction-specific overrepresentation analysis

## 11.1 Identifier mapping

This section assumes gene identifiers are either official human gene symbols or
Ensembl gene IDs. Version suffixes such as `.12` are removed from Ensembl IDs.
The detected identifier type and mapping statistics are exported. If the input
uses another identifier system, edit `detect_gene_id_type()` and `map_to_entrez()`.

```{r ora-functions}
detect_gene_id_type <- function(ids) {
  proportion_ensembl <- mean(str_detect(ids, "^ENSG[0-9]+(\\.[0-9]+)?$"))
  if (proportion_ensembl >= 0.5) "ENSEMBL" else "SYMBOL"
}

map_to_entrez <- function(ids, input_type) {
  clean_ids <- if (input_type == "ENSEMBL") str_remove(ids, "\\.[0-9]+$") else ids
  mapping <- AnnotationDbi::select(
    org.Hs.eg.db,
    keys = unique(clean_ids),
    keytype = input_type,
    columns = "ENTREZID"
  ) |>
    as_tibble() |>
    dplyr::rename(input_id_clean = all_of(input_type)) |>
    filter(!is.na(ENTREZID)) |>
    distinct(input_id_clean, ENTREZID)

  tibble(gene_id = ids, input_id_clean = clean_ids) |>
    left_join(mapping, by = "input_id_clean")
}

run_directional_ora <- function(results, model_name, output_folder) {
  id_type <- detect_gene_id_type(results$gene_id)
  mapped_all <- map_to_entrez(results$gene_id, id_type)
  universe_entrez <- unique(na.omit(mapped_all$ENTREZID))

  mapping_summary <- tibble(
    model = model_name,
    detected_input_type = id_type,
    tested_gene_ids = n_distinct(results$gene_id),
    tested_ids_mapped = n_distinct(mapped_all$gene_id[!is.na(mapped_all$ENTREZID)]),
    unique_universe_entrez_ids = length(universe_entrez),
    mapping_rate = tested_ids_mapped / tested_gene_ids
  )
  readr::write_csv(mapping_summary, file.path(output_folder, "ORA_mapping_summary.csv"))
  readr::write_csv(mapped_all, file.path(output_folder, "tested_gene_ID_mapping.csv"))

  ora_results <- list()
  for (direction_name in c("Fun-CIM higher", "Dys-CIM higher")) {
    direction_slug <- ifelse(direction_name == "Fun-CIM higher",
                             "Fun-CIM_higher", "Dys-CIM_higher")
    deg_ids <- results |>
      filter(significance == direction_name) |>
      pull(gene_id)
    deg_entrez <- mapped_all |>
      filter(gene_id %in% deg_ids, !is.na(ENTREZID)) |>
      pull(ENTREZID) |>
      unique()

    input_summary <- tibble(
      model = model_name, direction = direction_name,
      DEG_gene_ids = length(unique(deg_ids)),
      mapped_DEG_gene_ids = n_distinct(mapped_all$gene_id[
        mapped_all$gene_id %in% deg_ids & !is.na(mapped_all$ENTREZID)
      ]),
      unique_DEG_entrez_ids = length(deg_entrez),
      unique_universe_entrez_ids = length(universe_entrez)
    )
    readr::write_csv(
      input_summary,
      file.path(output_folder, paste0(direction_slug, "_ORA_input_summary.csv"))
    )

    if (length(deg_entrez) == 0) {
      warning("No mapped DEGs available for ", model_name, ": ", direction_name)
      next
    }

    go_fit <- clusterProfiler::enrichGO(
      gene = deg_entrez, universe = universe_entrez,
      OrgDb = org.Hs.eg.db, keyType = "ENTREZID", ont = "BP",
      pAdjustMethod = "BH", pvalueCutoff = 1, qvalueCutoff = 1,
      readable = TRUE
    )
    # Preserve the complete unsimplified output. GO semantic simplification can
    # be extremely slow when it is applied to every returned term, especially
    # because enrichGO() above intentionally uses permissive cutoffs. Restrict
    # simplification to statistically significant terms only.
    go_table <- as_tibble(as.data.frame(go_fit))
    go_fit_significant <- go_fit
    go_fit_significant@result <- go_fit_significant@result |>
      dplyr::filter(is.finite(p.adjust), p.adjust <= FDR_CUTOFF)

    if (nrow(go_fit_significant@result) > 1) {
      message("Simplifying significant GO BP terms for ", model_name,
              " / ", direction_name, " (n = ",
              nrow(go_fit_significant@result), ")")
      go_fit_simplified <- clusterProfiler::simplify(
        go_fit_significant, cutoff = 0.7, by = "p.adjust",
        select_fun = min, measure = "Wang"
      )
      go_simplified_table <- as_tibble(as.data.frame(go_fit_simplified))
    } else {
      go_simplified_table <- as_tibble(go_fit_significant@result)
    }

    hallmark_fit <- clusterProfiler::enricher(
      gene = as.character(deg_entrez), universe = as.character(universe_entrez),
      TERM2GENE = hallmark_sets, pAdjustMethod = "BH",
      pvalueCutoff = 1, qvalueCutoff = 1
    )
    hallmark_table <- as_tibble(as.data.frame(hallmark_fit))

    readr::write_csv(go_table, file.path(output_folder,
      paste0(direction_slug, "_GO_BP_all_results.csv")))
    readr::write_csv(go_simplified_table, file.path(output_folder,
      paste0(direction_slug, "_GO_BP_significant_simplified_results.csv")))
    readr::write_csv(hallmark_table, file.path(output_folder,
      paste0(direction_slug, "_Hallmark_all_results.csv")))

    ora_results[[direction_slug]] <- list(
      GO_BP = go_simplified_table,
      GO_BP_all = go_table,
      Hallmark = hallmark_table
    )
  }
  ora_results
}

prepare_ora_plot_data <- function(ora_table) {
  if (is.null(ora_table) || nrow(ora_table) == 0) return(tibble())
  ora_table |>
    filter(is.finite(p.adjust), p.adjust <= FDR_CUTOFF) |>
    arrange(p.adjust) |>
    slice_head(n = N_ORA_TERMS) |>
    mutate(
      Description = str_replace_all(Description, "_", " "),
      Description = str_wrap(Description, width = ORA_DESCRIPTION_WRAP),
      # First/top term is placed at the top after coord_flip().
      Description = factor(Description, levels = rev(unique(Description)))
    )
}

plot_ora_direction <- function(ora_results, direction_slug, collection, title) {
  direction_results <- ora_results[[direction_slug]]
  if (is.null(direction_results)) return(NULL)
  plot_data <- prepare_ora_plot_data(direction_results[[collection]])
  if (nrow(plot_data) == 0) return(NULL)

  ggplot(plot_data, aes(x = Description, y = Count)) +
    geom_col(aes(fill = p.adjust), color = "black", linewidth = 0.3) +
    coord_flip() +
    labs(
      title = title, x = NULL, y = "Gene Count", fill = "Adj p"
    ) +
    scale_fill_gradient(
      low = "red", high = "blue", trans = "reverse",
      labels = scales::scientific
    ) +
    theme_minimal(base_size = 8) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5),
      axis.text.y = element_text(size = 7, colour = "black", face = "bold"),
      axis.text.x = element_text(size = 7, colour = "black", face = "bold"),
      axis.title.x = element_text(size = 8, face = "bold"),
      legend.text = element_text(size = 6),
      legend.title = element_text(size = 7, face = "bold"),
      legend.key.size = grid::unit(0.3, "cm"),
      plot.title = element_text(size = 9, face = "bold", hjust = 0.5),
      plot.margin = margin(2, 2, 2, 2, "mm")
    )
}
```

```{r run-ora, fig.width=4.5, fig.height=3.2}
# Build Hallmark TERM2GENE once and reuse it for both DEG directions.
hallmark_sets <- msigdbr::msigdbr(species = "Homo sapiens", category = "H") |>
  dplyr::select(gs_name, entrez_gene) |>
  filter(!is.na(entrez_gene)) |>
  transmute(term = gs_name, gene = as.character(entrez_gene)) |>
  distinct()

cim_group_ora_results <- run_directional_ora(
  cim_group_model$results, "CIM-group model", dir_map$model_ora
)

ora_plot_jobs <- tidyr::crossing(
  direction_slug = c("Fun-CIM_higher", "Dys-CIM_higher"),
  collection = c("GO_BP", "Hallmark")
) |>
  mutate(
    results = map(direction_slug, ~ cim_group_ora_results),
    folder = dir_map$model_ora,
    group_label = if_else(direction_slug == "Fun-CIM_higher", "Fun-CIM", "Dys-CIM"),
    collection_label = if_else(collection == "GO_BP", "GO BP", "Hallmark"),
    title = paste0(group_label, ": Top 15 ", collection_label, " Terms"),
    filename = paste(direction_slug, collection, "CIM_group_model", "top15", sep = "_")
  )

for (i in seq_len(nrow(ora_plot_jobs))) {
  job <- ora_plot_jobs[i, ]
  p <- plot_ora_direction(
    job$results[[1]], job$direction_slug, job$collection, job$title
  )
  if (!is.null(p)) {
    save_plot(
      p, job$filename, job$folder,
      ORA_FIGURE_WIDTH, ORA_FIGURE_HEIGHT
    )
  }
}

p_go_fun <- plot_ora_direction(
  cim_group_ora_results, "Fun-CIM_higher", "GO_BP",
  "Fun-CIM: Top 15 GO BP Terms"
)
if (!is.null(p_go_fun)) p_go_fun
```

# 11. Analysis summary

```{r final-summary}
analysis_summary <- tibble(
  item = c(
    "Baseline samples", "Dys-CIM samples", "Fun-CIM samples",
    "NKI samples", "SMC samples", "Genes tested",
    "CIM-group model Fun-CIM-higher DEGs",
    "CIM-group model Dys-CIM-higher DEGs"
  ),
  value = c(
    nrow(baseline_metadata), sum(baseline_metadata$CIM_group == "Dys-CIM"),
    sum(baseline_metadata$CIM_group == "Fun-CIM"),
    sum(baseline_metadata$cohort == "nki"), sum(baseline_metadata$cohort == "smc"),
    nrow(filtered_matrix),
    sum(cim_group_model$results$significance == "Fun-CIM higher"),
    sum(cim_group_model$results$significance == "Dys-CIM higher")
  )
)

readr::write_csv(analysis_summary, file.path(dir_map$logs, "analysis_summary.csv"))
knitr::kable(analysis_summary)
```

Interpret the CIM-group model as the overall observed baseline difference
between Fun-CIM and Dys-CIM. Because cohort is not included as a covariate,
inspect cohort balance when evaluating whether cohort structure could contribute
to the observed expression differences.

# 12. Reproducibility records

```{r save-parameters-and-session-info}
parameter_lines <- c(
  paste("Expression file:", EXPRESSION_FILE),
  paste("Cluster file:", CLUSTER_FILE),
  paste("Output directory:", OUTPUT_DIR),
  paste("Gene identifier column:", GENE_COLUMN),
  paste("Minimum log expression:", MIN_LOG_EXPRESSION),
  paste("Minimum sample fraction:", MIN_SAMPLE_FRACTION),
  paste("Minimum variance:", MIN_VARIANCE),
  paste("FDR cutoff:", FDR_CUTOFF),
  paste("Absolute log2FC cutoff:", LOG2FC_CUTOFF),
  paste("Random seed:", RANDOM_SEED),
  paste("Analysis date:", Sys.Date())
)
writeLines(parameter_lines, file.path(dir_map$logs, "analysis_parameters.txt"))
writeLines(capture.output(sessionInfo()), file.path(dir_map$logs, "session_info.txt"))

sessionInfo()
```
