---
title: "CIMIC Clinical Analysis and NKI PAM50 Classification"
author: "Gbadamosi Lab - IL/MG"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
params:
  data_dir: "C:/Users/yasmi/OneDrive - University of Florida/Gbadamosi Lab/Iasmim Lopes/01.CIMIC/Clinical_data/data"
  results_dir: "C:/Users/yasmi/OneDrive - University of Florida/Gbadamosi Lab/Iasmim Lopes/01.CIMIC/Clinical_data/results"
  monte_carlo_B: 1000000
  random_seed: 20260812
---

<!--
PURPOSE
  1. Calculate research-use PAM50 subtypes for NKI biopsy (_B) and surgery (_S)
     samples from log2(TPM + 1) RNA-seq expression using genefu.
  2. Add the NKI PAM50 predictions to a new NKI master file.
  3. Construct validated one-row-per-patient NKI, SMC, and combined datasets.
  4. Generate cohort-specific demographics, combined demographics, and the
     Fun-CIM versus Dys-CIM clinical-characteristics table.
  5. Generate a separate CSV documenting every statistical test.
  6. Generate manuscript figures using the established CIMIC colors/aesthetics.

IMPORTANT
  - _B and _S are biological timepoints and are never discarded before a
    canonical patient ID and timepoint have been stored and validated.
  - SMC _1 identifies biopsy and _3 identifies surgery when an explicit _B/_S
    suffix is absent. When both are present, they must agree.
  - The original nki_master.csv is never overwritten.
  - PAM50 predictions are research-use classifications, not Prosigna results.
-->

# Analysis overview

This report uses the following inputs:

- `nki_master.csv`
- `smc_master.csv`
- `nki_smc_combine_TPM_lognorm.csv`
- `nki_smc_clusters.csv`

The expression matrix is expected to contain HGNC gene symbols in `gene_id`,
samples in columns, and log2(TPM + 1) values. NKI expression columns are expected
to follow the form `2046_B` and `2046_S` (optional `_Rep1` is accepted).

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.align = "center",
  dpi = 300
)

suppressPackageStartupMessages({
  library(data.table)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(tibble)
  library(ggplot2)
  library(scales)
  library(genefu)
  library(AnnotationDbi)
  library(org.Hs.eg.db)
  library(knitr)
})

set.seed(params$random_seed)

data_dir <- params$data_dir
results_dir <- params$results_dir

input_paths <- list(
  nki = file.path(data_dir, "nki_master.csv"),
  smc = file.path(data_dir, "smc_master.csv"),
  expression = file.path(data_dir, "nki_smc_combine_TPM_lognorm.csv"),
  clusters = file.path(data_dir, "nki_smc_clusters.csv")
)

output_dirs <- list(
  tables = file.path(results_dir, "tables"),
  figures = file.path(results_dir, "figures"),
  qc = file.path(results_dir, "QC")
)

invisible(lapply(
  c(results_dir, unname(unlist(output_dirs))),
  dir.create,
  recursive = TRUE,
  showWarnings = FALSE
))

missing_files <- unlist(input_paths)[!file.exists(unlist(input_paths))]
if (length(missing_files) > 0L) {
  stop("Missing input file(s):\n", paste(missing_files, collapse = "\n"))
}

cluster_levels <- c("Fun-CIM", "Dys-CIM")
pam50_levels <- c("Basal", "Her2", "LumA", "LumB", "Normal")
grade_levels <- c("Grade 1", "Grade 2", "Grade 3")

# Colors reproduce the supplied panels, with only publication-quality refinements.
cluster_colors <- c("Fun-CIM" = "#4FB5AD", "Dys-CIM" = "#ED8588")
pam50_colors <- c(
  "Basal" = "#C94F55",
  "Her2" = "#D98A9B",
  "LumA" = "#0072B2",
  "LumB" = "#5A9145",
  "Normal" = "#8E5BA6"
)
grade_colors <- c(
  "Grade 1" = "#F7BFC2",
  "Grade 2" = "#E77E82",
  "Grade 3" = "#CF5055"
)
```

# Helper functions

```{r helpers}
normalize_missing <- function(x) {
  x <- str_squish(as.character(x))
  x[x %in% c("", "NA", "N/A", "na", "n/a", "Unknown", "unknown", "999")] <- NA_character_
  x
}

first_unique <- function(x, variable, patient) {
  x <- unique(na.omit(normalize_missing(x)))
  if (length(x) > 1L) {
    stop("Conflicting ", variable, " values for patient ", patient, ": ", paste(x, collapse = ", "))
  }
  if (length(x) == 0L) NA_character_ else x[[1L]]
}

time_from_suffix <- function(x) {
  case_when(
    str_detect(as.character(x), "_B(?:_Rep\\d+)?$") ~ "B",
    str_detect(as.character(x), "_S(?:_Rep\\d+)?$") ~ "S",
    TRUE ~ NA_character_
  )
}

smc_stage_from_id <- function(x) {
  # Legacy SMC collection index: _1 = biopsy; _3 = surgery.
  case_when(
    str_detect(as.character(x), "_1(?:_[BS](?:_Rep\\d+)?)?$") ~ "B",
    str_detect(as.character(x), "_3(?:_[BS](?:_Rep\\d+)?)?$") ~ "S",
    TRUE ~ NA_character_
  )
}

validate_timepoint <- function(explicit, suffix, legacy, sample_id, cohort) {
  candidates <- unique(na.omit(c(explicit, suffix, legacy)))
  if (length(candidates) != 1L || !candidates %in% c("B", "S")) {
    stop(
      "Timepoint conflict or missing timepoint for ", cohort, " sample ", sample_id,
      ". Values observed: ", paste(candidates, collapse = ", ")
    )
  }
  candidates[[1L]]
}

combine_suffix_timepoints <- function(x, y, sample_id) {
  candidates <- unique(na.omit(c(x, y)))
  if (length(candidates) > 1L) {
    stop(
      "Conflicting _B/_S suffixes across SMC identifiers for sample ",
      sample_id, ": ", paste(candidates, collapse = ", ")
    )
  }
  if (length(candidates) == 0L) NA_character_ else candidates[[1L]]
}

clean_pam50 <- function(x) {
  z <- str_to_lower(str_replace_all(normalize_missing(x), "[^a-zA-Z0-9]", ""))
  case_when(
    z %in% c("basal", "basallike") ~ "Basal",
    z %in% c("her2", "her2enriched", "her2e") ~ "Her2",
    z %in% c("luma", "luminala") ~ "LumA",
    z %in% c("lumb", "luminalb") ~ "LumB",
    z %in% c("normal", "normallike") ~ "Normal",
    TRUE ~ NA_character_
  )
}

clean_grade <- function(x) {
  z <- str_to_lower(normalize_missing(x))
  case_when(
    str_detect(z, "(^|[^0-9])1([^0-9]|$)") ~ "Grade 1",
    str_detect(z, "(^|[^0-9])2([^0-9]|$)") ~ "Grade 2",
    str_detect(z, "(^|[^0-9])3([^0-9]|$)") ~ "Grade 3",
    TRUE ~ NA_character_
  )
}

clean_clinical_subtype <- function(x) {
  z <- str_to_upper(normalize_missing(x))
  case_when(
    str_detect(z, "TN|TRIPLE") ~ "TNBC",
    str_detect(z, "ER") ~ "ER+",
    TRUE ~ NA_character_
  )
}

extract_number <- function(x) {
  suppressWarnings(as.numeric(str_extract(normalize_missing(x), "-?[0-9]+(?:\\.[0-9]+)?")))
}

clean_recurrence <- function(x) {
  z <- str_to_lower(normalize_missing(x))
  case_when(
    str_detect(z, "1$|yes|present|positive") ~ "Yes",
    str_detect(z, "0$|no|absent|negative") ~ "No",
    TRUE ~ NA_character_
  )
}

fmt_pct <- function(x) {
  ifelse(abs(x - round(x)) < 0.05, sprintf("%.0f", x), sprintf("%.1f", x))
}

fmt_p <- function(p) {
  case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

# Type 5 (Hazen) quartiles reproduce the previously reported cohort summaries.
median_hazen <- function(x, digits = 2L) {
  x <- suppressWarnings(as.numeric(x))
  x <- x[is.finite(x)]
  if (!length(x)) return(NA_character_)
  q <- quantile(x, c(0.25, 0.75), type = 5, names = FALSE)
  paste0(
    formatC(median(x), format = "f", digits = digits), " (",
    formatC(q[[1L]], format = "f", digits = digits), "–",
    formatC(q[[2L]], format = "f", digits = digits), ")"
  )
}

theme_cimic <- function(base_size = 9) {
  theme_classic(base_size = base_size) +
    theme(
      axis.text = element_text(color = "black", face = "bold"),
      axis.title = element_text(color = "black", face = "bold"),
      axis.line = element_line(color = "black", linewidth = 0.55),
      axis.ticks = element_line(color = "black", linewidth = 0.55),
      legend.text = element_text(color = "black", face = "bold"),
      legend.title = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.margin = margin(5.5, 8, 5.5, 5.5)
    )
}

save_figure <- function(plot, filename, width, height) {
  ggsave(
    file.path(output_dirs$figures, paste0(filename, ".pdf")),
    plot, width = width, height = height, units = "in", device = cairo_pdf
  )
  ggsave(
    file.path(output_dirs$figures, paste0(filename, ".tiff")),
    plot, width = width, height = height, units = "in",
    dpi = 600, compression = "lzw"
  )
}
```

# Read inputs and validate schemas

```{r read-inputs}
nki_raw <- fread(input_paths$nki, data.table = FALSE, check.names = FALSE)
smc_raw <- fread(input_paths$smc, data.table = FALSE, check.names = FALSE)
expr_raw <- fread(input_paths$expression, data.table = FALSE, check.names = FALSE)
clusters_raw <- fread(input_paths$clusters, data.table = FALSE, check.names = FALSE)

required_nki <- c(
  "patient_id", "sample_id", "timepoint", "Stage", "NAC_Response",
  "Recurrence", "clinical_subtype_ihc", "chemotherapy", "tumor_grade",
  "node_status_bin", "mib1_pre", "os_event", "rfs_event"
)
required_smc <- c(
  "sample_id", "timepoint", "cluster_match_id", "age", "histological_grade",
  "nuclear_grade", "tnm_stage_clinical", "lymphovascular_invasion",
  "clinical_subtype_ihc", "subtype_pam50", "til_density",
  "proliferative_index", "cyt_score", "chemotherapy", "node_status_bin"
)
required_clusters <- c("sample_id", "cluster_assignments", "cohort")

stopifnot(all(required_nki %in% names(nki_raw)))
stopifnot(all(required_smc %in% names(smc_raw)))
stopifnot(all(required_clusters %in% names(clusters_raw)))
stopifnot("gene_id" %in% names(expr_raw))

cluster_map <- clusters_raw %>%
  transmute(
    cohort = str_to_upper(str_trim(cohort)),
    cluster_source_id = as.character(sample_id),
    cluster_number = as.integer(cluster_assignments),
    cluster = case_when(
      cluster_number == 1L ~ "Dys-CIM",
      cluster_number == 2L ~ "Fun-CIM",
      TRUE ~ NA_character_
    )
  )

if (anyNA(cluster_map$cluster)) stop("Only cluster assignments 1 and 2 are supported.")
if (anyDuplicated(cluster_map[c("cohort", "cluster_source_id")])) {
  stop("Duplicated patient entries in nki_smc_clusters.csv.")
}
```

# Sample identifier crosswalk

## NKI identifiers

```{r nki-crosswalk}
nki_crosswalk <- nki_raw %>%
  transmute(
    cohort = "NKI",
    sample_id_original = as.character(sample_id),
    patient_id_canonical = as.character(patient_id),
    explicit_timepoint = str_to_upper(str_trim(timepoint)),
    suffix_timepoint = time_from_suffix(sample_id_original)
  ) %>%
  rowwise() %>%
  mutate(
    timepoint = validate_timepoint(
      explicit_timepoint, suffix_timepoint, NA_character_,
      sample_id_original, cohort
    ),
    sample_id_canonical = paste0(patient_id_canonical, "_", timepoint),
    cluster_source_id = patient_id_canonical
  ) %>%
  ungroup() %>%
  left_join(
    cluster_map %>% filter(cohort == "NKI") %>% dplyr::select(cluster_source_id, cluster),
    by = "cluster_source_id"
  )

if (anyNA(nki_crosswalk$cluster)) stop("At least one NKI clinical sample did not match the cluster file.")
if (anyDuplicated(nki_crosswalk[c("patient_id_canonical", "timepoint")])) {
  stop("Duplicated NKI patient/timepoint records detected.")
}
```

## SMC identifiers

The stable SMC patient identifier is derived from `cluster_match_id`. For
example, `15_0139_B_Rep1` and `15_0139_S_Rep1` both map to patient `15_0139`.
The cluster file identifier `OB_0139` is joined through a validated four-digit
case suffix. This is safe here only because the suffix is unique in the supplied
16-patient SMC cohort; uniqueness is tested before joining.

```{r smc-crosswalk}
smc_cluster_key <- cluster_map %>%
  filter(cohort == "SMC") %>%
  mutate(case_suffix4 = str_extract(cluster_source_id, "[0-9]{4}$"))

if (anyNA(smc_cluster_key$case_suffix4) || anyDuplicated(smc_cluster_key$case_suffix4)) {
  stop("SMC four-digit case suffixes are missing or not unique in the cluster file.")
}

smc_crosswalk <- smc_raw %>%
  transmute(
    cohort = "SMC",
    sample_id_original = as.character(sample_id),
    cluster_match_id_original = as.character(cluster_match_id),
    explicit_timepoint = str_to_upper(str_trim(timepoint)),
    suffix_timepoint = time_from_suffix(cluster_match_id_original),
    full_id_timepoint = time_from_suffix(sample_id_original),
    legacy_timepoint = smc_stage_from_id(str_remove(sample_id_original, "_[BS](?:_Rep\\d+)?$")),
    patient_id_canonical = str_remove(cluster_match_id_original, "_[BS](?:_Rep\\d+)?$"),
    case_suffix4 = str_extract(patient_id_canonical, "[0-9]{4}$")
  ) %>%
  rowwise() %>%
  mutate(
    timepoint = validate_timepoint(
      explicit_timepoint,
      combine_suffix_timepoints(
        suffix_timepoint, full_id_timepoint, sample_id_original
      ),
      legacy_timepoint,
      sample_id_original,
      cohort
    ),
    sample_id_canonical = paste0(patient_id_canonical, "_", timepoint)
  ) %>%
  ungroup() %>%
  left_join(
    smc_cluster_key %>% dplyr::select(case_suffix4, cluster_source_id, cluster),
    by = "case_suffix4"
  )

if (anyNA(smc_crosswalk$cluster)) stop("At least one SMC sample did not match the cluster file.")
if (anyDuplicated(smc_crosswalk[c("patient_id_canonical", "timepoint")])) {
  stop("Duplicated SMC patient/timepoint records detected.")
}

sample_crosswalk <- bind_rows(
  nki_crosswalk %>%
    dplyr::select(cohort, sample_id_original, sample_id_canonical,
           patient_id_canonical, timepoint, cluster_source_id, cluster),
  smc_crosswalk %>%
    dplyr::select(cohort, sample_id_original, sample_id_canonical,
           patient_id_canonical, timepoint, cluster_source_id, cluster)
) %>%
  arrange(cohort, patient_id_canonical, timepoint)

write.csv(
  sample_crosswalk,
  file.path(output_dirs$qc, "sample_ID_crosswalk.csv"),
  row.names = FALSE,
  na = ""
)

crosswalk_counts <- sample_crosswalk %>%
  count(cohort, timepoint) %>%
  pivot_wider(names_from = timepoint, values_from = n, values_fill = 0)

stopifnot(
  n_distinct(nki_crosswalk$patient_id_canonical) == 20L,
  n_distinct(smc_crosswalk$patient_id_canonical) == 16L,
  all(crosswalk_counts$B == c(20L, 16L)),
  all(crosswalk_counts$S == c(20L, 16L))
)

kable(crosswalk_counts, caption = "Validated biopsy and surgery sample counts")
```

# NKI PAM50 classification

PAM50 is calculated only for NKI samples. The analysis uses all NKI biopsy and
surgery samples together for robust gene-wise scaling, while retaining a
separate prediction for every sample. Gene symbols are mapped to Entrez IDs for
the `genefu` PAM50 centroids.

```{r pam50-preparation}
expression_columns <- setdiff(names(expr_raw), "gene_id")

expression_manifest <- tibble(expression_sample_id = expression_columns) %>%
  mutate(
    expression_sample_clean = str_remove(expression_sample_id, "_Rep\\d+$"),
    timepoint = time_from_suffix(expression_sample_clean),
    patient_id_canonical = str_remove(expression_sample_clean, "_[BS]$"),
    sample_id_canonical = paste0(patient_id_canonical, "_", timepoint)
  )

nki_expression_manifest <- expression_manifest %>%
  semi_join(
    nki_crosswalk %>% dplyr::select(sample_id_canonical),
    by = "sample_id_canonical"
  ) %>%
  left_join(
    nki_crosswalk %>%
      dplyr::select(sample_id_canonical, patient_id_canonical, timepoint, cluster),
    by = "sample_id_canonical",
    suffix = c("_expression", "")
  )

if (nrow(nki_expression_manifest) != 40L) {
  stop("Expected 40 matched NKI expression samples; found ", nrow(nki_expression_manifest), ".")
}
if (anyDuplicated(nki_expression_manifest$sample_id_canonical)) {
  stop("Duplicated canonical NKI expression sample IDs detected.")
}

nki_expr <- expr_raw %>%
  dplyr::select(gene_id, all_of(nki_expression_manifest$expression_sample_id)) %>%
  mutate(
    gene_id = str_trim(as.character(gene_id)),
    across(-gene_id, ~ suppressWarnings(as.numeric(.x)))
  ) %>%
  filter(!is.na(gene_id), gene_id != "")

duplicate_symbols <- nki_expr %>% count(gene_id) %>% filter(n > 1L)
write.csv(
  duplicate_symbols,
  file.path(output_dirs$qc, "NKI_duplicate_gene_symbols.csv"),
  row.names = FALSE
)

if (nrow(duplicate_symbols) > 0L) {
  nki_expr <- nki_expr %>%
    group_by(gene_id) %>%
    summarise(across(everything(), ~ median(.x, na.rm = TRUE)), .groups = "drop")
}

symbol_to_entrez <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = unique(nki_expr$gene_id),
  keytype = "SYMBOL",
  column = "ENTREZID",
  multiVals = "first"
)

expression_annotation <- tibble(
  probe = names(symbol_to_entrez),
  EntrezGene.ID = unname(as.character(symbol_to_entrez))
) %>%
  filter(!is.na(EntrezGene.ID), EntrezGene.ID != "")

nki_expr_mapped <- nki_expr %>%
  semi_join(expression_annotation, by = c("gene_id" = "probe"))

expr_matrix <- nki_expr_mapped %>%
  column_to_rownames("gene_id") %>%
  as.matrix()
storage.mode(expr_matrix) <- "numeric"

# genefu requires samples in rows and genes/probes in columns.
expr_genefu <- t(expr_matrix)

annot_genefu <- expression_annotation %>%
  filter(probe %in% colnames(expr_genefu)) %>%
  distinct(probe, .keep_all = TRUE) %>%
  as.data.frame()
rownames(annot_genefu) <- annot_genefu$probe
annot_genefu <- annot_genefu[colnames(expr_genefu), , drop = FALSE]

stopifnot(identical(colnames(expr_genefu), rownames(annot_genefu)))
```

```{r run-pam50}
data("pam50.robust", package = "genefu")

pam50_fit <- genefu::intrinsic.cluster.predict(
  sbt.model = pam50.robust,
  data = expr_genefu,
  annot = annot_genefu,
  do.mapping = TRUE,
  do.prediction.strength = FALSE,
  verbose = TRUE
)

pam50_mapped_genes <- as.data.frame(pam50_fit$centroids.map)
pam50_gene_n <- nrow(pam50_mapped_genes)

write.csv(
  pam50_mapped_genes,
  file.path(output_dirs$qc, "NKI_PAM50_mapped_genes.csv"),
  row.names = FALSE
)

if (pam50_gene_n < 45L) {
  stop(
    "Only ", pam50_gene_n,
    " PAM50 genes mapped. Review NKI_PAM50_mapped_genes.csv before interpretation."
  )
}

pam50_correlations <- as.data.frame(pam50_fit$cor) %>%
  rownames_to_column("expression_sample_id") %>%
  rename_with(~ paste0("pam50_cor_", .x), -expression_sample_id)

pam50_predictions <- tibble(
  expression_sample_id = names(pam50_fit$subtype),
  pam50_subtype = clean_pam50(unname(pam50_fit$subtype))
) %>%
  left_join(pam50_correlations, by = "expression_sample_id") %>%
  left_join(
    nki_expression_manifest %>%
      dplyr::select(expression_sample_id, sample_id_canonical,
             patient_id_canonical, timepoint, cluster),
    by = "expression_sample_id"
  ) %>%
  relocate(patient_id_canonical, timepoint, sample_id_canonical,
           expression_sample_id, cluster, pam50_subtype)

if (anyNA(pam50_predictions$pam50_subtype)) {
  stop("At least one NKI PAM50 prediction is missing.")
}

write.csv(
  pam50_predictions,
  file.path(output_dirs$tables, "NKI_PAM50_predictions.csv"),
  row.names = FALSE,
  na = ""
)

kable(
  pam50_predictions %>% count(timepoint, pam50_subtype),
  caption = paste0("NKI PAM50 predictions; ", pam50_gene_n, " PAM50 genes mapped")
)
```

# Add PAM50 to a new NKI master

```{r augment-nki-master}
nki_augmented <- nki_raw %>%
  left_join(
    nki_crosswalk %>%
      dplyr::transmute(
        sample_id_original,
        patient_id_canonical,
        timepoint_validated = timepoint,
        sample_id_canonical,
        cluster
      ),
    by = c("sample_id" = "sample_id_original")
  ) %>%
  # Preserve the original clinical timepoint for provenance, but use the
  # crosswalk-validated value for all downstream matching.
  mutate(
    timepoint_original = timepoint,
    timepoint = timepoint_validated
  ) %>%
  dplyr::select(-timepoint_validated) %>%
  left_join(
    pam50_predictions %>%
      dplyr::select(patient_id_canonical, timepoint, pam50_subtype,
             starts_with("pam50_cor_")),
    by = c("patient_id_canonical", "timepoint")
  )

if (anyNA(nki_augmented$pam50_subtype)) {
  stop("PAM50 did not merge into every NKI clinical row.")
}

nki_augmented_path <- file.path(data_dir, "nki_master_with_genefu_PAM50.csv")
fwrite(nki_augmented, nki_augmented_path, na = "")
```

# Construct one-row-per-patient datasets

## NKI patient dataset

```{r nki-patient-data}
nki_long <- nki_augmented %>%
  mutate(
    tumor_grade_clean = clean_grade(tumor_grade),
    tumor_stage_clean = as.character(extract_number(Stage)),
    clinical_subtype_clean = clean_clinical_subtype(clinical_subtype_ihc),
    nri_value = extract_number(str_extract(NAC_Response, "[0-9]+(?:\\.[0-9]+)?$")),
    recurrence_clean = clean_recurrence(Recurrence),
    chemotherapy_clean = str_squish(chemotherapy),
    node_status_clean = normalize_missing(node_status_bin),
    mib1_pre_numeric = suppressWarnings(as.numeric(mib1_pre))
  )

nki_static <- nki_long %>%
  group_by(patient_id_canonical) %>%
  summarise(
    cohort = "NKI",
    cluster = first_unique(cluster, "cluster", cur_group_id()),
    tumor_grade = first_unique(tumor_grade_clean, "tumor grade", cur_group_id()),
    tumor_stage = first_unique(tumor_stage_clean, "tumor stage", cur_group_id()),
    clinical_subtype_ihc = first_unique(clinical_subtype_clean, "clinical subtype", cur_group_id()),
    NRI_value = suppressWarnings(as.numeric(first_unique(nri_value, "NRI", cur_group_id()))),
    recurrence = first_unique(recurrence_clean, "recurrence", cur_group_id()),
    chemotherapy = first_unique(chemotherapy_clean, "chemotherapy", cur_group_id()),
    node_status = first_unique(node_status_clean, "node status", cur_group_id()),
    mib1_pre = suppressWarnings(as.numeric(first_unique(mib1_pre_numeric, "MIB1", cur_group_id()))),
    os_event = suppressWarnings(as.integer(first_unique(os_event, "OS event", cur_group_id()))),
    rfs_event = suppressWarnings(as.integer(first_unique(rfs_event, "RFS event", cur_group_id()))),
    .groups = "drop"
  )

nki_timepoint <- nki_long %>%
  dplyr::select(patient_id_canonical, timepoint, pam50_subtype) %>%
  pivot_wider(names_from = timepoint, values_from = pam50_subtype,
              names_prefix = "pam50_")

nki_patient <- nki_static %>%
  left_join(nki_timepoint, by = "patient_id_canonical")

stopifnot(nrow(nki_patient) == 20L, !anyNA(nki_patient$pam50_B), !anyNA(nki_patient$pam50_S))
```

## SMC patient dataset

```{r smc-patient-data}
smc_long <- smc_raw %>%
  left_join(
    smc_crosswalk %>%
      dplyr::select(sample_id_original, patient_id_canonical, timepoint,
             sample_id_canonical, cluster),
    by = c("sample_id" = "sample_id_original"),
    suffix = c("_source", "")
  ) %>%
  mutate(
    tumor_grade_clean = clean_grade(histological_grade),
    nuclear_grade_clean = clean_grade(nuclear_grade),
    clinical_subtype_clean = clean_clinical_subtype(clinical_subtype_ihc),
    pam50_clean = clean_pam50(subtype_pam50),
    chemotherapy_clean = str_squish(chemotherapy),
    node_status_clean = normalize_missing(node_status_bin)
  )

smc_static <- smc_long %>%
  group_by(patient_id_canonical) %>%
  summarise(
    cohort = "SMC",
    cluster = first_unique(cluster, "cluster", cur_group_id()),
    age = suppressWarnings(as.numeric(first_unique(age, "age", cur_group_id()))),
    tumor_grade = first_unique(tumor_grade_clean, "histological grade", cur_group_id()),
    nuclear_grade = first_unique(nuclear_grade_clean, "nuclear grade", cur_group_id()),
    tumor_stage = first_unique(tnm_stage_clinical, "clinical TNM stage", cur_group_id()),
    lymphovascular_invasion = first_unique(lymphovascular_invasion, "LVI", cur_group_id()),
    clinical_subtype_ihc = first_unique(clinical_subtype_clean, "clinical subtype", cur_group_id()),
    chemotherapy = first_unique(chemotherapy_clean, "chemotherapy", cur_group_id()),
    node_status = first_unique(node_status_clean, "node status", cur_group_id()),
    .groups = "drop"
  )

smc_timepoint <- smc_long %>%
  dplyr::select(
    patient_id_canonical, timepoint, pam50_clean, til_density,
    proliferative_index, cyt_score
  ) %>%
  pivot_wider(
    names_from = timepoint,
    values_from = c(pam50_clean, til_density, proliferative_index, cyt_score),
    names_glue = "{.value}_{timepoint}"
  )

required_smc_wide_columns <- c(
  "pam50_clean_B", "pam50_clean_S",
  "til_density_B", "til_density_S",
  "proliferative_index_B", "proliferative_index_S",
  "cyt_score_B", "cyt_score_S"
)

missing_smc_wide_columns <- setdiff(required_smc_wide_columns, names(smc_timepoint))
if (length(missing_smc_wide_columns) > 0L) {
  stop(
    "SMC biopsy/surgery pivot did not create the expected columns: ",
    paste(missing_smc_wide_columns, collapse = ", "),
    ". Check the validated timepoint column before continuing."
  )
}

smc_timepoint <- smc_timepoint %>%
  dplyr::rename(
    pam50_B = pam50_clean_B,
    pam50_S = pam50_clean_S
  )

smc_patient <- smc_static %>%
  left_join(smc_timepoint, by = "patient_id_canonical")

stopifnot(nrow(smc_patient) == 16L, !anyNA(smc_patient$pam50_B), !anyNA(smc_patient$pam50_S))
```

## Combined patient dataset

```{r combined-patient-data}
common_variables <- c(
  "patient_id_canonical", "cohort", "cluster", "tumor_grade",
  "clinical_subtype_ihc", "chemotherapy", "node_status", "pam50_B", "pam50_S"
)

clinical_patient <- bind_rows(
  nki_patient %>% dplyr::select(all_of(common_variables)),
  smc_patient %>% dplyr::select(all_of(common_variables))
) %>%
  mutate(
    cluster = factor(cluster, levels = cluster_levels),
    tumor_grade = factor(tumor_grade, levels = grade_levels),
    pam50_B = factor(pam50_B, levels = pam50_levels),
    pam50_S = factor(pam50_S, levels = pam50_levels),
    clinical_subtype_ihc = factor(clinical_subtype_ihc, levels = c("ER+", "TNBC"))
  )

stopifnot(
  nrow(clinical_patient) == 36L,
  n_distinct(clinical_patient$patient_id_canonical) == 36L,
  sum(clinical_patient$cluster == "Fun-CIM") == 25L,
  sum(clinical_patient$cluster == "Dys-CIM") == 11L
)

write.csv(
  clinical_patient,
  file.path(output_dirs$tables, "patient_level_clinical_data.csv"),
  row.names = FALSE,
  na = ""
)

kable(
  clinical_patient %>% count(cohort, cluster),
  caption = "Validated patient counts by cohort and CIM state"
)
```

# Descriptive table functions

```{r table-helpers}
categorical_rows <- function(df, variable, label, include_missing = FALSE,
                             level_order = NULL) {
  x <- as.character(df[[variable]])
  if (include_missing) x[is.na(x) | x == ""] <- "Missing"
  denominator <- if (include_missing) length(x) else sum(!is.na(x) & x != "")
  x <- x[!is.na(x) & x != ""]

  out <- tibble(Level = x) %>%
    count(Level, name = "n") %>%
    mutate(
      denominator = denominator,
      percent = 100 * n / denominator,
      Value = paste0(n, " (", fmt_pct(percent), "%)")
    )

  if (!is.null(level_order)) {
    out <- out %>%
      mutate(Level = factor(Level, levels = level_order)) %>%
      arrange(Level) %>%
      mutate(Level = as.character(Level))
  }

  out %>% transmute(Variable = label, Level, Value, Available_N = denominator)
}

continuous_row <- function(df, variable, label) {
  x <- suppressWarnings(as.numeric(df[[variable]]))
  tibble(
    Variable = label,
    Level = "Median (Q1–Q3)",
    Value = median_hazen(x),
    Available_N = sum(is.finite(x))
  )
}

event_row <- function(df, variable, label, event = "Yes") {
  x <- as.character(df[[variable]])
  denominator <- sum(!is.na(x))
  n <- sum(x == event, na.rm = TRUE)
  tibble(
    Variable = label,
    Level = event,
    Value = paste0(n, " (", fmt_pct(100 * n / denominator), "%)"),
    Available_N = denominator
  )
}
```

# Table 1: NKI patient demographics

```{r table-1-nki}
table_1_nki <- bind_rows(
  categorical_rows(nki_patient, "tumor_grade", "Tumor Grade, n (%)", TRUE,
                   c(grade_levels, "Missing")),
  categorical_rows(nki_patient, "tumor_stage", "Tumor Stage, n (%)", FALSE,
                   c("1", "2", "3", "4")),
  categorical_rows(nki_patient, "clinical_subtype_ihc", "Clinical Subtype, n (%)", FALSE,
                   c("ER+", "TNBC")),
  categorical_rows(nki_patient, "pam50_B", "PAM50 Subtype – Biopsy, n (%)", FALSE,
                   pam50_levels),
  categorical_rows(nki_patient, "pam50_S", "PAM50 Subtype – Surgery, n (%)", FALSE,
                   pam50_levels),
  categorical_rows(nki_patient, "NRI_value", "NRI Value, n (%)"),
  event_row(nki_patient, "recurrence", "Recurrence, n (%)"),
  categorical_rows(nki_patient, "chemotherapy", "Chemotherapy Regimen, n (%)"),
  categorical_rows(nki_patient, "node_status", "Node Status, n (%)", FALSE, c("N0", "N+")),
  continuous_row(nki_patient, "mib1_pre", "MIB1/Ki67 Pretreatment")
) %>%
  mutate(Cohort = "NKI cohort; N = 20 (100%)", .before = 1)

write.csv(
  table_1_nki,
  file.path(output_dirs$tables, "Table_1_NKI_demographics.csv"),
  row.names = FALSE,
  na = ""
)

kable(table_1_nki, caption = "Table 1. Patient demographics within the NKI cohort")
```

# Table 2: SMC patient demographics

The values previously labeled “Ki67” are generated by `proliferative_index`.
They are therefore reported here using the accurate label **Proliferative
Index**. The mixed-format `ki67_diagnosis` and `ki67_surgery` variables are not
converted without a prespecified conversion rule.

```{r table-2-smc}
table_2_smc <- bind_rows(
  continuous_row(smc_patient, "age", "Age (years)"),
  categorical_rows(smc_patient, "tumor_grade", "Histological Grade, n (%)", FALSE,
                   grade_levels),
  categorical_rows(smc_patient, "nuclear_grade", "Nuclear Grade, n (%)"),
  categorical_rows(smc_patient, "tumor_stage", "Tumor Stage, n (%)"),
  categorical_rows(smc_patient, "lymphovascular_invasion", "Lymphovascular Invasion, n (%)",
                   FALSE, c("0", "1")),
  categorical_rows(smc_patient, "clinical_subtype_ihc", "Clinical Subtype, n (%)", FALSE,
                   c("ER+", "TNBC")),
  categorical_rows(smc_patient, "pam50_B", "PAM50 Subtype – Biopsy, n (%)", FALSE,
                   pam50_levels),
  categorical_rows(smc_patient, "pam50_S", "PAM50 Subtype – Surgery, n (%)", FALSE,
                   pam50_levels),
  continuous_row(smc_patient, "til_density_B", "TIL Density – Biopsy"),
  continuous_row(smc_patient, "til_density_S", "TIL Density – Surgery"),
  continuous_row(smc_patient, "proliferative_index_B", "Proliferative Index – Biopsy"),
  continuous_row(smc_patient, "proliferative_index_S", "Proliferative Index – Surgery"),
  continuous_row(smc_patient, "cyt_score_B", "Cytolytic Score – Biopsy"),
  continuous_row(smc_patient, "cyt_score_S", "Cytolytic Score – Surgery"),
  categorical_rows(smc_patient, "chemotherapy", "Chemotherapy Regimen, n (%)"),
  categorical_rows(smc_patient, "node_status", "Node Status, n (%)", FALSE, c("N0", "N+"))
) %>%
  mutate(Cohort = "SMC cohort; N = 16 (100%)", .before = 1)

write.csv(
  table_2_smc,
  file.path(output_dirs$tables, "Table_2_SMC_demographics.csv"),
  row.names = FALSE,
  na = ""
)

kable(table_2_smc, caption = "Table 2. Patient demographics within the SMC cohort")
```

# Table 3: Combined cohort demographics

Only variables with directly harmonized definitions in both cohorts are pooled.
NKI T stage is not pooled with SMC full cTNM stage; NRI, recurrence, TIL density,
and cytolytic score remain cohort-specific.

```{r table-3-combined}
combined_variables <- tribble(
  ~variable, ~label, ~levels,
  "cluster", "CIM State, n (%)", list(cluster_levels),
  "clinical_subtype_ihc", "Clinical Subtype, n (%)", list(c("ER+", "TNBC")),
  "pam50_B", "PAM50 Subtype – Biopsy, n (%)", list(pam50_levels),
  "pam50_S", "PAM50 Subtype – Surgery, n (%)", list(pam50_levels),
  "tumor_grade", "Tumor Grade, n (%)", list(grade_levels),
  "node_status", "Node Status, n (%)", list(c("N0", "N+")),
  "chemotherapy", "Chemotherapy Regimen, n (%)", list(NULL)
)

make_combined_section <- function(variable, label, levels) {
  overall <- categorical_rows(clinical_patient, variable, label, FALSE, levels) %>%
    dplyr::select(Variable, Level, Overall = Value)
  nki <- categorical_rows(clinical_patient %>% filter(cohort == "NKI"),
                          variable, label, FALSE, levels) %>%
    dplyr::select(Level, NKI = Value)
  smc <- categorical_rows(clinical_patient %>% filter(cohort == "SMC"),
                          variable, label, FALSE, levels) %>%
    dplyr::select(Level, SMC = Value)
  full_join(overall, nki, by = "Level") %>%
    full_join(smc, by = "Level") %>%
    mutate(Variable = coalesce(Variable, label))
}

table_3_combined <- bind_rows(lapply(seq_len(nrow(combined_variables)), function(i) {
  make_combined_section(
    combined_variables$variable[[i]],
    combined_variables$label[[i]],
    combined_variables$levels[[i]]
  )
}))

write.csv(
  table_3_combined,
  file.path(output_dirs$tables, "Table_3_combined_demographics.csv"),
  row.names = FALSE,
  na = ""
)

kable(
  table_3_combined,
  caption = "Table 3. Combined demographics: overall, NKI, and SMC"
)
```

# Statistical analysis: Fun-CIM versus Dys-CIM

Categorical associations are evaluated using two-sided Fisher’s exact tests for
2 × 2 tables and Fisher–Freeman–Halton exact tests for larger tables. Exact
enumeration is attempted first. Monte Carlo estimation is used only when exact
enumeration fails computationally. The final Monte Carlo analysis uses the
configured fixed seed and 1,000,000 replicates by default.

```{r statistical-functions}
cramers_v <- function(tab) {
  chi <- suppressWarnings(chisq.test(tab, correct = FALSE)$statistic)
  n <- sum(tab)
  denom <- min(nrow(tab) - 1L, ncol(tab) - 1L)
  if (denom <= 0L) return(NA_real_)
  as.numeric(sqrt(chi / (n * denom)))
}

serialize_table <- function(tab) {
  paste(
    apply(as.matrix(tab), 1, function(x) paste(x, collapse = ",")),
    collapse = ";"
  )
}

run_categorical_test <- function(df, variable, label) {
  test_df <- df %>%
    filter(!is.na(cluster), !is.na(.data[[variable]])) %>%
    transmute(
      cluster = droplevels(factor(cluster, levels = cluster_levels)),
      value = droplevels(factor(.data[[variable]]))
    )

  tab <- table(test_df$cluster, test_df$value)
  if (nrow(tab) < 2L || ncol(tab) < 2L) stop(label, " is not testable.")

  expected <- suppressWarnings(chisq.test(tab, correct = FALSE)$expected)
  monte_carlo_used <- FALSE
  B_used <- NA_integer_
  seed_used <- NA_integer_

  if (identical(dim(tab), c(2L, 2L))) {
    fit <- fisher.test(tab, alternative = "two.sided")
    test_used <- "Two-sided Fisher's exact test"
    effect_name <- "Odds ratio"
    effect_value <- unname(fit$estimate)
    ci_low <- fit$conf.int[[1L]]
    ci_high <- fit$conf.int[[2L]]
    statistic_name <- NA_character_
    statistic_value <- NA_real_
    reason <- "A 2 x 2 table with at least one expected count below 5."
  } else {
    exact_fit <- tryCatch(
      fisher.test(tab, hybrid = FALSE, workspace = 2e7),
      error = function(e) e
    )

    if (inherits(exact_fit, "error")) {
      set.seed(params$random_seed)
      fit <- fisher.test(
        tab,
        simulate.p.value = TRUE,
        B = params$monte_carlo_B
      )
      monte_carlo_used <- TRUE
      B_used <- params$monte_carlo_B
      seed_used <- params$random_seed
      test_used <- "Fisher-Freeman-Halton test; Monte Carlo p-value"
      reason <- paste0(
        "Sparse ", nrow(tab), " x ", ncol(tab),
        " table; exact enumeration failed computationally."
      )
    } else {
      fit <- exact_fit
      test_used <- "Fisher-Freeman-Halton exact test"
      reason <- paste0(
        "Sparse ", nrow(tab), " x ", ncol(tab),
        " table; exact conditional inference was feasible."
      )
    }

    effect_name <- "Cramer's V"
    effect_value <- cramers_v(tab)
    ci_low <- NA_real_
    ci_high <- NA_real_
    statistic_name <- NA_character_
    statistic_value <- NA_real_
  }

  tibble(
    analysis_id = paste0("CIM_", variable),
    analysis_group = "Clinical characteristics",
    variable = variable,
    variable_label = label,
    comparison = "Fun-CIM versus Dys-CIM",
    total_n = nrow(test_df),
    fun_cim_n = sum(test_df$cluster == "Fun-CIM"),
    dys_cim_n = sum(test_df$cluster == "Dys-CIM"),
    missing_n = nrow(df) - nrow(test_df),
    table_dimensions = paste(nrow(tab), "x", ncol(tab)),
    category_levels = paste(colnames(tab), collapse = " | "),
    contingency_table_row_order = paste(rownames(tab), collapse = " | "),
    contingency_table_counts = serialize_table(tab),
    minimum_expected_count = min(expected),
    cells_expected_below_5_n = sum(expected < 5),
    cells_expected_below_5_percent = 100 * mean(expected < 5),
    test_used = test_used,
    test_reason = reason,
    statistic_name = statistic_name,
    statistic_value = statistic_value,
    effect_size_name = effect_name,
    effect_size_value = effect_value,
    ci_95_lower = ci_low,
    ci_95_upper = ci_high,
    p_value_raw = unname(fit$p.value),
    monte_carlo_used = if_else(monte_carlo_used, "Yes", "No"),
    monte_carlo_replicates = B_used,
    random_seed = seed_used
  )
}
```

```{r run-statistical-tests}
test_specifications <- tribble(
  ~variable, ~label,
  "cohort", "Cohort",
  "clinical_subtype_ihc", "Clinical Subtype (IHC)",
  "pam50_B", "PAM50 Subtype (Biopsy)",
  "pam50_S", "PAM50 Subtype (Surgery)",
  "tumor_grade", "Tumor Grade",
  "node_status", "Node Status",
  "chemotherapy", "Chemotherapy Regimen"
)

statistical_tests <- bind_rows(lapply(seq_len(nrow(test_specifications)), function(i) {
  run_categorical_test(
    clinical_patient,
    test_specifications$variable[[i]],
    test_specifications$label[[i]]
  )
})) %>%
  mutate(p_value_BH = p.adjust(p_value_raw, method = "BH"))

write.csv(
  statistical_tests,
  file.path(output_dirs$tables, "Table_4_statistical_tests.csv"),
  row.names = FALSE,
  na = ""
)

kable(
  statistical_tests %>%
    dplyr::select(variable_label, total_n, table_dimensions, test_used,
           effect_size_name, effect_size_value, p_value_raw, p_value_BH),
  digits = 4,
  caption = "Statistical audit for Fun-CIM versus Dys-CIM analyses"
)
```

# Table 4: Clinical characteristics of Fun-CIM and Dys-CIM tumors

P-values are displayed on the main characteristic row and refer to the entire
contingency table. Test-specific markers and notes are generated from the tests
actually used rather than hard-coded in advance.

```{r table-4}
test_marker_key <- statistical_tests %>%
  distinct(test_used, monte_carlo_replicates, random_seed) %>%
  arrange(test_used) %>%
  mutate(test_marker = seq_len(n()))

statistical_tests <- statistical_tests %>%
  left_join(test_marker_key, by = c("test_used", "monte_carlo_replicates", "random_seed"))

make_table4_section <- function(variable_name, label) {
  test_row <- statistical_tests %>%
    dplyr::filter(.data$variable == .env$variable_name)

  if (nrow(test_row) != 1L) {
    stop(
      "Could not resolve exactly one Table 4 test for ", variable_name,
      ". Matches found: ", nrow(test_row), "."
    )
  }

  counts <- clinical_patient %>%
    dplyr::filter(!is.na(cluster), !is.na(.data[[variable_name]])) %>%
    dplyr::count(cluster, Level = .data[[variable_name]], name = "n") %>%
    group_by(cluster) %>%
    mutate(
      denominator = sum(n),
      percent = 100 * n / denominator,
      display = paste0(n, " (", fmt_pct(percent), "%)")
    ) %>%
    ungroup() %>%
    dplyr::select(cluster, Level, display) %>%
    pivot_wider(names_from = cluster, values_from = display, values_fill = "0 (0%)")

  header <- tibble(
    Variable = paste0(label, " [", test_row$test_marker, "]"),
    Level = "",
    `Fun-CIM` = "",
    `Dys-CIM` = "",
    `P-value` = fmt_p(test_row$p_value_raw),
    Available_N = test_row$total_n
  )

  levels <- counts %>%
    transmute(
      Variable = "",
      Level = as.character(Level),
      `Fun-CIM`,
      `Dys-CIM`,
      `P-value` = "",
      Available_N = test_row$total_n
    )

  bind_rows(header, levels)
}

table_4_body <- bind_rows(lapply(seq_len(nrow(test_specifications)), function(i) {
  make_table4_section(
    test_specifications$variable[[i]],
    test_specifications$label[[i]]
  )
}))

fun_n <- sum(clinical_patient$cluster == "Fun-CIM")
dys_n <- sum(clinical_patient$cluster == "Dys-CIM")
total_n <- fun_n + dys_n

names(table_4_body)[names(table_4_body) == "Fun-CIM"] <- paste0(
  "Functional-CIM; N = ", fun_n, " (", fmt_pct(100 * fun_n / total_n), "%)"
)
names(table_4_body)[names(table_4_body) == "Dys-CIM"] <- paste0(
  "Dysfunctional-CIM; N = ", dys_n, " (", fmt_pct(100 * dys_n / total_n), "%)"
)

fun_col <- names(table_4_body)[str_detect(names(table_4_body), "^Functional-CIM")]
dys_col <- names(table_4_body)[str_detect(names(table_4_body), "^Dysfunctional-CIM")]

test_footnotes <- test_marker_key %>%
  transmute(
    Variable = paste0("Footnote [", test_marker, "]"),
    Level = case_when(
      str_detect(test_used, "Monte Carlo") ~ paste0(
        test_used, "; ", format(monte_carlo_replicates, scientific = FALSE),
        " replicates; seed = ", random_seed, "."
      ),
      TRUE ~ paste0(test_used, ".")
    ),
    !!fun_col := "",
    !!dys_col := "",
    `P-value` = "",
    Available_N = NA_integer_
  )

general_notes <- tibble(
  Variable = c("General note", "General note"),
  Level = c(
    "Data are n (%). Percentages use the available nonmissing denominator within each CIM group.",
    "P-values are raw, two-sided values. BH-adjusted values and complete diagnostics are in Table_4_statistical_tests.csv."
  ),
  !!fun_col := "",
  !!dys_col := "",
  `P-value` = "",
  Available_N = NA_integer_
)

table_4_export <- bind_rows(table_4_body, test_footnotes, general_notes)

write.csv(
  table_4_export,
  file.path(output_dirs$tables, "Table_4_Fun_CIM_vs_Dys_CIM.csv"),
  row.names = FALSE,
  na = ""
)

kable(
  table_4_export,
  caption = "Table 4. Clinical characteristics of Fun-CIM and Dys-CIM tumors"
)
```

# Clinical figures generation

figure aesthetics:

- 5 × 5.5 inch canvas;
- 100% stacked bars with width 0.85;
- 18-point bold axes and legends;
- 20-point bold centered titles;
- count and percentage labels within every nonzero segment;
- sample size below each bar at y = -8;
- p-value centered at y = 110;
- y-axis range from -15 to 117;
- black panel border and bold black axes;
- the original CIM, PAM50, and grade palettes.

Quality improvements are limited to explicit namespaces, modern `linewidth`
arguments, consistent p-value formatting, and vector/600-dpi export.

## Plot helpers matching the original style

```{r plot-function}
manual_plot_width <- 5
manual_plot_height <- 5.5
manual_plot_dpi <- 600

# Original manuscript colors supplied with the reference plots.
cim_plot_colors <- c("Fun-CIM" = "#4DB6AC", "Dys-CIM" = "#F08A8A")
pam50_plot_colors <- c(
  "Basal" = "#C44E52",
  "Her2" = "#DD8DA6",
  "LumA" = "#0072B2",
  "LumB" = "#5C8C45",
  "Normal" = "#9467BD"
)
grade_plot_colors <- c(
  "Grade 1" = "#FFC5C5",
  "Grade 2" = "#F08A8A",
  "Grade 3" = "#D95F5F"
)

plot_p_value <- function(variable_name) {
  value <- statistical_tests %>%
    dplyr::filter(.data$variable == .env$variable_name) %>%
    dplyr::pull(p_value_raw)
  if (length(value) != 1L) {
    stop(
      "Could not resolve exactly one p-value for ", variable_name,
      ". Matches found: ", length(value), "."
    )
  }
  value
}

format_plot_p <- function(p) {
  if (is.na(p)) return("P = NA")
  if (p < 0.001) return("P < 0.001")
  paste0("P = ", format(round(p, 4), nsmall = 4, scientific = FALSE))
}

create_stack_plot_matched_style <- function(
    df,
    outcome_var,
    title,
    colors,
    category_levels = NULL,
    x_title = "") {

  clean_data <- df %>%
    dplyr::select(cluster, dplyr::all_of(outcome_var)) %>%
    dplyr::filter(
      !is.na(cluster),
      !is.na(.data[[outcome_var]]),
      as.character(.data[[outcome_var]]) != "",
      as.character(.data[[outcome_var]]) != "NA"
    ) %>%
    dplyr::mutate(
      cluster = factor(cluster, levels = cluster_levels),
      category = as.character(.data[[outcome_var]])
    )

  if (!is.null(category_levels)) {
    clean_data <- clean_data %>%
      dplyr::mutate(category = factor(category, levels = category_levels))
  }

  if (nrow(clean_data) == 0L) return(NULL)

  tab_n <- table(clean_data$cluster, clean_data$category)
  if (nrow(tab_n) < 2L || ncol(tab_n) < 2L) return(NULL)

  plot_data <- as.data.frame(prop.table(tab_n, margin = 1) * 100)
  names(plot_data) <- c("Cluster", "Category", "Percentage")

  count_data <- as.data.frame(tab_n)
  names(count_data) <- c("Cluster", "Category", "n")

  plot_data <- plot_data %>%
    dplyr::left_join(count_data, by = c("Cluster", "Category")) %>%
    dplyr::mutate(
      Percentage = round(Percentage, 1),
      label = dplyr::if_else(
        Percentage > 0,
        paste0(n, " (", fmt_pct(Percentage), "%)"),
        ""
      ),
      Cluster = factor(Cluster, levels = cluster_levels),
      Category = if (is.null(category_levels)) {
        factor(Category)
      } else {
        factor(Category, levels = category_levels)
      }
    )

  labels_df <- clean_data %>%
    dplyr::count(cluster, name = "N") %>%
    dplyr::transmute(
      x = factor(cluster, levels = cluster_levels),
      y = -8,
      label = paste0("N = ", N)
    )

  p_lab <- format_plot_p(plot_p_value(outcome_var))

  ggplot(plot_data, aes(x = Cluster, y = Percentage, fill = Category)) +
    geom_col(width = 0.85) +
    geom_text(
      aes(label = label),
      position = position_stack(vjust = 0.5),
      size = 5,
      color = "black",
      fontface = "bold"
    ) +
    geom_text(
      data = labels_df,
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE,
      size = 5,
      color = "black",
      fontface = "bold"
    ) +
    annotate(
      "text", x = 1.5, y = 110, label = p_lab,
      size = 5, fontface = "bold"
    ) +
    scale_fill_manual(values = colors, drop = FALSE) +
    scale_y_continuous(
      limits = c(-15, 117),
      breaks = seq(0, 100, 20),
      expand = expansion(mult = c(0, 0))
    ) +
    scale_x_discrete(expand = expansion(mult = c(0.55, 0.55))) +
    labs(x = x_title, y = "Percentage (%)", fill = NULL, title = title) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
      axis.text = element_text(color = "black", size = 18, face = "bold"),
      axis.title = element_text(color = "black", size = 18, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.ticks = element_line(linewidth = 1.5),
      legend.text = element_text(color = "black", size = 18, face = "bold"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      legend.position = "right",
      legend.direction = "vertical"
    ) +
    guides(fill = guide_legend(
      title = NULL,
      keywidth = grid::unit(1, "cm"),
      keyheight = grid::unit(0.8, "cm")
    ))
}

create_cim_by_clinical_subtype_plot <- function(
    df,
    title = "CIM Cluster Distribution by Clinical Subtype - Biopsy") {

  clean_data <- df %>%
    dplyr::filter(
      !is.na(cluster),
      !is.na(clinical_subtype_ihc),
      cluster %in% cluster_levels
    ) %>%
    dplyr::mutate(
      Clinical_Subtype = factor(clinical_subtype_ihc, levels = c("ER+", "TNBC")),
      Cluster = factor(cluster, levels = cluster_levels)
    )

  tab_n <- table(clean_data$Clinical_Subtype, clean_data$Cluster)
  if (nrow(tab_n) < 2L || ncol(tab_n) < 2L) return(NULL)

  plot_data <- as.data.frame(prop.table(tab_n, margin = 1) * 100)
  names(plot_data) <- c("Clinical_Subtype", "Cluster", "Percentage")

  count_data <- as.data.frame(tab_n)
  names(count_data) <- c("Clinical_Subtype", "Cluster", "n")

  plot_data <- plot_data %>%
    dplyr::left_join(count_data, by = c("Clinical_Subtype", "Cluster")) %>%
    dplyr::mutate(
      Percentage = round(Percentage, 1),
      label = dplyr::if_else(
        Percentage > 0,
        paste0(n, " (", fmt_pct(Percentage), "%)"),
        ""
      ),
      Clinical_Subtype = factor(Clinical_Subtype, levels = c("ER+", "TNBC")),
      Cluster = factor(Cluster, levels = cluster_levels)
    )

  labels_df <- clean_data %>%
    dplyr::count(Clinical_Subtype, name = "N") %>%
    dplyr::transmute(
      x = Clinical_Subtype,
      y = -8,
      label = paste0("N = ", N)
    )

  p_lab <- format_plot_p(plot_p_value("clinical_subtype_ihc"))

  ggplot(
    plot_data,
    aes(x = Clinical_Subtype, y = Percentage, fill = Cluster)
  ) +
    geom_col(width = 0.85) +
    geom_text(
      aes(label = label),
      position = position_stack(vjust = 0.5),
      size = 5,
      color = "black",
      fontface = "bold"
    ) +
    geom_text(
      data = labels_df,
      aes(x = x, y = y, label = label),
      inherit.aes = FALSE,
      size = 5,
      color = "black",
      fontface = "bold"
    ) +
    annotate(
      "text", x = 1.5, y = 110, label = p_lab,
      size = 5, fontface = "bold"
    ) +
    scale_fill_manual(values = cim_plot_colors, drop = FALSE) +
    scale_y_continuous(
      limits = c(-15, 117),
      breaks = seq(0, 100, 20),
      expand = expansion(mult = c(0, 0))
    ) +
    scale_x_discrete(expand = expansion(mult = c(0.55, 0.55))) +
    labs(
      x = "Clinical Subtype", y = "Percentage (%)",
      fill = NULL, title = title
    ) +
    theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
      axis.text = element_text(color = "black", size = 18, face = "bold"),
      axis.title = element_text(color = "black", size = 18, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 10)),
      axis.ticks = element_line(linewidth = 1.5),
      legend.text = element_text(color = "black", size = 18, face = "bold"),
      panel.border = element_rect(color = "black", fill = NA, linewidth = 1),
      legend.position = "top",
      legend.direction = "horizontal",
      legend.margin = margin(t = 0, b = 5),
      legend.box.margin = margin(0)
    ) +
    guides(fill = guide_legend(
      title = NULL,
      nrow = 1,
      keywidth = grid::unit(1, "cm"),
      keyheight = grid::unit(0.8, "cm"),
      override.aes = list(color = "black", linewidth = 1.2)
    ))
}

save_manual_plot <- function(plot, filename,
                             width = manual_plot_width,
                             height = manual_plot_height) {
  if (is.null(plot)) return(invisible(NULL))

  ggsave(
    file.path(output_dirs$figures, paste0(filename, ".pdf")),
    plot = plot, width = width, height = height,
    units = "in", device = cairo_pdf
  )
  ggsave(
    file.path(output_dirs$figures, paste0(filename, ".tiff")),
    plot = plot, width = width, height = height,
    units = "in", dpi = manual_plot_dpi, compression = "lzw"
  )
  ggsave(
    file.path(output_dirs$figures, paste0(filename, ".png")),
    plot = plot, width = width, height = height,
    units = "in", dpi = manual_plot_dpi, bg = "white"
  )

  invisible(plot)
}
```

## Dedicated manual plot creation

This chunk creates named plot objects without automatically saving them. Run
the individual assignment and `print()` lines interactively when only one plot
needs to be regenerated.

```{r create-manual-plots}
# CIM distribution within ER+ and TNBC (the supplied reversed-orientation plot).
p_cim_by_ihc_biopsy <- create_cim_by_clinical_subtype_plot(
  clinical_patient,
  "CIM Cluster Distribution by Clinical Subtype - Biopsy"
)

# Clinical subtype composition within each CIM group.
p_ihc_by_cim <- create_stack_plot_matched_style(
  clinical_patient,
  outcome_var = "clinical_subtype_ihc",
  title = "Clinical Subtype (IHC) - Biopsy",
  colors = c("ER+" = "#0072B2", "TNBC" = "#C44E52"),
  category_levels = c("ER+", "TNBC")
)

p_pam50_biopsy <- create_stack_plot_matched_style(
  clinical_patient,
  outcome_var = "pam50_B",
  title = "PAM50 Subtype - Biopsy",
  colors = pam50_plot_colors,
  category_levels = pam50_levels
)

p_pam50_surgery <- create_stack_plot_matched_style(
  clinical_patient,
  outcome_var = "pam50_S",
  title = "PAM50 Subtype - Surgery",
  colors = pam50_plot_colors,
  category_levels = pam50_levels
)

p_tumor_grade <- create_stack_plot_matched_style(
  clinical_patient,
  outcome_var = "tumor_grade",
  title = "Tumor Grade",
  colors = grade_plot_colors,
  category_levels = grade_levels
)

p_cohort <- create_stack_plot_matched_style(
  clinical_patient,
  outcome_var = "cohort",
  title = "Cohort Distribution Across Clusters",
  colors = c("NKI" = "#E69F00", "SMC" = "#56B4E9"),
  category_levels = c("NKI", "SMC")
)
```

## Display plots manually

```{r display-manual-plots, fig.show='hold'}
print(p_cim_by_ihc_biopsy)
print(p_ihc_by_cim)
print(p_pam50_biopsy)
print(p_pam50_surgery)
print(p_tumor_grade)
print(p_cohort)
```

## Save all plots at high quality

Comment out any line for a plot that should not be exported. Each call produces
PDF, 600-dpi TIFF, and 600-dpi PNG versions at 5 × 5.5 inches.

```{r save-manual-plots}
save_manual_plot(p_cim_by_ihc_biopsy, "cim_by_clinical_subtype_biopsy")
save_manual_plot(p_ihc_by_cim, "clinical_subtype_ihc_by_CIM")
save_manual_plot(p_pam50_biopsy, "PAM50_biopsy_by_CIM")
save_manual_plot(p_pam50_surgery, "PAM50_surgery_by_CIM")
save_manual_plot(p_tumor_grade, "tumor_grade_by_CIM")
save_manual_plot(p_cohort, "cohort_by_CIM")
```

# Methods summary

Research-use PAM50 intrinsic subtypes were predicted for NKI biopsy and surgery
samples from log2(TPM + 1)-transformed bulk RNA-sequencing expression using the
robust PAM50 model implemented in the `genefu` package. HGNC gene symbols were
mapped to Entrez Gene identifiers using `org.Hs.eg.db`. Predictions were
generated jointly across the NKI biopsy and surgery samples while retaining a
separate classification and centroid correlation for every sample. PAM50 calls
were merged with clinical records using a validated canonical patient ID and
the explicit biopsy (`B`) or surgery (`S`) timepoint.

Categorical characteristics are reported as counts and percentages. Percentages
in the Fun-CIM versus Dys-CIM analysis use the nonmissing denominator within
each CIM group. Continuous cohort characteristics are reported as the median
and first and third quartiles; quartiles were calculated using the Hazen/type-5
estimator to reproduce the established summaries. Associations between CIM
state and 2 × 2 categorical characteristics were evaluated using two-sided
Fisher’s exact tests. Larger categorical tables were evaluated using the
Fisher–Freeman–Halton exact test. Exact enumeration was attempted first; when
it was computationally infeasible, the conditional p-value was estimated by
Monte Carlo simulation using `r format(params$monte_carlo_B, scientific = FALSE)`
replicates and seed `r params$random_seed`. Raw two-sided p-values are displayed
in the clinical-characteristics table. Benjamini–Hochberg-adjusted p-values are
provided in the separate statistical-audit CSV. Odds ratios with exact 95%
confidence intervals are reported for 2 × 2 tables, and Cramér’s V is reported
for larger tables.

# Output manifest

```{r output-manifest}
output_manifest <- tibble(
  Output = c(
    "Augmented NKI master",
    "Sample ID crosswalk",
    "NKI PAM50 predictions",
    "Patient-level combined dataset",
    "NKI demographics",
    "SMC demographics",
    "Combined demographics",
    "Fun-CIM versus Dys-CIM table",
    "Statistical audit",
    "Figures"
  ),
  Location = c(
    nki_augmented_path,
    file.path(output_dirs$qc, "sample_ID_crosswalk.csv"),
    file.path(output_dirs$tables, "NKI_PAM50_predictions.csv"),
    file.path(output_dirs$tables, "patient_level_clinical_data.csv"),
    file.path(output_dirs$tables, "Table_1_NKI_demographics.csv"),
    file.path(output_dirs$tables, "Table_2_SMC_demographics.csv"),
    file.path(output_dirs$tables, "Table_3_combined_demographics.csv"),
    file.path(output_dirs$tables, "Table_4_Fun_CIM_vs_Dys_CIM.csv"),
    file.path(output_dirs$tables, "Table_4_statistical_tests.csv"),
    output_dirs$figures
  )
)

write.csv(
  output_manifest,
  file.path(results_dir, "output_manifest.csv"),
  row.names = FALSE
)

kable(output_manifest)
```

# Reproducibility

```{r session-info}
writeLines(capture.output(sessionInfo()), file.path(results_dir, "sessionInfo.txt"))
sessionInfo()
```
