---
title: "NEO Immune Deconvolution Pipeline"
author: "Gbadamosi Lab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
params:
  project_root: ""
  dataset_subdir: "Datasets/NEO"
  results_subdir: "Results/neo_immunodeconv_res"
  expression_file: "neo_tpm.csv"
  cluster_file: "neo_clusters.csv"
  cibersortx_file: "CIBERSORTx_NEO_rel1.txt"
  run_immunedeconv: true
  run_cibersortx_import: true
  make_plots: true
  make_heatmaps: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  error = FALSE,
  fig.align = "center"
)
```

# Overview

This R Markdown document runs the immune-deconvolution workflow for the NEO paired neoadjuvant dataset. It follows the same analysis and output conventions used by the CIM NKI/SMC immunedeconvolution pipeline while retaining NEO-specific sample-ID and cluster handling.

The workflow loads non-log TPM expression, joins NEO cluster assignments, runs the same immunedeconv methods, optionally imports CIBERSORTx, creates original and T-cell-aggregated patient-level outputs, performs unpaired cluster comparisons at biopsy and surgery, performs paired biopsy-to-surgery analyses within cluster, and generates the corresponding aggregated plots and cluster-delta heatmap.

# Repository layout

```text
oncoimmunology_paper/
├── Datasets/
│   └── NEO/
│       ├── neo_tpm.csv
│       ├── neo_clusters.csv
│       └── CIBERSORTx_NEO_rel1.txt
├── figure_generation_scripts/
│   └── NEO_Immunedeconv_pipeline_v2.1.Rmd
└── Results/
    └── neo_immunodeconv_res/
        ├── figures/
        ├── logs/
        ├── tmp/
        └── bioc_cache/
```

No machine-specific absolute path is stored in the Rmd. The parent directory of `figure_generation_scripts/` is treated as the project root.

# Analysis pipeline

```{r neo-immune-deconvolution-pipeline}
# ==============================================================================
# IMMUNE INFILTRATION ANALYSIS PIPELINE — NEO DATASET
# immunedeconv + CIBERSORTx
#
# Updated: 08/17/2026
# ==============================================================================

# ==============================================================================
# 0) REPRODUCIBILITY + GLOBAL OPTIONS
# ==============================================================================

set.seed(2025)
options(stringsAsFactors = FALSE, scipen = 999)

# ==============================================================================
# 1) PROJECT CONFIGURATION + INPUT PATHS
# ==============================================================================

# Repository layout expected:
# oncoimmunology_paper/
# ├── Datasets/
# │   └── NEO/
# │       ├── neo_tpm.csv
# │       ├── neo_clusters.csv
# │       └── CIBERSORTx_NEO_rel1.txt
# ├── figure_generation_scripts/
# │   └── NEO_Immunedeconv_pipeline_v2.1.Rmd
# └── Results/
#     └── neo_immunodeconv_res/
#
# The Rmd derives PROJECT_DIR from its own location. No user-specific path is
# required. project_root is only an optional fallback for unusual render contexts.

get_rmd_path <- function() {
  # 1. knitr render context
  p <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NA_character_)
  if (length(p) == 1 && !is.na(p) && nzchar(p) && file.exists(p)) {
    return(normalizePath(p, winslash = "/", mustWork = TRUE))
  }

  # 2. RStudio source editor context
  if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    p <- tryCatch(rstudioapi::getSourceEditorContext()$path, error = function(e) "")
    if (nzchar(p) && file.exists(p)) {
      return(normalizePath(p, winslash = "/", mustWork = TRUE))
    }
  }

  NA_character_
}

RMD_PATH <- get_rmd_path()

if (!is.null(params$project_root) &&
    length(params$project_root) == 1 &&
    !is.na(params$project_root) &&
    nzchar(params$project_root)) {
  PROJECT_DIR <- normalizePath(params$project_root, winslash = "/", mustWork = TRUE)
} else {
  if (is.na(RMD_PATH)) {
    stop(
      "Could not determine the Rmd location automatically. ",
      "Render the Rmd from the repository or provide ",
      "params = list(project_root = '/path/to/oncoimmunology_paper').",
      call. = FALSE
    )
  }

  SCRIPT_DIR <- dirname(RMD_PATH)
  PROJECT_DIR <- normalizePath(file.path(SCRIPT_DIR, ".."), winslash = "/", mustWork = TRUE)
}

DATA_DIR <- file.path(PROJECT_DIR, params$dataset_subdir)
RESULTS_DIR <- file.path(PROJECT_DIR, params$results_subdir)

NEO_EXP_FILE <- file.path(DATA_DIR, params$expression_file)
NEO_EXPR_FILE <- NEO_EXP_FILE
NEO_CLUSTER_CSV <- file.path(DATA_DIR, params$cluster_file)
CIBERSORT_NEO_FILE <- file.path(DATA_DIR, params$cibersortx_file)

required_input_files <- c(NEO_EXP_FILE, NEO_CLUSTER_CSV)
if (isTRUE(params$run_cibersortx_import)) {
  required_input_files <- c(required_input_files, CIBERSORT_NEO_FILE)
}

missing_input_files <- required_input_files[!file.exists(required_input_files)]
if (length(missing_input_files) > 0) {
  stop(
    paste0(
      "Required input file(s) not found:\n  ",
      paste(missing_input_files, collapse = "\n  "),
      "\n\nExpected data directory: ", DATA_DIR
    ),
    call. = FALSE
  )
}

# ==============================================================================
# 2) METHODS & PARAMETERS
# ==============================================================================

METHODS_TO_RUN <- c(
  "quantiseq",
  "epic",
  "xcell",
  "mcp_counter",
  "estimate",
  "timer",
  "consensus_tme"
)

TUMOR_INDICATION <- "BRCA"

GENE_COL_CANDIDATES <- c(
  "gene", "gene_id", "hgnc_symbol", "symbol", "Gene", "GeneSymbol"
)

METHOD_ORDER <- c(
  "CIBERSORTx",
  "quantiseq",
  "epic",
  "mcp_counter",
  "estimate",
  "xcell",
  "timer",
  "consensus_tme"
)

# ==============================================================================
# 3) T CELL AGGREGATION
# ==============================================================================

AGGREGATE_XCELL_TCELLS <- TRUE
AGGREGATE_CIBERSORT_TCELLS <- TRUE
CIBERSORT_AGGREGATION_STRATEGY <- "refined"

# ==============================================================================
# 4) COLORS
# ==============================================================================

METHOD_COLORS <- c(
  "quantiseq"     = "#1F77B4",
  "epic"          = "#FF7F0E",
  "xcell"         = "#2CA02C",
  "mcp_counter"   = "#D62728",
  "estimate"      = "#9467BD",
  "timer"         = "#8C564B",
  "consensus_tme" = "#E377C2",
  "CIBERSORTx"    = "#FFD700"
)

condition_colors <- c(
  "Biopsy" = "#2166AC",
  "Surgery" = "#762A83"
)

cluster_colors <- c(
  "1" = "#E41A1C",
  "2" = "#377EB8",
  "3" = "#4DAF4A",
  "4" = "#984EA3",
  "5" = "#FF7F00",
  "6" = "#FFFF33",
  "7" = "#A65628",
  "8" = "#F781BF"
)

# ==============================================================================
# 5) DIRECTORIES + TEMP/CACHE
# ==============================================================================

DIRS <- list(
  project = PROJECT_DIR,
  data = DATA_DIR,
  results = RESULTS_DIR,
  plots = file.path(RESULTS_DIR, "figures"),
  logs = file.path(RESULTS_DIR, "logs"),
  tmp = file.path(RESULTS_DIR, "tmp"),
  cache = file.path(RESULTS_DIR, "bioc_cache")
)

invisible(lapply(
  DIRS[c("results", "plots", "logs", "tmp", "cache")],
  dir.create, recursive = TRUE, showWarnings = FALSE
))

Sys.setenv(TMPDIR = DIRS$tmp)
Sys.setenv(TEMP = DIRS$tmp)
Sys.setenv(TMP = DIRS$tmp)
options(ExperimentHub.cache = DIRS$cache)

# ==============================================================================
# 6) LOGGING
# ==============================================================================

LOG_FILE <- file.path(
  DIRS$logs,
  paste0("run_log_NEO_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".txt")
)

sink(LOG_FILE, split = TRUE)

cat("=== Immune infiltration pipeline started: NEO ===\n")
cat("Timestamp:", format(Sys.time()), "\n")
cat("TMPDIR:", Sys.getenv("TMPDIR"), "\n")
cat("ExperimentHub.cache:", getOption("ExperimentHub.cache"), "\n\n")

# ==============================================================================
# 7) PACKAGES
# ==============================================================================

cat("Installing/loading required packages...\n")

if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

cran_pkgs <- c(
  "data.table", "dplyr", "tidyr", "tibble", "stringr", "purrr",
  "ggplot2", "readr", "ComplexHeatmap", "circlize", "withr",
  "conflicted", "FSA"
)

for (p in cran_pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    if (p %in% c("ComplexHeatmap", "circlize")) {
      if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
      }
      BiocManager::install(p, ask = FALSE, update = FALSE)
    } else {
      install.packages(p, dependencies = TRUE)
    }
  }
}

bioc_pkgs <- c(
  "HDF5Array",
  "DelayedArray",
  "BiocFileCache",
  "ExperimentHub"
)

if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

for (p in bioc_pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) {
    BiocManager::install(p, ask = FALSE, update = FALSE)
  }
}

if (!requireNamespace("immunedeconv", quietly = TRUE)) {
  cat("Installing immunedeconv from GitHub...\n")
  remotes::install_github("omnideconv/immunedeconv")
}

suppressPackageStartupMessages({
  library(data.table)
  library(immunedeconv)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(readr)
  library(ComplexHeatmap)
  library(circlize)
  library(withr)
  library(conflicted)
  library(FSA)
})

conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
conflict_prefer("mutate", "dplyr")
conflict_prefer("arrange", "dplyr")
conflict_prefer("slice", "dplyr")
conflicts_prefer(ComplexHeatmap::pheatmap)

cat("Namespace conflicts resolved: dplyr prioritized\n\n")

# ==============================================================================
# 8) PLOT AESTHETICS + HELPER FUNCTIONS
# ==============================================================================

AXIS_TEXT_SIZE <- 16
AXIS_TITLE_SIZE <- 20
LEGEND_TITLE_SIZE <- 16
LEGEND_TEXT_SIZE <- 16
TITLE_SIZE <- 20
SUBTITLE_SIZE <- 16
STRIP_TEXT_SIZE <- 18

theme_pub <- function() {
  theme_classic() +
    theme(
      plot.title = element_text(hjust = 0.5, size = TITLE_SIZE, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = SUBTITLE_SIZE, face = "bold"),
      axis.text = element_text(size = AXIS_TEXT_SIZE, face = "bold"),
      axis.title = element_text(size = AXIS_TITLE_SIZE, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 15)),
      axis.ticks = element_line(size = 1.1),
      axis.line = element_line(size = 1),
      panel.border = element_rect(color = "black", fill = NA, size = 1),
      legend.title = element_text(size = LEGEND_TITLE_SIZE, face = "bold"),
      legend.text = element_text(size = LEGEND_TEXT_SIZE),
      strip.text = element_text(size = STRIP_TEXT_SIZE, face = "bold"),
      strip.background = element_rect(fill = "white", color = "black", size = 1),
      plot.margin = margin(t = 10, r = 10, b = 50, l = 10)
    )
}

sanitize_filename <- function(x) {
  gsub("[^A-Za-z0-9]+", "_", x)
}

method_value_type <- function(m) {
  switch(
    m,
    "quantiseq" = "Cell Fraction",
    "epic" = "Cell Fraction",
    "mcp_counter" = "Abundance Score",
    "xcell" = "Enrichment Score",
    "timer" = "Infiltration Score",
    "consensus_tme" = "ssGSEA Score",
    "estimate" = "ESTIMATE Score",
    "CIBERSORTx" = "Cell Fraction",
    "Score"
  )
}

auto_ylims <- function(values, ymin_floor = 0) {
  v <- values[is.finite(values)]
  
  if (!length(v)) {
    return(c(ymin_floor, ymin_floor + 0.01))
  }
  
  ymin <- ymin_floor
  ymax <- max(v, na.rm = TRUE)
  yr <- ymax - ymin
  
  if (yr <= 0) {
    yr <- max(abs(ymax) * 0.15, 0.01)
  }
  
  c(ymin, ymax + 0.18 * yr)
}

PLOT_WIDTH_UNPAIRED <- 7
PLOT_HEIGHT_UNPAIRED <- 5
PLOT_WIDTH_PAIRED <- 6
PLOT_HEIGHT_PAIRED <- 5
PLOT_DPI <- 300

# Unpaired boxplots use the same displayed P-value convention as the CIM pipeline.
PLOT_P_COLUMN <- "p_value"
# PLOT_P_COLUMN <- "FDR"   # Use this instead to display BH-adjusted P values.

# ==============================================================================
# 9) ID PARSING — NEO DATASET
# ==============================================================================

# Clean and harmonize sample IDs from all sources.
#
# NEO source files use two equivalent naming conventions:
#   TPM:        NEO11 _Patient_11_T1_B_Rep1 / NEO11 _Patient_11_T4_S_Rep1
#   CIBERSORTx: NEO11 _Patient_11_B_Rep1    / NEO11 _Patient_11_S_Rep1
#
# The T1/T4 labels identify the same biopsy/surgery samples and are removed
# here so every downstream function receives one canonical naming scheme.
# Accidental whitespace is also removed.
clean_sample_ids <- function(x) {
  x %>%
    stringr::str_trim() %>%
    stringr::str_replace_all("\\s+", "") %>%
    stringr::str_replace("_T\\d+_(B|S)_Rep", "_\\1_Rep")
}

get_timepoint <- function(sample_id) {
  sample_id <- clean_sample_ids(sample_id)
  
  dplyr::case_when(
    stringr::str_detect(sample_id, "_B_Rep\\d+$") ~ "Biopsy",
    stringr::str_detect(sample_id, "_S_Rep\\d+$") ~ "Surgery",
    TRUE ~ NA_character_
  )
}

get_patient_id_with_cohort <- function(sample_id) {
  sample_id <- clean_sample_ids(sample_id)
  
  sample_id %>%
    stringr::str_replace("_(B|S)_Rep\\d+$", "") %>%
    stringr::str_replace("_delta$", "")
}

get_patient_key_from_sample <- function(sample_id) {
  sample_id <- clean_sample_ids(sample_id)
  
  sample_id %>%
    stringr::str_replace("_(B|S)_Rep\\d+$", "") %>%
    stringr::str_replace("_Patient_", "_")
}

get_patient_key_from_delta <- function(delta_id) {
  delta_id <- clean_sample_ids(delta_id)
  
  delta_id %>%
    stringr::str_replace("_delta$", "") %>%
    stringr::str_replace("_Patient_", "_")
}

# ==============================================================================
# 10) LOAD NEO CLUSTER TABLE
# ==============================================================================

cat("Loading NEO cluster table...\n")

clusters_raw <- readr::read_csv(NEO_CLUSTER_CSV, show_col_types = FALSE)

if (!all(c("sample_id", "cluster_assignments") %in% colnames(clusters_raw))) {
  stop("NEO cluster CSV must have columns: sample_id, cluster_assignments")
}

clusters <- clusters_raw %>%
  transmute(
    delta_sample_id = clean_sample_ids(as.character(sample_id)),
    patient_key = get_patient_key_from_delta(delta_sample_id),
    cluster = as.character(cluster_assignments)
  ) %>%
  mutate(cluster = factor(cluster))

dup_pat <- clusters %>%
  count(patient_key) %>%
  filter(n > 1)

if (nrow(dup_pat) > 0) {
  stop("NEO cluster file has duplicate patient_key rows. Please deduplicate.")
}

k <- nlevels(clusters$cluster)

cat("Detected k =", k, "clusters\n\n")

# ==============================================================================
# 11) LOAD EXPRESSION MATRIX — NEO
# ==============================================================================

load_expr_matrix <- function(path, dataset_label) {
  cat("Loading expression data for", dataset_label, "from:\n  ", path, "\n")
  
  if (!file.exists(path)) {
    stop("Expression file not found: ", path)
  }
  
  df <- tryCatch(
    fread(path, data.table = FALSE, check.names = FALSE),
    error = function(e) NULL
  )
  
  if (!is.null(df)) {
    # Remove accidental row-index columns commonly introduced by CSV export
    # (e.g. pandas "Unnamed: 0" or base-R "X"). These are not samples.
    index_cols <- names(df)[grepl("^Unnamed:|^X$", names(df))]
    if (length(index_cols) > 0) {
      cat("  Removing non-biological index column(s):",
          paste(index_cols, collapse = ", "), "\n")
      df[index_cols] <- NULL
    }
    
    gene_col <- intersect(GENE_COL_CANDIDATES, colnames(df))
    
    if (length(gene_col)) {
      genes <- make.unique(as.character(df[[gene_col[1]]]))
      df[[gene_col[1]]] <- NULL
      
      # Fail before coercion if any remaining sample column is not numeric.
      non_numeric_cols <- names(df)[!vapply(df, is.numeric, logical(1))]
      if (length(non_numeric_cols) > 0) {
        stop(
          "Non-numeric column(s) remain after removing the gene column: ",
          paste(non_numeric_cols, collapse = ", "),
          ". Check neo_tpm.csv formatting."
        )
      }
      
      mat <- as.matrix(df)
      rownames(mat) <- genes
      storage.mode(mat) <- "numeric"
      
      colnames(mat) <- clean_sample_ids(colnames(mat))
      
      if (anyDuplicated(colnames(mat))) {
        dup_ids <- unique(colnames(mat)[duplicated(colnames(mat))])
        stop(
          "Duplicate sample IDs after NEO ID harmonization: ",
          paste(dup_ids, collapse = ", ")
        )
      }
      
      cat("  Loaded matrix:", nrow(mat), "genes x", ncol(mat), "samples\n\n")
      return(mat)
    }
  }
  
  cat("  Treating first column as rownames...\n")
  
  df2 <- read.csv(path, row.names = 1, check.names = FALSE)
  mat <- as.matrix(df2)
  storage.mode(mat) <- "numeric"
  
  colnames(mat) <- clean_sample_ids(colnames(mat))
  
  if (anyDuplicated(colnames(mat))) {
    dup_ids <- unique(colnames(mat)[duplicated(colnames(mat))])
    stop(
      "Duplicate sample IDs after NEO ID harmonization: ",
      paste(dup_ids, collapse = ", ")
    )
  }
  
  cat("  Loaded matrix:", nrow(mat), "genes x", ncol(mat), "samples\n\n")
  
  mat
}

check_nonlog_tpm <- function(mat, label) {
  mx <- suppressWarnings(max(mat, na.rm = TRUE))
  mn <- suppressWarnings(min(mat, na.rm = TRUE))
  
  if (mn < 0) {
    warning(label, ": negative values present. Unexpected for TPM.", call. = FALSE)
  }
  
  if (mx < 30) {
    warning(label, ": max < 30; data may be log-scale. immunedeconv expects non-log TPM.", call. = FALSE)
  }
  
  invisible(TRUE)
}

expr_neo <- load_expr_matrix(NEO_EXPR_FILE, "NEO")
check_nonlog_tpm(expr_neo, "NEO TPM")

# ==============================================================================
# 12) BUILD METADATA FROM NEO SAMPLE IDS + CLUSTERS
# ==============================================================================

build_metadata_from_samples <- function(sample_ids, dataset_label = "NEO") {
  tibble(sample_id = clean_sample_ids(sample_ids)) %>%
    mutate(
      dataset = dataset_label,
      Timepoint = factor(get_timepoint(sample_id), levels = c("Biopsy", "Surgery")),
      Patient_ID = get_patient_id_with_cohort(sample_id),
      patient_key = get_patient_key_from_sample(sample_id)
    ) %>%
    left_join(clusters, by = "patient_key") %>%
    mutate(cluster = factor(cluster, levels = levels(clusters$cluster))) %>%
    select(sample_id, dataset, Patient_ID, patient_key, Timepoint, cluster)
}

meta_neo <- build_metadata_from_samples(colnames(expr_neo), "NEO")

cat("Metadata matching summary:\n")
cat(
  "  NEO samples with cluster + timepoint:",
  sum(!is.na(meta_neo$cluster) & !is.na(meta_neo$Timepoint)),
  "/", nrow(meta_neo), "\n"
)
cat(
  "  NEO unique patients:",
  dplyr::n_distinct(meta_neo$Patient_ID), "\n\n"
)

# Preflight diagnostics: make ID mismatches explicit before deconvolution.
unmatched_meta <- meta_neo %>%
  filter(is.na(cluster) | is.na(Timepoint))

if (nrow(unmatched_meta) > 0) {
  cat("Preflight: unmatched TPM sample IDs detected:\n")
  print(
    unmatched_meta %>%
      select(sample_id, patient_key, Timepoint, cluster)
  )
  cat("\nAvailable cluster patient keys:\n")
  print(clusters %>% select(patient_key, cluster))
  cat("\n")
}

# The current NEO dataset is paired and should resolve to 19 patients / 38 samples.
# This is reported as a diagnostic rather than hard-coded as an analytical filter.
cat(
  "Preflight canonical IDs:",
  n_distinct(meta_neo$sample_id), "samples;",
  n_distinct(meta_neo$Patient_ID), "patients;",
  sum(meta_neo$Timepoint == "Biopsy", na.rm = TRUE), "biopsy;",
  sum(meta_neo$Timepoint == "Surgery", na.rm = TRUE), "surgery\n\n"
)

keep_neo <- meta_neo %>%
  filter(!is.na(cluster), !is.na(Timepoint)) %>%
  pull(sample_id)

if (length(keep_neo) == 0) {
  stop("No NEO samples matched cluster + timepoint metadata.")
}

expr_neo <- expr_neo[, keep_neo, drop = FALSE]

meta_neo <- meta_neo %>%
  filter(sample_id %in% colnames(expr_neo)) %>%
  arrange(match(sample_id, colnames(expr_neo)))

stopifnot(identical(meta_neo$sample_id, colnames(expr_neo)))

cat("Samples retained:\n")
cat("  NEO:", ncol(expr_neo), "\n\n")

# ==============================================================================
# 13) immunedeconv RUNNER — NEO
# ==============================================================================

make_indications <- function(mat, method) {
  if (method %in% c("timer", "consensus_tme")) {
    setNames(rep(tolower(TUMOR_INDICATION), ncol(mat)), colnames(mat))
  } else {
    NULL
  }
}

deconv_to_long <- function(res, method, dataset) {
  if (is.null(res)) {
    return(NULL)
  }
  
  if (method == "estimate") {
    long <- res %>%
      as.data.frame() %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id")
    
    long %>%
      mutate(sample_id = clean_sample_ids(sample_id)) %>%
      pivot_longer(
        cols = -sample_id,
        names_to = "Cell_Type",
        values_to = "Value"
      ) %>%
      mutate(Method = method, dataset = dataset, Value = as.numeric(Value))
    
  } else {
    if (!"cell_type" %in% colnames(res)) {
      return(NULL)
    }
    
    res %>%
      column_to_rownames("cell_type") %>%
      t() %>%
      as.data.frame() %>%
      rownames_to_column("sample_id") %>%
      mutate(sample_id = clean_sample_ids(sample_id)) %>%
      pivot_longer(
        cols = -sample_id,
        names_to = "Cell_Type",
        values_to = "Value"
      ) %>%
      mutate(Method = method, dataset = dataset, Value = as.numeric(Value))
  }
}

run_methods <- function(expr, meta, dataset_label = "NEO") {
  cat("\n========================================\n")
  cat("Running deconvolution for", dataset_label, "\n")
  cat("========================================\n")
  
  out <- list()
  log <- list()
  
  for (m in METHODS_TO_RUN) {
    cat("\n  Method:", m, "\n")
    
    ind <- make_indications(expr, m)
    res <- NULL
    convergence_warning <- FALSE
    
    res <- withr::with_dir(DIRS$tmp, {
      if (m == "estimate") {
        cat("    Running ESTIMATE...\n")
        
        tryCatch(
          immunedeconv::deconvolute_estimate(expr),
          error = function(e) {
            cat("    ERROR:", e$message, "\n")
            NULL
          }
        )
        
      } else if (m == "epic") {
        cat("    Running EPIC (tumor=TRUE, arrays=FALSE)...\n")
        
        tryCatch(
          withCallingHandlers(
            immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE),
            warning = function(w) {
              if (grepl("didn't fully converge", w$message, ignore.case = TRUE)) {
                convergence_warning <<- TRUE
                cat("    NOTE: EPIC partial convergence\n")
              }
              invokeRestart("muffleWarning")
            }
          ),
          error = function(e) {
            cat("    ERROR:", e$message, "\n")
            NULL
          }
        )
        
      } else if (m == "quantiseq") {
        cat("    Running quanTIseq (tumor=TRUE, arrays=FALSE)...\n")
        
        tryCatch(
          immunedeconv::deconvolute(expr, m, tumor = TRUE, arrays = FALSE),
          error = function(e) {
            cat("    ERROR:", e$message, "\n")
            NULL
          }
        )
        
      } else if (!is.null(ind)) {
        cat("    Running with BRCA indication...\n")
        
        tryCatch(
          immunedeconv::deconvolute(expr, m, indications = ind, arrays = FALSE),
          error = function(e) {
            cat("    ERROR:", e$message, "\n")
            NULL
          }
        )
        
      } else {
        cat("    Running standard deconvolution (arrays=FALSE)...\n")
        
        tryCatch(
          immunedeconv::deconvolute(expr, m, arrays = FALSE),
          error = function(e) {
            cat("    ERROR:", e$message, "\n")
            NULL
          }
        )
      }
    })
    
    if (m == "epic" && !is.null(res) && "cell_type" %in% colnames(res)) {
      n_samples_returned <- ncol(res) - 1
      rate <- n_samples_returned / ncol(expr)
      
      cat(
        "    EPIC results for ",
        n_samples_returned, "/", ncol(expr),
        " samples (", round(rate * 100, 1), "%)\n",
        sep = ""
      )
      
      if (rate < 0.3) {
        cat("    WARNING: <30% samples converged, discarding EPIC output\n")
        res <- NULL
      }
    }
    
    long <- deconv_to_long(res, m, dataset_label)
    
    if (!is.null(long)) {
      long <- long %>%
        left_join(
          meta %>% select(sample_id, Patient_ID, Timepoint, cluster),
          by = "sample_id"
        )
      
      miss_meta <- sum(is.na(long$Patient_ID) | is.na(long$cluster))
      
      if (miss_meta) {
        cat("    WARNING:", miss_meta, "rows missing metadata after join\n")
      }
      
      cat("    SUCCESS:", nrow(long), "rows\n")
    }
    
    out[[m]] <- long
    
    log[[m]] <- tibble(
      dataset = dataset_label,
      Method = m,
      status = ifelse(is.null(long), "FAILED", "OK"),
      n_samples = ifelse(is.null(long), 0, n_distinct(long$sample_id)),
      n_celltypes = ifelse(is.null(long), 0, n_distinct(long$Cell_Type)),
      notes = case_when(
        is.null(long) ~ "Method failed or temp/cache issue",
        m == "epic" & convergence_warning ~ "Partial convergence",
        TRUE ~ "Success"
      )
    )
  }
  
  list(
    immune_long = bind_rows(out),
    run_log = bind_rows(log)
  )
}

if (isTRUE(params$run_immunedeconv)) {
  res_neo <- run_methods(expr_neo, meta_neo, "NEO")
  immune_long_raw <- res_neo$immune_long
  run_summary <- res_neo$run_log
} else {
  immune_long_raw <- tibble()
  run_summary <- tibble()
  cat("\nSkipping immunedeconv because params$run_immunedeconv = FALSE.\n")
}

# ==============================================================================
# 14) CIBERSORTx IMPORT — NEO
# ==============================================================================

read_cibersort_noqc_long <- function(path, dataset_label = "NEO") {
  cat("\nLoading CIBERSORTx for", dataset_label, "from:\n  ", path, "\n")
  
  if (!file.exists(path)) {
    warning("CIBERSORTx file not found: ", path)
    return(NULL)
  }
  
  cs <- fread(path, data.table = FALSE, check.names = FALSE)
  
  if (!("Mixture" %in% colnames(cs))) {
    stop("CIBERSORTx file must contain a column named 'Mixture'.")
  }
  
  cs_long <- cs %>%
    mutate(Mixture = clean_sample_ids(Mixture)) %>%
    rename(sample_id = Mixture) %>%
    pivot_longer(
      cols = -sample_id,
      names_to = "Cell_Type",
      values_to = "Value"
    ) %>%
    mutate(
      Method = "CIBERSORTx",
      dataset = dataset_label,
      Value = as.numeric(Value)
    )
  
  meta <- build_metadata_from_samples(unique(cs_long$sample_id), dataset_label)
  
  cs_long %>%
    left_join(
      meta %>% select(sample_id, Patient_ID, Timepoint, cluster),
      by = "sample_id"
    )
}

ciber_neo_long <- NULL
if (isTRUE(params$run_cibersortx_import)) {
  ciber_neo_long <- read_cibersort_noqc_long(CIBERSORT_NEO_FILE, "NEO")
}

if (!is.null(ciber_neo_long)) {
  immune_long_raw <- bind_rows(immune_long_raw, ciber_neo_long)
  
  run_summary <- bind_rows(
    run_summary,
    tibble(
      dataset = "NEO",
      Method = "CIBERSORTx",
      status = "OK",
      n_samples = n_distinct(ciber_neo_long$sample_id),
      n_celltypes = n_distinct(ciber_neo_long$Cell_Type),
      notes = "Imported from CIBERSORTx web output; Mixture sample names cleaned"
    )
  )
}

saveRDS(
  immune_long_raw,
  file.path(DIRS$results, "NEO_immune_long_raw_all_methods.rds")
)

write.csv(
  run_summary,
  file.path(DIRS$results, "NEO_run_summary_methods.csv"),
  row.names = FALSE
)

cat("\nStep 14 complete: immunedeconv + CIBERSORTx imported for NEO.\n")


# ==============================================================================
# 15) PATIENT-LEVEL AGGREGATION — NEO
# ==============================================================================

immune_long_clean <- immune_long_raw %>%
  filter(
    !is.na(dataset),
    !is.na(Patient_ID),
    !is.na(Timepoint),
    !is.na(cluster),
    is.finite(Value)
  )

immune_patient <- immune_long_clean %>%
  group_by(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster) %>%
  summarise(
    Value = mean(Value, na.rm = TRUE),
    n_replicates = n(),
    .groups = "drop"
  )

saveRDS(
  immune_patient,
  file.path(DIRS$results, "NEO_immune_patient_level_ORIGINAL.rds")
)

write.csv(
  immune_patient,
  file.path(DIRS$results, "NEO_immune_patient_level_ORIGINAL.csv"),
  row.names = FALSE
)

method_ymin_original <- immune_patient %>%
  group_by(Method) %>%
  summarise(
    ymin = ifelse(min(Value, na.rm = TRUE) < 0, min(Value, na.rm = TRUE), 0),
    .groups = "drop"
  )


# ==============================================================================
# 16) T CELL AGGREGATION — xCell + CIBERSORTx
# ==============================================================================

aggregate_xcell_tcells <- function(immune_patient) {
  if (!AGGREGATE_XCELL_TCELLS) {
    return(list(
      immune_patient_aggregated = immune_patient,
      aggregation_details = NULL
    ))
  }
  
  xcell_data <- immune_patient %>%
    filter(Method == "xcell")
  
  non_xcell <- immune_patient %>%
    filter(Method != "xcell")
  
  if (nrow(xcell_data) == 0) {
    return(list(
      immune_patient_aggregated = immune_patient,
      aggregation_details = NULL
    ))
  }
  
  cd4_pattern <- "T cell CD4\\+"
  cd8_pattern <- "T cell CD8\\+"
  
  cd4_aggregated <- xcell_data %>%
    filter(grepl(cd4_pattern, Cell_Type)) %>%
    group_by(dataset, Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD4+ (aggregated)")
  
  cd8_aggregated <- xcell_data %>%
    filter(grepl(cd8_pattern, Cell_Type)) %>%
    group_by(dataset, Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = "T cell CD8+ (aggregated)")
  
  other_xcell <- xcell_data %>%
    filter(
      !grepl(cd4_pattern, Cell_Type),
      !grepl(cd8_pattern, Cell_Type)
    )
  
  immune_patient_aggregated <- bind_rows(
    non_xcell,
    other_xcell,
    cd4_aggregated %>%
      select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_aggregated %>%
      select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )
  
  aggregation_details <- bind_rows(
    cd4_aggregated %>%
      select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_aggregated %>%
      select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )
  
  list(
    immune_patient_aggregated = immune_patient_aggregated,
    aggregation_details = aggregation_details
  )
}


aggregate_cibersort_tcells <- function(df, strategy = "refined") {
  if (!AGGREGATE_CIBERSORT_TCELLS) {
    return(list(aggregated = df, details = NULL))
  }
  
  if (nrow(df) == 0) {
    return(list(aggregated = df, details = NULL))
  }
  
  if (strategy == "refined") {
    cd4_subtypes <- c(
      "T cells CD4 naive",
      "T cells CD4 memory resting",
      "T cells CD4 memory activated"
    )
    
    cd8_subtypes <- c("T cells CD8")
    
    cd4_label <- "T cells CD4+ (general)"
    cd8_label <- "T cells CD8+ (total)"
    
  } else if (strategy == "simple") {
    cd4_subtypes <- c(
      "T cells CD4 naive",
      "T cells CD4 memory resting",
      "T cells CD4 memory activated",
      "T cells follicular helper"
    )
    
    cd8_subtypes <- c("T cells CD8")
    
    cd4_label <- "T cells CD4+ (total)"
    cd8_label <- "T cells CD8+ (total)"
    
  } else {
    stop("CIBERSORT_AGGREGATION_STRATEGY must be 'refined' or 'simple'.")
  }
  
  cd4_agg <- df %>%
    filter(Cell_Type %in% cd4_subtypes) %>%
    group_by(dataset, Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd4_label)
  
  cd8_agg <- df %>%
    filter(Cell_Type %in% cd8_subtypes) %>%
    group_by(dataset, Method, Patient_ID, Timepoint, cluster) %>%
    summarise(
      Value = sum(Value, na.rm = TRUE),
      n_subtypes = n(),
      original_subtypes = paste(sort(unique(Cell_Type)), collapse = " + "),
      .groups = "drop"
    ) %>%
    mutate(Cell_Type = cd8_label)
  
  aggregated_subtypes <- c(cd4_subtypes, cd8_subtypes)
  
  other_cells <- df %>%
    filter(!Cell_Type %in% aggregated_subtypes)
  
  aggregated <- bind_rows(
    other_cells,
    cd4_agg %>%
      select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value),
    cd8_agg %>%
      select(dataset, Method, Cell_Type, Patient_ID, Timepoint, cluster, Value)
  )
  
  details <- bind_rows(
    cd4_agg %>%
      select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes),
    cd8_agg %>%
      select(dataset, Patient_ID, Timepoint, cluster, Cell_Type, n_subtypes, original_subtypes)
  )
  
  list(
    aggregated = aggregated,
    details = details
  )
}


xcell_agg <- aggregate_xcell_tcells(immune_patient)

immune_after_xcell <- xcell_agg$immune_patient_aggregated

if (!is.null(xcell_agg$aggregation_details)) {
  write.csv(
    xcell_agg$aggregation_details,
    file.path(DIRS$results, "NEO_xcell_tcell_aggregation_details.csv"),
    row.names = FALSE
  )
}

ciber_df <- immune_after_xcell %>%
  filter(Method == "CIBERSORTx")

non_ciber <- immune_after_xcell %>%
  filter(Method != "CIBERSORTx")

ciber_agg <- aggregate_cibersort_tcells(
  ciber_df,
  CIBERSORT_AGGREGATION_STRATEGY
)

immune_patient_aggregated <- bind_rows(
  non_ciber,
  ciber_agg$aggregated
)

if (!is.null(ciber_agg$details)) {
  write.csv(
    ciber_agg$details,
    file.path(DIRS$results, "NEO_cibersort_tcell_aggregation_details.csv"),
    row.names = FALSE
  )
}

saveRDS(
  immune_patient_aggregated,
  file.path(DIRS$results, "NEO_immune_patient_level_AGGREGATED.rds")
)

write.csv(
  immune_patient_aggregated,
  file.path(DIRS$results, "NEO_immune_patient_level_AGGREGATED.csv"),
  row.names = FALSE
)

method_ymin_aggregated <- immune_patient_aggregated %>%
  group_by(Method) %>%
  summarise(
    ymin = ifelse(min(Value, na.rm = TRUE) < 0, min(Value, na.rm = TRUE), 0),
    .groups = "drop"
  )


# ==============================================================================
# 16.5) CLUSTER LABEL HANDLING — NEO
# ==============================================================================

# Keep original cluster labels from the NEO cluster file.
USE_CLUSTER_NAMES <- FALSE

apply_cluster_names <- function(df, cluster_col = "cluster") {
  df %>%
    mutate(
      !!cluster_col := factor(as.character(!!sym(cluster_col)))
    )
}

immune_patient <- apply_cluster_names(immune_patient)
immune_patient_aggregated <- apply_cluster_names(immune_patient_aggregated)

NEO_CLUSTER_ORDER <- levels(factor(immune_patient_aggregated$cluster))

cluster_colors <- cluster_colors[
  names(cluster_colors) %in% NEO_CLUSTER_ORDER
]

cat("NEO cluster labels retained as:\n")
print(NEO_CLUSTER_ORDER)

# ==============================================================================
# 17) STATS (k-flexible) — UNPAIRED at Biopsy and Surgery
#    Global: Wilcoxon (k=2) / Kruskal (k>=3)
#    Posthoc: Dunn (k>=3) with BH-adjusted p (from FSA)
#    Also writes a comprehensive "skipped" report.
# ==============================================================================

# ---- helpers (safe summaries + safe empty tibbles) ----
summarise_by_cluster <- function(df) {
  df %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(
      n    = dplyr::n(),
      mean = mean(Value, na.rm = TRUE),
      sd   = sd(Value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    dplyr::mutate(cluster = as.character(cluster))
}

empty_skip_tbl <- function() {
  tibble::tibble(
    analysis = character(),
    Method = character(),
    Cell_Type = character(),
    Timepoint = character(),
    reason = character(),
    k = integer(),
    n_total = integer(),
    n_unique_value = integer(),
    min_cluster_n = integer()
  )
}

empty_global_tbl <- function() {
  tibble::tibble(
    Method = character(),
    Cell_Type = character(),
    Timepoint = character(),
    test = character(),
    p_value = numeric(),
    n_clusters = integer(),
    note = character(),
    summary = list()
  )
}

empty_dunn_tbl <- function() {
  tibble::tibble(
    Method = character(),
    Cell_Type = character(),
    Timepoint = character(),
    comparison = character(),
    cluster1 = character(),
    cluster2 = character(),
    n1 = integer(), mean1 = numeric(), sd1 = numeric(),
    n2 = integer(), mean2 = numeric(), sd2 = numeric(),
    z = numeric(),
    p_unadj = numeric(),
    fdr = numeric()
  )
}

make_skip_row <- function(Method, Cell_Type, Timepoint,
                          analysis = c("unpaired_global", "unpaired_dunn"),
                          reason,
                          k = NA_integer_,
                          n_total = NA_integer_,
                          n_unique_value = NA_integer_,
                          min_cluster_n = NA_integer_) {
  analysis <- match.arg(analysis)
  tibble::tibble(
    analysis = analysis,
    Method = as.character(Method),
    Cell_Type = as.character(Cell_Type),
    Timepoint = as.character(Timepoint),
    reason = as.character(reason),
    k = as.integer(k),
    n_total = as.integer(n_total),
    n_unique_value = as.integer(n_unique_value),
    min_cluster_n = as.integer(min_cluster_n)
  )
}

# ---- one-group global test (returns list(out, skip)) ----
unpaired_global_one <- function(df, Method, Cell_Type, Timepoint) {
  df <- df %>% dplyr::filter(is.finite(Value), !is.na(cluster))
  kk <- dplyr::n_distinct(df$cluster)
  n_total <- nrow(df)
  n_unique <- dplyr::n_distinct(df$Value)
  min_n <- if (kk > 0) min(as.integer(table(df$cluster))) else NA_integer_
  
  if (kk < 2) {
    return(list(
      out  = NULL,
      skip = make_skip_row(Method, Cell_Type, Timepoint,
                           analysis = "unpaired_global",
                           reason = "k<2",
                           k = kk, n_total = n_total, n_unique_value = n_unique, min_cluster_n = min_n)
    ))
  }
  
  summ <- summarise_by_cluster(df)
  
  # no variation → don't run test
  if (n_unique < 2) {
    return(list(
      out = tibble::tibble(
        Method = as.character(Method),
        Cell_Type = as.character(Cell_Type),
        Timepoint = as.character(Timepoint),
        test = ifelse(kk == 2, "wilcox", "kruskal"),
        p_value = NA_real_,
        n_clusters = kk,
        note = "skipped_no_variation",
        summary = list(summ)
      ),
      skip = make_skip_row(Method, Cell_Type, Timepoint,
                           analysis = "unpaired_global",
                           reason = "no_variation",
                           k = kk, n_total = n_total, n_unique_value = n_unique, min_cluster_n = min_n)
    ))
  }
  
  p <- if (kk == 2) {
    tryCatch(stats::wilcox.test(Value ~ cluster, data = df, exact = FALSE)$p.value,
             error = function(e) NA_real_)
  } else {
    tryCatch(stats::kruskal.test(Value ~ cluster, data = df)$p.value,
             error = function(e) NA_real_)
  }
  
  list(
    out = tibble::tibble(
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      Timepoint = as.character(Timepoint),
      test = ifelse(kk == 2, "wilcox", "kruskal"),
      p_value = p,
      n_clusters = kk,
      note = NA_character_,
      summary = list(summ)
    ),
    skip = NULL
  )
}

# ---- one-group Dunn posthoc (returns list(out, skip)) ----
unpaired_dunn_one <- function(df, Method, Cell_Type, Timepoint, drop_small_groups_n = 2) {
  df <- df %>% dplyr::filter(is.finite(Value), !is.na(cluster))
  
  # optional: remove tiny groups that can cause undefined Z
  df2 <- df %>%
    dplyr::group_by(cluster) %>%
    dplyr::filter(dplyr::n() >= drop_small_groups_n) %>%
    dplyr::ungroup()
  
  kk <- dplyr::n_distinct(df2$cluster)
  n_total <- nrow(df2)
  n_unique <- dplyr::n_distinct(df2$Value)
  min_n <- if (kk > 0) min(as.integer(table(df2$cluster))) else NA_integer_
  
  if (kk < 3) {
    return(list(
      out = NULL,
      skip = make_skip_row(Method, Cell_Type, Timepoint,
                           analysis = "unpaired_dunn",
                           reason = "k<3_after_filter",
                           k = kk, n_total = n_total, n_unique_value = n_unique, min_cluster_n = min_n)
    ))
  }
  
  if (n_unique < 2) {
    return(list(
      out = NULL,
      skip = make_skip_row(Method, Cell_Type, Timepoint,
                           analysis = "unpaired_dunn",
                           reason = "no_variation",
                           k = kk, n_total = n_total, n_unique_value = n_unique, min_cluster_n = min_n)
    ))
  }
  
  cl_summ <- summarise_by_cluster(df2) %>%
    dplyr::rename(n_group = n, mean_group = mean, sd_group = sd)
  
  dt <- tryCatch(
    suppressWarnings(FSA::dunnTest(Value ~ cluster, data = df2, method = "bh")$res),
    error = function(e) NULL
  )
  
  if (is.null(dt)) {
    return(list(
      out = NULL,
      skip = make_skip_row(Method, Cell_Type, Timepoint,
                           analysis = "unpaired_dunn",
                           reason = "dunn_error",
                           k = kk, n_total = n_total, n_unique_value = n_unique, min_cluster_n = min_n)
    ))
  }
  
  out <- tibble::as_tibble(dt) %>%
    dplyr::rename(
      comparison = Comparison,
      z = Z,
      p_unadj = P.unadj,
      fdr = P.adj
    ) %>%
    dplyr::mutate(
      Method = as.character(Method),
      Cell_Type = as.character(Cell_Type),
      Timepoint = as.character(Timepoint),
      cluster1 = sub(" -.*$", "", comparison),
      cluster2 = sub("^.*- ", "", comparison),
      z = suppressWarnings(as.numeric(z))
    ) %>%
    dplyr::left_join(
      cl_summ %>% dplyr::rename(cluster1 = cluster, n1 = n_group, mean1 = mean_group, sd1 = sd_group),
      by = "cluster1"
    ) %>%
    dplyr::left_join(
      cl_summ %>% dplyr::rename(cluster2 = cluster, n2 = n_group, mean2 = mean_group, sd2 = sd_group),
      by = "cluster2"
    ) %>%
    dplyr::relocate(
      Method, Cell_Type, Timepoint, comparison, cluster1, cluster2,
      n1, mean1, sd1, n2, mean2, sd2, z, p_unadj, fdr
    )
  
  list(out = out, skip = NULL)
}

# ---- main wrapper ----
compute_unpaired_all <- function(immune_patient_df, out_prefix, drop_small_groups_n = 2) {
  # Ensure required columns exist
  stopifnot(all(c("Method","Cell_Type","Timepoint","cluster","Value") %in% colnames(immune_patient_df)))
  
  # Apply cluster names if enabled
  immune_patient_df <- apply_cluster_names(immune_patient_df)
  
  keys <- immune_patient_df %>%
    dplyr::distinct(Method, Cell_Type, Timepoint) %>%
    dplyr::arrange(Timepoint, Method, Cell_Type)
  
  global_rows <- list()
  dunn_rows   <- list()
  skip_rows   <- list()
  
  for (i in seq_len(nrow(keys))) {
    k <- keys[i, ]
    df_sub <- immune_patient_df %>%
      dplyr::filter(Method == k$Method, Cell_Type == k$Cell_Type, Timepoint == k$Timepoint)
    
    # global
    g <- unpaired_global_one(df_sub, k$Method, k$Cell_Type, k$Timepoint)
    if (!is.null(g$out))  global_rows[[length(global_rows)+1]] <- g$out
    if (!is.null(g$skip)) skip_rows[[length(skip_rows)+1]] <- g$skip
    
    # dunn
    d <- unpaired_dunn_one(df_sub, k$Method, k$Cell_Type, k$Timepoint,
                           drop_small_groups_n = drop_small_groups_n)
    if (!is.null(d$out))  dunn_rows[[length(dunn_rows)+1]] <- d$out
    if (!is.null(d$skip)) skip_rows[[length(skip_rows)+1]] <- d$skip
  }
  
  global_tbl <- if (length(global_rows)) dplyr::bind_rows(global_rows) else empty_global_tbl()
  dunn_tbl   <- if (length(dunn_rows))   dplyr::bind_rows(dunn_rows)   else empty_dunn_tbl()
  skipped    <- if (length(skip_rows))   dplyr::bind_rows(skip_rows)   else empty_skip_tbl()
  
  # FDR for global tests (across all rows in this run)
  global_tbl <- global_tbl %>%
    dplyr::mutate(FDR = p.adjust(p_value, method = "BH"))
  
  # Wide cluster summary for reporting - preserve cluster names
  global_wide <- global_tbl %>%
    tidyr::unnest(summary) %>%
    tidyr::pivot_wider(
      id_cols = c(Method, Cell_Type, Timepoint, test, p_value, FDR, n_clusters, note),
      names_from = cluster,
      values_from = c(n, mean, sd),
      names_glue = "{cluster}_{.value}"
    ) %>%
    dplyr::arrange(Timepoint, Method, Cell_Type)
  
  # Arrange only if present (safe even if empty)
  if (nrow(dunn_tbl) > 0) {
    dunn_tbl <- dunn_tbl %>% dplyr::arrange(Timepoint, Method, Cell_Type, comparison)
  }
  if (nrow(skipped) > 0) {
    skipped <- skipped %>% dplyr::arrange(analysis, Timepoint, Method, Cell_Type)
  }
  
  # Write
  utils::write.csv(global_wide, file.path(DIRS$results, paste0(out_prefix, "_unpaired_global.csv")), row.names = FALSE)
  utils::write.csv(dunn_tbl,   file.path(DIRS$results, paste0(out_prefix, "_unpaired_dunn.csv")),   row.names = FALSE)
  utils::write.csv(skipped,    file.path(DIRS$results, paste0(out_prefix, "_unpaired_skipped.csv")),row.names = FALSE)
  
  list(global = global_wide, dunn = dunn_tbl, skipped = skipped)
}

# ==============================================================================
# Usage
# ==============================================================================
unpaired_original   <- compute_unpaired_all(immune_patient,           "ORIGINAL",   drop_small_groups_n = 2)
unpaired_aggregated <- compute_unpaired_all(immune_patient_aggregated,"AGGREGATED", drop_small_groups_n = 2)

# ==============================================================================
# 18) PAIRED STATS (k-flexible) — within each cluster
#   - Paired Wilcoxon (Surgery vs Biopsy) within each cluster
#   - Robust skip logging (n_pairs<3, no variation, missing pairs)
#   - Outputs: _pairedstats.csv, paired_patient_deltas.csv, *_paired_skipped.csv
# ==============================================================================

make_skip_row_paired <- function(Method, Cell_Type, cluster,
                                 reason,
                                 n_pairs = NA_integer_,
                                 n_unique_delta = NA_integer_) {
  tibble::tibble(
    analysis = "paired",
    Method = Method,
    Cell_Type = Cell_Type,
    cluster = as.character(cluster),
    reason = reason,
    n_pairs = n_pairs,
    n_unique_delta = n_unique_delta
  )
}

paired_one <- function(df, Method, Cell_Type, cluster_label) {
  # df here is already filtered to one Method/Cell_Type/cluster
  wide <- df %>%
    filter(is.finite(Value), !is.na(Timepoint), !is.na(Patient_ID)) %>%
    group_by(Patient_ID, Timepoint) %>%
    summarise(Value = mean(Value), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = Timepoint, values_from = Value)
  
  # require both timepoints
  wide <- wide %>% filter(is.finite(Biopsy) & is.finite(Surgery))
  
  if (nrow(wide) == 0) {
    return(list(
      stats = NULL,
      wide = NULL,
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, reason = "no_complete_pairs", n_pairs = 0)
    ))
  }
  
  wide <- wide %>% mutate(Delta = Surgery - Biopsy)
  n_pairs <- nrow(wide)
  n_unique_delta <- dplyr::n_distinct(wide$Delta)
  
  if (n_pairs < 3) {
    return(list(
      stats = tibble::tibble(
        Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label),
        n_pairs = n_pairs,
        mean_biopsy = mean(wide$Biopsy), sd_biopsy = sd(wide$Biopsy),
        mean_surgery = mean(wide$Surgery), sd_surgery = sd(wide$Surgery),
        mean_delta = mean(wide$Delta), sd_delta = sd(wide$Delta),
        p_value = NA_real_,
        note = "skipped_n_pairs<3"
      ),
      wide = wide %>% mutate(Method=Method, Cell_Type=Cell_Type, cluster=as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, reason="n_pairs<3", n_pairs=n_pairs, n_unique_delta=n_unique_delta)
    ))
  }
  
  if (n_unique_delta < 2) {
    return(list(
      stats = tibble::tibble(
        Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label),
        n_pairs = n_pairs,
        mean_biopsy = mean(wide$Biopsy), sd_biopsy = sd(wide$Biopsy),
        mean_surgery = mean(wide$Surgery), sd_surgery = sd(wide$Surgery),
        mean_delta = mean(wide$Delta), sd_delta = sd(wide$Delta),
        p_value = NA_real_,
        note = "skipped_no_variation_in_delta"
      ),
      wide = wide %>% mutate(Method=Method, Cell_Type=Cell_Type, cluster=as.character(cluster_label)),
      skip = make_skip_row_paired(Method, Cell_Type, cluster_label, reason="no_variation_in_delta", n_pairs=n_pairs, n_unique_delta=n_unique_delta)
    ))
  }
  
  p <- tryCatch(
    stats::wilcox.test(wide$Surgery, wide$Biopsy, paired = TRUE, exact = FALSE)$p.value,
    error = function(e) NA_real_
  )
  
  stats_out <- tibble::tibble(
    Method = Method, Cell_Type = Cell_Type, cluster = as.character(cluster_label),
    n_pairs = n_pairs,
    mean_biopsy = mean(wide$Biopsy), sd_biopsy = sd(wide$Biopsy),
    mean_surgery = mean(wide$Surgery), sd_surgery = sd(wide$Surgery),
    mean_delta = mean(wide$Delta), sd_delta = sd(wide$Delta),
    p_value = p,
    note = NA_character_
  )
  
  list(
    stats = stats_out,
    wide  = wide %>% mutate(Method=Method, Cell_Type=Cell_Type, cluster=as.character(cluster_label)),
    skip  = NULL
  )
}

compute_paired_all <- function(immune_patient_df, out_prefix) {
  stopifnot(all(c("Method","Cell_Type","Timepoint","cluster","Patient_ID","Value") %in% colnames(immune_patient_df)))
  
  # Apply cluster names if enabled
  immune_patient_df <- apply_cluster_names(immune_patient_df)
  
  # split by Method/Cell_Type/cluster
  groups <- immune_patient_df %>%
    filter(!is.na(cluster)) %>%
    group_by(Method, Cell_Type, cluster) %>%
    group_split(.keep = TRUE)
  
  stats_tbl <- purrr::map(groups, function(g) {
    Method <- g$Method[[1]]; Cell_Type <- g$Cell_Type[[1]]; cl <- g$cluster[[1]]
    paired_one(g, Method, Cell_Type, cl)$stats
  }) %>% dplyr::bind_rows()
  
  wide_tbl <- purrr::map(groups, function(g) {
    Method <- g$Method[[1]]; Cell_Type <- g$Cell_Type[[1]]; cl <- g$cluster[[1]]
    paired_one(g, Method, Cell_Type, cl)$wide
  }) %>% dplyr::bind_rows()
  
  skip_tbl <- purrr::map(groups, function(g) {
    Method <- g$Method[[1]]; Cell_Type <- g$Cell_Type[[1]]; cl <- g$cluster[[1]]
    paired_one(g, Method, Cell_Type, cl)$skip
  }) %>% dplyr::bind_rows()
  
  # FDR correction within Method x cluster (matches your previous philosophy)
  stats_tbl <- stats_tbl %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(Method, cluster, Cell_Type)
  
  if (nrow(skip_tbl) == 0) {
    skip_tbl <- tibble::tibble(
      analysis=character(), Method=character(), Cell_Type=character(), cluster=character(),
      reason=character(), n_pairs=integer(), n_unique_delta=integer()
    )
  } else {
    skip_tbl <- skip_tbl %>% arrange(Method, cluster, Cell_Type)
  }
  
  utils::write.csv(stats_tbl, file.path(DIRS$results, paste0(out_prefix, "_paired_stats.csv")), row.names = FALSE)
  utils::write.csv(wide_tbl,  file.path(DIRS$results, paste0(out_prefix, "_paired_patient_deltas.csv")), row.names = FALSE)
  utils::write.csv(skip_tbl,  file.path(DIRS$results, paste0(out_prefix, "_paired_skipped.csv")), row.names = FALSE)
  
  list(stats = stats_tbl, wide = wide_tbl, skipped = skip_tbl)
}

paired_original     <- compute_paired_all(immune_patient,            "ORIGINAL")
paired_aggregated   <- compute_paired_all(immune_patient_aggregated, "AGGREGATED")

# ==============================================================================
# 19) PLOTS — UNPAIRED + PAIRED SPLIT (ONE FILE PER CLUSTER)
#     - saves companion CSV(s) per plot so stats and plots stay linked
#     - shared y-axis across clusters for same cell/method
#     - cluster colors for unpaired boxplots
#     - significance brackets for unpaired comparisons
#     - n_pairs in subtitle for paired plots
# ==============================================================================

safe_stem <- function(...) paste(..., sep = "_") |> sanitize_filename()

# -----------------------------
# Helper: compute shared y-limits across all clusters
# -----------------------------
compute_shared_ylims <- function(df, method_ymin) {
  method_name <- unique(df$Method)[1]
  method_min <- method_ymin %>% filter(Method == method_name) %>% pull(ymin)
  if (!length(method_min) || is.na(method_min)) method_min <- 0
  
  # Get all finite values across all clusters
  all_values <- df %>%
    filter(is.finite(Value), !is.na(cluster)) %>%
    pull(Value)
  
  auto_ylims(all_values, ymin_floor = method_min)
}

# -----------------------------
# Helper: format p-values for display
# -----------------------------
format_pval <- function(p) {
  if (is.na(p)) return("ns")
  if (p < 0.001) return("***")
  if (p < 0.01) return("**")
  if (p < 0.05) return("*")
  return("ns")
}

# -----------------------------
# Helper: add significance brackets to plot
# -----------------------------
add_significance_brackets <- function(p, dunn_data, ylim) {
  if (is.null(dunn_data) || nrow(dunn_data) == 0) return(p)
  
  # Filter for significant comparisons (p < 0.05)
  sig_comps <- dunn_data %>%
    filter(fdr < 0.05) %>%
    arrange(fdr)
  
  if (nrow(sig_comps) == 0) return(p)
  
  # Calculate bracket positions
  y_range <- ylim[2] - ylim[1]
  y_start <- ylim[2] - 0.12 * y_range  # Start below title area
  y_step <- 0.04 * y_range  # Space between brackets
  
  # Add brackets for each significant comparison (limit to top 5)
  for (i in seq_len(min(nrow(sig_comps), 5))) {
    row <- sig_comps[i, ]
    y_pos <- y_start - (i - 1) * y_step
    
    # Get cluster positions (handle both numeric and named clusters)
    cluster_levels <- levels(factor(dunn_data$cluster1))
    x1 <- match(as.character(row$cluster1), cluster_levels)
    x2 <- match(as.character(row$cluster2), cluster_levels)
    
    if (is.na(x1) || is.na(x2)) next
    
    p <- p +
      # Left vertical line
      annotate("segment",
               x = x1, xend = x1,
               y = y_pos, yend = y_pos - 0.01 * y_range,
               size = 0.7, color = "black") +
      # Horizontal line
      annotate("segment",
               x = x1, xend = x2,
               y = y_pos, yend = y_pos,
               size = 0.7, color = "black") +
      # Right vertical line
      annotate("segment",
               x = x2, xend = x2,
               y = y_pos, yend = y_pos - 0.01 * y_range,
               size = 0.7, color = "black") +
      # P-value label
      annotate("text",
               x = (x1 + x2) / 2,
               y = y_pos + 0.005 * y_range,
               label = format_pval(row$fdr),
               size = 4.5, fontface = "bold")
  }
  
  return(p)
}

# -----------------------------
# UNPAIRED single plot (with colors and p-values)
# -----------------------------
plot_unpaired_one <- function(
    df,
    shared_ylim,
    unpaired_stats,
    p_column = PLOT_P_COLUMN
) {

  df <- df %>%
    filter(is.finite(Value), !is.na(cluster))

  if (n_distinct(df$cluster) < 2) {
    return(NULL)
  }

  method_name <- unique(df$Method)[1]
  cell_type   <- unique(df$Cell_Type)[1]
  timepoint   <- as.character(unique(df$Timepoint)[1])

  # Retain the original NEO cluster identities/colors, but use the same
  # boxplot geometry, typography, labeling and P-value display as CIM.
  cluster_levels_present <- NEO_CLUSTER_ORDER[
    NEO_CLUSTER_ORDER %in% as.character(unique(df$cluster))
  ]

  if (length(cluster_levels_present) < 2) {
    cluster_levels_present <- sort(unique(as.character(df$cluster)))
  }

  df <- df %>%
    mutate(
      cluster = factor(cluster, levels = cluster_levels_present)
    )

  # Retrieve the global unpaired P value already calculated for this
  # Method / Cell_Type / Timepoint combination.
  stat_row <- unpaired_stats %>%
    filter(
      Method == method_name,
      Cell_Type == cell_type,
      as.character(Timepoint) == timepoint
    )

  if (p_column %in% names(stat_row) && nrow(stat_row) > 0) {
    p_value <- suppressWarnings(as.numeric(stat_row[[p_column]][1]))
  } else {
    p_value <- NA_real_
  }

  p_label <- if (is.na(p_value)) {
    "P = N/A"
  } else if (p_value < 0.0001) {
    "P < 0.0001"
  } else {
    paste0("P = ", formatC(p_value, format = "f", digits = 4))
  }

  y_position <- shared_ylim[2] - 0.05 * diff(shared_ylim)
  x_position <- (length(cluster_levels_present) + 1) / 2

  ggplot(df, aes(x = cluster, y = Value, fill = cluster)) +
    geom_boxplot(
      outlier.shape = NA,
      alpha = 0.7
    ) +
    geom_jitter(
      width = 0.15,
      alpha = 0.65,
      size = 2,
      color = "gray20"
    ) +
    annotate(
      "text",
      x = x_position,
      y = y_position,
      label = p_label,
      size = 5,
      fontface = "bold"
    ) +
    scale_fill_manual(
      values = cluster_colors,
      name = "Cluster",
      drop = FALSE
    ) +
    theme_pub() +
    coord_cartesian(
      ylim = shared_ylim,
      clip = "off"
    ) +
    labs(
      title = paste0(cell_type, " — ", timepoint),
      x = "",
      y = method_value_type(method_name)
    ) +
    theme(
      legend.position = "none"
    )
}

# -----------------------------
# UNPAIRED save loop
# Same figure aesthetics/output logic as the CIM immunedeconv pipeline.
# Global and Dunn tables are still retained as companion CSVs for NEO.
# -----------------------------
save_unpaired_plots <- function(
    immune_patient_df,
    method_ymin,
    version_label,
    unpaired_obj,
    p_column = PLOT_P_COLUMN
) {

  stopifnot(!is.null(unpaired_obj))
  if (is.null(unpaired_obj$global)) {
    stop("unpaired_obj$global is missing")
  }

  immune_patient_df <- apply_cluster_names(immune_patient_df) %>%
    mutate(cluster = droplevels(cluster))

  out_dir <- file.path(
    DIRS$plots,
    version_label,
    "unpaired"
  )

  dir.create(
    out_dir,
    recursive = TRUE,
    showWarnings = FALSE
  )

  combinations <- immune_patient_df %>%
    distinct(Method, Cell_Type, Timepoint)

  purrr::pwalk(
    combinations,
    function(Method, Cell_Type, Timepoint) {

      timepoint_chr <- as.character(Timepoint)

      plot_data <- immune_patient_df %>%
        filter(
          Method == !!Method,
          Cell_Type == !!Cell_Type,
          as.character(Timepoint) == timepoint_chr
        )

      if (
        nrow(plot_data) == 0 ||
        n_distinct(
          plot_data$cluster[is.finite(plot_data$Value)]
        ) < 2
      ) {
        return(invisible(NULL))
      }

      shared_ylim <- compute_shared_ylims(
        plot_data,
        method_ymin
      )

      plot_object <- plot_unpaired_one(
        df = plot_data,
        shared_ylim = shared_ylim,
        unpaired_stats = unpaired_obj$global,
        p_column = p_column
      )

      if (is.null(plot_object)) {
        return(invisible(NULL))
      }

      stem <- safe_stem(
        "Unpaired",
        Method,
        as.character(Timepoint),
        Cell_Type
      )

      file_png <- file.path(
        out_dir,
        paste0(stem, ".png")
      )

      # Keep plot-linked statistical outputs for traceability.
      global_row <- unpaired_obj$global %>%
        filter(
          Method == !!Method,
          Cell_Type == !!Cell_Type,
          as.character(Timepoint) == timepoint_chr
        )

      if (nrow(global_row) > 0) {
        write.csv(
          global_row,
          file.path(out_dir, paste0(stem, "_stats_global.csv")),
          row.names = FALSE
        )
      }

      if (!is.null(unpaired_obj$dunn)) {
        dunn_rows <- unpaired_obj$dunn %>%
          filter(
            Method == !!Method,
            Cell_Type == !!Cell_Type,
            as.character(Timepoint) == timepoint_chr
          )

        if (nrow(dunn_rows) > 0) {
          write.csv(
            dunn_rows,
            file.path(out_dir, paste0(stem, "_stats_dunn.csv")),
            row.names = FALSE
          )
        }
      }

      ggsave(
        file_png,
        plot_object,
        width = PLOT_WIDTH_UNPAIRED,
        height = PLOT_HEIGHT_UNPAIRED,
        dpi = PLOT_DPI
      )
    }
  )

  invisible(TRUE)
}

# -----------------------------
# Helper: compute shared y-limits for paired data across clusters
# -----------------------------
compute_shared_ylims_paired <- function(paired_wide_df, method, cell_type, method_ymin) {
  method_min <- method_ymin %>% filter(Method == method) %>% pull(ymin)
  if (!length(method_min) || is.na(method_min)) method_min <- 0
  
  # Get all values (Biopsy + Surgery) across all clusters for this method/cell combo
  all_values <- paired_wide_df %>%
    filter(Method == !!method, Cell_Type == !!cell_type) %>%
    select(Biopsy, Surgery) %>%
    unlist() %>%
    as.numeric()
  
  auto_ylims(all_values, ymin_floor = method_min)
}

# -----------------------------
# PAIRED split plot (one cluster) with shared ylim and n_pairs in subtitle
# -----------------------------
plot_paired_split_one <- function(paired_wide_df, method, cell_type, cluster_level, shared_ylim) {
  d <- paired_wide_df %>%
    filter(Method == !!method, Cell_Type == !!cell_type, cluster == !!cluster_level)
  
  if (nrow(d) < 3) return(NULL)
  
  pval <- tryCatch(
    wilcox.test(d$Surgery, d$Biopsy, paired = TRUE, exact = FALSE)$p.value,
    error = function(e) NA_real_
  )
  
  p_label <- if (is.na(pval)) {
    "P = N/A"
  } else if (pval < 0.001) {
    "P < 0.001"
  } else {
    sprintf("P = %.4f", pval)
  }
  
  plot_df <- d %>%
    select(Patient_ID, Biopsy, Surgery) %>%
    pivot_longer(cols = c(Biopsy, Surgery), names_to = "Timepoint", values_to = "Value") %>%
    mutate(Timepoint = factor(Timepoint, levels = c("Biopsy", "Surgery")))
  
  ggplot(plot_df, aes(x = Timepoint, y = Value, group = Patient_ID)) +
    geom_line(color = "gray65", alpha = 0.7, size = 0.9) +
    geom_point(aes(color = Timepoint), size = 2.8, alpha = 0.85) +
    scale_color_manual(values = condition_colors, name = "Condition") +
    theme_pub() +
    coord_cartesian(ylim = shared_ylim, clip = "off") +
    labs(
      title = paste0(cell_type, " — ", as.character(cluster_level)),
      subtitle = paste0("n = ", nrow(d), " pairs"),
      x = "",
      y = method_value_type(method)
    ) +
    annotate("text", x = 1.5, y = shared_ylim[2] * 0.97,
             label = p_label, size = 6, fontface = "bold")
}

# -----------------------------
# PAIRED save loop (+ CSV per plot)
# Pre-compute shared y-limits per Method/Cell_Type across clusters
# -----------------------------
save_paired_split_plots <- function(paired_obj, method_ymin, version_label) {
  stopifnot(!is.null(paired_obj$wide), !is.null(paired_obj$stats))
  
  out_dir <- file.path(DIRS$plots, version_label, "paired_split")
  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  
  wide  <- paired_obj$wide
  stats <- paired_obj$stats
  
  # PRE-COMPUTE shared y-limits for each Method/Cell_Type combination (across all clusters)
  method_cell_combos <- wide %>% distinct(Method, Cell_Type)
  ylims_by_method_cell <- method_cell_combos %>%
    rowwise() %>%
    mutate(
      shared_ylim = list(compute_shared_ylims_paired(wide, Method, Cell_Type, method_ymin))
    ) %>%
    ungroup()
  
  combos <- wide %>% distinct(Method, Cell_Type, cluster)
  
  purrr::pwalk(combos, function(Method, Cell_Type, cluster) {
    # Get the pre-computed shared ylim for this Method/Cell_Type
    shared_ylim <- ylims_by_method_cell %>%
      filter(Method == !!Method, Cell_Type == !!Cell_Type) %>%
      pull(shared_ylim) %>%
      .[[1]]
    
    p <- plot_paired_split_one(wide, Method, Cell_Type, cluster, shared_ylim)
    
    stem <- safe_stem("Paired", Method, Cell_Type, paste0("cluster", gsub("[^A-Za-z0-9]+", "_", cluster)))
    fn_csv <- file.path(out_dir, paste0(stem, "_stats.csv"))
    fn_png <- file.path(out_dir, paste0(stem, ".png"))
    
    if (is.null(p)) {
      # Write skip CSV
      st <- NULL
      if (!is.null(paired_obj$skipped)) {
        st <- paired_obj$skipped %>%
          filter(Method == !!Method, Cell_Type == !!Cell_Type, cluster == !!cluster) %>%
          head(1)
      }
      if (is.null(st) || nrow(st) == 0) {
        st <- tibble(
          Method = Method, Cell_Type = Cell_Type, cluster = cluster,
          reason = "plot_skipped_n_pairs<3_or_missing_pairs",
          n_pairs = NA_integer_, p_value = NA_real_, FDR = NA_real_, mean_delta = NA_real_
        )
      }
      write.csv(st, fn_csv, row.names = FALSE)
      return(invisible(NULL))
    }
    
    # Get stats row for this plot
    st <- stats %>% filter(Method == !!Method, Cell_Type == !!Cell_Type, cluster == !!cluster)
    if (nrow(st) == 0) {
      st <- tibble(
        Method = Method, Cell_Type = Cell_Type, cluster = cluster,
        reason = "no_stats_row_found",
        n_pairs = NA_integer_, p_value = NA_real_, FDR = NA_real_, mean_delta = NA_real_
      )
    }
    write.csv(st, fn_csv, row.names = FALSE)
    
    # Save plot
    ggsave(fn_png, p, width = PLOT_WIDTH_PAIRED, height = PLOT_HEIGHT_PAIRED, dpi = PLOT_DPI)
  })
  
  invisible(TRUE)
}

# ==============================================================================
# RUN PLOTTING
# ==============================================================================
if (isTRUE(params$make_plots)) {
  # Match the CIM pipeline output logic: statistics are produced for both
  # ORIGINAL and AGGREGATED data, while the primary figure set uses the
  # AGGREGATED representation.
  save_unpaired_plots(
    immune_patient_aggregated,
    method_ymin_aggregated,
    "AGGREGATED",
    unpaired_aggregated
  )

  save_paired_split_plots(
    paired_aggregated,
    method_ymin_aggregated,
    "AGGREGATED"
  )
}

# ==============================================================================
# 20) HEATMAPS — NEO SAFE VERSION
# ==============================================================================

# Safely calculate z-scores.
# If there is no variation, return zeros instead of NA/NaN.
safe_zscore <- function(x) {
  x <- as.numeric(x)
  s <- sd(x, na.rm = TRUE)
  
  if (is.na(s) || s == 0) {
    return(rep(0, length(x)))
  }
  
  as.numeric((x - mean(x, na.rm = TRUE)) / s)
}


# Keep original NEO cluster labels.
# This does NOT rename clusters to Fun-CIM/Dys-CIM.
apply_cluster_names_safe <- function(df) {
  df %>%
    mutate(
      cluster = factor(as.character(cluster), levels = NEO_CLUSTER_ORDER)
    )
}


# Calculate cluster-level mean deltas and z-scored mean deltas.
# Delta = Surgery - Biopsy.
calculate_zscore_cluster_mean_deltas <- function(paired_wide_df) {
  
  required_cols <- c("Method", "Cell_Type", "cluster", "Biopsy", "Surgery", "Delta")
  
  missing_cols <- setdiff(required_cols, colnames(paired_wide_df))
  if (length(missing_cols) > 0) {
    stop("Missing required columns: ", paste(missing_cols, collapse = ", "))
  }
  
  paired_wide_df <- apply_cluster_names_safe(paired_wide_df)
  
  if (all(is.na(paired_wide_df$cluster))) {
    stop("All cluster values are NA after formatting. Check NEO_CLUSTER_ORDER and cluster labels.")
  }
  
  mean_deltas <- paired_wide_df %>%
    filter(!is.na(cluster)) %>%
    mutate(cluster = as.character(cluster)) %>%
    group_by(Method, Cell_Type, cluster) %>%
    summarise(
      n_pairs      = n(),
      mean_biopsy  = mean(Biopsy, na.rm = TRUE),
      sd_biopsy    = sd(Biopsy, na.rm = TRUE),
      mean_surgery = mean(Surgery, na.rm = TRUE),
      sd_surgery   = sd(Surgery, na.rm = TRUE),
      mean_delta   = mean(Delta, na.rm = TRUE),
      sd_delta     = sd(Delta, na.rm = TRUE),
      p_value = if (n() >= 3) {
        tryCatch(
          wilcox.test(Surgery, Biopsy, paired = TRUE, exact = FALSE)$p.value,
          error = function(e) NA_real_
        )
      } else {
        NA_real_
      },
      .groups = "drop"
    ) %>%
    filter(n_pairs >= 3) %>%
    group_by(Method, cluster) %>%
    mutate(FDR = p.adjust(p_value, method = "BH")) %>%
    ungroup()
  
  mean_deltas %>%
    group_by(Method) %>%
    mutate(zscore_delta = safe_zscore(mean_delta)) %>%
    ungroup()
}


# Convert long-format z-score results into heatmap matrices.
prepare_method_cell_by_cluster_matrix <- function(zscore_deltas,
                                                  fill_na = FALSE,
                                                  cluster_order = NEO_CLUSTER_ORDER) {
  
  zscore_deltas <- zscore_deltas %>%
    mutate(
      CellType_Method = paste0(Method, ": ", Cell_Type),
      cluster = as.character(cluster)
    ) %>%
    filter(!is.na(cluster))
  
  if (nrow(zscore_deltas) == 0) {
    stop("zscore_deltas has no rows after removing NA clusters.")
  }
  
  present_clusters <- unique(zscore_deltas$cluster)
  clusts_display <- cluster_order[cluster_order %in% present_clusters]
  
  if (length(clusts_display) == 0) {
    stop(
      "No matching clusters found. Present clusters are: ",
      paste(present_clusters, collapse = ", "),
      ". NEO_CLUSTER_ORDER is: ",
      paste(cluster_order, collapse = ", ")
    )
  }
  
  mat <- zscore_deltas %>%
    select(CellType_Method, cluster, zscore_delta) %>%
    tidyr::pivot_wider(names_from = cluster, values_from = zscore_delta) %>%
    tibble::column_to_rownames("CellType_Method") %>%
    as.matrix()
  
  sig_mat <- zscore_deltas %>%
    select(CellType_Method, cluster, FDR) %>%
    tidyr::pivot_wider(names_from = cluster, values_from = FDR) %>%
    tibble::column_to_rownames("CellType_Method") %>%
    as.matrix()
  
  mat <- mat[, clusts_display, drop = FALSE]
  sig_mat <- sig_mat[, clusts_display, drop = FALSE]
  
  if (fill_na) {
    mat[is.na(mat)] <- 0
  }
  
  method_anno <- data.frame(
    CellType_Method = rownames(mat),
    Method = sub(":.*$", "", rownames(mat)),
    stringsAsFactors = FALSE
  )
  
  method_anno$Method <- factor(method_anno$Method, levels = METHOD_ORDER)
  ord <- order(method_anno$Method, method_anno$CellType_Method)
  
  list(
    mat  = mat[ord, , drop = FALSE],
    sig  = sig_mat[ord, , drop = FALSE],
    anno = method_anno[ord, , drop = FALSE]
  )
}


# Plot and save cluster-level heatmap.
plot_cluster_delta_heatmap <- function(zscore_deltas,
                                       output_file,
                                       show_significance = FALSE,
                                       cluster_rows = TRUE,
                                       fill_na = FALSE,
                                       na_col = "grey90",
                                       cluster_order = NEO_CLUSTER_ORDER) {
  
  prep <- prepare_method_cell_by_cluster_matrix(
    zscore_deltas = zscore_deltas,
    fill_na = fill_na,
    cluster_order = cluster_order
  )
  
  mat <- prep$mat
  sig <- prep$sig
  anno <- prep$anno
  
  if (nrow(mat) == 0 || ncol(mat) == 0) {
    stop("Matrix is empty after preparation.")
  }
  
  if (all(is.na(mat))) {
    stop("All values in matrix are NA.")
  }
  
  max_abs <- max(abs(mat), na.rm = TRUE)
  
  if (!is.finite(max_abs) || max_abs == 0) {
    max_abs <- 2
  }
  
  max_abs <- max(max_abs, 2)
  
  col_fun <- circlize::colorRamp2(
    c(-max_abs, -1, 0, 1, max_abs),
    c("#0571B0", "#92C5DE", "white", "#F4A582", "#CA0020")
  )
  
  methods_present <- unique(as.character(anno$Method))
  method_colors_subset <- METHOD_COLORS[methods_present]
  method_colors_subset <- method_colors_subset[!is.na(method_colors_subset)]
  
  row_ha <- ComplexHeatmap::rowAnnotation(
    Method = anno$Method,
    col = list(Method = method_colors_subset),
    annotation_name_gp = grid::gpar(fontsize = 11, fontface = "bold")
  )
  
  cell_fun <- NULL
  
  if (show_significance) {
    cell_fun <- function(j, i, x, y, width, height, fill) {
      fdr <- sig[i, j]
      
      if (is.na(fdr)) return(NULL)
      
      if (fdr < 0.001) {
        grid::grid.text("***", x, y, gp = grid::gpar(fontsize = 8, fontface = "bold"))
      } else if (fdr < 0.01) {
        grid::grid.text("**", x, y, gp = grid::gpar(fontsize = 8, fontface = "bold"))
      } else if (fdr < 0.05) {
        grid::grid.text("*", x, y, gp = grid::gpar(fontsize = 8, fontface = "bold"))
      }
    }
  }
  
  display_names <- stringr::str_remove(rownames(mat), "^[^:]+: ")
  
  ht <- ComplexHeatmap::Heatmap(
    mat,
    name = "Z-score\n(mean Δ)",
    col = col_fun,
    na_col = na_col,
    cell_fun = cell_fun,
    row_names_side = "left",
    row_labels = display_names,
    row_names_gp = grid::gpar(fontsize = 9),
    cluster_rows = cluster_rows,
    show_row_dend = cluster_rows,
    cluster_columns = FALSE,
    column_names_gp = grid::gpar(fontsize = 14, fontface = "bold"),
    column_names_rot = 0,
    column_title = "NEO cluster-mean zΔ (Surgery - Biopsy)",
    column_title_gp = grid::gpar(fontsize = 16, fontface = "bold"),
    left_annotation = row_ha,
    border = TRUE
  )
  
  plot_height <- max(10, nrow(mat) * 0.15 + 3)
  
  grDevices::png(
    output_file,
    width = 12,
    height = plot_height,
    units = "in",
    res = PLOT_DPI
  )
  
  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "bottom"
  )
  
  grDevices::dev.off()
  
  output_file
}


# ==============================================================================
# RUN HEATMAP ANALYSIS — NEO
# ==============================================================================

if (isTRUE(params$make_heatmaps)) {

  paired_aggregated_named <- paired_aggregated

  paired_aggregated_named$wide <- paired_aggregated_named$wide %>%
    mutate(cluster = factor(as.character(cluster), levels = NEO_CLUSTER_ORDER))

  cat("\nDEBUG: Unique cluster values in paired_aggregated_named$wide:\n")
  print(unique(paired_aggregated_named$wide$cluster))

  cat("\nDEBUG: Cluster counts:\n")
  print(table(paired_aggregated_named$wide$cluster, useNA = "ifany"))

  zscore_deltas_agg <- calculate_zscore_cluster_mean_deltas(
    paired_aggregated_named$wide
  )

  cat("\nDEBUG: Unique cluster values in zscore_deltas_agg:\n")
  print(unique(zscore_deltas_agg$cluster))

  cat("\nDEBUG: First few rows of zscore_deltas_agg:\n")
  print(head(zscore_deltas_agg))

  write.csv(
    zscore_deltas_agg,
    file.path(DIRS$results, "NEO_AGGREGATED_zscore_cluster_mean_deltas.csv"),
    row.names = FALSE
  )

  prep_tmp <- prepare_method_cell_by_cluster_matrix(
    zscore_deltas_agg,
    fill_na = FALSE,
    cluster_order = NEO_CLUSTER_ORDER
  )

  write.csv(
    prep_tmp$mat,
    file.path(DIRS$results, "NEO_AGGREGATED_heatmap_matrix_zscore_delta.csv")
  )

  write.csv(
    prep_tmp$sig,
    file.path(DIRS$results, "NEO_AGGREGATED_heatmap_matrix_FDR.csv")
  )

  heat_file <- file.path(
    DIRS$plots,
    "NEO_AGGREGATED_heatmap_cluster_mean_zDelta.png"
  )

  plot_cluster_delta_heatmap(
    zscore_deltas_agg,
    heat_file,
    show_significance = TRUE,
    cluster_rows = TRUE,
    fill_na = FALSE,
    na_col = "grey90",
    cluster_order = NEO_CLUSTER_ORDER
  )

  cat("\nHeatmap saved to:", heat_file, "\n")


}

# ==============================================================================
# 21) REPRODUCIBILITY RECORD
# ==============================================================================

cat("\nWriting reproducibility metadata...\n")

input_manifest <- tibble::tibble(
  input = c("NEO expression TPM", "CIBERSORTx", "NEO cluster assignments"),
  file = c(NEO_EXP_FILE, CIBERSORT_NEO_FILE, NEO_CLUSTER_CSV),
  exists = file.exists(c(NEO_EXP_FILE, CIBERSORT_NEO_FILE, NEO_CLUSTER_CSV)),
  size_bytes = file.info(c(NEO_EXP_FILE, CIBERSORT_NEO_FILE, NEO_CLUSTER_CSV))$size,
  modified = as.character(
    file.info(c(NEO_EXP_FILE, CIBERSORT_NEO_FILE, NEO_CLUSTER_CSV))$mtime
  )
)

utils::write.csv(
  input_manifest,
  file.path(DIRS$logs, "NEO_input_manifest.csv"),
  row.names = FALSE
)

loaded_packages <- c(
  "data.table", "immunedeconv", "dplyr", "tidyr", "tibble", "stringr",
  "purrr", "ggplot2", "readr", "ComplexHeatmap", "circlize", "withr",
  "conflicted", "FSA"
)

package_versions <- tibble::tibble(
  package = loaded_packages,
  version = vapply(
    loaded_packages,
    function(p) as.character(utils::packageVersion(p)),
    character(1)
  )
)

utils::write.csv(
  package_versions,
  file.path(DIRS$logs, "NEO_package_versions.csv"),
  row.names = FALSE
)

capture.output(
  sessionInfo(),
  file = file.path(DIRS$logs, "NEO_sessionInfo.txt")
)

cat("Project directory:", PROJECT_DIR, "\n")
cat("Reproducibility files written to:", DIRS$logs, "\n")
cat("=== Immune infiltration pipeline completed: NEO ===\n")
cat("Timestamp:", format(Sys.time()), "\n")

# Close the run log sink opened in Section 6.
sink()


```
